---
title: "1 · Course overview, install party, onboarding"
subtitle: "Tuesday, August 25, 2026"
pagetitle: "AEM 6850 · Session 1 · Overview & onboarding"
execute:
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
  revealjs:
    output-file: 01-overview-slides.html
    theme: [default, slides.scss]
    footer: "AEM 6850 · Session 1"
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    code-overflow: scroll
    history: false
    link-external-newwindow: true
---

```{r}
#| purl: false
#| include: false
knitr::opts_chunk$set(
  comment = "#>",
  collapse = FALSE,
  fig.width = 8,
  fig.height = 4,
  fig.align = "center",
  dpi = 150
)
```

::: {.content-hidden when-format="revealjs"}

[Slides](01-overview-slides.html) · [Live script](01-overview.R) ·
[Source](https://github.com/ArielOrtizBobea/aem6850/blob/main/fall-2026/sessions/01-overview.qmd)

This page is the manual for the session. The slides are the same material in
the shape it takes in the room; the live script is every line of code below,
in order, ready to run. All three come out of one file, so they cannot
disagree with each other.

:::

## Objectives {#objectives}

By the end of this session you can:

- State the course thesis and read the course map.
- Run a script and find its outputs.
- Predict what a simple script prints, then reconcile the prediction with what
  actually happened.
- Accept a GitHub Classroom assignment and commit a change in the browser.

::: {.content-hidden when-format="revealjs"}

## Before class {#before-class}

Nothing is required. If you already have R and RStudio, make sure you can
launch both.

The install party runs at very different speeds on different machines. If
yours finishes early, start the onboarding section; if it is still churning,
do the onboarding anyway — it happens entirely in a browser and needs nothing
installed. If your laptop is locked down, borrowed, or eight years old, come
anyway: the fallback path — a free browser-based R — is in the [install
party](#install-check), and it is a perfectly good way to get through the
first weeks while you sort out a local install. The demo that closes the
session is driven from the front of the room; following along on your own
machine is better, watching is fine, and the script is on this page either
way.

:::

## The thesis {#thesis}

::: {.content-hidden when-format="revealjs"}

Every empirical paper you have ever read rests on a dataset somebody had to
find, fetch, clean, check, and reshape. Your methods courses begin *after*
that work is finished. This course is that work.

Two things about it have changed recently.

The first is the raw material. The evidence economists now use arrives as
text, PDFs and scans, photographs, satellite tiles, audio, and maps — not just
as tables. Nightlights measure growth where the national accounts are thin;
scraped listings measure prices at daily frequency; a scanned census from 1890
measures things nobody digitized until someone did.

The second is that a machine will now write a competent first draft of the
code that handles all of it. This does not make the work easier. It moves
where the work is. If plausible code takes seconds to produce, the scarce
skill is no longer producing it — it is knowing whether the result is right.

:::

::: {.content-visible when-format="revealjs"}

- The data economists use now arrives as text, documents, images, audio, maps
  — not just tables.
- A model will write a competent first draft of the code that handles it.
- That does not make the work easier. It moves where the work is.

:::

> **Models draft. You verify.**
>
> "It runs" is not "it's right." A wrong number that runs clean is the
> characteristic failure of this era, and it is the one this course trains you
> to catch.

::: {.content-hidden when-format="revealjs"}

So every homework this semester ends the same way: with a number checked
against something outside the computer. A published total. A figure printed on
the page the table came from. A calendar fact. A place you have been. That
check is not a formality at the end of the assignment — it is the assignment.

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
Do not rush this. Everything else in the course is downstream of the students
believing this claim. Ask the room: who has already had a model hand them code
that ran fine and was wrong?
:::
:::

## The course map {#map}

Twenty-eight meetings, two parts.

::: {.content-visible when-format="revealjs"}

- **Part I — foundations** (1–12). The whole pipeline, on ordinary tables.
  By hand through session 7; AI enters at session 8.
- **Part II — applications** (13–28). One family of unconventional data per
  session, each through the same pipeline.

:::

::: {.content-hidden when-format="revealjs"}

**Part I — foundations** (sessions 1–12, weeks 1–6). The whole pipeline, on
ordinary tables: R, diagnostic plots, wrangling, joins, functions and
debugging, git, AI, verification strategy, reproducible projects, graphics for
readers. You work by hand through session 7; AI enters the toolkit at session
8, once you can tell whether its output is any good.

**Part II — applications** (sessions 13–28, weeks 7–15). One family of
unconventional data per session, each walked through the same pipeline: what
the family is good for, where to get it, how to process it, and the checks
that particular family demands.

The [full schedule](../index.qmd#schedule) lists all twenty-eight sessions.
The [syllabus](../syllabus.qmd) has the grading, the AI policy, and the
homework calendar.

:::

## The six families {#families}

**messy tables · text · documents (PDFs & scans) · images · audio · maps &
rasters**

Cutting across them are the *channels* — the ways data reaches you: packaged
downloads, APIs, scraping, licensed vendors, replication archives, and data
you collect yourself.

**What this course does not cover.** No causal inference, no econometric
theory, no estimation strategy — that is AEM 6851 in the spring, and this
course is its prerequisite in spirit. No machine-learning theory: we *use*
models, we do not derive them. No video, no network data.

::: {.content-hidden when-format="revealjs"}

And no package worship: the point is never `dplyr`, it is what `dplyr` is for.

:::

## Install party {#install}

Do these in order. Order matters: RStudio looks for R when it starts, so R has
to be there first.

**1 — Install R.** [cran.r-project.org](https://cran.r-project.org) → your
operating system.

- *macOS:* there are two builds. Apple silicon (M1–M4) needs the `arm64`
  installer; older Intel Macs need the `x86_64` one. Apple menu → About This
  Mac tells you which you have.
- *Windows:* click "base", then the download link. You do not need Rtools.

**2 — Install RStudio.**
[posit.co/download/rstudio-desktop](https://posit.co/download/rstudio-desktop)
→ the free Desktop version. Quarto ships inside it; there is nothing else to
install.

## Check it worked {#install-check}

**3 —** Open RStudio — not R — and type this into the Console, bottom left:

```{r}
#| label: check-install
#| script: "Check the install"
#| eval: false
R.version.string
```

You should see version 4-point-something. If you see an error, or RStudio says
it cannot find an R installation, raise your hand.

**4 — One setting, right now.** Tools → Global Options → General: uncheck
**"Restore .RData into workspace at startup"**, and set **"Save workspace to
.RData on exit"** to **Never**.

::: {.content-hidden when-format="revealjs"}

This takes fifteen seconds and prevents a whole category of misery. By default
RStudio saves your invisible leftovers between sessions, so a script can
appear to work because of an object you made an hour ago and have since
deleted from the file. With the setting off, your script is the only record of
what you did — which is the entire premise of a reproducible project.

**Four panes, and which one matters today.** The Console (bottom left) runs
code the moment you press Return. The Source pane (top left) holds scripts:
code you keep. The Environment tab (top right) lists the objects you have
made. Plots and Files (bottom right) is where figures and your folder show up.
For now: type in the Console, keep anything worth keeping in Source.

::: {.column-margin}
**If your laptop refuses.** Make a free account at
[posit.cloud](https://posit.cloud) — it is RStudio in a browser, and
everything in the first weeks works there. Then bring the machine to office
hours and we will fix the local install properly.
:::

:::

::: {.content-visible when-format="revealjs"}

Laptop refusing? [posit.cloud](https://posit.cloud) is RStudio in a browser.
Free, works today, fix the local install later.

:::

## Onboarding — get your repo {#onboarding}

Not graded. Everyone finishes this today, in the room. It happens entirely in
a browser, so a half-installed laptop is no obstacle.

**1 — Get a GitHub account.** [github.com/signup](https://github.com/signup).

::: {.content-hidden when-format="revealjs"}

Any email address works; you can add your Cornell address later to claim the
free GitHub Education benefits. Pick a username you would be willing to put on
a CV, because for many of you this account will eventually *be* your CV.
Already have an account? Sign in with it. Do not make a second one.

:::

**2 — Click the invite link.** It is on Canvas and on the screen.

**3 — Accept the assignment.** Authorize GitHub Classroom when it asks, find
your name in the roster list, and click through. Wait for "You're ready to
go!" and follow the link to your new repository.

That repository is private, it belongs to you, and the instructor can see it.
Nothing you put there is visible to the rest of the class.

## Onboarding — your first commit {#onboarding-commit}

**4 — Edit the README in the browser.** Click `README.md`, then the pencil
icon. Fill in the three lines:

```
Name:
Program:
One dataset or question I would like to be able to handle by December:
```

**5 — Commit.** Green **"Commit changes…"** button, top right. Type a short
message — `add my intro` — leave "Commit directly to the `main` branch"
selected, and click **Commit changes**.

**6 — Look at what you did.** Reload the repository's front page. Your text is
there, and above it your message, `add my intro`, with a timestamp.

## You just used version control {#onboarding-reveal}

::: {.callout-note appearance="simple"}
**What just happened.** That was a commit — a permanent, timestamped,
attributed snapshot of your work, the thing version control exists to produce.
You made one without installing git, opening a terminal, or learning a single
git command, because the browser did it for you.
:::

That is how the first three homeworks get handed in: edit or drag files onto
the repository page, click Commit. Session 7 pulls the curtain back and shows
you what you have been doing for three weeks — doing it first and
understanding it second is deliberate.

**7 — Check in.** Paste your repository URL into the poll. That is today's
attendance, and it is also the verification step: it proves forty repositories
exist, not that forty people nodded.

::: {.content-hidden when-format="revealjs"}

**If something goes wrong.**

| Symptom | Fix |
|:--|:--|
| Your name is not in the roster list | Use "Skip to the next step", then email the instructor your GitHub username |
| The link 404s or says the invitation expired | Raise your hand — links get regenerated in seconds |
| You accepted with the wrong GitHub account | Sign out, sign in as the right one, click the link again |
| No pencil icon on the README | You are looking at someone else's repository, or you are signed out |

:::

## Predict before you run {#predict}

::: {.content-hidden when-format="revealjs"}

On January 7, 2025, two fires started in Los Angeles County — the Palisades
fire near the coast, the Eaton fire above Altadena. Between them they
destroyed something like sixteen thousand structures. We will come back to
those fires all semester; they are the running example the homeworks are built
on. Today they are just a reason to pull one small file.

:::

::: {.content-visible when-format="revealjs"}

January 7, 2025: the Palisades and Eaton fires start in Los Angeles County.
The semester's running example. Today, just a reason to pull one file.

:::

We are about to download daily weather for downtown Los Angeles — December 1,
2024 through February 28, 2025, ninety days, six columns — and plot the
strongest wind gust recorded each day.

**Before any of that, write down three predictions.** On paper, or in a
comment at the top of your script. Written down, not held in your head.

1. Which date in that window has the largest daily maximum gust?
2. Roughly how many times bigger is that gust than a typical day's?
3. How much rain fell on Los Angeles in December 2024?

::: {.content-hidden when-format="revealjs"}

You are allowed — encouraged — to reason from what you already know about
those fires and about Southern California in winter. That is the difference
between a prediction and a guess.

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
Take answers out loud before running anything. Someone will say Jan 7 for Q1;
ask them why they believe it. Q2 gets wild spread — good, that is the point.
Q3 is the one nobody predicts well.
:::
:::

## Step 1 — Pull the data {#demo-pull}

The data comes from [Open-Meteo](https://open-meteo.com), which serves
historical weather as a plain CSV over a URL. No account, no key, no package.

```{r}
#| label: demo-pull
#| script: "Pull the data"
#| message: false
url <- paste0(
  "https://archive-api.open-meteo.com/v1/archive",
  "?latitude=34.05&longitude=-118.24",
  "&start_date=2024-12-01&end_date=2025-02-28",
  "&daily=wind_gusts_10m_max,wind_speed_10m_max,",
  "temperature_2m_max,relative_humidity_2m_min,precipitation_sum",
  "&timezone=America%2FLos_Angeles&format=csv"
)

la <- read.csv(url, skip = 3)
names(la) <- c("date", "gust", "wind", "tmax", "rh_min", "precip")
la$date <- as.Date(la$date)
```

::: {.content-hidden when-format="revealjs"}

Three things in there are worth a sentence each.

`read.csv()` takes a URL as happily as a filename. `skip = 3` is there because
the first three lines of the file are about the *location* — latitude,
longitude, elevation, time zone — not about the weather; the real header is on
line four. Open the URL in a browser tab and you can see it. And the file
ships column names like `wind_gusts_10m_max..km.h.`, which is what happens
when R sanitizes `wind_gusts_10m_max (km/h)`. Renaming them yourself is not
cosmetic: you will type these names two hundred times today.

::: {.callout-tip collapse="true"}
## If the download fails

Campus wifi, a rate limit, a sleeping API — it happens. A copy of the exact
response is saved with this session, so the rest of the demo still works:

```r
la <- read.csv(paste0(
  "https://arielortizbobea.github.io/aem6850/fall-2026/sessions/",
  "data/la-weather-dec2024-feb2025.csv"
), skip = 3)
```

Committing the raw response so a pipeline can be re-run identically later is a
habit we formalize in session 14.
:::

:::

## Step 2 — Look at what arrived {#demo-look}

Never compute on a file you have not looked at. Three lines, every time:

```{r}
#| label: demo-look
#| script: "Look at what arrived"
dim(la)
str(la)
summary(la$gust)
```

::: {.content-hidden when-format="revealjs"}

`dim()` says ninety rows and six columns — which is what ninety days of daily
data should be, and if it said 89 or 91 we would stop and find out why.
`str()` says `date` is a Date and the rest are numbers, so nothing arrived as
text by accident. `summary()` gives the range: if the largest gust were 4 or
4000, something would be wrong with the units before we ever made a plot.

None of this is ceremony. Session 3 turns it into a named habit — arrival
checks — and every homework from here on expects it.

:::

## Step 3 — One plot {#demo-plot}

```{r}
#| label: demo-plot
#| script: "One plot"
#| fig-alt: "Daily maximum wind gust in downtown Los Angeles, December 2024 through February 2025. Most days sit near 20 km/h. A single spike reaches about 66 km/h on January 7, 2025, with a second, smaller spike the following day."
plot(la$date, la$gust,
     type = "h", col = "grey30",
     xlab = "", ylab = "Maximum wind gust (km/h)",
     main = "Downtown Los Angeles, daily maximum gust")

abline(h = median(la$gust), lty = 2, col = "grey60")
abline(v = as.Date("2025-01-07"), col = "#b31b1b", lwd = 2)
text(as.Date("2025-01-07"), max(la$gust), "  Jan 7",
     col = "#b31b1b", adj = c(0, 1))
```

::: {.content-hidden when-format="revealjs"}

That is base R graphics, which is what session 3 is about: plots you make in
one line to interrogate a column, not plots you make for readers. `type = "h"`
draws each day as a vertical spike, which is the right shape for the question
"how unusual was this one day."

:::

## Step 4 — Reconcile {#demo-reconcile}

Now compare the screen against what you wrote down.

```{r}
#| label: demo-reconcile
#| script: "Reconcile against the predictions"
la$date[which.max(la$gust)]                 # 1. the windiest day
max(la$gust) / median(la$gust)              # 2. how unusual it was
sum(la$precip[format(la$date, "%Y-%m") == "2024-12"])  # 3. December rain, in mm
```

::: {.content-hidden when-format="revealjs"}

**The windiest day of the winter is the day the fires started.** Not the week
— the day. The data and the world agree, and they agree on something specific
enough that the agreement means something.

**The gust is 3.3 times a normal day.** Most winter days in downtown LA top
out around 20 km/h; January 7 hit 66.

**December brought half a millimetre of rain.** Not half an inch. Two days in
the entire month recorded anything at all. Wind does not burn a city on its
own; wind arriving on three months of dry fuel does.

:::

::: {.content-visible when-format="revealjs"}

- The windiest day of the winter is the day the fires started.
- It was 3.3× a normal day.
- December's total rainfall: half a millimetre.

:::

## Where the number came from {#measurement}

66 km/h is about 41 mph. The National Weather Service was warning about that
windstorm in far more dramatic terms than 41 mph, and it was right to. Both
numbers are correct.

This file reports **one model grid cell over downtown Los Angeles, ten metres
above flat ground.** The fires started in canyons twenty kilometres away,
where Santa Ana winds funnel and accelerate.

::: {.callout-important appearance="simple"}
A dataset is not the world. It is a measurement of the world, taken somewhere
specific, by some specific method, at some specific resolution. Half the
mistakes in empirical work are made by people who forgot to ask where their
number came from.
:::

::: {.content-hidden when-format="revealjs"}

**Your turn:** find the peak gust the National Weather Service actually
recorded during that windstorm, and where the station was. Write both numbers
down and bring them Thursday. That is a reconciliation — the smallest possible
version of what every homework in this course asks you to do.

:::

::: {.content-visible when-format="revealjs"}

**Before Thursday:** find the NWS's recorded peak gust for that windstorm, and
the station that recorded it. Bring both numbers.

:::

## You try {#you-try}

Five minutes. The stub is in the live script.

```{r}
#| label: you-try
#| exercise: true
#| script: "You try (5 minutes)"
#| eval: false
# 1. Pick a place you care about. Look up its latitude and longitude, put them
#    into the URL above, and re-run everything from the top.
#
# 2. Plot `tmax` instead of `gust`. What units is it in? (If you are not sure,
#    look at line 4 of the raw file -- open the URL in a browser.)
#
# 3. Add one comment saying, in a sentence, what your plot shows.
```

## Verify {#verify}

Each session in this course ends with the verification habit it installs.
This is the first one, and everything later is a variation on it.

> **Predict, then run, then reconcile.**

::: {.content-hidden when-format="revealjs"}

Write the prediction down before you run the code. A prediction you keep in
your head quietly reshapes itself to match whatever appears on the screen —
you will remember having expected exactly that. A written one cannot do that.

When the prediction and the output disagree, exactly one of three things is
true:

- **Your prediction was wrong** — you learned something about the world.
- **The code is wrong** — you learned something about your code.
- **The data is not what you thought it was** — you learned something about
  the data, which is usually the most valuable of the three.

All three are wins. The only losing move is not predicting, because then the
output has nothing to disagree with and you will believe it.

:::

::: {.content-visible when-format="revealjs"}

Prediction ≠ output means one of three things, all of them useful:

- your prediction was wrong → you learned about the world
- the code is wrong → you learned about your code
- the data is not what you thought → you learned about the data

No prediction, no disagreement, no learning.

:::

## Quick reference {#quick-reference}

::: {.content-visible when-format="revealjs"}

Everything from today lives on the session page:

- the install checklist
- every command we used, with a one-line description
- the keyboard shortcuts
- this deck, and the runnable script

:::

::: {.content-hidden when-format="revealjs"}

**Install checklist**

| Step | Where | Done when |
|:--|:--|:--|
| R | [cran.r-project.org](https://cran.r-project.org) — match your chip on macOS | `R.version.string` prints 4.x |
| RStudio | [posit.co/download/rstudio-desktop](https://posit.co/download/rstudio-desktop) | It opens and finds R |
| Workspace setting | Tools → Global Options → General | Restore is off, Save is Never |
| GitHub | [github.com/signup](https://github.com/signup) | You can sign in |
| Course repo | The Classroom invite link | Your commit shows on the repo page |
| Fallback | [posit.cloud](https://posit.cloud) | Only if the laptop lost |

**Today's R, in one screen**

| Code | What it does |
|:--|:--|
| `R.version.string` | Which R you are running |
| `read.csv(url, skip = 3)` | Read a CSV from a file or a URL, ignoring the first 3 lines |
| `names(d) <- c(...)` | Rename all the columns |
| `as.Date(x)` | Turn text like `"2025-01-07"` into a date |
| `dim(d)`, `str(d)`, `summary(d$x)` | The three arrival checks |
| `d$x` | One column, by name |
| `d$x[which.max(d$y)]` | The value of `x` in the row where `y` is largest |
| `plot(x, y, type = "h")` | Scatter, or spikes with `type = "h"` |
| `abline(h =, v =, lty =, col =)` | A horizontal or vertical reference line |
| `median(x)`, `max(x)`, `sum(x)` | The obvious three |

**Keyboard**

| Keys | What |
|:--|:--|
| `Cmd`/`Ctrl` + `Return` | Run the current line and move down |
| `Cmd`/`Ctrl` + `Shift` + `Return` | Run the whole script |
| `Ctrl` + `L` | Clear the console |
| `Tab` | Complete a name |
| `↑` | The last thing you typed |

**Links**

- Session materials: [slides](01-overview-slides.html) ·
  [live script](01-overview.R) ·
  [snapshot of today's data](data/la-weather-dec2024-feb2025.csv)
- [Syllabus](../syllabus.qmd) · [full schedule](../index.qmd#schedule)
- Open-Meteo historical archive: [open-meteo.com](https://open-meteo.com)

:::

## After class {#after-class}

**Before Thursday**, four things:

1. R and RStudio installed and opening, with the workspace setting off.
2. Your course repository accepted, with your commit on it.
3. Today's script run once on your own machine, start to finish.
4. The two NWS numbers from the reconciliation above.

::: {.content-hidden when-format="revealjs"}

**Thursday** is R essentials — objects, vectors, types, data frames, reading a
real and slightly broken file, subsetting. It is the densest session of the
semester, and the live script for it is the safety net: everything we do in
the room, you can re-run at home, one line at a time.

**HW1 posts Thursday after class** and is due the following Thursday at 8:40
am. It uses EPA air-quality data from the same fires. You hand it in the way
you handed in today's README: edit or drag files onto the repository page,
click Commit.

**No textbook.** These pages are the manual. They stay at these URLs, and next
year's version of the course lives somewhere else, so a link you save today
keeps working.

**Stuck?** Office hours, or ask in class. Nobody in this room is expected to
have done this before — that is what "no prerequisites" means, and it is
enforced.

:::

::: {.content-visible when-format="revealjs"}

Thursday: R essentials. HW1 posts after class.

:::

```{r}
#| purl: false
#| include: false
# Single-source build: the .R live script is generated from this file every
# time the page renders, so it cannot fall out of step with the prose.
source("_purl.R")
purl_session("01-overview.qmd", "01-overview.R")
```
