---
title: "2 · R essentials"
subtitle: "Thursday, August 27, 2026"
pagetitle: "AEM 6850 · Session 2 · R essentials"
execute:
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
  revealjs:
    output-file: 02-r-essentials-slides.html
    theme: [default, slides.scss]
    template-partials:
      - title-slide.html
    footer: "AEM 6850 · Session 2"
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    code-overflow: scroll
    history: false
    link-external-newwindow: true
---

```{r}
#| purl: false
#| include: false
knitr::opts_chunk$set(
  comment = "#>",
  collapse = FALSE,
  fig.width = 8,
  fig.height = 4,
  fig.align = "center",
  dpi = 150
)
```

::: {.content-hidden when-format="revealjs"}

[Slides](02-r-essentials-slides.html) · [Live script](02-r-essentials.R) ·
[Source](https://github.com/ArielOrtizBobea/aem6850/blob/main/fall-2026/sessions/02-r-essentials.qmd)

This is the densest session of the semester. Everything below is in the live
script. Run it line by line at home, as many times as it takes. You do not
need to memorize anything today.

:::

## Objectives {#objectives}

By the end of this session you can:

- Build vectors and do arithmetic on them.
- Say what R does when a vector mixes classes, and why that turns a column
  of numbers into text.
- Read a CSV and check its size, its column classes, and its ranges.
- Subset rows and columns with logical conditions.
- Protect an identifier column so R does not turn it into a number.
- Recognize factors, matrices, and lists, and say what each is for.

::: {.content-hidden when-format="revealjs"}

This is a lot for one session, and that is deliberate. Nobody absorbs it all
in the room. See it once today, then re-run the script until it sticks.

## Before class {#before-class}

Have R and RStudio open and working, and bring the two National Weather
Service numbers from [session 1](01-overview.qmd#measurement).

If your install still does not work, sit next to someone whose install does,
and use [posit.cloud](https://posit.cloud) today. Do not spend this session
troubleshooting. Bring the machine to office hours instead.

:::

## R is a language and a program {#language-and-program}

:::: {.columns}
::: {.column width="74%"}

- **R is a language.** A grammar for saying what to do with data.
- **R is also a program.** The thing you installed, which reads that
  grammar and runs it.
- **CRAN is where the program comes from.** The Comprehensive R Archive
  Network: mirrored servers holding identical copies of R and of the
  roughly 25,000 packages people have written for it.
- **RStudio is neither.** It is an editor with an R console attached.
- Delete RStudio and R still works. Delete R and nothing works.

:::
::: {.column width="26%"}

![](figures/Rlogo.svg){fig-alt="The R logo" width="100%"}

:::
::::

::: {.content-hidden when-format="revealjs"}

The distinction tells you who to ask when something breaks. "R does not have
that function" is a language question, and the answer is a package or a
different function. "R will not start" is a program question, and the answer
is a reinstall. Most people report the second when they mean the first.

You will not touch CRAN directly today. Session 4 installs the course's
first package from it, with `install.packages()`.

[R logo](https://www.r-project.org/logo/) © The R Foundation, used
unmodified under
[CC-BY-SA 4.0](https://creativecommons.org/licenses/by-sa/4.0/).

:::

## Everything that exists is an object {#everything-is-an-object}

```{r}
#| label: everything-is-an-object
#| script: "Everything that exists is an object"
x <- c(1, 2, 3)
f <- function(v) v * 2

class(x)
class(f)        # a function is an object too
class(`+`)      # so is the plus sign
```

::: {.content-hidden when-format="revealjs"}

Not just data. A function is an object with a class, a name you gave it, and
a place in your environment, exactly like a vector. You can store one in a
list, pass one to another function, or print it by typing its name without
parentheses. `class()` says `function` for all three above.

Underneath, the classes differ: `typeof(f)` is `closure` and `` typeof(`+`) ``
is `builtin`. That is the class-and-type distinction again.

:::

## Everything that happens is a function call {#everything-is-a-call}

```{r}
#| label: everything-is-a-call
#| script: "Everything that happens is a function call"
x + 1
`+`(x, 1)      # the same call, written plainly
`[`(x, 2)      # the square bracket is a function too
```

::: {.content-hidden when-format="revealjs"}

The two headings above are
[John Chambers'](https://adv-r.hadley.nz/functions.html), one of R's
designers. They are a pair, and the second is the half people garble: the
function is the object, and the *call* is what happens. `x + 1` is a call to
the function object named `+`. The backticks let you write that name where R
would otherwise read an operator.

You will never write code this way. Knowing it is true explains why R is
consistent: there is one mechanism underneath, not a list of special cases.

:::

## What "object-oriented" means {#oop}

```{r}
#| label: dispatch
#| script: "One function, many classes"
summary(c(1, 2, 3, 100))
summary(c("a", "b", "a"))
```

::: {.content-visible when-format="revealjs"}

- One name, two behaviors. R reads the object's class and picks a method to
  match. The word for this is **polymorphism**.
- You learn `summary()`, `print()` and `plot()` once, and they keep working
  on things you have not met yet.

:::

::: {.content-hidden when-format="revealjs"}

`summary()` is one name and two different behaviors. On numbers it returns
six statistics. On text it reports length, class and mode, because quartiles
of "a" mean nothing. R reads the object's class and picks a method to match.
The technical word is **polymorphism**: one interface, many implementations,
chosen by the class of what you hand it.

That is the whole advantage, and it is worth being concrete. You learn
`summary()`, `print()` and `plot()` once. Install a package next month that
defines some new kind of object, and those three will usually work on it
straight away, because the package author wrote methods for the generics you
already know. You are not learning a fresh vocabulary per package. Without
this you would need `summary_vector()`, `summary_data_frame()`,
`summary_regression()`, and a new name for every type anyone ever invents.

**Two ways to arrange it.** Python and Java put the methods *inside* the
object, so calls read `object.method(arg)`. R puts them inside the
*function*: `summary()` is a generic that owns a set of methods and chooses
between them, so calls read `generic(object, arg)`. Same idea, opposite
arrangement. R's version is why R code looks like ordinary function calls
and never like `x.summary()`.

**So which languages are object-oriented?** Python thoroughly, in the
methods-inside-the-object style. R thoroughly, in the
methods-inside-the-function style, and it has four systems for it: S3, S4 and
Reference Classes in base R, plus R6 from a package. This course uses S3 and
never says so, because S3 asks nothing of you: an object carries a `class`
attribute, and a generic like `summary()` dispatches to `summary.factor()` or
`summary.data.frame()`.

Stata is the interesting case. The command language you type is not
object-oriented: `regress y x` is an instruction, not a message to an object.
But Stata's two *programming* languages both provide classes, `[P] class` for
ado and `[M-2] class` for Mata. The distinction is between what a user types
and what a package author writes, and it is a good reminder that
"object-oriented" describes how code is organised, not how good it is.

For the formal version, see the
[OOP chapter](https://adv-r.hadley.nz/oo.html) of *Advanced R*, linked from
the [Resources](../resources.qmd) page.

:::

::: {.content-hidden when-format="revealjs"}

## Class and type {#class-and-type}

Two words that sound alike and are not.

```{r}
#| label: class-and-type
#| script: "Class and type"
#| purl: false
class(as.Date("2025-01-07"))
typeof(as.Date("2025-01-07"))

class(factor("a"))
typeof(factor("a"))
```

**Class** is what an object announces, and what functions dispatch on.
**Type** is how R stores it underneath. Usually the two agree, and you can
ignore the difference. When they disagree you get the traps in this session.
A Date is a number with a class attached, counting days from 1970. A factor
is a set of integers with a class attached.

This page says class almost everywhere, because `class()` is the function you
will actually run. Type comes back once, at the factor trap, which is where
the difference costs you something.

:::

## R next to Stata and Python {#r-vs-others}

::: {.tight-cols}

| | Stata | R | Python |
|:--|:--|:--|:--|
| Built for | data analysis | data analysis | general programming |
| Data in memory | one active dataset; more via `frames` | many named objects | many named objects |
| Naming the data | implicit: `regress y x` | explicit: `lm(y ~ x, data = d)` | explicit |
| Where tables come from | the dataset *is* the workspace | objects you create and name | from libraries |
| Strongest at | a clean panel to a published table | statistical methods and graphics | pipelines, APIs, ML, production code |
| Who uses it | economists, social scientists | statisticians, many fields | programmers, CS, machine learning |

:::

Economics is a Stata field. That is measured, not folklore.

::: {.content-hidden when-format="revealjs"}

Stata works on one active dataset, and commands act on it without naming it:
`regress y x` needs no `data =` because there is only one thing it could
mean. Since Stata 16 the `frames` command holds several datasets at once,
but one of them is still the active one, and that shapes how Stata code
reads.

R has no ambient dataset. Every object has a name and every function says
which one it works on. The result of `lm()` is itself an object you can
store, print, and take apart.

That is what the last row means. In Stata the table is the workspace: a
variable is a column of the dataset, not a thing you can hold on its own. In
R a vector or a data frame is an object like any other. You make it, name
it, hand it to a function, and keep ten more beside it.

Grammar makes the difference concrete. A Stata command drops its subject
because the subject is understood. R names every noun, and lets several be
in play at once. That is why R needs `data = d` and Stata does not, and why R
can hold your raw file, your cleaned file, and three models at the same time.

Python is a general-purpose language that acquired data tools later. R was
built for data analysis from the start, so vectors, data frames, and formulas
like `y ~ x` belong to the language rather than to a library.

The last row of the table needs a word. `c()` and `data.frame()` work the
moment R starts, because vectors and tables are part of R itself. Python has
neither. You get them from **libraries**: collections of code somebody else
wrote, which you install once and load in each script. Two matter for data
work. **numpy** adds numeric arrays and fast arithmetic over them.
**pandas** adds a table type called a DataFrame, which is R's data frame
under a different name. A Python analyst loads both at the top of nearly
every script.

That is not a defect, it is what "general-purpose" means: Python keeps a
small core and you add what the job needs. R made the opposite trade, which
is why R feels quick for a table and a model and clumsy for building an
application. R has libraries too, called packages, and you install your first
one in session 4. The difference is what you get before you install anything.

:::

::: {.content-hidden when-format="revealjs"}

**What each is strongest at.** Stata gives you the shortest path from a clean
panel to a published table: one documented way to do most things, and
versioned, so old code keeps running. R has the widest library of statistical
methods and the strongest tools for turning a messy source into a publication
figure. Python is general programming: pipelines, APIs, scraping, machine
learning, and software other people run.

Lars Vilhuber's annual [Report of the AEA Data
Editor](https://www.aeaweb.org/articles?id=10.1257/pandp.114.878) (*AEA
Papers and Proceedings* 114, 2024, 878-90) tabulates the software found in
every replication package the AEA journals accept. Stata appears in the large
majority of them. MATLAB and R follow, each in a substantial minority, and
Python has gone from almost absent before 2016 to a meaningful share. The
trend worth noticing is that a large and rising fraction of packages use more
than one language: the question is becoming which tool for which step, not
which tool.

Across disciplines the picture is different. Robert Muenchen has tracked
scholarly use by software for more than a decade in [The Popularity of Data
Science Software](https://r4stats.com/articles/popularity/). R and Python
dominate the overall counts, while Stata's use is concentrated in economics
and parts of political science, sociology, and epidemiology. That
concentration is why Stata can look universal from inside economics and
marginal from outside it, and why a paper you want to reuse may well arrive
in a language your field does not default to.

None of this makes one of them correct. It does mean that the tool you know
determines whose code you can read, and that reading other people's code is
most of what reuse means.

:::

## Objects {#objects}

Everything in R is an object with a name you give it and a class R assigns.

```{r}
#| label: objects
#| script: "Objects and assignment"
x <- c(1, 2, 3, 4, 5)   # the arrow points from the value into the name
x = c(1, 2, 3, 4, 5)    # also assigns; you will meet it in other code
c(1, 2, 3, 4, 5) -> x   # legal, and points the other way

x
class(x)
length(x)
```

::: {.content-hidden when-format="revealjs"}

`<-` is the assignment arrow, and it is the convention in R. `=` assigns
too, and you will meet it in other people's code and in code an AI writes
for you. `->` assigns rightward. It is legal and rare.

Two reasons this course writes `<-`. The arrow shows which way the value
moves, and `=` does not. And `=` already means something else inside a
function call: in `seq(from = 1)`, the `=` sets an argument, it does not
create an object called `from`.

Do not confuse `=` with `==`. One assigns. The other tests two things for
equality and returns `TRUE` or `FALSE`. You meet it in the next section.

R prints nothing when you assign; that is normal. Type the object's name on
its own line to see it.

:::

## Your environment {#environment}

```{r}
#| label: environment
#| script: "Your environment"
ls()            # everything you have named, right now

5 + 3           # this prints, and is gone
ls()            # nothing new: it was never given a name
```

::: {.content-hidden when-format="revealjs"}

`ls()` lists what is in your environment. RStudio shows the same list in the
Environment pane, top right, and that pane is worth watching while you work:
if an object is not in it, no line of your script created it.

The second half matters more than it looks. A result you do not assign is
printed and discarded. `5 + 3` shows you `8` and leaves nothing behind, so
the next line cannot use it. This is the most common reason a script "worked"
in the console and then fails when you re-run it from the top: the object it
needs was never actually stored, only displayed.

`rm(x)` removes one object; `rm(list = ls())` empties the environment. Do
that before a final re-run, so you find out whether the script really builds
everything it uses.

:::

## Asking R for help {#help}

Every function ships with a help page. Both of these open it, from the
console:

```{r}
#| label: help
#| script: "Asking R for help"
#| eval: false
?seq             # the help page for seq()
help(seq)        # the same thing
??regression     # search the help system when you do not know the name
```

::: {.content-hidden when-format="revealjs"}

The page opens in RStudio's Help pane. Read the **Usage** block first: it
lists the arguments and their defaults. Then go to **Examples** at the bottom
and run one. The prose in between is written for people who already know the
function.

`?` needs the exact name. `??` searches titles and keywords, so reach for it
when you know what you want but not what it is called.

:::

## Vectors {#vectors}

A vector is an ordered sequence of values that are all the same class:
all numbers, or all text, or all `TRUE`/`FALSE`. It is the basic unit of
R: the single number `5` is a vector of length one.

```{r}
#| label: vectors
#| script: "Building vectors"
1:10
seq(0, 100, by = 25)
seq(0, 1, length.out = 5)
rep(c("a", "b"), times = 3)
rep(c("a", "b"), each = 3)
```

## Vector arithmetic {#vector-math}

Arithmetic applies to every element at once. You almost never write a loop for
this:

```{r}
#| label: vector-math
#| script: "Vector arithmetic"
x * 2
x + c(10, 20, 30, 40, 50)
x + c(0, 100)      # lengths do not match -- R recycles, and warns
```

## Summaries {#summaries}

One number out of a whole vector.

```{r}
#| label: summaries
#| script: "Summaries"
sum(x)
mean(x)
min(x)
max(x)
range(x)      # both ends at once
```

## Exercise 1 {#exercise-1}

::: {.panel-tabset}

### Exercise

```{r}
#| label: exercise-1
#| exercise: true
#| script: "Exercise 1"
#| eval: false
# One line, two commands: what is the mean of 0, 25, 50, 75, 100?
```

### Solution

```{r}
#| label: exercise-1-solution
#| purl: false
mean(seq(0, 100, by = 25))
```

:::

::: {.content-hidden when-format="revealjs"}

R reads from the inside out. `seq()` runs first and hands its answer to
`mean()`. Nesting is how you combine commands today, and it is worth getting
comfortable with: most real lines do two or three things at once. Session 4
gives you a second way to write the same chain, left to right.

:::

## Classes {#classes}

```{r}
#| label: types
#| script: "Types"
class(1)
class("a")
class(TRUE)
class(as.Date("2025-01-07"))
```

::: {.content-hidden when-format="revealjs"}

Four classes carry almost all of empirical work: **numeric** (numbers),
**character** (text), **logical** (`TRUE`/`FALSE`), and **Date**. A vector
holds exactly one of them. That constraint sets up the next section.

:::

## Silent conversion {#coercion}

A vector can only hold **one class**. So when you mix classes, R does not
stop or warn. It silently converts everything to a class that can hold them
all.

```{r}
#| label: coercion
#| script: "Silent conversion"
c(1, 2, 3)
c(1, 2, "three")          # one text value converts the whole vector
class(c(1, 2, "three"))
```

## TRUE is 1 {#logical-arithmetic}

```{r}
#| label: logical-arithmetic
#| script: "TRUE is 1"
sum(c(TRUE, FALSE, TRUE))   # TRUE becomes 1, FALSE becomes 0
mean(c(TRUE, FALSE, TRUE))  # so the mean of a logical is a proportion
```

::: {.content-hidden when-format="revealjs"}

This one is useful rather than dangerous, and you will use it all semester.
Counting how many rows meet a condition is a `sum()` over a logical vector.
The share of rows that meet it is a `mean()` over the same vector.

:::

## Text sorts like text {#coercion-sort}

```{r}
#| label: coercion-sort
#| script: "Text sorts like text"
sort(c(10, 9, 100))
sort(c("10", "9", "100"))   # character by character
```

::: {.content-visible when-format="revealjs"}
::: {.notes}
Ask before running the second line: "where does 9 go?" Let them be wrong out
loud. This is the single most important slide in the session.
:::
:::

::: {.content-hidden when-format="revealjs"}

Text sorts character by character, so `"100"` lands before `"9"` because
`"1"` comes before `"9"`. Nothing errors. A column of years, ZIP codes, or
account numbers read as text will sort into an order that looks almost
right.
This returns twice today: at the factor trap, and again when the months in
session 4 print February before January.

:::

## What it costs you {#coercion-matters}

The convenience has a cost. Here is what one stray text entry does to a
column of numbers:

```{r}
#| label: coercion-cost
#| script: "What silent conversion costs you"
readings <- c(12.4, 18.1, 9.7, "n/a", 22.3)
class(readings)
mean(readings)
```

::: {.content-hidden when-format="revealjs"}

`mean()` did not stop. It returned `NA` with a warning you will scroll past,
and if that value had gone into a table nobody would have seen a red flag.
This is the failure mode the whole course trains you to catch: **it ran, and
it was wrong.**

:::

Repair it, then check the repair:

```{r}
#| label: coercion-fix
#| script: "The repair"
readings <- as.numeric(readings)   # "n/a" cannot convert -- it becomes NA
readings
mean(readings, na.rm = TRUE)
```

## NA: not available {#na}

`NA` stands for **not available**. It is not zero, and it is not the text
`"NA"`. It marks a value R does not have, and R enforces that: any
computation touching an unknown returns an unknown.

```{r}
#| label: na
#| script: "Missing values"
temps <- c(21.0, NA, 19.4, 23.8)
mean(temps)
mean(temps, na.rm = TRUE)

is.na(temps)
sum(is.na(temps))
temps == NA        # == tests for equality; this is not how you test for NA
```

::: {.content-hidden when-format="revealjs"}

`==` is a test, not an assignment. It asks whether two things are equal and
returns `TRUE` or `FALSE` for each element.

Testing against `NA` returns `NA` rather than `FALSE`, and that is
consistent rather than broken. R does not know whether an unknown value
equals an unknown value, so it declines to guess. This is why `is.na()`
exists: it asks a different question, "is this value missing," which R can
always answer.

:::

## Indexing {#indexing}

Square brackets pull elements out by position.

```{r}
#| label: indexing
#| script: "Indexing"
x <- c(10, 20, 30, 40, 50)
x[1]
x[c(1, 3)]
x[-1]              # everything except the first
x[length(x)]       # the last, however long it is
```

## Logical subsetting {#logical-subsetting}

Get comfortable with this one: every filter you write for the rest of the
semester works this way.

```{r}
#| label: logical-subsetting
#| script: "Logical subsetting"
x > 25             # a TRUE/FALSE for every element
x[x > 25]          # keep the elements where it is TRUE
sum(x > 25)        # count them -- TRUE is 1
x[x > 25 & x < 45] # & is and, | is or
```

::: {.content-hidden when-format="revealjs"}

Read `x[x > 25]` aloud as "x, where x is greater than 25." The condition
inside the brackets produces a logical vector, and R keeps the positions that
are `TRUE`. Once you see it that way, `sum(x > 25)` is not a trick: it
counts the `TRUE`s.

:::

## Exercise 2 {#exercise-2}

::: {.panel-tabset}

### Exercise

```{r}
#| label: exercise-2
#| exercise: true
#| script: "Exercise 2"
#| eval: false
# w <- c(12, 45, 7, 33, 88, 21)
#
# One line each: how many values are above 20?
#                what is the mean of just those?
```

### Solution

```{r}
#| label: exercise-2-solution
#| purl: false
w <- c(12, 45, 7, 33, 88, 21)
sum(w > 20)
mean(w[w > 20])
```

:::

::: {.content-hidden when-format="revealjs"}

The second line nests too: the subset runs first, then `mean()` gets what
survives. This pair is the single most reused pattern in the course. Session
4 writes it as `filter()` and `summarize()`, over the same idea.

:::

## Data frames {#data-frames}

A data frame is a set of equal-length vectors standing side by side: columns
of possibly different classes, rows that line up. Every dataset in this course
is one.

```{r}
#| label: data-frames
#| script: "Data frames"
d <- data.frame(
  site = c("Compton", "Compton", "Reseda"),
  date = c("2025-01-01", "2025-01-02", "2025-01-01"),
  pm25 = c(53.2, 33.6, 47.0)
)
d
str(d)
```

## Rows and columns {#df-subsetting}

Subsetting takes two coordinates: `d[rows, columns]`. Leave one blank to
mean "all of them."

```{r}
#| label: data-frame-subsetting
#| script: "Getting at rows and columns"
d$pm25            # one column, by name
d[1, ]            # first row, all columns
d[, "pm25"]       # all rows, one column
d[d$pm25 > 40, ]  # the rows where a condition is TRUE
```

## Sorting by position {#order}

`sort()` returns values. `order()` returns the positions that would put a
vector in order.

```{r}
#| label: order
#| script: "Sorting by position"
v <- c(10, 50, 30)
order(v)              # positions, not values
v[order(v)]           # which is what sort() does
d[order(d$pm25), ]    # so this sorts the whole frame by one column
```

::: {.content-hidden when-format="revealjs"}

Positions are what you need when the values are one column of a table.
`sort()` puts a column in order but cannot bring the other columns with it.
`order()` hands you the row numbers instead, and the brackets do the rest.
`order(-x)` reverses it.

This is the third way to subset, after position and condition, and it is the
one you compute rather than type. It comes back at the end of today, when we
ask which days were worst, and again in session 4 as `arrange()`.

:::

## Factors {#factors}

A factor is how R stores a **categorical** column: the values live as
numeric codes, with a lookup table of labels called *levels*.

```{r}
#| label: factors
#| script: "Factors"
site <- factor(c("Compton", "Reseda", "Compton", "Compton"))
site
levels(site)
table(site)
```

::: {.content-hidden when-format="revealjs"}

You will rarely create a factor on purpose this early. You will run into
them: some functions return them, some files load as them, and from
session 4 on, grouping and counting use them. For now, know what one is,
and know the trap in the next section.

:::

## The factor trap {#factor-trap}

Remember `"10"` sorting before `"9"`? Factors do the same thing, and it
is harder to see.

```{r}
#| label: factor-trap
#| script: "The factor trap"
f <- factor(c("10", "9", "100"))   # numbers, stored as categories
as.numeric(f)                      # NOT the numbers -- the level codes
as.numeric(as.character(f))        # the repair: text first, then number
```

::: {.content-hidden when-format="revealjs"}

`as.numeric()` on a factor returns the internal codes (the position
of each value in the alphabetically sorted level table), not the
numbers the labels spell. You get no error, no warning, and plausible
output. The repair is always the same two steps: `as.character()`
first, then `as.numeric()`.

This is the one place today where class and type come apart. A factor's
*class* is `factor`; its *type*, the way R stores it, is integer.
`as.numeric()` reaches past the class down to the storage, which is why you
get codes. `as.character()` first asks the class for its labels, and then
there are real numbers to convert.

:::

## Matrices and lists {#matrices-lists}

You will meet two more containers later. Today you only need their
names and what each one is for.

```{r}
#| label: matrices-lists
#| script: "Matrices and lists"
m <- matrix(1:6, nrow = 2)   # a vector with dimensions
m
m[2, 3]                      # row 2, column 3

l <- list(site = "Compton", readings = c(53.2, 33.6), clean = TRUE)
str(l)                       # a container of anything
l$readings
```

::: {.content-hidden when-format="revealjs"}

A **matrix** is one vector with dimensions: every element has the same
type, and you address it as `m[row, column]`. Matrices come back in Part
II: a photo is a matrix of pixels, and a weather grid is a matrix of
temperatures (sessions 19 and 24).

A **list** holds anything, including other lists. A data frame is in
fact a list of column vectors, which is why `d$pm25` and `l$readings`
look alike. `lapply()` takes a list and returns one when you run a
function over every file in a folder; that is session 6.

Neither needs practice today. The sessions that use them reintroduce
them.

:::

## Where am I? {#working-directory}

R always has a working directory. Paths you write are relative to it, and
"file not found" is almost always a disagreement about where you are.

```{r}
#| label: working-directory
#| script: "Where am I"
#| eval: false
getwd()
list.files()
```

::: {.content-hidden when-format="revealjs"}

In HW1 the data ships inside your repository, so you will write
`read.csv("data/epa_airdata_pm25_060371103_2025.csv")`, a path relative to
`getwd()`. If that errors, run `list.files()` and `list.files("data")` before
you change anything: nine times in ten the file is fine and you are somewhere
else. Session 11 makes this problem go away permanently.

:::

## Read a real file {#read}

Today's file comes from EPA, the US Environmental Protection Agency: daily
PM2.5 from the **Compton** monitor in Los Angeles County, calendar year 2025,
exactly as EPA's "Download Daily Data" tool ships it.

Download [the file](data/epa_pm25_compton_2025.csv) into a `data/` folder
next to your script (the same layout HW1 uses), then:

```{r}
#| label: read
#| script: "Read the file"
pm <- read.csv("data/epa_pm25_compton_2025.csv")
```

::: {.content-hidden when-format="revealjs"}

```{r}
#| label: read-fetch
#| script: "If you do not have the file yet"
#| eval: false
# Uncomment and run once. Same file, fetched instead of clicked.
# dir.create("data", showWarnings = FALSE)
# download.file(paste0("https://arielortizbobea.github.io/aem6850/",
#                      "fall-2026/sessions/data/epa_pm25_compton_2025.csv"),
#               "data/epa_pm25_compton_2025.csv")
```

If `read.csv()` says it cannot open the connection, you are not in the folder
you think you are. Run `getwd()` and `list.files()` before changing anything.

:::

::: {.callout-important appearance="simple"}
**Write down what the data should show.** The Eaton and Palisades fires
started January 7, 2025. Compton sits about 25 km from the Eaton fire,
roughly downwind. Which calendar day between January 1 and February 28
should show the worst air at this monitor? Write the date down.
:::

## Arrival checks {#arrival-checks}

The three questions from session 1 apply again, now to a file with
twenty-two columns.

```{r}
#| label: arrival-checks
#| script: "Arrival checks"
dim(pm)
names(pm)
```

## What `str()` tells you {#str-check}

`Daily.Mean.PM2.5.Concentration` is correct and unbearable to type. Rename it
first, then look:

```{r}
#| label: str-check
#| script: "Rename, then str()"
names(pm)[names(pm) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"

str(pm[, 1:6])
summary(pm$pm25)
```

::: {.content-hidden when-format="revealjs"}

Note what `read.csv()` did to the column names. The file's header says
`Daily Mean PM2.5 Concentration`; R does not allow spaces in a name, so it
substituted periods.

:::

## The identifier trap {#identifiers}

Look at what `str()` said about `Site.ID`.

```{r}
#| label: identifiers
#| script: "The identifier trap"
pm$Site.ID[1]
class(pm$Site.ID)
pm$State.FIPS.Code[1]
pm$County.FIPS.Code[1]
```

::: {.content-hidden when-format="revealjs"}

The file says `"060371302"`, `"06"`, `"037"`. `read.csv()` read the digits as
numbers and dropped the leading zeros. The site is now `60371302` and
California is state `6`.

This is not cosmetic. Census tract GEOIDs, FIPS codes, and ZIP codes (three
US systems for identifying places), CUSIPs (security identifiers), and
account numbers are **labels that happen to be spelled with digits**. Strip a
leading zero and the label no longer matches anything you join it to. The
join will not error; it will silently match fewer rows. Session 5 shows what
that costs.

:::

## Protect them at read time {#identifiers-fix}

The fix goes at read time, not after. Once a leading zero is gone, you
cannot get it back.

```{r}
#| label: identifiers-fix
#| script: "Protect identifiers at read time"
pm <- read.csv("data/epa_pm25_compton_2025.csv",
               colClasses = c("Site.ID"          = "character",
                              "State.FIPS.Code"  = "character",
                              "County.FIPS.Code" = "character"))
names(pm)[names(pm) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"

pm$Site.ID[1]
pm$County.FIPS.Code[1]
class(pm$pm25)     # the concentration is still a number, as it should be
```

## One site, two instruments {#instruments}

One monitoring site does not mean one row per day.

```{r}
#| label: instruments
#| script: "One site, two instruments"
table(pm$POC, pm$AQS.Parameter.Code)
```

::: {.content-hidden when-format="revealjs"}

Compton runs two instruments. POC 1 is the regulatory FRM sampler reporting
parameter 88101. POC 3 is a continuous instrument reporting mostly 88502, a
non-regulatory series, plus seven stray rows under 88101. Filtering on
`POC == 1` alone leaves the wrong set; so does filtering on the parameter
alone. You need both conditions.

:::

## Dates {#dates}

R read the date as text, in the US order: month/day/year. One line converts
it. Do it immediately: text dates sort wrong, exactly like `"10"` and `"9"`.

```{r}
#| label: dates
#| script: "Dates, one line deep"
pm$Date[1]
class(pm$Date)

pm$date <- as.Date(pm$Date, format = "%m/%d/%Y")
class(pm$date)
range(pm$date)
```

::: {.content-hidden when-format="revealjs"}

`%m/%d/%Y` is a format string describing where the pieces are: two-digit
month, slash, two-digit day, slash, four-digit year. Get it wrong and you get
`NA`s rather than an error, so check `range()` afterwards. Session 22 deals
with time zones, daylight saving, and the other ways timestamps go wrong;
this one line covers most of what you need until then.

:::

## Subset to the window {#subset}

Now use everything at once: keep the rows in the window, from the right
instrument.

```{r}
#| label: subset
#| script: "Subset to the window"
win <- pm[pm$date <= as.Date("2025-02-28"), ]
nrow(win)                      # rows...
length(unique(win$date))       # ...for how many calendar days?

one <- win[win$POC == 1 & win$AQS.Parameter.Code == 88101, ]
nrow(one)
nrow(one) == length(unique(one$date))   # one row per day now?
```

::: {.content-hidden when-format="revealjs"}

118 rows for 59 days: every day was in there twice, once per instrument. Any
mean you computed before filtering would have been a blend of two different
measurement methods.

The last line is the check. January 1 to February 28 is 59 calendar days, we
have 59 rows, and no date appears twice, so this monitor has no gaps in the
window. The monitor you get in HW1 does have gaps. The same check returns a
different answer there; that is why you run it rather than assume.

:::

## Predict, then reconcile {#reconcile}

```{r}
#| label: reconcile
#| script: "Predict, then reconcile"
one$date[which.max(one$pm25)]
max(one$pm25)
median(one$pm25)

head(one[order(-one$pm25), c("date", "pm25")], 5)
```

::: {.content-visible when-format="revealjs"}

The worst day at this monitor is **January 1**, not January 7.

:::

::: {.content-hidden when-format="revealjs"}

The worst air of the winter at this monitor came on **January 1**, a week
before the fires started.

The fire is unmistakably in there: January 9, 8 and 10 are the next three
worst days, all far above the median of 15.3. But the single worst day belongs
to something else, and if your script had simply reported "the peak," you
would have written a sentence about wildfire smoke that was about New Year's
Eve fireworks.

:::

## Was it real? {#was-it-real}

A spike on one instrument on one day has three explanations: a real event, an
instrument problem, or a data-processing artifact. You cannot tell from this
file alone.

Two checks would settle it:

- **Do other monitors show it?** An instrument fault is local to one machine;
  a county-wide event is not. Five other Los Angeles monitors also peak on
  January 1.
- **Does it repeat?** Fireworks come back every year on the same date. Smoke
  does not.

::: {.content-hidden when-format="revealjs"}

Nearly every verification this semester has that shape: work out what the
number implies, then check that implication against a source that could
contradict it. Session 10 makes it systematic.

:::

## You try {#you-try}

Five minutes. Stubs are in the live script; `one` is the data frame you just
built.

```{r}
#| label: you-try
#| exercise: true
#| script: "You try (5 minutes)"
#| eval: false
# 1. How many days in the window were above 35 ug/m3 -- the EPA 24-hour
#    standard? (One line. TRUE is 1.)
#
# 2. Compare January's mean with February's. format(one$date, "%Y-%m") gives
#    you "2025-01" and "2025-02" to subset on.
#
# 3. The site's OTHER instrument is POC 3, parameter 88502. Pull its rows for
#    the window and compare its January mean with POC 1's. Two instruments,
#    one site, same air -- how close do they agree?
```

## Verify {#verify}

::: {.content-hidden when-format="revealjs"}

Session 1 installed one habit: write down what the data should show, run,
reconcile. This session adds the habit that goes in front of it:

:::

> **Run `str()` before you compute. Every file, every time.**

::: {.content-hidden when-format="revealjs"}

Today `str()` would have told you, before you did any arithmetic, that the
site identifier had become a number and the date had not become a date. At
that point each had a one-line repair. Three joins later, each would have
been expensive.

Run the four arrival checks in this order:

| Check | Question it answers |
|:--|:--|
| `dim()` | Did I get the number of rows I expected? |
| `str()` | Is every column the class it should be? |
| `head()` | Does row one look like a row of data, or like a stray header? |
| `summary()` | Is anything impossible, or off by a factor of a thousand? |

:::

::: {.content-visible when-format="revealjs"}

`dim()`: the right number of rows?<br>
`str()`: is every column the class it should be?<br>
`head()`: does row one look like data?<br>
`summary()`: is anything impossible, or off by 1000×?

:::

## Practice {#practice}

::: {.content-visible when-format="revealjs"}

Nineteen more on the session page, with solutions behind a click. One or
two lines each, in the order we covered them. Work through them before
Tuesday.

:::

::: {.content-hidden when-format="revealjs"}

Nineteen exercises, one or two lines each, in the order the session covered
them. They are in the live script too. Try each one before you open the
solutions.

```{r}
#| label: practice
#| exercise: true
#| script: "Practice: 19 exercises (solutions on the session page)"
#| eval: false
# Vectors and sequences
#  1. Build the whole numbers 1 to 20.
#  2. Build the even numbers from 2 to 20.
#  3. Repeat "yes" and "no", alternating, four times each.
#  4. How many odd numbers are there between 1 and 99? Two commands, one line.
#
# Classes and conversion
#  5. Predict first, then check: what class is c(1, 2, "3")?
#  6. x <- c("4.5", "2.1", "8.8") arrived as text. Get its mean.
#  7. Sort c("5", "10", "9") into true numeric order, in one line.
#  8. y <- c(3, NA, 7, NA, 12). How many values are missing?
#  9. Same y: what is its mean, ignoring the gaps?
#
# Subsetting
# 10. z <- c(10, 25, 3, 47, 18, 60). Keep only the values above 20.
# 11. Same z: how many values are above 20?
# 12. Same z: what is the mean of the values above 20?
# 13. Same z: drop the first and the last value, however long z is.
# 14. Same z: put it in order without using sort().
#
# Data frames and factors
# 15. Build a data frame: city = Ithaca, Buffalo; pop = 30, 275.
# 16. From it, print the row where pop is above 100.
# 17. f <- factor(c("2019", "2007", "2013")). Put the years in order,
#     as numbers.
# 18. Same f: what does as.numeric(f) give instead, and why?
#
# Dates
# 19. How many days are there between 2025-01-07 and 2025-02-28?
```

::: {.callout-tip collapse="true"}
## Solutions

```{r}
#| label: practice-solutions
#| purl: false
#| collapse: true
# 1
1:20
# 2
seq(2, 20, by = 2)
# 3
rep(c("yes", "no"), times = 4)
# 4
length(seq(1, 99, by = 2))
# 5  one text value converts the whole vector
class(c(1, 2, "3"))
# 6
ex <- c("4.5", "2.1", "8.8")
mean(as.numeric(ex))
# 7
sort(as.numeric(c("5", "10", "9")))
# 8
y <- c(3, NA, 7, NA, 12)
sum(is.na(y))
# 9
mean(y, na.rm = TRUE)
# 10
z <- c(10, 25, 3, 47, 18, 60)
z[z > 20]
# 11
sum(z > 20)
# 12
mean(z[z > 20])
# 13
z[-c(1, length(z))]
# 14
z[order(z)]
# 15
cities <- data.frame(city = c("Ithaca", "Buffalo"), pop = c(30, 275))
cities
# 16
cities[cities$pop > 100, ]
# 17
f <- factor(c("2019", "2007", "2013"))
sort(as.numeric(as.character(f)))
# 18  the level codes, not the years: levels sort alphabetically as
#     "2007" "2013" "2019", so 2019 is code 3, 2007 is 1, 2013 is 2
as.numeric(f)
# 19
as.Date("2025-02-28") - as.Date("2025-01-07")
```
:::

:::

## Quick reference {#quick-reference}

::: {.content-visible when-format="revealjs"}

The session page has every command from today in one table, plus the four
arrival checks and the format strings for dates.

:::

::: {.content-hidden when-format="revealjs"}

**Making and inspecting things**

| Code | What it does |
|:--|:--|
| `x <- c(1, 2, 3)` | Assign a vector |
| `1:10`, `seq(0, 100, by = 25)` | Regular sequences |
| `rep(x, times =)`, `rep(x, each =)` | Repetition, two different ways |
| `length(x)`, `class(x)` | How many, and what class |
| `sum(x)`, `mean(x)`, `min(x)`, `max(x)`, `range(x)` | One number out of a vector |
| `ls()` | What is in the workspace |
| `?mean` | The help page for a function |
| `factor(x)`, `levels(f)` | A categorical column and its categories |
| `matrix(1:6, nrow = 2)`, `m[2, 3]` | A vector with dimensions |
| `list(a = 1, b = "x")`, `l$a` | A container of anything |

**Types and missingness**

| Code | What it does |
|:--|:--|
| `as.numeric(x)`, `as.character(x)` | Convert on purpose |
| `is.na(x)`, `sum(is.na(x))` | Test for missing; count it |
| `mean(x, na.rm = TRUE)` | Compute over what is there |
| `as.Date(x, format = "%m/%d/%Y")` | Text to date; `%Y-%m-%d` for ISO |

**Subsetting**

| Code | What it does |
|:--|:--|
| `x[1]`, `x[c(1, 3)]`, `x[-1]` | By position; `-` drops |
| `order(x)`, `x[order(x)]` | The positions that sort it; `-x` reverses |
| `x[x > 25]` | "x, where x is greater than 25" |
| `sum(x > 25)`, `which(x > 25)` | Count them; find where they are |
| `d$col` | One column by name |
| `d[rows, cols]` | Blank means "all of them" |
| `d[d$col > 40, ]` | The rows where a condition holds |

**Reading files**

| Code | What it does |
|:--|:--|
| `getwd()`, `list.files()` | Where am I, what is here |
| `read.csv(path)` | Path or URL, either one |
| `read.csv(p, colClasses = c("Site.ID" = "character"))` | Keep the leading zeros on read |
| `names(d)[names(d) == "old"] <- "new"` | Rename one column |
| `dim()`, `str()`, `head()`, `summary()` | The arrival checks |

**Things that bite**

| Symptom | Cause |
|:--|:--|
| A numeric column is `chr` | One stray text value converted the whole thing |
| `"100"` sorts before `"9"` | It is text; text sorts character by character |
| An ID lost its leading zero | `read.csv()` read a label as a number |
| `mean()` returns `NA` | There are `NA`s; use `na.rm = TRUE` |
| `x == NA` gives `NA` | Use `is.na(x)` |
| `as.numeric()` on a factor gives codes | `as.character()` first, then `as.numeric()` |
| More rows than days | More than one instrument, or a bad join later |

**Links**

- Session materials: [slides](02-r-essentials-slides.html) ·
  [live script](02-r-essentials.R) ·
  [today's data](data/epa_pm25_compton_2025.csv) ·
  [where it came from](data/SOURCE.md)
- EPA AirData: [Download Daily Data](https://www.epa.gov/outdoor-air-quality-data/download-daily-data)
- [Syllabus](../syllabus.qmd) · [full schedule](../index.qmd#schedule)

:::

## After class {#after-class}

**HW1 is out: "How bad was the air?"** It is due Thursday, September 3, at
8:40 am. It uses the same tool on a different monitor: downtown Los Angeles,
December through February, five numbers in a `results.csv`.

::: {.content-hidden when-format="revealjs"}

You did everything it asks for at least once today. It ships with the data
already in the repository, a `results_spec.md` that states the format
exactly, and CHECK values for three of the answers so you can tell whether
you are on track without guessing. Hand it in the way you handed in your
README: edit or drag files onto the repository page, then commit.

**Re-run today's script before you start.** Do not just read it; run it,
line by line, and change things to see what breaks. The session went fast on
purpose; the script is where you slow down.

**Tuesday** is *Look at your data*: the one-line plots that catch the errors
`str()` and `summary()` miss. It uses this same file, so bring it.

:::

::: {.content-visible when-format="revealjs"}

Re-run today's script at home, line by line. Nineteen practice exercises,
with solutions, are on the session page. Tuesday: *Look at your data*.

:::

```{r}
#| purl: false
#| include: false
source("_purl.R")
purl_session("02-r-essentials.qmd", "02-r-essentials.R")
```
