---
title: "2 · R essentials"
subtitle: "Thursday, August 27, 2026"
pagetitle: "AEM 6850 · Session 2 · R essentials"
execute:
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
  revealjs:
    output-file: 02-r-essentials-slides.html
    theme: [default, slides.scss]
    footer: "AEM 6850 · Session 2"
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    code-overflow: scroll
    history: false
    link-external-newwindow: true
---

```{r}
#| purl: false
#| include: false
knitr::opts_chunk$set(
  comment = "#>",
  collapse = FALSE,
  fig.width = 8,
  fig.height = 4,
  fig.align = "center",
  dpi = 150
)
```

::: {.content-hidden when-format="revealjs"}

[Slides](02-r-essentials-slides.html) · [Live script](02-r-essentials.R) ·
[Source](https://github.com/ArielOrtizBobea/aem6850/blob/main/fall-2026/sessions/02-r-essentials.qmd)

The densest session of the semester. Everything below is in the live script —
run it line by line at home, as many times as it takes. Nothing here needs to
be memorized today.

:::

## Objectives {#objectives}

By the end of this session you can:

- Build and manipulate vectors.
- Predict what R does when types mix, and name what it costs you.
- Read a CSV and inspect what actually arrived.
- Subset rows and columns with logical conditions.
- Protect an identifier column from being turned into a number.
- Recognize the other containers — factors, matrices, lists — and say
  what each is for.

::: {.content-hidden when-format="revealjs"}

This is a lot for one session, and it is meant to be. You are not expected to
absorb it in the room — you are expected to *see* it once, then re-run the
script until it is yours.

## Before class {#before-class}

Have R and RStudio open and working, and bring the two National Weather
Service numbers from [session 1](01-overview.qmd#measurement).

If your install still is not working, sit next to someone whose is, and use
[posit.cloud](https://posit.cloud) today. Do not spend this session
troubleshooting — the material does not wait.

:::

## Where we left off {#recap}

Session 1 ended with a reconciliation: our file said the peak gust in downtown
LA was 66 km/h, about 41 mph, and the National Weather Service was describing
that same windstorm in much more alarming terms. Both were right, because they
were measuring different places.

> **A dataset is a measurement of the world, taken somewhere specific, by some
> specific method.**

Today we do that again, with a messier file and a bigger surprise.

::: {.content-visible when-format="revealjs"}
::: {.notes}
Collect the two NWS numbers out loud — 30 seconds, no more. If nobody did it,
say so plainly and move on; the point lands again at the end of today.
:::
:::

## Objects {#objects}

Everything in R is an object with a name you chose and a type you did not.

```{r}
#| label: objects
#| script: "Objects and assignment"
x <- c(1, 2, 3, 4, 5)
x
class(x)
length(x)

ls()          # what is in your workspace right now
```

::: {.content-hidden when-format="revealjs"}

`<-` is the assignment arrow. `=` also works and you will see it in other
people's code, but `<-` is the convention in R and it is what this course
uses. Nothing is printed when you assign — that is normal. Type the object's
name on its own line to see it.

:::

## Vectors {#vectors}

A vector is an ordered run of values of **one type**. It is the basic unit of
R — the single number `5` is a vector of length one.

```{r}
#| label: vectors
#| script: "Building vectors"
1:10
seq(0, 100, by = 25)
seq(0, 1, length.out = 5)
rep(c("a", "b"), times = 3)
rep(c("a", "b"), each = 3)
```

## Vector arithmetic {#vector-math}

Arithmetic applies to every element at once. You almost never write a loop for
this:

```{r}
#| label: vector-math
#| script: "Vector arithmetic"
x * 2
x + c(10, 20, 30, 40, 50)
x + c(0, 100)      # lengths do not match -- R recycles, and warns
```

## Types {#types}

```{r}
#| label: types
#| script: "Types"
class(1)
class("a")
class(TRUE)
class(as.Date("2025-01-07"))
```

::: {.content-hidden when-format="revealjs"}

Four types carry almost all of empirical work: **numeric** (numbers),
**character** (text), **logical** (`TRUE`/`FALSE`), and **Date**. A vector
holds exactly one of them. That constraint is where the next slide comes from.

:::

## Silent conversion {#coercion}

A vector can only hold one type. So when you mix types, R does not complain —
it quietly converts everything to a type that can hold them all.

```{r}
#| label: coercion
#| script: "Silent type conversion"
c(1, 2, 3)
c(1, 2, "three")          # one text value converts the whole vector
class(c(1, 2, "three"))

sum(c(TRUE, FALSE, TRUE))  # TRUE becomes 1, FALSE becomes 0
mean(c(TRUE, FALSE, TRUE)) # ... so the mean of a logical is a proportion

sort(c(10, 9, 100))
sort(c("10", "9", "100"))  # text sorts character by character
```

::: {.content-visible when-format="revealjs"}
::: {.notes}
Ask before running the last pair: "where does 9 go?" Let them be wrong out
loud. This is the single most important slide in the session.
:::
:::

## What it costs you {#coercion-matters}

That behaviour is convenient right up until the moment it is expensive. One
stray text entry in a column of numbers:

```{r}
#| label: coercion-cost
#| script: "What silent conversion costs you"
readings <- c(12.4, 18.1, 9.7, "n/a", 22.3)
class(readings)
mean(readings)
```

::: {.content-hidden when-format="revealjs"}

`mean()` did not stop. It returned `NA` with a warning you will scroll past,
and if that value had gone into a table nobody would have seen a red flag.
This is the failure mode this whole course is organized around: **it ran, and
it was wrong.**

:::

The repair, and the check:

```{r}
#| label: coercion-fix
#| script: "The repair"
readings <- as.numeric(readings)   # "n/a" cannot convert -- it becomes NA
readings
mean(readings, na.rm = TRUE)
```

## NA {#na}

`NA` is not zero and it is not text. It means *we do not know*, and R is
deliberately stubborn about it: anything computed from an unknown is unknown.

```{r}
#| label: na
#| script: "Missing values"
temps <- c(21.0, NA, 19.4, 23.8)
mean(temps)
mean(temps, na.rm = TRUE)

is.na(temps)
sum(is.na(temps))
temps == NA        # not how you test for NA -- use is.na()
```

## Indexing {#indexing}

Square brackets pull elements out by position.

```{r}
#| label: indexing
#| script: "Indexing"
x <- c(10, 20, 30, 40, 50)
x[1]
x[c(1, 3)]
x[-1]              # everything except the first
x[length(x)]       # the last, however long it is
```

## Logical subsetting {#logical-subsetting}

This is the one to get comfortable with, because every filter you write for
the rest of the semester is this.

```{r}
#| label: logical-subsetting
#| script: "Logical subsetting"
x > 25             # a TRUE/FALSE for every element
x[x > 25]          # keep the elements where it is TRUE
sum(x > 25)        # count them -- TRUE is 1
x[x > 25 & x < 45] # & is and, | is or
```

::: {.content-hidden when-format="revealjs"}

Read `x[x > 25]` aloud as "x, where x is greater than 25." The condition
inside the brackets produces a logical vector, and R keeps the positions that
are `TRUE`. Once you see it that way, `sum(x > 25)` counting things stops
being a trick and starts being obvious.

:::

## Data frames {#data-frames}

A data frame is a set of equal-length vectors standing side by side: columns
of possibly different types, rows that line up. It is what every dataset in
this course arrives as.

```{r}
#| label: data-frames
#| script: "Data frames"
d <- data.frame(
  site = c("Compton", "Compton", "Reseda"),
  date = c("2025-01-01", "2025-01-02", "2025-01-01"),
  pm25 = c(53.2, 33.6, 47.0)
)
d
str(d)
```

## Rows and columns {#df-subsetting}

Two coordinates: `d[rows, columns]`. Leave one blank to mean "all of them."

```{r}
#| label: data-frame-subsetting
#| script: "Getting at rows and columns"
d$pm25            # one column, by name
d[1, ]            # first row, all columns
d[, "pm25"]       # all rows, one column
d[d$pm25 > 40, ]  # the rows where a condition is TRUE
```

## Factors {#factors}

A factor is how R stores a **categorical** column: the values live as
numeric codes, with a lookup table of labels called *levels*.

```{r}
#| label: factors
#| script: "Factors"
site <- factor(c("Compton", "Reseda", "Compton", "Compton"))
site
levels(site)
table(site)
```

::: {.content-hidden when-format="revealjs"}

You rarely *make* factors on purpose this early; you *meet* them —
some functions return them, some files arrive as them, and from
session 4 on, grouping and counting lean on them. For now, know what
one is and know its one famous bite, on the next slide.

:::

## The factor trap {#factor-trap}

Remember `"10"` sorting before `"9"`? Factors have a sneakier version.

```{r}
#| label: factor-trap
#| script: "The factor trap"
f <- factor(c("10", "9", "100"))   # numbers, stored as categories
as.numeric(f)                      # NOT the numbers -- the level codes
as.numeric(as.character(f))        # the repair: text first, then number
```

::: {.content-hidden when-format="revealjs"}

`as.numeric()` on a factor returns the internal codes — the position
of each value in the (alphabetically sorted) level table — not the
numbers the labels spell. No error, no warning, plausible output:
silent conversion's meaner cousin. The repair is always the same two
steps: `as.character()` first, then `as.numeric()`.

:::

## Matrices and lists {#matrices-lists}

Two more containers you will meet later — today just their names and
their one-line reasons to exist.

```{r}
#| label: matrices-lists
#| script: "Matrices and lists, at introduction depth"
m <- matrix(1:6, nrow = 2)   # a vector with dimensions
m
m[2, 3]                      # row 2, column 3

l <- list(site = "Compton", readings = c(53.2, 33.6), clean = TRUE)
str(l)                       # a container of anything
l$readings
```

::: {.content-hidden when-format="revealjs"}

A **matrix** is one vector wearing dimensions — every element the same
type, addressed as `m[row, column]`. It returns in Part II: a photo is
a matrix of pixels, a weather grid is a matrix of temperatures
(sessions 19 and 24).

A **list** holds anything, including other lists — and a data frame is
secretly a list of column vectors, which is why `d$pm25` and `l$readings`
look alike. Lists are what `lapply()` eats and returns when you run a
function over every file in a folder; that's session 6.

Neither needs practice today. When they matter, they will be
reintroduced where they are used.

:::

## Where am I? {#working-directory}

R always has a working directory. Paths you write are relative to it, and
"file not found" is almost always a disagreement about where you are.

```{r}
#| label: working-directory
#| script: "Where am I"
#| eval: false
getwd()
list.files()
```

::: {.content-hidden when-format="revealjs"}

In HW1 the data ships inside your repository, so you will write
`read.csv("data/epa_airdata_pm25_060371103_2025.csv")` — a path relative to
`getwd()`. If that errors, run `list.files()` and `list.files("data")` before
you change anything: nine times in ten the file is fine and you are somewhere
else. Session 11 makes this problem go away permanently.

:::

## Read a real file {#read}

Today's file is a real EPA download: daily PM2.5 from the **Compton** monitor
in Los Angeles County, calendar year 2025, exactly as EPA's "Download Daily
Data" tool ships it.

Download it from [this session's data
folder](data/epa_pm25_compton_2025.csv) into a `data/` folder sitting next to
your script — the same shape HW1 arrives in — then:

```{r}
#| label: read
#| script: "Read the file"
pm <- read.csv("data/epa_pm25_compton_2025.csv")
```

::: {.content-hidden when-format="revealjs"}

```{r}
#| label: read-fetch
#| script: "If you do not have the file yet"
#| eval: false
# Uncomment and run once. Same file, fetched instead of clicked.
# dir.create("data", showWarnings = FALSE)
# download.file(paste0("https://arielortizbobea.github.io/aem6850/",
#                      "fall-2026/sessions/data/epa_pm25_compton_2025.csv"),
#               "data/epa_pm25_compton_2025.csv")
```

If `read.csv()` says it cannot open the connection, you are not in the folder
you think you are. Run `getwd()` and `list.files()` before changing anything.

:::

::: {.callout-important appearance="simple"}
**Predict before you look.** The Eaton and Palisades fires started January 7,
2025. Compton is 25 km downwind-ish of the Eaton fire. Write down: which
calendar day between January 1 and February 28 had the worst air at this
monitor?
:::

## Arrival checks {#arrival-checks}

Same three questions as session 1, now on a file with twenty-two columns.

```{r}
#| label: arrival-checks
#| script: "Arrival checks"
dim(pm)
names(pm)
```

## What `str()` tells you {#str-check}

`Daily.Mean.PM2.5.Concentration` is correct and unbearable to type. Rename it
first, then look:

```{r}
#| label: str-check
#| script: "Rename, then str()"
names(pm)[names(pm) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"

str(pm[, 1:6])
summary(pm$pm25)
```

::: {.content-hidden when-format="revealjs"}

Note what `read.csv()` did to the column names. The file's header says
`Daily Mean PM2.5 Concentration`; R does not allow spaces in a name, so it
substituted periods.

:::

## The identifier trap {#identifiers}

Look at what `str()` said about `Site.ID`.

```{r}
#| label: identifiers
#| script: "The identifier trap"
pm$Site.ID[1]
class(pm$Site.ID)
pm$State.FIPS.Code[1]
pm$County.FIPS.Code[1]
```

::: {.content-hidden when-format="revealjs"}

The file says `"060371302"`, `"06"`, `"037"`. R saw digits, decided they were
numbers, and ate the leading zeros. The site is now `60371302` and California
is state `6`.

This is not cosmetic. Census tract GEOIDs, FIPS codes, ZIP codes, CUSIPs and
account numbers are **labels that happen to be spelled with digits**. Strip a
leading zero and the label no longer matches anything you try to join it to —
and the join will not error, it will just silently match fewer rows. Session 5
is where that bill comes due.

:::

## Protect them at read time {#identifiers-fix}

The fix goes at read time, not after — once a leading zero is gone, it is gone.

```{r}
#| label: identifiers-fix
#| script: "Protect identifiers at read time"
pm <- read.csv("data/epa_pm25_compton_2025.csv",
               colClasses = c("Site.ID"          = "character",
                              "State.FIPS.Code"  = "character",
                              "County.FIPS.Code" = "character"))
names(pm)[names(pm) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"

pm$Site.ID[1]
pm$County.FIPS.Code[1]
class(pm$pm25)     # the concentration is still a number, as it should be
```

## One site, two instruments {#instruments}

One monitoring site does not mean one row per day.

```{r}
#| label: instruments
#| script: "One site, two instruments"
table(pm$POC, pm$AQS.Parameter.Code)
```

::: {.content-hidden when-format="revealjs"}

Compton runs two instruments. POC 1 is the regulatory FRM sampler reporting
parameter 88101. POC 3 is a continuous instrument reporting mostly 88502, a
non-regulatory series — plus seven stray rows under 88101. So filtering on
`POC == 1` alone, or on the parameter alone, both leave you with the wrong
set. You need both conditions.

:::

## Dates {#dates}

The date arrived as text in American order. One line converts it, and it is
worth doing immediately — text dates sort wrong, exactly like `"10"` and
`"9"`.

```{r}
#| label: dates
#| script: "Dates, one line deep"
pm$Date[1]
class(pm$Date)

pm$date <- as.Date(pm$Date, format = "%m/%d/%Y")
class(pm$date)
range(pm$date)
```

::: {.content-hidden when-format="revealjs"}

`%m/%d/%Y` is a format string describing where the pieces are: two-digit
month, slash, two-digit day, slash, four-digit year. Get it wrong and you get
`NA`s rather than an error, so check `range()` afterwards. Session 22 does
time zones, daylight saving, and the ways clocks lie; this one line covers
most of what you need until then.

:::

## Subset to the window {#subset}

Now everything at once: rows in the window, from the right instrument.

```{r}
#| label: subset
#| script: "Subset to the window"
win <- pm[pm$date <= as.Date("2025-02-28"), ]
nrow(win)                      # rows...
length(unique(win$date))       # ...for how many calendar days?

one <- win[win$POC == 1 & win$AQS.Parameter.Code == 88101, ]
nrow(one)
nrow(one) == length(unique(one$date))   # one row per day now?
```

::: {.content-hidden when-format="revealjs"}

118 rows for 59 days: every day was in there twice, once per instrument. Any
mean you computed before filtering would have been a blend of two different
measurement methods.

The last line is the check. January 1 to February 28 is 59 calendar days, we
have 59 rows, and no date appears twice — so this monitor has no gaps in the
window. The monitor you get in HW1 does have gaps. Same check, different
answer; that is the point of running it rather than assuming.

:::

## Predict, then reconcile {#reconcile}

```{r}
#| label: reconcile
#| script: "Predict, then reconcile"
one$date[which.max(one$pm25)]
max(one$pm25)
median(one$pm25)

head(one[order(-one$pm25), c("date", "pm25")], 5)
```

::: {.content-visible when-format="revealjs"}

The worst day at this monitor is **January 1** — not January 7.

:::

::: {.content-hidden when-format="revealjs"}

The worst air of the winter at this monitor was **January 1**, a week before
the fires started.

The fire is unmistakably in there — January 9, 8 and 10 are the next three
worst days, all far above the median of 15.3. But the single worst day belongs
to something else, and if your script had simply reported "the peak," you
would have written a sentence about wildfire smoke that was about New Year's
Eve fireworks.

:::

## Was it real? {#was-it-real}

A spike on one instrument on one day has three explanations: a real event, an
instrument problem, or a data-processing artifact. You cannot tell from this
file alone.

Two checks you could run, both of which you will have the tools for shortly:

- **Do other monitors see it?** An instrument fault is local to one machine; a
  county-wide event is not. Five other Los Angeles monitors also peak on
  January 1.
- **Does it repeat?** Fireworks come back every year on the same date. Smoke
  does not.

::: {.content-hidden when-format="revealjs"}

That is the shape of nearly every verification you will do this semester: find
something the number implies about the world, then go look at whether the
world agrees. Session 10 turns this into a strategy rather than a reflex.

:::

## You try {#you-try}

Five minutes. Stubs are in the live script; `one` is the data frame you just
built.

```{r}
#| label: you-try
#| exercise: true
#| script: "You try (5 minutes)"
#| eval: false
# 1. How many days in the window were above 35 ug/m3 -- the EPA 24-hour
#    standard? (One line. TRUE is 1.)
#
# 2. Compare January's mean with February's. format(one$date, "%Y-%m") gives
#    you "2025-01" and "2025-02" to subset on.
#
# 3. The site's OTHER instrument is POC 3, parameter 88502. Pull its rows for
#    the window and compare its January mean with POC 1's. Two instruments,
#    one site, same air -- how close do they agree?
```

## Verify {#verify}

::: {.content-hidden when-format="revealjs"}

Session 1 installed *predict, then run, then reconcile*. This session adds the
habit that goes in front of it:

:::

> **Run `str()` before you compute. Every file, every time.**

::: {.content-hidden when-format="revealjs"}

Today `str()` would have told you, before you did any arithmetic, that the
site identifier had become a number and the date had not become a date. Both
were repairable in one line at that point. Both would have been expensive
three joins later.

The four arrival checks, in the order worth running them:

| Check | Question it answers |
|:--|:--|
| `dim()` | Did I get the number of rows I expected? |
| `str()` | Is every column the type it should be? |
| `head()` | Does row one look like a row of data, or like a stray header? |
| `summary()` | Is anything impossible, or off by a factor of a thousand? |

:::

::: {.content-visible when-format="revealjs"}

`dim()` — the right number of rows?<br>
`str()` — is every column the type it should be?<br>
`head()` — does row one look like data?<br>
`summary()` — is anything impossible, or off by 1000×?

:::

## Quick reference {#quick-reference}

::: {.content-visible when-format="revealjs"}

The session page has every command from today in one table, plus the four
arrival checks and the format strings for dates.

:::

::: {.content-hidden when-format="revealjs"}

**Making and inspecting things**

| Code | What it does |
|:--|:--|
| `x <- c(1, 2, 3)` | Assign a vector |
| `1:10`, `seq(0, 100, by = 25)` | Regular sequences |
| `rep(x, times =)`, `rep(x, each =)` | Repetition, two different ways |
| `length(x)`, `class(x)` | How many, and what type |
| `ls()` | What is in the workspace |
| `?mean` | The help page for a function |
| `factor(x)`, `levels(f)` | A categorical column and its categories |
| `matrix(1:6, nrow = 2)`, `m[2, 3]` | A vector with dimensions |
| `list(a = 1, b = "x")`, `l$a` | A container of anything |

**Types and missingness**

| Code | What it does |
|:--|:--|
| `as.numeric(x)`, `as.character(x)` | Convert on purpose |
| `is.na(x)`, `sum(is.na(x))` | Test for missing; count it |
| `mean(x, na.rm = TRUE)` | Compute over what is there |
| `as.Date(x, format = "%m/%d/%Y")` | Text to date; `%Y-%m-%d` for ISO |

**Subsetting**

| Code | What it does |
|:--|:--|
| `x[1]`, `x[c(1, 3)]`, `x[-1]` | By position; `-` drops |
| `x[x > 25]` | "x, where x is greater than 25" |
| `sum(x > 25)`, `which(x > 25)` | Count them; find where they are |
| `d$col` | One column by name |
| `d[rows, cols]` | Blank means "all of them" |
| `d[d$col > 40, ]` | The rows where a condition holds |

**Reading files**

| Code | What it does |
|:--|:--|
| `getwd()`, `list.files()` | Where am I, what is here |
| `read.csv(path)` | Path or URL, either one |
| `read.csv(p, colClasses = c("Site.ID" = "character"))` | Stop R eating leading zeros |
| `names(d)[names(d) == "old"] <- "new"` | Rename one column |
| `dim()`, `str()`, `head()`, `summary()` | The arrival checks |

**Things that bite**

| Symptom | Cause |
|:--|:--|
| A numeric column is `chr` | One stray text value converted the whole thing |
| `"100"` sorts before `"9"` | It is text; text sorts character by character |
| An ID lost its leading zero | `read.csv()` decided a label was a number |
| `mean()` returns `NA` | There are `NA`s; use `na.rm = TRUE` |
| `x == NA` gives `NA` | Use `is.na(x)` |
| `as.numeric()` on a factor gives codes | `as.character()` first, then `as.numeric()` |
| More rows than days | More than one instrument, or a bad join later |

**Links**

- Session materials: [slides](02-r-essentials-slides.html) ·
  [live script](02-r-essentials.R) ·
  [today's data](data/epa_pm25_compton_2025.csv) ·
  [where it came from](data/SOURCE.md)
- EPA AirData: [Download Daily Data](https://www.epa.gov/outdoor-air-quality-data/download-daily-data)
- [Syllabus](../syllabus.qmd) · [full schedule](../index.qmd#schedule)

:::

## After class {#after-class}

**HW1 is out — "How bad was the air?"** Due Thursday, September 3, 8:40 am.
Same tool, different monitor: downtown Los Angeles, December through February,
five numbers in a `results.csv`.

::: {.content-hidden when-format="revealjs"}

Everything it asks for, you did today at least once. It ships with the data
already in the repository, a `results_spec.md` that states the format exactly,
and CHECK values for three of the six numbers so you can tell whether you are
on track without guessing. Hand it in the way you handed in your README: edit
or drag files onto the repository page, click Commit.

**Re-run today's script before you start.** Not read it — run it, line by
line, and change things to see what breaks. This session went fast on purpose;
the script is where it goes slow.

**Tuesday** is *Look at your data*: the one-line plots that catch the errors
`str()` and `summary()` miss. It uses this same file, so bring it.

:::

::: {.content-visible when-format="revealjs"}

Re-run today's script at home, line by line. Tuesday: *Look at your data*.

:::

```{r}
#| purl: false
#| include: false
source("_purl.R")
purl_session("02-r-essentials.qmd", "02-r-essentials.R")
```
