---
title: "2 · R essentials"
subtitle: "Thursday, August 27, 2026"
pagetitle: "AEM 6850 · Session 2 · R essentials"
execute:
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
  revealjs:
    output-file: 02-r-essentials-slides.html
    theme: [default, slides.scss]
    template-partials:
      - title-slide.html
    footer: "AEM 6850 · Session 2"
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    code-overflow: scroll
    history: false
    link-external-newwindow: true
---

```{r}
#| purl: false
#| include: false
knitr::opts_chunk$set(
  comment = "#>",
  collapse = FALSE,
  fig.width = 8,
  fig.height = 4,
  fig.align = "center",
  dpi = 150
)
```

::: {.content-hidden when-format="revealjs"}

[Slides](02-r-essentials-slides.html) · [Live script](02-r-essentials.R) ·
[Source](https://github.com/ArielOrtizBobea/aem6850/blob/main/fall-2026/sessions/02-r-essentials.qmd)

This is the densest session of the semester. Everything below is in the live
script. Run it line by line at home, as many times as it takes. You do not
need to memorize anything today.

:::

## Objectives {#objectives}

By the end of this session you can:

- Build vectors and do arithmetic on them.
- Say what R does when a vector mixes types, and why that turns a column
  of numbers into text.
- Read a CSV and check its size, its column types, and its ranges.
- Subset rows and columns with logical conditions.
- Protect an identifier column so R does not turn it into a number.
- Recognize factors, matrices, and lists, and say what each is for.

::: {.content-hidden when-format="revealjs"}

This is a lot for one session, and that is deliberate. Nobody absorbs it all
in the room. See it once today, then re-run the script until it sticks.

## Before class {#before-class}

Have R and RStudio open and working, and bring the two National Weather
Service numbers from [session 1](01-overview.qmd#measurement).

If your install still does not work, sit next to someone whose install does,
and use [posit.cloud](https://posit.cloud) today. Do not spend this session
troubleshooting. Bring the machine to office hours instead.

:::

## Objects {#objects}

Everything in R is an object with a name you give it and a type R assigns.

```{r}
#| label: objects
#| script: "Objects and assignment"
x <- c(1, 2, 3, 4, 5)
x
class(x)
length(x)

ls()          # what is in your workspace right now
```

::: {.content-hidden when-format="revealjs"}

`<-` is the assignment arrow. `=` also works and you will see it in other
people's code, but `<-` is the convention in R and it is what this course
uses. R prints nothing when you assign; that is normal. Type the object's
name on its own line to see it.

:::

## Vectors {#vectors}

A vector is an ordered sequence of values of **one type**. It is the basic unit of
R: the single number `5` is a vector of length one.

```{r}
#| label: vectors
#| script: "Building vectors"
1:10
seq(0, 100, by = 25)
seq(0, 1, length.out = 5)
rep(c("a", "b"), times = 3)
rep(c("a", "b"), each = 3)
```

## Vector arithmetic {#vector-math}

Arithmetic applies to every element at once. You almost never write a loop for
this:

```{r}
#| label: vector-math
#| script: "Vector arithmetic"
x * 2
x + c(10, 20, 30, 40, 50)
x + c(0, 100)      # lengths do not match -- R recycles, and warns
```

## Types {#types}

```{r}
#| label: types
#| script: "Types"
class(1)
class("a")
class(TRUE)
class(as.Date("2025-01-07"))
```

::: {.content-hidden when-format="revealjs"}

Four types carry almost all of empirical work: **numeric** (numbers),
**character** (text), **logical** (`TRUE`/`FALSE`), and **Date**. A vector
holds exactly one of them. That constraint sets up the next section.

:::

## Silent conversion {#coercion}

A vector can only hold one type. So when you mix types, R does not stop or
warn. It silently converts everything to a type that can hold them all.

```{r}
#| label: coercion
#| script: "Silent type conversion"
c(1, 2, 3)
c(1, 2, "three")          # one text value converts the whole vector
class(c(1, 2, "three"))

sum(c(TRUE, FALSE, TRUE))  # TRUE becomes 1, FALSE becomes 0
mean(c(TRUE, FALSE, TRUE)) # ... so the mean of a logical is a proportion

sort(c(10, 9, 100))
sort(c("10", "9", "100"))  # text sorts character by character
```

::: {.content-visible when-format="revealjs"}
::: {.notes}
Ask before running the last pair: "where does 9 go?" Let them be wrong out
loud. This is the single most important slide in the session.
:::
:::

## What it costs you {#coercion-matters}

The convenience has a cost. Here is what one stray text entry does to a
column of numbers:

```{r}
#| label: coercion-cost
#| script: "What silent conversion costs you"
readings <- c(12.4, 18.1, 9.7, "n/a", 22.3)
class(readings)
mean(readings)
```

::: {.content-hidden when-format="revealjs"}

`mean()` did not stop. It returned `NA` with a warning you will scroll past,
and if that value had gone into a table nobody would have seen a red flag.
This is the failure mode the whole course trains you to catch: **it ran, and
it was wrong.**

:::

Repair it, then check the repair:

```{r}
#| label: coercion-fix
#| script: "The repair"
readings <- as.numeric(readings)   # "n/a" cannot convert -- it becomes NA
readings
mean(readings, na.rm = TRUE)
```

## NA {#na}

`NA` is not zero and it is not text. It means *we do not know*, and R
enforces that: any computation that touches an unknown returns an unknown.

```{r}
#| label: na
#| script: "Missing values"
temps <- c(21.0, NA, 19.4, 23.8)
mean(temps)
mean(temps, na.rm = TRUE)

is.na(temps)
sum(is.na(temps))
temps == NA        # not how you test for NA -- use is.na()
```

## Indexing {#indexing}

Square brackets pull elements out by position.

```{r}
#| label: indexing
#| script: "Indexing"
x <- c(10, 20, 30, 40, 50)
x[1]
x[c(1, 3)]
x[-1]              # everything except the first
x[length(x)]       # the last, however long it is
```

## Logical subsetting {#logical-subsetting}

Get comfortable with this one: every filter you write for the rest of the
semester works this way.

```{r}
#| label: logical-subsetting
#| script: "Logical subsetting"
x > 25             # a TRUE/FALSE for every element
x[x > 25]          # keep the elements where it is TRUE
sum(x > 25)        # count them -- TRUE is 1
x[x > 25 & x < 45] # & is and, | is or
```

::: {.content-hidden when-format="revealjs"}

Read `x[x > 25]` aloud as "x, where x is greater than 25." The condition
inside the brackets produces a logical vector, and R keeps the positions that
are `TRUE`. Once you see it that way, `sum(x > 25)` is not a trick: it
counts the `TRUE`s.

:::

## Data frames {#data-frames}

A data frame is a set of equal-length vectors standing side by side: columns
of possibly different types, rows that line up. Every dataset in this course
is one.

```{r}
#| label: data-frames
#| script: "Data frames"
d <- data.frame(
  site = c("Compton", "Compton", "Reseda"),
  date = c("2025-01-01", "2025-01-02", "2025-01-01"),
  pm25 = c(53.2, 33.6, 47.0)
)
d
str(d)
```

## Rows and columns {#df-subsetting}

Subsetting takes two coordinates: `d[rows, columns]`. Leave one blank to
mean "all of them."

```{r}
#| label: data-frame-subsetting
#| script: "Getting at rows and columns"
d$pm25            # one column, by name
d[1, ]            # first row, all columns
d[, "pm25"]       # all rows, one column
d[d$pm25 > 40, ]  # the rows where a condition is TRUE
```

## Factors {#factors}

A factor is how R stores a **categorical** column: the values live as
numeric codes, with a lookup table of labels called *levels*.

```{r}
#| label: factors
#| script: "Factors"
site <- factor(c("Compton", "Reseda", "Compton", "Compton"))
site
levels(site)
table(site)
```

::: {.content-hidden when-format="revealjs"}

You will rarely create a factor on purpose this early. You will run into
them: some functions return them, some files load as them, and from
session 4 on, grouping and counting use them. For now, know what one is,
and know the trap in the next section.

:::

## The factor trap {#factor-trap}

Remember `"10"` sorting before `"9"`? Factors do the same thing, and it
is harder to see.

```{r}
#| label: factor-trap
#| script: "The factor trap"
f <- factor(c("10", "9", "100"))   # numbers, stored as categories
as.numeric(f)                      # NOT the numbers -- the level codes
as.numeric(as.character(f))        # the repair: text first, then number
```

::: {.content-hidden when-format="revealjs"}

`as.numeric()` on a factor returns the internal codes (the position
of each value in the alphabetically sorted level table), not the
numbers the labels spell. You get no error, no warning, and plausible
output. The repair is always the same two steps: `as.character()`
first, then `as.numeric()`.

:::

## Matrices and lists {#matrices-lists}

You will meet two more containers later. Today you only need their
names and what each one is for.

```{r}
#| label: matrices-lists
#| script: "Matrices and lists"
m <- matrix(1:6, nrow = 2)   # a vector with dimensions
m
m[2, 3]                      # row 2, column 3

l <- list(site = "Compton", readings = c(53.2, 33.6), clean = TRUE)
str(l)                       # a container of anything
l$readings
```

::: {.content-hidden when-format="revealjs"}

A **matrix** is one vector with dimensions: every element has the same
type, and you address it as `m[row, column]`. Matrices come back in Part
II: a photo is a matrix of pixels, and a weather grid is a matrix of
temperatures (sessions 19 and 24).

A **list** holds anything, including other lists. A data frame is in
fact a list of column vectors, which is why `d$pm25` and `l$readings`
look alike. `lapply()` takes a list and returns one when you run a
function over every file in a folder; that is session 6.

Neither needs practice today. The sessions that use them reintroduce
them.

:::

## Where am I? {#working-directory}

R always has a working directory. Paths you write are relative to it, and
"file not found" is almost always a disagreement about where you are.

```{r}
#| label: working-directory
#| script: "Where am I"
#| eval: false
getwd()
list.files()
```

::: {.content-hidden when-format="revealjs"}

In HW1 the data ships inside your repository, so you will write
`read.csv("data/epa_airdata_pm25_060371103_2025.csv")`, a path relative to
`getwd()`. If that errors, run `list.files()` and `list.files("data")` before
you change anything: nine times in ten the file is fine and you are somewhere
else. Session 11 makes this problem go away permanently.

:::

## Read a real file {#read}

Today's file comes from EPA, the US Environmental Protection Agency: daily
PM2.5 from the **Compton** monitor in Los Angeles County, calendar year 2025,
exactly as EPA's "Download Daily Data" tool ships it.

Download [the file](data/epa_pm25_compton_2025.csv) into a `data/` folder
next to your script (the same layout HW1 uses), then:

```{r}
#| label: read
#| script: "Read the file"
pm <- read.csv("data/epa_pm25_compton_2025.csv")
```

::: {.content-hidden when-format="revealjs"}

```{r}
#| label: read-fetch
#| script: "If you do not have the file yet"
#| eval: false
# Uncomment and run once. Same file, fetched instead of clicked.
# dir.create("data", showWarnings = FALSE)
# download.file(paste0("https://arielortizbobea.github.io/aem6850/",
#                      "fall-2026/sessions/data/epa_pm25_compton_2025.csv"),
#               "data/epa_pm25_compton_2025.csv")
```

If `read.csv()` says it cannot open the connection, you are not in the folder
you think you are. Run `getwd()` and `list.files()` before changing anything.

:::

::: {.callout-important appearance="simple"}
**Write down what the data should show.** The Eaton and Palisades fires
started January 7, 2025. Compton sits about 25 km from the Eaton fire,
roughly downwind. Which calendar day between January 1 and February 28
should show the worst air at this monitor? Write the date down.
:::

## Arrival checks {#arrival-checks}

The three questions from session 1 apply again, now to a file with
twenty-two columns.

```{r}
#| label: arrival-checks
#| script: "Arrival checks"
dim(pm)
names(pm)
```

## What `str()` tells you {#str-check}

`Daily.Mean.PM2.5.Concentration` is correct and unbearable to type. Rename it
first, then look:

```{r}
#| label: str-check
#| script: "Rename, then str()"
names(pm)[names(pm) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"

str(pm[, 1:6])
summary(pm$pm25)
```

::: {.content-hidden when-format="revealjs"}

Note what `read.csv()` did to the column names. The file's header says
`Daily Mean PM2.5 Concentration`; R does not allow spaces in a name, so it
substituted periods.

:::

## The identifier trap {#identifiers}

Look at what `str()` said about `Site.ID`.

```{r}
#| label: identifiers
#| script: "The identifier trap"
pm$Site.ID[1]
class(pm$Site.ID)
pm$State.FIPS.Code[1]
pm$County.FIPS.Code[1]
```

::: {.content-hidden when-format="revealjs"}

The file says `"060371302"`, `"06"`, `"037"`. `read.csv()` read the digits as
numbers and dropped the leading zeros. The site is now `60371302` and
California is state `6`.

This is not cosmetic. Census tract GEOIDs, FIPS codes, and ZIP codes (three
US systems for identifying places), CUSIPs (security identifiers), and
account numbers are **labels that happen to be spelled with digits**. Strip a
leading zero and the label no longer matches anything you join it to. The
join will not error; it will silently match fewer rows. Session 5 shows what
that costs.

:::

## Protect them at read time {#identifiers-fix}

The fix goes at read time, not after. Once a leading zero is gone, you
cannot get it back.

```{r}
#| label: identifiers-fix
#| script: "Protect identifiers at read time"
pm <- read.csv("data/epa_pm25_compton_2025.csv",
               colClasses = c("Site.ID"          = "character",
                              "State.FIPS.Code"  = "character",
                              "County.FIPS.Code" = "character"))
names(pm)[names(pm) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"

pm$Site.ID[1]
pm$County.FIPS.Code[1]
class(pm$pm25)     # the concentration is still a number, as it should be
```

## One site, two instruments {#instruments}

One monitoring site does not mean one row per day.

```{r}
#| label: instruments
#| script: "One site, two instruments"
table(pm$POC, pm$AQS.Parameter.Code)
```

::: {.content-hidden when-format="revealjs"}

Compton runs two instruments. POC 1 is the regulatory FRM sampler reporting
parameter 88101. POC 3 is a continuous instrument reporting mostly 88502, a
non-regulatory series, plus seven stray rows under 88101. Filtering on
`POC == 1` alone leaves the wrong set; so does filtering on the parameter
alone. You need both conditions.

:::

## Dates {#dates}

R read the date as text, in the US order: month/day/year. One line converts
it. Do it immediately: text dates sort wrong, exactly like `"10"` and `"9"`.

```{r}
#| label: dates
#| script: "Dates, one line deep"
pm$Date[1]
class(pm$Date)

pm$date <- as.Date(pm$Date, format = "%m/%d/%Y")
class(pm$date)
range(pm$date)
```

::: {.content-hidden when-format="revealjs"}

`%m/%d/%Y` is a format string describing where the pieces are: two-digit
month, slash, two-digit day, slash, four-digit year. Get it wrong and you get
`NA`s rather than an error, so check `range()` afterwards. Session 22 deals
with time zones, daylight saving, and the other ways timestamps go wrong;
this one line covers most of what you need until then.

:::

## Subset to the window {#subset}

Now use everything at once: keep the rows in the window, from the right
instrument.

```{r}
#| label: subset
#| script: "Subset to the window"
win <- pm[pm$date <= as.Date("2025-02-28"), ]
nrow(win)                      # rows...
length(unique(win$date))       # ...for how many calendar days?

one <- win[win$POC == 1 & win$AQS.Parameter.Code == 88101, ]
nrow(one)
nrow(one) == length(unique(one$date))   # one row per day now?
```

::: {.content-hidden when-format="revealjs"}

118 rows for 59 days: every day was in there twice, once per instrument. Any
mean you computed before filtering would have been a blend of two different
measurement methods.

The last line is the check. January 1 to February 28 is 59 calendar days, we
have 59 rows, and no date appears twice, so this monitor has no gaps in the
window. The monitor you get in HW1 does have gaps. The same check returns a
different answer there; that is why you run it rather than assume.

:::

## Predict, then reconcile {#reconcile}

```{r}
#| label: reconcile
#| script: "Predict, then reconcile"
one$date[which.max(one$pm25)]
max(one$pm25)
median(one$pm25)

head(one[order(-one$pm25), c("date", "pm25")], 5)
```

::: {.content-visible when-format="revealjs"}

The worst day at this monitor is **January 1**, not January 7.

:::

::: {.content-hidden when-format="revealjs"}

The worst air of the winter at this monitor came on **January 1**, a week
before the fires started.

The fire is unmistakably in there: January 9, 8 and 10 are the next three
worst days, all far above the median of 15.3. But the single worst day belongs
to something else, and if your script had simply reported "the peak," you
would have written a sentence about wildfire smoke that was about New Year's
Eve fireworks.

:::

## Was it real? {#was-it-real}

A spike on one instrument on one day has three explanations: a real event, an
instrument problem, or a data-processing artifact. You cannot tell from this
file alone.

Two checks would settle it:

- **Do other monitors show it?** An instrument fault is local to one machine;
  a county-wide event is not. Five other Los Angeles monitors also peak on
  January 1.
- **Does it repeat?** Fireworks come back every year on the same date. Smoke
  does not.

::: {.content-hidden when-format="revealjs"}

Nearly every verification this semester has that shape: work out what the
number implies, then check that implication against a source that could
contradict it. Session 10 makes it systematic.

:::

## You try {#you-try}

Five minutes. Stubs are in the live script; `one` is the data frame you just
built.

```{r}
#| label: you-try
#| exercise: true
#| script: "You try (5 minutes)"
#| eval: false
# 1. How many days in the window were above 35 ug/m3 -- the EPA 24-hour
#    standard? (One line. TRUE is 1.)
#
# 2. Compare January's mean with February's. format(one$date, "%Y-%m") gives
#    you "2025-01" and "2025-02" to subset on.
#
# 3. The site's OTHER instrument is POC 3, parameter 88502. Pull its rows for
#    the window and compare its January mean with POC 1's. Two instruments,
#    one site, same air -- how close do they agree?
```

## Verify {#verify}

::: {.content-hidden when-format="revealjs"}

Session 1 installed one habit: write down what the data should show, run,
reconcile. This session adds the habit that goes in front of it:

:::

> **Run `str()` before you compute. Every file, every time.**

::: {.content-hidden when-format="revealjs"}

Today `str()` would have told you, before you did any arithmetic, that the
site identifier had become a number and the date had not become a date. At
that point each had a one-line repair. Three joins later, each would have
been expensive.

Run the four arrival checks in this order:

| Check | Question it answers |
|:--|:--|
| `dim()` | Did I get the number of rows I expected? |
| `str()` | Is every column the type it should be? |
| `head()` | Does row one look like a row of data, or like a stray header? |
| `summary()` | Is anything impossible, or off by a factor of a thousand? |

:::

::: {.content-visible when-format="revealjs"}

`dim()`: the right number of rows?<br>
`str()`: is every column the type it should be?<br>
`head()`: does row one look like data?<br>
`summary()`: is anything impossible, or off by 1000×?

:::

## Quick reference {#quick-reference}

::: {.content-visible when-format="revealjs"}

The session page has every command from today in one table, plus the four
arrival checks and the format strings for dates.

:::

::: {.content-hidden when-format="revealjs"}

**Making and inspecting things**

| Code | What it does |
|:--|:--|
| `x <- c(1, 2, 3)` | Assign a vector |
| `1:10`, `seq(0, 100, by = 25)` | Regular sequences |
| `rep(x, times =)`, `rep(x, each =)` | Repetition, two different ways |
| `length(x)`, `class(x)` | How many, and what type |
| `ls()` | What is in the workspace |
| `?mean` | The help page for a function |
| `factor(x)`, `levels(f)` | A categorical column and its categories |
| `matrix(1:6, nrow = 2)`, `m[2, 3]` | A vector with dimensions |
| `list(a = 1, b = "x")`, `l$a` | A container of anything |

**Types and missingness**

| Code | What it does |
|:--|:--|
| `as.numeric(x)`, `as.character(x)` | Convert on purpose |
| `is.na(x)`, `sum(is.na(x))` | Test for missing; count it |
| `mean(x, na.rm = TRUE)` | Compute over what is there |
| `as.Date(x, format = "%m/%d/%Y")` | Text to date; `%Y-%m-%d` for ISO |

**Subsetting**

| Code | What it does |
|:--|:--|
| `x[1]`, `x[c(1, 3)]`, `x[-1]` | By position; `-` drops |
| `x[x > 25]` | "x, where x is greater than 25" |
| `sum(x > 25)`, `which(x > 25)` | Count them; find where they are |
| `d$col` | One column by name |
| `d[rows, cols]` | Blank means "all of them" |
| `d[d$col > 40, ]` | The rows where a condition holds |

**Reading files**

| Code | What it does |
|:--|:--|
| `getwd()`, `list.files()` | Where am I, what is here |
| `read.csv(path)` | Path or URL, either one |
| `read.csv(p, colClasses = c("Site.ID" = "character"))` | Keep the leading zeros on read |
| `names(d)[names(d) == "old"] <- "new"` | Rename one column |
| `dim()`, `str()`, `head()`, `summary()` | The arrival checks |

**Things that bite**

| Symptom | Cause |
|:--|:--|
| A numeric column is `chr` | One stray text value converted the whole thing |
| `"100"` sorts before `"9"` | It is text; text sorts character by character |
| An ID lost its leading zero | `read.csv()` read a label as a number |
| `mean()` returns `NA` | There are `NA`s; use `na.rm = TRUE` |
| `x == NA` gives `NA` | Use `is.na(x)` |
| `as.numeric()` on a factor gives codes | `as.character()` first, then `as.numeric()` |
| More rows than days | More than one instrument, or a bad join later |

**Links**

- Session materials: [slides](02-r-essentials-slides.html) ·
  [live script](02-r-essentials.R) ·
  [today's data](data/epa_pm25_compton_2025.csv) ·
  [where it came from](data/SOURCE.md)
- EPA AirData: [Download Daily Data](https://www.epa.gov/outdoor-air-quality-data/download-daily-data)
- [Syllabus](../syllabus.qmd) · [full schedule](../index.qmd#schedule)

:::

## After class {#after-class}

**HW1 is out: "How bad was the air?"** It is due Thursday, September 3, at
8:40 am. It uses the same tool on a different monitor: downtown Los Angeles,
December through February, five numbers in a `results.csv`.

::: {.content-hidden when-format="revealjs"}

You did everything it asks for at least once today. It ships with the data
already in the repository, a `results_spec.md` that states the format
exactly, and CHECK values for three of the answers so you can tell whether
you are on track without guessing. Hand it in the way you handed in your
README: edit or drag files onto the repository page, then commit.

**Re-run today's script before you start.** Do not just read it; run it,
line by line, and change things to see what breaks. The session went fast on
purpose; the script is where you slow down.

**Tuesday** is *Look at your data*: the one-line plots that catch the errors
`str()` and `summary()` miss. It uses this same file, so bring it.

:::

::: {.content-visible when-format="revealjs"}

Re-run today's script at home, line by line. Tuesday: *Look at your data*.

:::

```{r}
#| purl: false
#| include: false
source("_purl.R")
purl_session("02-r-essentials.qmd", "02-r-essentials.R")
```
