---
title: "3 · Look at your data"
subtitle: "Tuesday, September 1, 2026"
pagetitle: "AEM 6850 · Session 3 · Look at your data"
execute:
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
  revealjs:
    output-file: 03-look-at-data-slides.html
    theme: [default, slides.scss]
    template-partials:
      - title-slide.html
    footer: "AEM 6850 · Session 3"
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    code-overflow: scroll
    history: false
    link-external-newwindow: true
---

```{r}
#| purl: false
#| include: false
knitr::opts_chunk$set(
  comment = "#>",
  collapse = FALSE,
  fig.width = 8,
  fig.height = 4,
  fig.align = "center",
  dpi = 150
)
```

::: {.content-hidden when-format="revealjs"}

[Slides](03-look-at-data-slides.html) · [Live script](03-look-at-data.R) ·
[Source](https://github.com/ArielOrtizBobea/aem6850/blob/main/fall-2026/sessions/03-look-at-data.qmd)

This session adds the checks `str()` and `summary()` cannot do: four
base-R plots, one line each, used as instruments rather than
illustrations. It reuses the EPA file from session 2, so have your `data/`
folder in place.

:::

## Objectives {#objectives}

By the end of this session you can:

- Produce one-line diagnostic plots with base R.
- Choose the plot that matches a suspicion about a column.
- Read a distribution for skew, outliers, and impossible values.
- Say what a plot shows that a summary table cannot.

::: {.content-hidden when-format="revealjs"}

## Before class {#before-class}

Have Thursday's `data/` folder in place. Today reuses
`data/epa_pm25_compton_2025.csv` and brings back session 1's weather file. If
either file is missing, the [rebuild section](#rebuild) has a download
fallback that takes thirty seconds.

**HW1 is due Thursday at 8:40 am.** If you have not started, start tonight.
You did everything it asks for at least once on [session
2's page](02-r-essentials.qmd). Today adds the instruments that let you
*see* whether your answers are right before you hand them in.

:::

## What a plot is for {#thesis}

> **A summary answers the question you asked. A plot answers questions you
> did not think to ask.**

Four instruments, one line each, over the data frames you already know.

## Rebuild the table {#rebuild}

Thursday this took forty minutes. Today it takes four statements. That
speed is the payoff from Thursday's work.

```{r}
#| label: rebuild
#| script: "Rebuild session 2's data frame"
pm <- read.csv("data/epa_pm25_compton_2025.csv",
               colClasses = c("Site.ID"          = "character",
                              "State.FIPS.Code"  = "character",
                              "County.FIPS.Code" = "character"))
names(pm)[names(pm) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"
pm$date <- as.Date(pm$Date, format = "%m/%d/%Y")

one <- pm[pm$date <= as.Date("2025-02-28") &
          pm$POC == 1 & pm$AQS.Parameter.Code == 88101, ]
nrow(one)   # 59 -- one row per day, as established Thursday
```

::: {.content-hidden when-format="revealjs"}

The four statements protect the identifiers at read time, rename the long
column, convert the date, and cut the window to one instrument. The final
subset joins its two conditions with `&`. If any line surprises you, reread
session 2 before Thursday's homework deadline.

::: {.callout-tip collapse="true"}
## If `data/` is missing a file

Run these once and you are caught up. They fetch the same snapshots these
pages use:

```{r}
#| label: data-fallback
#| script: "If data/ is missing a file"
#| eval: false
# dir.create("data", showWarnings = FALSE)
# download.file(paste0("https://arielortizbobea.github.io/aem6850/",
#                      "fall-2026/sessions/data/epa_pm25_compton_2025.csv"),
#               "data/epa_pm25_compton_2025.csv")
# download.file(paste0("https://arielortizbobea.github.io/aem6850/",
#                      "fall-2026/sessions/data/la-weather-dec2024-feb2025.csv"),
#               "data/la-weather-dec2024-feb2025.csv")
```

[The data folder's provenance notes](data/SOURCE.md) document where both
files came from, byte for byte.
:::

:::

## Four instruments {#instruments}

::: {.tight-cols}

| The instrument | What it draws |
|:--|:--|
| `hist(x)` | The shape of one numeric column |
| `plot(x, y)` | Two columns against each other, or one through time |
| `boxplot(y ~ g)` | One column, split by the groups in another |
| `barplot(table(g))` | How often each category shows up |

:::

::: {.content-hidden when-format="revealjs"}

Call them instruments, not figures. A figure serves readers and takes an
hour to polish. That is session 12. An instrument serves you. It takes one
line and ten seconds, it can stay ugly, and it never leaves your screen.
You make dozens of them on a normal working day.

Nothing new sits underneath: every one of these takes the vectors and
data-frame columns from session 2. The only new notation is the `~` in
`boxplot()`. We will read it aloud when we get there.

:::

::: {.content-visible when-format="revealjs"}

These are instruments for you, not figures for readers. Each takes one line
and ten seconds, and it never leaves your screen. Reader-ready figures come
in session 12.

:::

## `hist()`: the shape of one column {#hist}

```{r}
#| label: hist-default
#| script: "The first instrument: hist()"
#| fig-alt: "Histogram of daily PM2.5 at Compton for January and February 2025, drawn with default settings. Most days fall between 5 and 25 micrograms per cubic meter, and a long right tail stretches past 50."
hist(one$pm25)
```

::: {.content-hidden when-format="revealjs"}

The call takes one line, default title and all. R chops the column into
bins and draws how many days fell in each. You now see all fifty-nine days
at once. The sort showed you five rows, and `summary()` compressed the rest
into six numbers.

Do not clean this plot up. It is an instrument reading, and it can stay
ugly. The moment you polish, you are making a figure. The cost jumps from
seconds to minutes, and you will stop making these plots.

:::

## Reading the shape {#hist-read}

```{r}
#| label: hist-breaks
#| script: "Turning up the resolution"
#| fig-alt: "The same histogram with twenty bins and a dashed red line at 35 micrograms per cubic meter. The bulk of days sits between 5 and 25, and a separated island of four days lies beyond 40."
hist(one$pm25, breaks = 20,
     main = "", xlab = "Daily PM2.5 (ug/m3)")
abline(v = 35, col = "#b31b1b", lty = 2)   # the EPA 24-hour standard
```

::: {.content-hidden when-format="revealjs"}

`breaks = 20` asks for more bins, which means more resolution on the same
column. Read three things off this plot, in this order:

- **Where the mass sits.** Most winter days at Compton land between 5 and 25
  µg/m³. That is what ordinary air looks like at this monitor.
- **The long right tail.** The distribution is *right-skewed*: the column
  cannot go below zero, so it piles up low and stretches right. Skew is why
  the mean (17.6) sits above the median (15.3). A few bad days pull the
  average up. You used that on Thursday without naming it.
- **The island.** Four separated days sit past the red line, the US EPA's
  24-hour standard of 35 µg/m³: January 1, and January 8, 9, 10. A gap
  between the mass and a clump of stragglers signals *events*, distinct
  from the ordinary process that made the rest.

:::

::: {.content-visible when-format="revealjs"}

- **The mass:** ordinary days, 5–25 µg/m³.
- **The tail:** right skew; mean 17.6 sits above median 15.3.
- **The island past 35:** four days. January 1, then 8, 9, 10.

:::

## The sentinel {#sentinel}

::: {.content-hidden when-format="revealjs"}

Our file comes from AQS, EPA's air-quality archive, and it marks a missing
day by *absence*: the row is not there. Plenty of files instead put a number
where the gap is: `-999`, `9999`, `0`. That number is called a **sentinel**.
EPA's own real-time feed, AirNow, ships `-999` to this day. Sooner or later
you will open a file with sentinels in it.

::: {.column-margin}
**Why −999?** Old data formats could not leave a field blank. Every cell had
to hold a number, so agencies picked one no real measurement could produce.
The formats died; the convention did not.
:::

Thursday we verified this file is clean. So we corrupt a copy ourselves,
and watch the failure happen before it costs you anything:

:::

```{r}
#| label: sentinel
#| script: "The sentinel"
bad <- one$pm25
bad[c(40, 45, 50, 55)] <- -999   # four missing days, the AirNow way
mean(bad)
summary(bad)
```

::: {.content-hidden when-format="revealjs"}

Thursday's `"n/a"` at least produced an `NA` and a warning. This produced
nothing. `-999` is a legitimate number, so R raises no error and no warning.
`mean()` ran clean and returned garbage. If −51 went into a difference or
a ratio instead of onto your screen, nothing downstream would flag it.

`summary()` does show the damage: `Min. -999`. But only if you read all six
numbers, and only if you know the physical floor is zero.

:::

::: {.content-visible when-format="revealjs"}

- No error, no warning, no `NA`. It ran clean and it is wrong.
- `summary()` shows it, if you read all six numbers and know the floor is
  zero.

:::

## The histogram catches it {#sentinel-hist}

```{r}
#| label: sentinel-hist
#| script: "The histogram catches it"
#| fig-alt: "Histogram of the corrupted column: a lone short bar at minus 1000, a wide empty gap, and all the real data crushed into a single tall bar near zero."
hist(bad, main = "", xlab = "Daily PM2.5 (ug/m3)")
```

```{r}
#| label: sentinel-repair
#| script: "The repair"
bad[bad == -999] <- NA        # sentinel to honest missing
mean(bad, na.rm = TRUE)       # back to 17.6
```

::: {.content-hidden when-format="revealjs"}

You do not need to know what PM2.5 is to see that something is wrong here.
One bar sits a thousand units from everything else. The axis stretched to
fit the garbage and crushed the real data into a single tower.
**A ruined histogram is itself a reading.** When everything you care about
collapses into one bin, the axis made room for a value that should not
exist. Go find that value.

The repair is one line of Thursday's logical subsetting, and the mean
returns to 17.6. The order matters: *look, then repair, then compute*.
Never go straight to the computing.

:::

## `plot()`: a column through time {#plot-time}

```{r}
#| label: plot-time
#| script: "A column through time"
#| fig-alt: "Daily PM2.5 at Compton drawn as vertical spikes from January through February 2025. The tallest spike stands alone on January 1. A red vertical line marks January 7, followed immediately by a three-day cluster of high readings. February is much calmer."
plot(one$date, one$pm25, type = "h", col = "grey30",
     xlab = "", ylab = "Daily PM2.5 (ug/m3)")
abline(v = as.Date("2025-01-07"), col = "#b31b1b", lwd = 2)
```

::: {.content-hidden when-format="revealjs"}

This repeats session 1's wind plot: `type = "h"` draws one spike per day.
When a column has a natural order, usually time, plot it in that order. The
whole winter then shows in one glance: a lone spike on January 1, the red
line at January 7, a broad three-day shoulder as the fires burn, then a
February back to normal.

Compare that with what sorting told us Thursday. The sorted list put January
9, 8, and 10 at ranks two, three, and four: three separate entries, with the
order destroyed. The plot shows what they are: **one event**, three days
wide, starting the day after ignition. The plot also shows January 1 the way
the sort never could. The spike stands *alone*, with quiet days on both
sides. Fireworks last one night; fires last days.

:::

::: {.content-visible when-format="revealjs"}

- The sort gave ranks 1, 2, 3, 4. The plot shows a one-day spike, then a
  **three-day event** starting the day after ignition.
- Fireworks last one night; fires last days.

:::

## `plot()`: two columns together {#plot-scatter}

Session 1's weather file comes back: the same ninety days, saved alongside
these pages.

```{r}
#| label: weather-load
#| script: "Bring back session 1's weather"
la <- read.csv("data/la-weather-dec2024-feb2025.csv", skip = 3)
names(la) <- c("date", "gust", "wind", "tmax", "rh_min", "precip")
la$date <- as.Date(la$date)
```

```{r}
#| label: plot-scatter
#| script: "Two columns together"
#| fig-alt: "Scatterplot of daily maximum gust against daily maximum sustained wind for ninety winter days in Los Angeles. The points form one tight upward ray. The windiest days sit far up the ray but on it."
plot(la$wind, la$gust,
     xlab = "Max sustained wind (km/h)", ylab = "Max gust (km/h)")
```

::: {.content-hidden when-format="revealjs"}

A scatterplot draws a *relationship*, and these two columns must have one.
A gust is a burst on top of the sustained wind, so gusts should run above
wind speed by a believable factor. They do: one tight ray, with gusts at
roughly two to four times the sustained speed, every day of the winter.

Look at where January 7 and 8 sit: far up the ray, *and on it*. Both columns
are extreme; the relationship is ordinary. A real windstorm looks like this.
**Extreme is not the same as wrong.** The point to worry about is never the
one far along the ray. It is the one *off* the ray.

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
Read it live: one ray, gusts at 2–4× the sustained wind, every day. Point
at Jan 7–8, far up the ray and ON it: extreme, but consistent. The point to
worry about is never far along the ray; it is off it. That sentence sets up
the next slide.
:::
:::

## The unit error {#unit-error}

::: {.content-hidden when-format="revealjs"}

This instrument exists to catch one classic failure. Weather feeds serve
wind in km/h, m/s, mph, or knots, and the choice sits in a header field or
a query parameter; Open-Meteo names its parameter `wind_speed_unit`. Stitch
together files pulled with different settings and you get one column with
two units inside. Again we corrupt a copy ourselves, to watch it happen:

:::

```{r}
#| label: unit-error
#| script: "The unit error"
mixed <- la$wind
jan   <- format(la$date, "%Y-%m") == "2025-01"
mixed[jan] <- round(la$wind[jan] / 3.6, 1)   # January delivered in m/s

summary(la$wind)    # the honest column
summary(mixed)      # the corrupted one
```

::: {.content-hidden when-format="revealjs"}

Read the corrupted summary and try to reject it. The minimum, 0.9, reads as
a calm day. The maximum, 21.8, reads as a windy one. The mean, 6.5, runs
low for winter, and you would need LA wind climatology in your head
to doubt it. **Every number in it is a plausible wind speed in km/h.**
There is no sentinel to spot and no impossible value to catch. The
corruption lives in how thirty-one values relate to the rest, and single
numbers cannot show a relationship.

:::

::: {.content-visible when-format="revealjs"}

Min 0.9, max 21.8, mean 6.5: **every number is a plausible wind speed.**
Nothing is impossible; the corruption lives in the *relationships*.

:::

## The scatterplot catches it {#unit-error-plot}

```{r}
#| label: unit-error-plot
#| script: "The scatterplot catches it"
#| fig-alt: "The same gust-versus-wind scatterplot after January's wind speeds are converted to meters per second: the single ray splits into two distinct rays with clearly different slopes."
plot(mixed, la$gust,
     xlab = "Wind speed, as delivered", ylab = "Max gust (km/h)")
```

::: {.content-hidden when-format="revealjs"}

The plot shows two rays. The right-hand ray is December and February, with
gusts at two to four times the wind, as before. The steep left-hand ray is
January: the same relationship, times 3.6, because its x-values shrank by
exactly the km/h-to-m/s factor. Thirty-one points do not land on a second
straight line by accident. **A ray that splits in two means two regimes in
one column**, and a clean split at a constant factor means units.

This is the strongest version of today's claim. `summary()` showed the
sentinel, barely. It never showed the unit error. No single number was
wrong, so no list of single numbers could catch it. Only the picture of the
*relationship* could.

:::

::: {.content-visible when-format="revealjs"}

- Two rays mean two regimes in one column. A clean constant factor means
  units.
- `summary()` had nothing to show: no individual number was wrong.
- Only the picture of the relationship could catch it.

:::

## `boxplot()`: a column by groups {#boxplot}

```{r}
#| label: boxplot-month
#| script: "A column by groups: boxplot()"
#| fig-alt: "Box plots of daily PM2.5 for January and February 2025. January's box sits higher and stretches wider, with a single outlier dot far above its whisker. February's box is lower and tighter, with no outliers."
one$month <- format(one$date, "%Y-%m")
boxplot(pm25 ~ month, data = one)
```

::: {.content-hidden when-format="revealjs"}

Read `pm25 ~ month` aloud as "pm25, *by* month". The `~` is R's word for
"split by," and `data = one` says which data frame the two names live in.
(Use `"%Y-%m"` rather than `"%m"` so the labels sort in time order when a
window crosses New Year. The weather file's December would otherwise sort
after February.)

One box per group, five readings per box: the heavy line marks the median;
the box holds the middle half of the days; the whiskers cover the ordinary
range; the plot draws anything beyond them as its own dot. A dot is a
candidate for a closer look, not automatically an error.

Now read this plot against the histogram, because the two instruments
disagree in an instructive way. The histogram showed an island of four days.
Here only one of them survives as a dot: January 1. The three fire days sit
at the top of January's whisker, *inside* the ordinary range. A boxplot
judges "beyond" against each group's own spread, and January's spread is
enormous, so three bad days read as the high end of normal. Neither
instrument is wrong. They answer different questions: "extreme for this
data?" versus "extreme for this group?"

The medians hold the quieter finding: 16.9 in January against 12.6 in
February. Take out the four event days and January's median still reads
14.5 to February's 12.6. The baseline shifted, not just a few days. The
plot does not say why. Lingering smoke from a fire month? Stronger winter
inversions? The plot raises the question and leaves it open. You knew about
the events. You probably did not know about the shifted baseline. That
second kind of finding is why you make these plots.

:::

::: {.content-visible when-format="revealjs"}

- `pm25 ~ month` reads "pm25, **by** month."
- One dot survives: Jan 1. The fire days land *inside* January's whisker,
  because a boxplot judges "extreme" against each group's own spread.
- The quieter finding: January's *median* (16.9) sits above February's (12.6).
  The baseline shifted, not just the events.

:::

## `barplot()`: counts of categories {#barplot}

Thursday's window was 59 days of a 662-row file. Zoom out to the whole
year: did the regulatory instrument run all of it?

```{r}
#| label: barplot-months
#| script: "Counts of categories: barplot()"
#| fig-alt: "Bar chart of days with a reading per month of 2025 for the regulatory instrument at Compton. Most months show about thirty days, but April shows two and May shows one."
year <- pm[pm$POC == 1 & pm$AQS.Parameter.Code == 88101, ]
barplot(table(format(year$date, "%m")),
        xlab = "Month of 2025", ylab = "Days with a reading")
```

::: {.content-hidden when-format="revealjs"}

`table()` counts, as you saw Thursday, and `barplot()` draws the counts.
The pairing is the whole trick: any categorical column, months here, goes
through `table()` and comes out as a picture of coverage.

And the picture has a hole in it. **April has two days. May has one.**
Nobody planted that. It sits in the file exactly as EPA ships it, and we
walked past it twice: session 2 never looked outside the winter window, and
the file's 662 rows *looked* like plenty.

:::

::: {.content-visible when-format="revealjs"}

**April: 2 days. May: 1.** The file ships this way from EPA. We walked
past it twice.

:::

## The missing spring {#missing-spring}

```{r}
#| label: missing-spring
#| script: "The missing spring: site or sampler?"
table(format(year$date, "%m"))    # the numbers behind the bars

other <- pm[pm$POC == 3 & pm$AQS.Parameter.Code == 88502, ]
table(format(other$date, "%m"))   # the site's OTHER instrument
```

::: {.content-hidden when-format="revealjs"}

This repeats Thursday's "was it real?" move: check against an independent
measurement. The site carries one. The continuous instrument, POC 3,
reported 30 days in April and 31 in May. The *site* was fine; the
*regulatory sampler* was down for two months. (AQS validation can also
invalidate readings after the fact. Either way, the regulatory series has
no spring.)

Here is what the gap costs. Compute "the 2025 annual mean" from the
regulatory series and nothing errors, nothing warns. The number silently
skips almost all of spring, a mild and cleaner season, so it lands biased.
Every mean is a mean *of the days that are there*. The count plot finds out
which days those are, and this one took two lines against a file we had
already "checked" for a week.

:::

::: {.content-visible when-format="revealjs"}

- POC 3 ran 30 and 31 days those months: the **sampler** was down, not the
  site.
- An "annual mean" of the regulatory series silently skips spring, and it
  lands biased.
- Every mean is a mean *of the days that are there.*

:::

## Choosing the instrument {#choosing}

Start from the suspicion, not from the plot type.

::: {.tight-cols}

| The instrument | The suspicion it answers |
|:--|:--|
| `hist(x)` | Anything impossible in here? Skew? A gap? A second hump? |
| `plot(d$date, y, type = "h")` | Did something happen in time: a spike, a break, a dead stretch? |
| `plot(x, y)` | Do these two columns relate the way they are supposed to? |
| `boxplot(y ~ g, data = d)` | Does one group differ in a way the pooled column averages away? |
| `barplot(table(g))` | Is every category, month, or site there, as often as it should be? |

:::

::: {.content-hidden when-format="revealjs"}

Suspicions come from three places. `str()` and `summary()` supply some: a
character column that should be numeric, a minimum that grazes an
impossible value. The data's own documentation supplies more, like the
two-instrument structure we found in the POC table. The rest come from the
world: you knew when the fires started before you opened the file, and that
knowledge made the January 1 spike worth chasing. The instruments do not
replace knowing things. They test the file against what you know.

:::

## Same summaries, different data {#anscombe}

```{r}
#| label: anscombe
#| script: "Same summaries, different data: Anscombe 1973"
#| fig-height: 5
#| fig-alt: "Anscombe's four datasets as four scatterplots: a loose linear cloud, a smooth arch, a tight line with one far outlier, and a vertical stack of identical x values with one point far to the right."
par(mfrow = c(2, 2), mar = c(4, 4, 1, 1))
plot(anscombe$x1, anscombe$y1)
plot(anscombe$x2, anscombe$y2)
plot(anscombe$x3, anscombe$y3)
plot(anscombe$x4, anscombe$y4)
par(mfrow = c(1, 1))
```

::: {.content-hidden when-format="revealjs"}

In 1973 the statistician Francis Anscombe built four small datasets to make
today's point, and they ship inside R as `anscombe`. All four pairs have
the same x mean, the same y mean, the same variances, and the same
correlation of 0.82. Every summary statistic you have met says these four
are the same data.

The plots disagree. The first is a loose linear cloud. The second is a
smooth curve with nothing linear about it. The third is a perfect line plus
one outlier that wrecks the fit. The fourth stacks ten identical x values,
and one far point creates the whole correlation. Fifty years on, this
is still the clearest demonstration that **a summary is a compression, and
compression discards structure you needed to see.**

(`par(mfrow = c(2, 2))` splits the plot window into panels, and the setting
sticks until you set it back. The last line resets it. `cor()` gets its
full introduction later; here it is one more summary that cannot tell these
apart.)

::: {.callout-tip collapse="true"}
## Check the claim

```{r}
#| label: anscombe-check
#| script: "Anscombe: check the claim"
colMeans(anscombe)
cor(anscombe$x1, anscombe$y1)
cor(anscombe$x2, anscombe$y2)
cor(anscombe$x3, anscombe$y3)
cor(anscombe$x4, anscombe$y4)
```
:::

:::

::: {.content-visible when-format="revealjs"}

Four datasets share the same means, variances, and correlation, to two
decimal places. Every summary you have met calls them identical.

:::

## You try {#you-try}

Five minutes, on the weather frame `la`. Stubs are in the live script.

```{r}
#| label: you-try
#| exercise: true
#| script: "You try (5 minutes)"
#| eval: false
# 1. hist(la$rh_min) -- each day's LOWEST relative humidity. Read it in one
#    sentence. Twelve days sit in single digits; la$date[la$rh_min < 10]
#    says when they were. What do those dates line up with?
#
# 2. plot(la$date, la$precip, type = "h") -- when did the drought break?
#    Now run summary(la$precip). Could you have read the break off that?
#
# 3. la$month <- format(la$date, "%Y-%m")
#    boxplot(gust ~ month, data = la)
#    Which month has the extremes -- and are they a shifted box, or dots
#    beyond the whisker? Those are different claims about January.
```

::: {.content-visible when-format="revealjs"}
::: {.notes}
Debriefs, one sentence each: (1) the single-digit days run Dec 10-Jan 24:
Santa Ana air, the fire weather itself; (2) Jan 26 breaks the drought with
17 mm after seven dry weeks, and the biggest storm is Feb 13 at 38 mm;
summary() shows a median of 0 and a max, never *when*; (3) January's
extremes are dots (Jan 7 at 66, far beyond the whisker) while its median
(22.7) barely beats February's (21.6): extreme days, not a windier month
throughout.
:::
:::

## Verify {#verify}

::: {.content-hidden when-format="revealjs"}

Session 1 installed the habit: write down what the data should show, run,
reconcile. Session 2 added: run `str()` before you compute. This session
completes the quick checks:

:::

> **Plot the column before you trust the column.**

::: {.content-hidden when-format="revealjs"}

Plot every column that feeds a number you will report, once, before you
believe anything computed from it. The habit costs ten seconds per column.
It is the only check in the kit that works when every value is plausible
and only the shape is wrong.

**From HW2 on, homeworks grade this habit.** They ask for diagnostic plots,
and each plot needs one written sentence that reads it, the way this page
read each plot above. An unread plot checks nothing.

The quick checks, now five:

| Check | Question it answers |
|:--|:--|
| `dim()` | Did I get the number of rows I expected? |
| `str()` | Is every column the class it should be? |
| `head()` | Does row one look like a row of data? |
| `summary()` | Is anything impossible, or off by a factor of a thousand? |
| `hist()` / `plot()` | Is anything wrong that no single number can show? |

:::

::: {.content-visible when-format="revealjs"}

The quick checks, now five:<br>
`dim()`: the rows I expected?<br>
`str()`: the classes I expected?<br>
`head()`: does row one look like data?<br>
`summary()`: anything impossible?<br>
`hist()` / `plot()`: anything wrong that **no single number can show?**

From HW2 on, every diagnostic plot needs one written sentence that reads
it. HW2 grades the habit.

:::

## Quick reference {#quick-reference}

::: {.content-visible when-format="revealjs"}

On the session page: the four instruments with their options, a
what-shapes-mean table, and every command from today.

:::

::: {.content-hidden when-format="revealjs"}

**The four instruments**

| Code | What it draws |
|:--|:--|
| `hist(x)` | The shape of one numeric column |
| `hist(x, breaks = 20)` | Same, with more resolution |
| `plot(d$date, y, type = "h")` | A column through time, one spike per day |
| `plot(x, y)` | Two columns against each other |
| `boxplot(y ~ g, data = d)` | `y` split by the groups in `g`; read `~` as "by" |
| `barplot(table(g))` | How often each category shows up |

**Dressing an instrument (rarely needed)**

| Code | What it does |
|:--|:--|
| `main = "", xlab = "", ylab = ""` | Title and axis labels |
| `abline(v = 35)`, `abline(h = 10)` | A vertical or horizontal reference line |
| `col =`, `lty =`, `lwd =` | Color, line type, line width |
| `par(mfrow = c(2, 2))` | Panels in one window; reset with `c(1, 1)` |

**What a shape is telling you**

::: {.tight-cols}

| You see | It usually means |
|:--|:--|
| A bar far from the mass | A sentinel or impossible value; go look at those rows |
| Everything crushed into one bin | The axis stretched to fit a bad value; go find it |
| Two humps | Two populations in one column: units, instruments, or groups |
| A long right tail | Skew; mean and median will disagree, and the mean is the fragile one |
| A ray that splits in two | Two regimes; a clean constant factor means units |
| Dots beyond the whiskers | Outliers to inspect: candidates, not confirmed errors |
| A short bar in a count plot | A coverage gap: days or categories missing |

:::

**Links**

- Session materials: [slides](03-look-at-data-slides.html) ·
  [live script](03-look-at-data.R) ·
  [EPA data](data/epa_pm25_compton_2025.csv) ·
  [weather data](data/la-weather-dec2024-feb2025.csv) ·
  [where they came from](data/SOURCE.md)
- [Session 2](02-r-essentials.qmd): the subsetting this session leans on
- [Syllabus](../syllabus.qmd) · [full schedule](../index.qmd#schedule)

:::

## After class {#after-class}

**HW1 is due Thursday at 8:40 am.** Before you submit, point today's
instruments at your own answers: `hist()` the concentration column, and
`plot()` it through time. If your peak day is not visibly the peak of the
plot, one of the two is wrong. Find that out before the grader does.

::: {.content-hidden when-format="revealjs"}

**Thursday** is Wrangling I: `dplyr`. It covers the verbs, the pipe, and
grouped summaries. The code will start to read like sentences.

**HW2 posts Thursday after class**: *What did the fire
destroy?*, due September 10. The data
comes from CAL FIRE, California's state fire agency: structure-by-structure
damage inspections from the Eaton and Palisades fires. Each row is an
inspected building, with its damage category and its year built. The
homework asks for three diagnostic plots, each read in one written
sentence. HW2 grades the habit from today. One of its columns has a
surprise in it that `summary()` will not show you. Plot first.

:::

::: {.content-visible when-format="revealjs"}

- Thursday: Wrangling I, with `dplyr`.
- HW2 posts Thursday: *What did the fire destroy?* It asks for three
  diagnostic plots, each read in one sentence. It grades today's habit.

:::

```{r}
#| purl: false
#| include: false
# Single-source build: the .R live script is generated from this file every
# time the page renders, so it cannot fall out of step with the prose.
source("_purl.R")
purl_session("03-look-at-data.qmd", "03-look-at-data.R")
```
