---
title: "3 · R essentials II"
subtitle: "Tuesday, September 1, 2026"
pagetitle: "AEM 6850 · Session 3 · R essentials II"
execute:
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
  revealjs:
    output-file: 03-r-essentials-ii-slides.html
    theme: [default, slides.scss]
    template-partials:
      - title-slide.html
    footer: "AEM 6850 · Session 3"
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    code-overflow: scroll
    history: false
    link-external-newwindow: true
---

```{r}
#| purl: false
#| include: false
knitr::opts_chunk$set(
  comment = "#>",
  collapse = FALSE,
  fig.width = 8,
  fig.height = 4,
  fig.align = "center",
  dpi = 150
)
```

::: {.content-hidden when-format="revealjs"}

[Slides](03-r-essentials-ii-slides.html) · [Live script](03-r-essentials-ii.R) ·
[Source](https://github.com/ArielOrtizBobea/aem6850/blob/main/fall-2026/sessions/03-r-essentials-ii.qmd)

Everything below is in the live script. Run it line by line at home, as many
times as it takes.

:::

## Objectives {#objectives}

By the end of this session you can:

- Build and index a matrix, and say what the one-type rule costs you.
- Build and index a list, and tell `[` from `[[`.
- Combine pieces: `rbind()` and `cbind()` on matrices, `unlist()` and
  `do.call()` on lists.
- Run one function over every row, column, group or file with `apply()`,
  `lapply()` and `sapply()`.
- Read a real CSV, protect an identifier column, and convert text dates.
- Turn a long file into a sites-by-months matrix and summarise both margins.
- Join two tables on a shared key, and check the row count before and after.

::: {.content-hidden when-format="revealjs"}

## Before class {#before-class}

The last part uses `data/epa_pm25_la_county_2025.csv`: daily PM2.5 from every
monitor in Los Angeles County, calendar year 2025, exactly as EPA's "Download
Daily Data" tool ships it. Download [the
file](data/epa_pm25_la_county_2025.csv) into a `data/` folder next to your
script. The [read section](#read) has a fallback that fetches it from R.

:::

## Matrices {#matrices}

A **matrix** is one vector with dimensions.

```{r}
#| label: matrices
#| script: "Matrices"
m <- matrix(1:6, nrow = 2)
m
dim(m)

matrix(1:6, nrow = 2, byrow = TRUE)   # filled across instead of down
```

::: {.content-hidden when-format="revealjs"}

R folds the vector down the columns unless you say `byrow = TRUE`. Get that
backwards and every value lands in the wrong cell, with no error.

:::

## Indexing a matrix {#matrix-indexing}

```{r}
#| label: matrix-indexing
#| script: "Indexing a matrix"
m[2, 3]     # row 2, column 3
m[2, ]      # all of row 2
m[, 3]      # all of column 3
```

::: {.content-hidden when-format="revealjs"}

Two coordinates, `m[row, column]`, and a blank means all of them. The same
rule you use on a data frame.

Note that `m[2, ]` came back as a plain vector, not a one-row matrix. R drops
the dimension when you ask for a single row or column. Add `drop = FALSE`
when you need the shape kept.

:::

## Names on the margins {#matrix-names}

```{r}
#| label: matrix-names
#| script: "Names on the margins"
temps <- matrix(c(12, 15, 19,
                  14, 17, 22), nrow = 2, byrow = TRUE)
rownames(temps) <- c("Compton", "Reseda")
colnames(temps) <- c("Jan", "Feb", "Mar")
temps

temps["Reseda", "Feb"]
rowMeans(temps)
colMeans(temps)
```

::: {.content-hidden when-format="revealjs"}

Names make a matrix readable and let you index by label instead of counting
positions. `rowMeans()`, `colMeans()`, `rowSums()` and `colSums()` are the
four summaries you reach for most.

:::

## Matrix arithmetic {#matrix-math}

```{r}
#| label: matrix-math
#| script: "Matrix arithmetic"
temps * 9/5 + 32     # every cell at once, Celsius to Fahrenheit
t(temps)             # flip rows and columns
```

::: {.content-hidden when-format="revealjs"}

Arithmetic runs cell by cell, exactly as it does on a vector, because
underneath a matrix is a vector.

This is why images and grids are stored as matrices. A photograph is a
matrix of pixel values, and making it brighter is one multiplication. A
weather grid is a matrix of temperatures, and converting the whole grid to
Fahrenheit is the line above.

:::

## Combining matrices {#matrix-combine}

```{r}
#| label: matrix-combine
#| script: "Combining matrices"
more <- matrix(c(9, 11, 16), nrow = 1,
               dimnames = list("Pasadena", c("Jan", "Feb", "Mar")))

rbind(temps, more)          # a new row: same columns
cbind(temps, Apr = c(21, 24))   # a new column: same rows
```

::: {.content-hidden when-format="revealjs"}

`rbind()` stacks matrices on top of each other and `cbind()` puts them side
by side. The shared margin has to agree: `rbind()` needs the same number of
columns, `cbind()` the same number of rows.

R matches columns by **position, not by name**. If the second matrix carries
the same three months in a different order, `rbind()` will stack them anyway
and put February's numbers under the January heading, with no warning.
Check the names agree before you stack, not after.

:::

## One matrix, one type {#matrix-type}

```{r}
#| label: matrix-type
#| script: "One matrix, one type"
rbind(c(1, 2, 3),
      c("a", "b", "c"))
```

::: {.content-hidden when-format="revealjs"}

One vector underneath means one class throughout. Add a row of text and
every number becomes text, silently, with no warning. It is the same
conversion that turns a numeric column into text when one entry is a word.

A data frame does not do this, because its columns are separate vectors.
That is the whole reason both containers exist: tables of mixed data are
data frames, and grids of one measurement are matrices.

:::

## When most of the cells are zero {#sparse}

```{r}
#| label: sparse
#| script: "Sparse matrices"
library(Matrix)

set.seed(1)
n <- 1000
d <- matrix(0, n, n)
d[sample(n * n, n * n * 0.01)] <- 1   # only 1% of cells are not zero

s <- Matrix(d, sparse = TRUE)

format(object.size(d), units = "MB")   # dense: every cell stored
format(object.size(s), units = "MB")   # sparse: only the non-zeros
```

::: {.content-hidden when-format="revealjs"}

A dense matrix stores every cell, including all the zeros. A **sparse**
matrix stores only the non-zero values and the positions they sit in. Same
numbers, same indexing, about sixty times less memory here.

You meet these the moment a matrix gets large and mostly empty: a
county-by-county adjacency matrix, a word-by-document count, a design matrix
of fixed effects with one 1 per row. At `n = 1000` the difference is a
convenience. At `n = 100,000` it is the difference between fitting in memory
and not.

`Matrix` ships with R, so there is nothing to install. Its matrices index and
multiply like ordinary ones; `as.matrix()` converts back when a function
insists on dense input.

:::

## Lists {#lists}

A **list** holds anything, in any mix, at any length.

```{r}
#| label: lists
#| script: "Lists"
l <- list(site = "Compton", readings = c(53.2, 33.6, 47.0), clean = TRUE)
str(l)
length(l)      # three elements...
lengths(l)     # ...of these lengths
```

::: {.content-hidden when-format="revealjs"}

Text, a numeric vector, and a logical, side by side in one object. A vector
could not do this: it would convert everything to text. A matrix could not
either: one type throughout, and every row the same length.

That freedom is the point. Lists are what R uses whenever the pieces do not
line up.

:::

## One bracket or two {#brackets}

```{r}
#| label: brackets
#| script: "One bracket or two"
l["readings"]     # a LIST of length one
l[["readings"]]   # the vector itself
l$readings        # the same thing, less typing

class(l["readings"])
class(l[["readings"]])
```

::: {.content-hidden when-format="revealjs"}

Single brackets keep the container: you asked a list for part of itself, so
you get a shorter list. Double brackets reach inside and hand you the
element.

The mistake to expect is arithmetic on a single-bracket result:
`mean(l["readings"])` fails, because you handed `mean()` a list. `$` and
`[[ ]]` do the same job; `[[ ]]` is the one that also takes a variable, which
matters as soon as the name is computed rather than typed.

:::

## Lists inside lists {#nested-lists}

```{r}
#| label: nested-lists
#| script: "Lists inside lists"
monitor <- list(
  id   = "060371302",
  site = list(name = "Compton", lat = 33.90, lon = -118.21),
  pm   = c(53.2, 33.6, 47.0)
)

monitor$site$name
monitor[["site"]][["lat"]]
str(monitor)
```

::: {.content-hidden when-format="revealjs"}

An element of a list can be another list, to any depth. Chain the operators
to walk down: `$` after `$`, or `[[ ]]` after `[[ ]]`.

This is exactly the shape of the JSON that web APIs return, which is why the
lesson pays off well beyond R. `str()` is how you find your way around one
you did not build.

:::

## Where lists come from {#lists-from-functions}

```{r}
#| label: lists-from-functions
#| script: "Where lists come from"
fit <- lm(c(1, 3, 2, 5) ~ c(1, 2, 3, 4))

class(fit)
names(fit)          # a list underneath
fit$coefficients
```

::: {.content-hidden when-format="revealjs"}

Most functions that return more than one thing return a list, because a list
is the only container that can hold a number, a vector and a data frame at
once.

A fitted model is a list with a class on it. `names()` shows you the parts
and `$` takes one out. Every modelling function you meet this semester
follows this pattern, so knowing lists means never being stuck at "what did
this thing give me?"

:::

## Combining list elements {#list-combine}

```{r}
#| label: list-combine
#| script: "Combining list elements"
pieces <- list(a = c(1, 2), b = c(3, 4), c = c(5, 6))

unlist(pieces)              # one flat vector, names kept
do.call(rbind, pieces)      # one matrix, one row per element
```

::: {.content-hidden when-format="revealjs"}

`unlist()` flattens a list into a single vector. It is convenient and it is
lossy: everything becomes one type, and any structure below the top level
disappears.

`do.call(f, list)` calls `f` with the list's elements as its arguments, so
`do.call(rbind, pieces)` is `rbind(pieces$a, pieces$b, pieces$c)` without
your having to know how many there are or what they are called. That is the
standard way to turn a list of rows into a matrix, or a list of data frames
into one data frame.

:::

## Exercise 1 {#exercise-1}

::: {.panel-tabset}

### Exercise

```{r}
#| label: exercise-1
#| exercise: true
#| script: "Exercise 1 (5 minutes)"
#| eval: false
# 1. Build a 3 x 4 matrix of the numbers 1 to 12, filled ACROSS the rows.
#    Name the rows A, B, C. What is the mean of row B?
#
# 2. Build a list holding your name, the numbers 1 to 5, and TRUE.
#    Pull the numeric vector out two different ways, then take its mean.
#
# 3. What does l["readings"] give you that l[["readings"]] does not?
#    Answer with class(), not from memory.
```

### Solution

```{r}
#| label: exercise-1-solution
#| purl: false
mm <- matrix(1:12, nrow = 3, byrow = TRUE)
rownames(mm) <- c("A", "B", "C")
mm
mean(mm["B", ])

l2 <- list(name = "Ariel", nums = 1:5, ok = TRUE)
mean(l2$nums)
mean(l2[["nums"]])

# single brackets keep the container, so this is a list of length one
class(l["readings"])
class(l[["readings"]])
```

:::

## `apply()`: once per row, or once per column {#apply}

```{r}
#| label: apply
#| script: "apply(): once per row, once per column"
temps
apply(temps, 1, mean)    # 1 = rows
apply(temps, 2, mean)    # 2 = columns
```

::: {.content-hidden when-format="revealjs"}

`apply(m, MARGIN, f)` runs `f` over one margin of a matrix and collects the
results.

The margin numbers are the bracket coordinates. You write `m[row, column]`,
so 1 is rows and 2 is columns. Margin 1 gives you one answer per row; margin
2 gives you one per column.

`rowMeans()` and `colMeans()` are fast special cases of exactly this.

:::

## Any function, not just `mean` {#apply-any}

```{r}
#| label: apply-any
#| script: "Any function, not just mean"
apply(temps, 2, max)
apply(temps, 2, range)                       # two numbers per column
apply(temps, 2, function(x) max(x) - min(x)) # write your own, inline
```

::: {.content-hidden when-format="revealjs"}

The function is an argument like any other. Pass it by name with no
parentheses, or write it inline with `function(x)` when it is too small to
deserve a name. Inside, `x` is one row or one column.

`range` returns two numbers per column, so `apply()` hands back a matrix
rather than a vector. What you get out depends on what your function
returns.

:::

## `split()`: one group per element {#split}

```{r}
#| label: split
#| script: "split(): one group per element"
x <- c(12, 15, 19, 14, 17, 22)
g <- c("Compton", "Compton", "Compton", "Reseda", "Reseda", "Reseda")

split(x, g)
```

::: {.content-hidden when-format="revealjs"}

`split()` cuts an object into a list, one element per group. Those are the
same six numbers that went into `temps`, arranged as a list of two vectors
instead of a two-row matrix.

That is the choice in miniature. A matrix needs every group to be the same
length; a list does not. Real groups are rarely the same size, which is why
`split()` returns a list.

:::

## `lapply()`: a list in, a list out {#lapply}

```{r}
#| label: lapply
#| script: "lapply(): a list in, a list out"
by_site <- split(x, g)

lapply(by_site, mean)
```

::: {.content-hidden when-format="revealjs"}

`lapply(x, f)` runs `f` on every element of `x` and returns a list of the
same length, with the same names. One call, one result per group, and no
loop written by hand.

Inside the function, the argument is one element: here, one site's readings.
You write the code once, for one group, and `lapply()` handles the
repetition.

:::

## `sapply()`: the same, simplified {#sapply}

```{r}
#| label: sapply
#| script: "sapply(): the same, simplified"
sapply(by_site, mean)      # a named vector, not a list
sapply(by_site, length)
sapply(by_site, range)     # two numbers each, so a matrix comes back

apply(temps, 1, mean)      # same answers, from the matrix
```

::: {.content-hidden when-format="revealjs"}

`sapply()` is `lapply()` plus one step: when every result is a single value,
it returns a named vector instead of a list. When every result is the same
longer length, it returns a matrix.

Use `sapply()` when you want something to print or compute with. Use
`lapply()` when each result is bigger than one value, or when you want the
shape guaranteed. `sapply()` decides what to return based on what it gets,
which is convenient at the console and a hazard inside a script.

The last line is the same question asked of the matrix. Two containers, two
verbs, one answer: `apply()` walks a margin, `sapply()` walks a list.

:::

## Exercise 2 {#exercise-2}

::: {.panel-tabset}

### Exercise

```{r}
#| label: exercise-2
#| exercise: true
#| script: "Exercise 2 (5 minutes)"
#| eval: false
# Everything here uses temps and by_site, built earlier in this script.
#
# 1. One line: the highest reading for each city. One line: the highest
#    for each month. Which margin number did each one need, and why?
#
# 2. Build a list of three numeric vectors of DIFFERENT lengths.
#    One line for how long each one is; one line for the mean of each.
#    Why could a matrix not hold this?
#
# 3. sapply(by_site, range) came back as a matrix, not a vector.
#    Use dim() to explain why.
```

### Solution

```{r}
#| label: exercise-2-solution
#| purl: false
apply(temps, 1, max)   # margin 1 = rows = cities
apply(temps, 2, max)   # margin 2 = columns = months

ragged <- list(a = 1:3, b = 1:7, c = 1:2)
lengths(ragged)
sapply(ragged, mean)   # a matrix needs every group the same length

dim(sapply(by_site, range))   # 2 x 2: two numbers per site, so a matrix
```

:::

## Read it in {#read}

```{r}
#| label: read
#| script: "Read the file"
la <- read.csv("data/epa_pm25_la_county_2025.csv")
dim(la)
```

```{r}
#| label: read-fallback
#| script: "If you do not have the file yet"
#| eval: false
# Uncomment and run once. Same file, fetched instead of clicked.
# dir.create("data", showWarnings = FALSE)
# download.file(paste0("https://arielortizbobea.github.io/aem6850/",
#                      "fall-2026/sessions/data/epa_pm25_la_county_2025.csv"),
#               "data/epa_pm25_la_county_2025.csv")
```

::: {.content-hidden when-format="revealjs"}

`read.csv()` takes a path relative to the working directory, which `getwd()`
prints and `list.files()` lets you check. "File not found" is almost always a
disagreement about where you are, not a missing file.

4,757 rows and 22 columns: every monitor in the county, every day of 2025.

:::

## A data frame is a list of columns {#df-is-a-list}

```{r}
#| label: df-is-a-list
#| script: "A data frame is a list of columns"
is.list(la)                             # a data frame IS a list
length(la)                              # of 22 columns

sapply(la, class)                       # so sapply walks the columns
```

::: {.content-hidden when-format="revealjs"}

A data frame is a list of equal-length vectors with a class on top. `$` is
the list's own operator, which is why the same notation reaches a column of a
table and an element of a list.

That is why `sapply()` works here with no extra effort: hand it a data frame
and it walks the columns. On a file with 22 columns this beats scrolling
through `str()` output, and it puts two problems on screen at once. `Site.ID`
came in as an integer. `Date` came in as character.

:::

## The identifier trap {#identifiers}

```{r}
#| label: identifiers
#| script: "The identifier trap"
la$Site.ID[1]        # what R holds
```

::: {.content-hidden when-format="revealjs"}

The file says `060370016`. R says `60370016`.

`read.csv()` looked at a column of digits and decided it was a number, and
the leading zero of a number is meaningless, so it is gone. The identifier is
now wrong, silently, and it will fail to match every other file that spells it
correctly.

This is not an EPA quirk. FIPS codes, ZIP codes, CUSIPs, phone numbers and
account numbers all lead with zeros that R will strip on the way in.

:::

## Protect them at read time {#identifiers-fix}

```{r}
#| label: identifiers-fix
#| script: "Protect identifiers at read time"
la <- read.csv("data/epa_pm25_la_county_2025.csv",
               colClasses = c("Site.ID" = "character"))

la$Site.ID[1]
length(unique(la$Site.ID))    # eleven monitors in the county
```

::: {.content-hidden when-format="revealjs"}

`colClasses` names the columns you want read a particular way. Everything you
do not name is guessed as before.

Fix it at read time, not afterwards. Once the zero is gone, `as.character()`
gives you back `"60370016"`, which is a different string from `"060370016"`
and matches nothing. The information left the building when the file came in.

:::

## Dates {#dates}

```{r}
#| label: dates
#| script: "Dates, one line deep"
la$Date[1]
class(la$Date)

la$date <- as.Date(la$Date, format = "%m/%d/%Y")
class(la$date)
range(la$date)

la$month <- format(la$date, "%m")   # pull pieces back out of a date
```

::: {.content-hidden when-format="revealjs"}

`%m/%d/%Y` is a format string describing where the pieces are: two-digit
month, slash, two-digit day, slash, four-digit year. Get it wrong and you get
`NA`s rather than an error, so check `range()` afterwards.

Text dates sort character by character, so `"10/01"` lands before `"9/01"`.
Convert on the way in and the problem never arises.

:::

## One instrument, one method {#one-instrument}

```{r}
#| label: one-instrument
#| script: "One instrument, one method"
frm <- la[la$AQS.Parameter.Code == 88101 & la$POC == 1, ]

nrow(frm)
length(unique(frm$Local.Site.Name))
```

::: {.content-hidden when-format="revealjs"}

A site can run several instruments, and the file gives each its own row on
the same day. Average without filtering and you blend two measurement
methods, which is a number that describes nothing.

Parameter 88101 is the federal reference method; `POC == 1` keeps the first
instrument at each site. Eight of the eleven monitors ran one.

:::

## Sites by months {#sites-by-months}

```{r}
#| label: sites-by-months
#| script: "Sites by months"
tb <- tapply(frm$Daily.Mean.PM2.5.Concentration,
             list(frm$Local.Site.Name, frm$month),
             mean)

class(tb)
dim(tb)
round(tb[, 1:6], 1)
```

::: {.content-hidden when-format="revealjs"}

`tapply()` splits a vector by one or more grouping vectors and applies a
function to each cell. Give it two groupings and what comes back is a
**matrix**: eight sites down, twelve months across, one mean in each cell.

A long file of 4,757 rows is now a grid of 96 numbers you can read. That is
the move: the data frame is how the file arrives, the matrix is the shape the
question has.

:::

## Both margins at once {#both-margins}

```{r}
#| label: both-margins
#| script: "Both margins at once"
round(sort(apply(tb, 1, mean), decreasing = TRUE), 1)   # per site, all year
round(apply(tb, 2, mean), 1)                            # per month, all sites
```

::: {.content-hidden when-format="revealjs"}

Now the two margins answer two different questions with the same verb.

Down the rows: which monitors sit in the dirtiest air all year. Lancaster, out
in the desert, reads a third of what the basin sites do.

Across the columns: what the county did month by month. January is the worst
month at nearly every site at once, which is the signature of something
county-wide rather than one instrument drifting.

:::

## Two tables, one key {#merge}

Results in one table, the thing you want to attach in another. `merge()`
matches them on a shared column.

```{r}
#| label: merge
#| script: "Two tables, one key"
means <- data.frame(site = rownames(tb),
                    pm25 = round(apply(tb, 1, mean), 1))

coords <- unique(frm[, c("Local.Site.Name", "Site.Latitude", "Site.Longitude")])
names(coords) <- c("site", "lat", "lon")

nrow(means); nrow(coords)          # count BEFORE
sites <- merge(means, coords, by = "site")
nrow(sites)                        # and after
head(sites, 3)
```

::: {.content-hidden when-format="revealjs"}

`by` names the column the two tables share. Rows whose keys match are glued
side by side; the result carries the columns of both.

Eight rows in, eight rows out, so nothing was lost. **That check is the whole
discipline of joining.** Write down the row count you expect before you run
it, and compare. A join has two silent failure modes and neither one errors.

:::

## What a join does silently {#merge-hazards}

```{r}
#| label: merge-hazards
#| script: "What a join does silently"
partial <- coords[coords$site != "Compton", ]   # pretend one site is missing

nrow(merge(means, partial, by = "site"))                # dropped
nrow(merge(means, partial, by = "site", all.x = TRUE))  # kept, with NA
```

::: {.content-hidden when-format="revealjs"}

**Rows can disappear.** By default `merge()` keeps only keys present in
both tables, so a site missing from the lookup takes its result with it.
Seven rows came back where you asked about eight, with no error and no
warning. `all.x = TRUE` keeps every row of the first table and fills the
missing side with `NA`, which is almost always what you want: an `NA` you
can see beats a row you cannot.

**Rows can also multiply.** If the second table has two rows for one site,
every matching row of the first is repeated to pair with each. Ask for eight
and get twelve. That is why `unique()` is in the line that built `coords`.

Keys have to be the same type, too. A site id read as a number will not
match the same id read as text, and that failure looks exactly like "no
matching rows".

:::

::: {.content-visible when-format="revealjs"}

- Default keeps only keys in **both** tables: rows vanish, silently.
- `all.x = TRUE` keeps the left table and fills with `NA`.
- A duplicated key on the right **multiplies** rows instead.
- Count rows before and after. Every time.

:::

## One line, every file {#lapply-files}

```{r}
#| label: lapply-files
#| script: "One line, every file"
files <- list.files("data", pattern = "\\.csv$")
files

sapply(files, function(f) nrow(read.csv(file.path("data", f))))
```

::: {.content-hidden when-format="revealjs"}

`list.files()` gives you the names, and `lapply()` or `sapply()` runs the same
call on each one.

This is the shape of most real data collection: one file per year, per state
or per month, and one function that handles any single file. You get it right
once, then hand the function the list. `do.call(rbind, ...)` on the result
stacks them into one frame.

:::

## Writing your answer out {#write}

```{r}
#| label: write
#| script: "Writing your answer out"
#| eval: false
answers <- data.frame(
  site  = rownames(tb),
  mean  = round(apply(tb, 1, mean), 1)
)

write.csv(answers, "results.csv", row.names = FALSE)
```

::: {.content-hidden when-format="revealjs"}

`write.csv()` is `read.csv()` run backwards. `row.names = FALSE` stops R
adding a column of row numbers you never asked for, which is the single most
common complaint about files written from R.

Round before you write, not after. A number carries every decimal it has
until you cut it, so a mean lands in the file as `19.386666666667` unless you
say `round(x, 1)` first.

:::

## Verify {#verify}

> **Ask what shape the question has, then put the data in it.**

::: {.content-hidden when-format="revealjs"}

Three containers, three jobs, and the cost of the wrong one:

| Container | Holds | Use it when |
|:--|:--|:--|
| Matrix | One type, two dimensions | Every cell is the same measurement |
| List | Anything, any length | The pieces do not line up |
| Data frame | A list of equal-length columns | Rows are observations, columns are variables |

The file arrived as a data frame because that is how files arrive. The
question was about sites and months, so the answer wanted a matrix, and
`tapply()` was the bridge. Getting that right is most of what separates two
lines from twenty.

And the check that costs nothing: run `sapply(d, class)` before you compute.
Today it would have told you the site identifier had become a number and the
date had not become a date, at a point where each was a one-line repair.

:::

::: {.content-visible when-format="revealjs"}

| Container | Holds | Use it when |
|:--|:--|:--|
| Matrix | One type, two dimensions | Every cell is the same measurement |
| List | Anything, any length | The pieces do not line up |
| Data frame | A list of columns | Rows are observations |

`sapply(d, class)` before you compute. Every file, every time.

:::

## Practice {#practice}

::: {.content-visible when-format="revealjs"}

Everything below is in the live script, with solutions.

:::

::: {.content-hidden when-format="revealjs"}

::: {.panel-tabset}

### Practice

```{r}
#| label: practice
#| exercise: true
#| script: "Practice"
#| eval: false
# Everything below uses la, frm and tb, built earlier in this script.

# Matrices
#  1. Build a 3 x 4 matrix of the numbers 1 to 12, filled across the rows.
#  2. Same matrix: what is the sum of each row? Of each column?
#  3. Same matrix: divide every cell by its own column's total.
#  4. Add a fourth row of zeros with rbind(). Then add a fifth column.
#  5. Run rbind(c(1, 2), c("a", "b")). What class is the result, and why?

# Lists
#  6. Build a list holding your name, the numbers 1 to 5, and TRUE.
#  7. Same list: pull the numeric vector out two different ways.
#  8. What class does l["readings"] return? Check with class().
#  9. Split "2025-01-07" on the dash and get back a vector of three pieces.
# 10. unlist() the list from question 6. What happened to the numbers?

# apply and friends
# 11. Which column of la came in as character? (One line, all 22 at once.)
# 12. How many rows does each site have in la? (One line.)
# 13. Which month had the highest county-wide mean, and what was it?
# 14. Which site had the single dirtiest month of 2025?
# 15. Rebuild tb with median instead of mean. Where do the two disagree
#     most, and what does that tell you about those days?
# 16. One line: the number of rows in every CSV in your data folder.
```

### Solutions

```{r}
#| label: practice-solutions
#| purl: false
#| eval: false
# 1
mm <- matrix(1:12, nrow = 3, byrow = TRUE)

# 2
apply(mm, 1, sum); apply(mm, 2, sum)

# 3
sweep(mm, 2, colSums(mm), "/")

# 4
mm <- rbind(mm, 0); mm <- cbind(mm, 0)

# 5  character: one vector underneath, so one type throughout
class(rbind(c(1, 2), c("a", "b")))

# 6
l3 <- list(name = "Ariel", nums = 1:5, ok = TRUE)

# 7
l3$nums; l3[["nums"]]

# 8  a list of length one, not the vector: single brackets stay wrapped
class(l["readings"])

# 9
strsplit("2025-01-07", "-")[[1]]

# 10  everything became text: one type, and "TRUE" is now a word
unlist(l3)

# 11  a data frame is a list of columns, so sapply walks them
names(which(sapply(la, class) == "character"))

# 12
table(la$Local.Site.Name)

# 13
month_means <- apply(tb, 2, mean, na.rm = TRUE)
month_means[which.max(month_means)]

# 14
which(tb == max(tb, na.rm = TRUE), arr.ind = TRUE)

# 15  the mean sits above the median wherever a few days ran very high
tb_med <- tapply(frm$Daily.Mean.PM2.5.Concentration,
                 list(frm$Local.Site.Name, frm$month), median)
round(sort(apply(tb - tb_med, 1, mean, na.rm = TRUE), decreasing = TRUE), 2)

# 16
sapply(list.files("data", pattern = "\\.csv$"),
       function(f) nrow(read.csv(file.path("data", f))))
```

:::

:::

## Quick reference {#quick-reference}

::: {.content-hidden when-format="revealjs"}

**Matrices**

| Code | What it does |
|:--|:--|
| `matrix(x, nrow =, byrow =)` | Build one; `byrow = TRUE` fills across |
| `dim(m)`, `t(m)` | Its shape; flip rows and columns |
| `m[i, j]`, `m[i, ]`, `m[, j]` | Cell, row, column; blank means all |
| `rownames(m)`, `colnames(m)` | Names on the margins |
| `rbind(a, b)`, `cbind(a, b)` | Stack; matched by position, not name |
| `rowMeans(m)`, `colSums(m)` | The fast special cases |
| `Matrix(m, sparse = TRUE)` | Store only the non-zeros |

**Lists**

| Code | What it does |
|:--|:--|
| `list(a = , b = )` | Build one; anything, any length |
| `l[["a"]]`, `l$a` | The element itself |
| `l["a"]` | A list of length one |
| `str(l)`, `lengths(l)` | What is in there; how long each part is |
| `unlist(l)` | Flatten to one vector, one type |
| `do.call(rbind, l)` | Stack the elements into a matrix or frame |

**One function, many things**

| Code | What it does |
|:--|:--|
| `apply(m, 1, f)` | Once per row; `2` for columns |
| `split(x, g)` | A list, one element per group |
| `lapply(l, f)` | A list in, a list of the same length out |
| `sapply(l, f)` | The same, simplified to a vector or matrix |
| `tapply(x, list(g1, g2), f)` | A matrix of `f` over two groupings |
| `sapply(d, class)` | The class of every column at once |

**Files**

| Code | What it does |
|:--|:--|
| `read.csv(f)` | Read a CSV, guessing every column's type |
| `colClasses = c(id = "character")` | Stop it guessing on identifiers |
| `as.Date(x, format = "%m/%d/%Y")` | Text to a real date |
| `write.csv(d, f, row.names = FALSE)` | Write one back out |
| `list.files(dir, pattern =)` | The names, ready for `lapply()` |
| `merge(a, b, by = "id")` | Glue two tables on a shared key |
| `merge(a, b, by =, all.x = TRUE)` | Keep every row of `a`; fill the rest with `NA` |

**Links**

- Session materials: [slides](03-r-essentials-ii-slides.html) ·
  [live script](03-r-essentials-ii.R) ·
  [EPA data](data/epa_pm25_la_county_2025.csv) ·
  [where it came from](data/SOURCE.md) ·
  [EPA's download tool](https://www.epa.gov/outdoor-air-quality-data/download-daily-data)
- [Syllabus](../syllabus.qmd) · [full schedule](../index.qmd#schedule)

:::

::: {.content-visible when-format="revealjs"}

| Code | What it does |
|:--|:--|
| `matrix(x, nrow =, byrow =)` | Build one |
| `m[i, j]` · `rbind()` · `cbind()` | Index; stack; widen |
| `l[["a"]]` vs `l["a"]` | The element vs a list of one |
| `unlist(l)` · `do.call(rbind, l)` | Flatten; stack |
| `apply(m, 1, f)` · `sapply(l, f)` | A margin; a list |
| `tapply(x, list(g1, g2), f)` | Two groupings, one matrix |
| `colClasses = c(id = "character")` | Protect identifiers at read time |

:::

```{r}
#| purl: false
#| include: false
source("_purl.R")
purl_session("03-r-essentials-ii.qmd", "03-r-essentials-ii.R")
```
