---
title: "3 · R essentials II"
subtitle: "Tuesday, September 1, 2026"
pagetitle: "AEM 6850 · Session 3 · R essentials II"
execute:
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
  revealjs:
    output-file: 03-r-essentials-ii-slides.html
    theme: [default, slides.scss]
    template-partials:
      - title-slide.html
    footer: "AEM 6850 · Session 3"
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    code-overflow: scroll
    history: false
    link-external-newwindow: true
---

```{r}
#| purl: false
#| include: false
knitr::opts_chunk$set(
  comment = "#>",
  collapse = FALSE,
  fig.width = 8,
  fig.height = 4,
  fig.align = "center",
  dpi = 150
)
```

::: {.content-hidden when-format="revealjs"}

[Slides](03-r-essentials-ii-slides.html) · [Live script](03-r-essentials-ii.R) ·
[Source](https://github.com/ArielOrtizBobea/aem6850/blob/main/fall-2026/sessions/03-r-essentials-ii.qmd)

Everything below is in the live script. Run it line by line at home, as many
times as it takes.

:::

## Objectives {#objectives}

By the end of this session you can:

- Build, index and combine matrices.
- Build and index lists, and tell `[` from `[[`.
- Run one function over every row, column, group or file.
- Read a real CSV, fix its types, and write results back out.
- Join two tables, and check the row count before and after.

## Matrices {#matrices}

A **matrix** is one vector with dimensions.

```{r}
#| label: matrices
#| script: "Matrices"
m <- matrix(1:6, nrow = 2)
m
c(dim(m), nrow(m), ncol(m))

matrix(1:6, nrow = 2, byrow = TRUE)   # filled across, not down
matrix(1:2, nrow = 2, ncol = 3)       # too few values: R recycles them
```

::: {.content-hidden when-format="revealjs"}

R folds the vector down the columns unless you say `byrow = TRUE`. Get that
backwards and every value lands in the wrong cell, with no error.

The last line is the one that bites. Give `matrix()` fewer values than cells
and it **recycles** them to fill the shape, silently, as long as the count
divides evenly. When it does not divide evenly you get a warning, which is
the friendlier case. Check `dim()` against what you meant.

:::

## Indexing a matrix {#matrix-indexing}

```{r}
#| label: matrix-indexing
#| script: "Indexing a matrix"
m[2, 3]     # row 2, column 3
m[2, ]      # all of row 2
m[, 3]      # all of column 3

class(m[, 3])                # a plain vector: the dimension was dropped
class(m[, 3, drop = FALSE])  # still a matrix

head(m, 1)                   # first rows; tail() for the last
```

::: {.content-hidden when-format="revealjs"}

Two coordinates, `m[row, column]`, and a blank means all of them. The same
rule you use on a data frame.

Note that `m[2, ]` came back as a plain vector, not a one-row matrix. R drops
the dimension when you ask for a single row or column, which is convenient
at the console and a bug inside a function that expected a matrix.
`drop = FALSE` keeps the shape.

:::

## Subsetting by position, name, and logical {#matrix-names}

```{r}
#| label: matrix-names
#| script: "Three ways to pick: position, name, logical"
temps <- matrix(c(12, 15, 19,
                  14, 17, 22), nrow = 2, byrow = TRUE)
rownames(temps) <- c("Compton", "Reseda")
colnames(temps) <- c("Jan", "Feb", "Mar")
temps

temps["Reseda", "Feb"]     # by name, not position
temps[, c("Jan", "Feb")]   # several at once
temps[temps > 15]          # by logical: returns a vector
```

::: {.content-hidden when-format="revealjs"}

Three notations, one operation. Position counts, names read, and a logical
test picks whatever passes.

Names are worth setting the moment a matrix means something. `temps[2, 2]`
is a number you have to decode; `temps["Reseda", "Feb"]` says what it is,
and it keeps working when the rows get reordered.

The logical form behaves differently from the other two: it hands back a
plain vector, not a matrix, because the cells that pass do not generally
form a rectangle. Ask for the positions instead with
`which(temps > 15, arr.ind = TRUE)` when you need to know *where* they were.

:::

## The diagonal {#diag}

```{r}
#| label: diag
#| script: "The diagonal"
sq <- matrix(1:9, nrow = 3)
sq

diag(sq)          # pull the diagonal out
diag(3)           # build a 3 x 3 identity matrix
diag(c(4, 5, 6))  # build a diagonal matrix from a vector
```

::: {.content-hidden when-format="revealjs"}

`diag()` does two jobs depending on what you hand it. Give it a matrix and
it extracts the diagonal. Give it a single number and it builds an identity
matrix of that size. Give it a vector and it builds a diagonal matrix with
that vector down the middle and zeros elsewhere.

Diagonal matrices come up constantly once you write estimators by hand:
weights, variances, and the identity that sits in a projection formula are
all diagonal. Most of a diagonal matrix is zeros, which is worth remembering
for later today.

:::

## Matrix arithmetic {#matrix-math}

```{r}
#| label: matrix-math
#| script: "Matrix arithmetic"
temps * 9/5 + 32     # every cell at once, Celsius to Fahrenheit
t(temps)             # flip rows and columns
```

::: {.content-hidden when-format="revealjs"}

Arithmetic runs cell by cell, exactly as it does on a vector, because
underneath a matrix is a vector.

This is why images and grids are stored as matrices. A photograph is a
matrix of pixel values, and making it brighter is one multiplication. A
weather grid is a matrix of temperatures, and converting the whole grid to
Fahrenheit is the line above.

:::

## Combining matrices {#matrix-combine}

```{r}
#| label: matrix-combine
#| script: "Combining matrices"
more <- matrix(c(9, 11, 16), nrow = 1,
               dimnames = list("Pasadena", c("Jan", "Feb", "Mar")))

rbind(temps, more)              # a new row: same columns
cbind(temps, Apr = c(21, 24))   # a new column: same rows

Matrix::bdiag(diag(2), matrix(1, 2, 2))   # blocks down the diagonal
```

::: {.content-hidden when-format="revealjs"}

`rbind()` stacks matrices on top of each other and `cbind()` puts them side
by side. The shared margin has to agree: `rbind()` needs the same number of
columns, `cbind()` the same number of rows.

R matches columns by **position, not by name**. If the second matrix carries
the same three months in a different order, `rbind()` will stack them anyway
and put February's numbers under the January heading, with no warning.
Check the names agree before you stack, not after.

:::

## A matrix holds one type {#matrix-type}

```{r}
#| label: matrix-type
#| script: "One matrix, one type"
rbind(c(1, 2, 3),
      c("a", "b", "c"))
```

::: {.content-hidden when-format="revealjs"}

One vector underneath means one class throughout. Add a row of text and
every number becomes text, silently, with no warning. It is the same
conversion that turns a numeric column into text when one entry is a word.

A data frame does not do this, because its columns are separate vectors.
That is the whole reason both containers exist: tables of mixed data are
data frames, and grids of one measurement are matrices.

:::

## Sparse matrices {#sparse}

Most of the cells are zero. Store only the ones that are not, and both the
memory and the arithmetic get cheaper.

```{r}
#| label: sparse
#| script: "Sparse matrices"
library(Matrix)
set.seed(1); n <- 2000
d <- matrix(0, n, n); d[sample(n * n, n * n * 0.01)] <- 1   # 1% non-zero
s <- Matrix(d, sparse = TRUE); v <- rnorm(n)

c(dense = format(object.size(d), units = "MB"),
  sparse = format(object.size(s), units = "MB"))

c(dense  = system.time(for (i in 1:20) d %*% v)[["elapsed"]],
  sparse = system.time(for (i in 1:20) s %*% v)[["elapsed"]])
```

::: {.content-hidden when-format="revealjs"}

A dense matrix stores every cell, including all the zeros. A **sparse**
matrix stores only the non-zero values and the positions they sit in. Same
numbers, same indexing, a fraction of the memory.

**The arithmetic gets faster too**, which is the part that pays. Multiplying
skips the zeros instead of multiplying by them, so the same product runs
more than ten times quicker here. Two things you will do constantly are matrix
multiplications underneath: **aggregating** values up to groups, which is a
matrix of weights times a vector of values, and any estimator written in
matrix form.

You meet these the moment a matrix gets large and mostly empty: a
county-by-county adjacency matrix, a word-by-document count, a design matrix
of fixed effects with one 1 per row, or extraction weights mapping grid cells
to regions. At `n = 2000` the saving is a convenience. At `n = 100,000` it
decides whether the code runs at all.

`Matrix` ships with R, so there is nothing to install. Its matrices index and
multiply like ordinary ones; `as.matrix()` converts back when a function
insists on dense input.

:::

## Lists {#lists}

A **list** holds anything, in any mix, at any length.

```{r}
#| label: lists
#| script: "Lists"
l <- list(site = "Compton", readings = c(53.2, 33.6, 47.0), clean = TRUE)
str(l)
length(l)      # three elements...
lengths(l)     # ...of these lengths
```

::: {.content-hidden when-format="revealjs"}

Text, a numeric vector, and a logical, side by side in one object. A vector
could not do this: it would convert everything to text. A matrix could not
either: one type throughout, and every row the same length.

That freedom is the point. Lists are what R uses whenever the pieces do not
line up.

:::

## Single and double brackets {#brackets}

```{r}
#| label: brackets
#| script: "One bracket or two"
l["readings"]     # a LIST of length one
l[["readings"]]   # the vector itself
l$readings        # the same thing, less typing

class(l["readings"])
class(l[["readings"]])
```


::: {.content-hidden when-format="revealjs"}



The mistake to expect is arithmetic on a single-bracket result:
`mean(l["readings"])` fails, because you handed `mean()` a list. `$` and
`[[ ]]` do the same job; `[[ ]]` is the one that also takes a variable, which
matters as soon as the name is computed rather than typed.

:::

## A list is a chest of drawers {#drawers}

Think of the list as a chest of drawers. **Single brackets take the drawer
out of the chest**, and a drawer is still a drawer. **Double brackets open
the drawer and hand you what is inside.**

```{=html}
<svg viewBox="0 0 620 128" role="img" aria-label="A chest of drawers: single brackets take a whole drawer out, double brackets hand you its contents."
     style="max-width:100%;height:auto;font-family:inherit">
  <g fill="none" stroke="currentColor" stroke-width="2">
    <rect x="10" y="14" width="92" height="100" rx="4"/>
    <rect x="20" y="24" width="72" height="24" rx="3"/>
    <rect x="20" y="54" width="72" height="24" rx="3"/>
    <rect x="20" y="84" width="72" height="24" rx="3"/>
    <line x1="47" y1="36" x2="65" y2="36"/><line x1="47" y1="66" x2="65" y2="66"/>
    <line x1="47" y1="96" x2="65" y2="96"/>
    <path d="M116 44 L162 44"/><path d="M155 39 L162 44 L155 49"/>
    <path d="M116 92 L162 92"/><path d="M155 87 L162 92 L155 97"/>
    <rect x="176" y="30" width="86" height="26" rx="3"/>
    <line x1="210" y1="43" x2="228" y2="43"/>
    <circle cx="192" cy="92" r="8"/><circle cx="219" cy="92" r="8"/><circle cx="246" cy="92" r="8"/>
  </g>
  <g font-size="12" fill="currentColor" font-family="ui-monospace,monospace" text-anchor="middle">
    <text x="139" y="36">l["a"]</text><text x="139" y="84">l[["a"]]</text>
  </g>
  <g font-size="12.5" fill="currentColor">
    <text x="278" y="47">the drawer, still a list</text>
    <text x="278" y="97">what was inside it</text>
  </g>
</svg>
```

## Lists inside lists {#nested-lists}

```{r}
#| label: nested-lists
#| script: "Lists inside lists"
monitor <- list(
  id   = "060371302",
  site = list(name = "Compton", lat = 33.90, lon = -118.21),
  pm   = c(53.2, 33.6, 47.0)
)

monitor$site$name
monitor[["site"]][["lat"]]
str(monitor)
```

::: {.content-hidden when-format="revealjs"}

An element of a list can be another list, to any depth. Chain the operators
to walk down: `$` after `$`, or `[[ ]]` after `[[ ]]`.

This is exactly the shape of the JSON that web APIs return, which is why the
lesson pays off well beyond R. `str()` is how you find your way around one
you did not build.

:::

## Functions often return lists {#lists-from-functions}

```{r}
#| label: lists-from-functions
#| script: "Where lists come from"
x <- c(1, 2, 3, 4)
y <- c(1, 3, 2, 5)
fit <- lm(y ~ x)

class(fit)
names(fit)          # a list underneath
fit$coefficients
```

```{r}
#| label: lists-from-functions-plot
#| script: "What the fitted list is describing"
#| fig-height: 3.6
#| fig-alt: "Four points scattered upward with the fitted least-squares line through them, and dashed vertical segments joining each point to the line to show its residual."
plot(x, y, pch = 16, xlab = "x", ylab = "y")
abline(fit, col = "#b31b1b", lwd = 2)
segments(x, y, x, fitted(fit), lty = 2, col = "grey55")   # the residuals
```

::: {.content-hidden when-format="revealjs"}

Most functions that return more than one thing return a list, because a list
is the only container that can hold a number, a vector and a data frame at
once.

A fitted model is a list with a class on it. `names()` shows you the parts
and `$` takes one out. Every modelling function you meet this semester
follows this pattern, so knowing lists means never being stuck at "what did
this thing give me?"

The plot is what those numbers describe. `fit$coefficients` is the line,
`fitted(fit)` is where the line sits above each `x`, and `fit$residuals` is
the length of each dashed segment. Three elements of one list, all visible
at once.

:::

## Combining list elements {#list-combine}

```{r}
#| label: list-combine
#| script: "Combining list elements"
pieces <- list(a = c(1, 2), b = c(3, 4), c = c(5, 6))

unlist(pieces)              # one flat vector, names kept
do.call(rbind, pieces)      # one matrix, one row per element
```

::: {.content-hidden when-format="revealjs"}

`unlist()` flattens a list into a single vector. It is convenient and it is
lossy: everything becomes one type, and any structure below the top level
disappears.

`do.call(f, list)` calls `f` with the list's elements as its arguments, so
`do.call(rbind, pieces)` is `rbind(pieces$a, pieces$b, pieces$c)` without
your having to know how many there are or what they are called. That is the
standard way to turn a list of rows into a matrix, or a list of data frames
into one data frame.

:::

## Exercise 1 {#exercise-1}

::: {.panel-tabset}

### Exercise

```{r}
#| label: exercise-1
#| exercise: true
#| script: "Exercise 1 (5 minutes)"
#| eval: false
# 1. Build a diagonal matrix with the numbers 1 to 5 down the middle.
#
# 2. From m (the 2 x 3 matrix), pull the second column two ways: once
#    as a vector, once still a matrix.
#
# 3. l["readings"] and l[["readings"]]: which one can you take a mean of?
#    Find out with class(), not from memory.
```

### Solution

```{r}
#| label: exercise-1-solution
#| purl: false
diag(1:5)
m[, 2]; m[, 2, drop = FALSE]
class(l["readings"]); class(l[["readings"]])   # list vs numeric
```

:::

## apply(): over rows and columns {#apply}

You built these containers and picked things out of them. Now the other
half: run one function over every row, column, group or file.

```{r}
#| label: apply
#| script: "apply(): once per row, once per column"
apply(temps, 1, mean)              # 1 = rows
apply(temps, 2, mean)              # 2 = columns

rowMeans(temps); colMeans(temps)   # same answers, named and faster
rowSums(temps)
```

::: {.content-hidden when-format="revealjs"}

`apply(m, MARGIN, f)` runs `f` over one margin of a matrix and collects the
results.

The margin numbers are the bracket coordinates. You write `m[row, column]`,
so 1 is rows and 2 is columns. Margin 1 gives you one answer per row; margin
2 gives you one per column.

`rowMeans()` and `colMeans()` are fast special cases of exactly this.

:::

## apply() with any function {#apply-any}

```{r}
#| label: apply-any
#| script: "Any function, not just mean"
apply(temps, 2, max)
apply(temps, 2, range)                       # two numbers per column
apply(temps, 2, function(x) max(x) - min(x)) # write your own, inline
```

::: {.content-hidden when-format="revealjs"}

The function is an argument like any other. Pass it by name with no
parentheses, or write it inline with `function(x)` when it is too small to
deserve a name. Inside, `x` is one row or one column.

`range` returns two numbers per column, so `apply()` hands back a matrix
rather than a vector. What you get out depends on what your function
returns.

:::

## `split()`: one group per element {#split}

```{r}
#| label: split
#| script: "split(): one group per element"
x <- c(12, 15, 19, 14, 17, 22)
g <- c("Compton", "Compton", "Compton", "Reseda", "Reseda", "Reseda")

split(x, g)
```

::: {.content-hidden when-format="revealjs"}

`split()` cuts an object into a list, one element per group. Those are the
same six numbers that went into `temps`, arranged as a list of two vectors
instead of a two-row matrix.

That is the choice in miniature. A matrix needs every group to be the same
length; a list does not. Real groups are rarely the same size, which is why
`split()` returns a list.

:::

## `lapply()`: a list in, a list out {#lapply}

```{r}
#| label: lapply
#| script: "lapply(): a list in, a list out"
by_site <- split(x, g)

lapply(by_site, mean)
```

::: {.content-hidden when-format="revealjs"}

`lapply(x, f)` runs `f` on every element of `x` and returns a list of the
same length, with the same names. One call, one result per group, and no
loop written by hand.

Inside the function, the argument is one element: here, one site's readings.
You write the code once, for one group, and `lapply()` handles the
repetition.

:::

## `sapply()`: the same, simplified {#sapply}

```{r}
#| label: sapply
#| script: "sapply(): the same, simplified"
sapply(by_site, mean)      # a named vector, not a list
sapply(by_site, length)
sapply(by_site, range)     # two numbers each, so a matrix comes back

apply(temps, 1, mean)      # same answers, from the matrix
```

::: {.content-hidden when-format="revealjs"}

`sapply()` is `lapply()` plus one step: when every result is a single value,
it returns a named vector instead of a list. When every result is the same
longer length, it returns a matrix.

Use `sapply()` when you want something to print or compute with. Use
`lapply()` when each result is bigger than one value, or when you want the
shape guaranteed. `sapply()` decides what to return based on what it gets,
which is convenient at the console and a hazard inside a script.

The last line is the same question asked of the matrix. Two containers, two
verbs, one answer: `apply()` walks a margin, `sapply()` walks a list.

:::

## Exercise 2 {#exercise-2}

::: {.panel-tabset}

### Exercise

```{r}
#| label: exercise-2
#| exercise: true
#| script: "Exercise 2 (5 minutes)"
#| eval: false
# 1. One line: the highest reading in each city. One line: the highest
#    in each month.
#
# 2. One line: the range of each site in by_site. Why does that come
#    back as a matrix?
```

### Solution

```{r}
#| label: exercise-2-solution
#| purl: false
apply(temps, 1, max)   # margin 1 = rows = cities
apply(temps, 2, max)   # margin 2 = columns = months

sapply(by_site, range)        # two numbers per site...
dim(sapply(by_site, range))   # ...so sapply stacks them into a matrix
```

:::

## Reading a CSV {#read}

```{r}
#| label: read
#| script: "Read the file"
la <- read.csv("data/epa_pm25_la_county_2025.csv")
dim(la)
```

```{r}
#| label: read-fallback
#| script: "If you do not have the file yet"
#| eval: false
# Uncomment and run once. Same file, fetched instead of clicked.
# dir.create("data", showWarnings = FALSE)
# download.file(paste0("https://arielortizbobea.github.io/aem6850/",
#                      "fall-2026/sessions/data/epa_pm25_la_county_2025.csv"),
#               "data/epa_pm25_la_county_2025.csv")
```

::: {.content-hidden when-format="revealjs"}

`read.csv()` takes a path relative to the working directory, which `getwd()`
prints and `list.files()` lets you check. "File not found" is almost always a
disagreement about where you are, not a missing file.

4,757 rows and 22 columns: every monitor in the county, every day of 2025.

:::

## A data frame is a list of columns {#df-is-a-list}

```{r}
#| label: df-is-a-list
#| script: "A data frame is a list of columns"
is.list(la)             # a data frame IS a list
length(la)              # of 22 columns
head(sapply(la, class)) # so sapply walks the columns
```

::: {.content-hidden when-format="revealjs"}

A data frame is a list of equal-length vectors with a class on top. `$` is
the list's own operator, which is why the same notation reaches a column of a
table and an element of a list.

That is why `sapply()` works here with no extra effort: hand it a data frame
and it walks the columns. On a file with 22 columns this beats scrolling
through `str()` output, and it puts two problems on screen at once. `Site.ID`
came in as an integer. `Date` came in as character.

:::

## Identifiers read as numbers {#identifiers}

```{r}
#| label: identifiers
#| script: "The identifier trap"
la$Site.ID[1]        # what R holds
```

::: {.content-hidden when-format="revealjs"}

The file says `060370016`. R says `60370016`.

`read.csv()` looked at a column of digits and decided it was a number, and
the leading zero of a number is meaningless, so it is gone. The identifier is
now wrong, silently, and it will fail to match every other file that spells it
correctly.

This is not an EPA quirk. FIPS codes, ZIP codes, CUSIPs, phone numbers and
account numbers all lead with zeros that R will strip on the way in.

:::

## colClasses: setting types at read time {#identifiers-fix}

```{r}
#| label: identifiers-fix
#| script: "Protect identifiers at read time"
la <- read.csv("data/epa_pm25_la_county_2025.csv",
               colClasses = c("Site.ID" = "character"))

la$Site.ID[1]

unique(la$Site.ID)            # unique() drops repeats
length(unique(la$Site.ID))    # ...so this counts monitors: eleven
```

::: {.content-hidden when-format="revealjs"}

`colClasses` names the columns you want read a particular way. Everything you
do not name is guessed as before.

Fix it at read time, not afterwards. Once the zero is gone, `as.character()`
gives you back `"60370016"`, which is a different string from `"060370016"`
and matches nothing. The information left the building when the file came in.

:::

## Dates {#dates}

```{r}
#| label: dates
#| script: "Dates, one line deep"
la$Date[1]
class(la$Date)

la$date <- as.Date(la$Date, format = "%m/%d/%Y")
class(la$date)
range(la$date)
```

## Pulling pieces out of a date {#dates-parts}

```{r}
#| label: dates-parts
#| script: "Pulling pieces back out of a date"
d <- as.Date("2025-01-07")

c(month = format(d, "%m"), name = format(d, "%b"),
  year = format(d, "%Y"), doy = format(d, "%j"),
  day = weekdays(d))

la$month <- format(la$date, "%m")
```

::: {.content-hidden when-format="revealjs"}

`format()` runs the parsing backwards: give it a code and it returns that
piece as text. `weekdays()` and `months()` are the two shortcuts worth
remembering, because they are what you want for seasonality and for
day-of-week effects.

:::

`%m/%d/%Y` is a format string describing where the pieces are:

```{=html}
<svg viewBox="0 0 560 132" role="img" aria-label="The date 01/07/2025 with each piece labelled by its format code: %m for month, %d for day, %Y for four-digit year."
     style="max-width:100%;height:auto;font-family:inherit">
  <g fill="none" stroke="currentColor" stroke-width="2">
    <rect x="16" y="16" width="86" height="96" rx="8"/>
    <line x1="16" y1="42" x2="102" y2="42"/>
    <line x1="42" y1="16" x2="42" y2="8"/><line x1="76" y1="16" x2="76" y2="8"/>
  </g>
  <text x="59" y="36" text-anchor="middle" font-size="12" fill="currentColor">JAN</text>
  <text x="59" y="88" text-anchor="middle" font-size="34" fill="currentColor">7</text>

  <text x="150" y="58" font-size="30" fill="currentColor" font-family="ui-monospace,monospace">01/07/2025</text>

  <g stroke="currentColor" stroke-width="1.5" fill="none" opacity="0.6">
    <path d="M171 70 L171 86"/><path d="M228 70 L228 86"/><path d="M312 70 L312 86"/>
  </g>
  <g font-size="14" fill="currentColor" font-family="ui-monospace,monospace" text-anchor="middle">
    <text x="171" y="103">%m</text><text x="228" y="103">%d</text><text x="312" y="103">%Y</text>
  </g>
  <g font-size="11.5" fill="currentColor" opacity="0.72" text-anchor="middle">
    <text x="171" y="121">month</text><text x="228" y="121">day</text><text x="312" y="121">year</text>
  </g>
  <g font-size="12.5" fill="currentColor">
    <text x="392" y="46">Get a code wrong and</text>
    <text x="392" y="66">you get NAs, not an error.</text>
    <text x="392" y="92" opacity="0.72">So check range() after.</text>
  </g>
</svg>
```

## Dates are numbers {#dates-arithmetic}

```{r}
#| label: dates-seq
#| script: "Dates are numbers, so they do arithmetic"
d + 30                                # thirty days later
as.Date("2025-03-01") - d             # how far apart
seq(d, by = "week", length.out = 4)   # a sequence of dates
seq(as.Date("2025-01-01"), as.Date("2025-12-01"), by = "month")
```

::: {.content-hidden when-format="revealjs"}

Underneath, a `Date` is a count of days since January 1, 1970. That is why
`+` and `-` work, why `seq()` can step by `"day"`, `"week"`, `"month"` or
`"year"`, and why sorting finally puts dates in time order. Text dates sort
character by character, so `"10/01"` lands before `"9/01"`.

:::

::: {.content-hidden when-format="revealjs"}



Two-digit month, slash, two-digit day, slash, four-digit year. Get a code
wrong and you get `NA`s rather than an error, so check `range()` afterwards.

:::

## Checking for duplicate rows {#one-instrument}

Before averaging anything, find out what one row is.

```{r}
#| label: one-instrument
#| script: "One row per what?"
key <- la[, c("Site.ID", "date")]

anyDuplicated(key)     # 0 means none; otherwise the first repeated row
sum(duplicated(key))   # how many rows repeat a site-day
```

```{r}
#| label: one-instrument-fix
#| script: "Keep one instrument per site"
frm <- la[la$AQS.Parameter.Code == 88101 & la$POC == 1, ]

anyDuplicated(frm[, c("Site.ID", "date")])   # 0: one row per site-day now
```

::: {.content-hidden when-format="revealjs"}

`duplicated()` returns `TRUE` for every row that repeats an earlier one, so
`sum()` counts them. `anyDuplicated()` is the fast version: it stops at the
first repeat and returns its position, or `0` if the rows are unique. Use
`anyDuplicated()` when you want a yes or no, and `duplicated()` when you
want to see the offenders: `key[duplicated(key), ]`.

**There are 1,270 repeated site-days here.** Nothing warned you: the file is
correct, it just is not one row per thing you think you are counting.

The reason is that a site can run several instruments, and the file gives
each its own row on the same day. Take a mean now and you average a site
with itself, weighted by how many instruments it happens to run. That is not
a number about air quality; it is a number about equipment.

The fix keeps one instrument per site. Parameter 88101 is the federal
reference method, and `POC == 1` takes the first instrument at each site.
Then the check passes, and every row is one site on one day.

Ask this of every file. "What is one row?" is the question that catches
duplicated keys, and it is the same check a join needs later.

:::

::: {.content-visible when-format="revealjs"}

- More rows than site-days: something is duplicated, with no warning.
- A site runs several instruments, each with a row on the same day.

:::

## tapply(): a matrix from two groupings {#sites-by-months}

Numbers to summarise, what to group them by, a function to run on each
group. Two groupings give one value per combination.

```{r}
#| label: sites-by-months
#| script: "Sites by months"
tb <- tapply(frm$Daily.Mean.PM2.5.Concentration,
             list(frm$Local.Site.Name, frm$month), mean)

dim(tb)
round(tb[, 1:6], 1)
```

::: {.content-hidden when-format="revealjs"}

Read the call left to right. The first argument is the column you want
summarised, here the concentration. The second says how to cut it up:
`list(site, month)` means group by site *and* by month, so every
site-and-month pair is its own group. The third is what to compute on each
group, here `mean`.

`tapply()` then does three things. It splits the vector into those groups,
runs the function on each one, and arranges the answers by the groupings it
was given. One grouping gives a named vector. Two give a **matrix**, with
the first grouping down the rows and the second across the columns: eight
sites by twelve months, one mean in each of the 96 cells.

Cells with no rows come back `NA` rather than zero, which is the honest
answer. A month a site never reported is missing, not clean air.

A long file of 4,757 rows is now a grid you can read at a glance. That is
the move worth taking away: the data frame is how the file arrives, and the
matrix is the shape the question has.

:::

## reshape(): long and wide {#reshape}

`tapply()` gave a matrix. `reshape()` does the same turn on a data frame,
and turns it back.

```{r}
#| label: reshape
#| script: "reshape(): long to wide and back"
long <- data.frame(site  = c("A", "A", "B", "B"),
                   month = c("01", "02", "01", "02"),
                   pm25  = c(21.3, 13.4, 11.6, 7.2))
long

wide <- reshape(long, direction = "wide",
                idvar = "site", timevar = "month", v.names = "pm25")
wide
```

## reshape(): back to long {#reshape-back}

```{r}
#| label: reshape-back
#| script: "reshape(): wide back to long"
reshape(wide, direction = "long", idvar = "site",
        varying = list(2:3), v.names = "pm25", times = c("01", "02"))
```

::: {.content-hidden when-format="revealjs"}

**Long** means one row per observation, with a column saying which group it
belongs to. **Wide** means one row per unit and one column per group. The
same numbers, two shapes, and most of the reshaping you will do in your
career is moving between them.

The arguments name the pieces: `idvar` is what stays a row, `timevar` is
what becomes columns, and `v.names` is the value being spread. Going back,
`varying` names the columns to stack and `times` says what to call them.

Base `reshape()` is famously awkward, and it is worth knowing that
`tidyr::pivot_wider()` and `pivot_longer()` do the same job with friendlier
arguments. Panels arrive long, regressions usually want long, and tables for
readers want wide.

:::

::: {.content-visible when-format="revealjs"}

- **Long**: one row per observation. **Wide**: one row per unit, one column
  per group.
- `idvar` stays a row, `timevar` becomes columns, `v.names` is the value.

:::

## apply() over both margins {#both-margins}

```{r}
#| label: both-margins
#| script: "Both margins at once"
round(sort(apply(tb, 1, mean), decreasing = TRUE), 1)   # per site, all year
round(apply(tb, 2, mean), 1)                            # per month, all sites
```

::: {.content-hidden when-format="revealjs"}

Read that first line from the **inside out**, which is how you write it too:
`apply(tb, 1, mean)` gives one number per site, `sort()` puts them in order,
`round()` tidies them for printing. Build the innermost piece, run it, look
at what came back, then wrap the next call around it. Typing the whole nest
in one go and hoping is how you end up debugging four functions at once.

Now the two margins answer two different questions with the same verb.

Down the rows: which monitors sit in the dirtiest air all year. Lancaster, out
in the desert, reads a third of what the basin sites do.

Across the columns: what the county did month by month. January is the worst
month at nearly every site at once, which is the signature of something
county-wide rather than one instrument drifting.

:::

## merge(): joining two tables {#merge}

Results in one table, the thing you want to attach in another. `merge()`
matches them on a shared column.

```{r}
#| label: merge-setup
#| script: "Two tables, one key"
#| echo: false
means <- data.frame(site = rownames(tb),
                    pm25 = round(apply(tb, 1, mean), 1))
coords <- unique(frm[, c("Local.Site.Name", "Site.Latitude", "Site.Longitude")])
names(coords) <- c("site", "lat", "lon")
```

```{r}
#| label: merge-setup-shown
#| eval: false
means  <- data.frame(site = rownames(tb), pm25 = round(apply(tb, 1, mean), 1))
coords <- unique(frm[, c("Local.Site.Name", "Site.Latitude", "Site.Longitude")])
names(coords) <- c("site", "lat", "lon")
```

```{r}
#| label: merge
#| script: "Merge, counting rows before and after"
nrow(means); nrow(coords)                    # count BEFORE
sites <- merge(means, coords, by = "site")
nrow(sites)                                  # and after
```

::: {.content-hidden when-format="revealjs"}

`by` names the column the two tables share. Rows whose keys match are glued
side by side; the result carries the columns of both.

Eight rows in, eight rows out, so nothing was lost. **That check is the whole
discipline of joining.** Write down the row count you expect before you run
it, and compare. A join has two silent failure modes and neither one errors.

:::

## How merge() drops and duplicates rows {#merge-hazards}

```{r}
#| label: merge-hazards
#| script: "What a join does silently"
partial <- coords[coords$site != "Compton", ]   # pretend one site is missing

nrow(merge(means, partial, by = "site"))                # dropped
nrow(merge(means, partial, by = "site", all.x = TRUE))  # kept, with NA
```

::: {.content-hidden when-format="revealjs"}

**Rows can disappear.** By default `merge()` keeps only keys present in
both tables, so a site missing from the lookup takes its result with it.
Seven rows came back where you asked about eight, with no error and no
warning. `all.x = TRUE` keeps every row of the first table and fills the
missing side with `NA`, which is almost always what you want: an `NA` you
can see beats a row you cannot.

**Rows can also multiply.** If the second table has two rows for one site,
every matching row of the first is repeated to pair with each. Ask for eight
and get twelve. That is why `unique()` is in the line that built `coords`.

Keys have to be the same type, too. A site id read as a number will not
match the same id read as text, and that failure looks exactly like "no
matching rows".

:::

::: {.content-visible when-format="revealjs"}

- Default keeps only keys in **both** tables: rows vanish, silently.
- `all.x = TRUE` keeps the left table and fills with `NA`.
- A duplicated key on the right **multiplies** rows instead.
- Count rows before and after. Every time.

:::

## Reading many files {#lapply-files}

```{r}
#| label: lapply-files
#| script: "One line, every file"
files <- list.files("data", pattern = "\\.csv$")
files

sapply(files, function(f) nrow(read.csv(file.path("data", f))))
```

::: {.content-hidden when-format="revealjs"}

`list.files()` gives you the names, and `lapply()` or `sapply()` runs the same
call on each one.

This is the shape of most real data collection: one file per year, per state
or per month, and one function that handles any single file. You get it right
once, then hand the function the list. `do.call(rbind, ...)` on the result
stacks them into one frame.

:::

## Writing files to disk {#write}

```{r}
#| label: write
#| script: "Writing files to disk"
#| eval: false
answers <- data.frame(site = rownames(tb),
                      mean = round(apply(tb, 1, mean), 1))

write.csv(answers, "results.csv", row.names = FALSE)   # text, anything reads it
saveRDS(answers, "results.rds")                        # R's own format
readRDS("results.rds")
```

::: {.content-hidden when-format="revealjs"}

`write.csv()` is `read.csv()` run backwards. `row.names = FALSE` stops R
adding a column of row numbers you never asked for, which is the most common
complaint about files written from R. Round before you write, not after: a
mean lands in the file as `19.386666666667` unless you say `round(x, 1)`.

**CSV forgets what things were.** It is text, so a `Date` comes back as
character, a factor comes back as character, and every column has to be
guessed again on the way in. That is fine for a results table someone else
will open, and wrong for an intermediate you will read back into R.

`saveRDS()` writes one R object exactly as it is, compressed. Dates stay
dates, factors keep their levels, and `readRDS()` hands it back unchanged.
Use it for anything you computed and do not want to recompute. `save()` and
`load()` do the same for several objects at once, but they restore names
into your workspace instead of returning a value, so `saveRDS()` is easier
to reason about.

**Other people's formats.** Applied economics runs on Stata files, and the
`haven` package reads and writes them: `read_dta()`, `write_dta()`, plus
`read_sav()` for SPSS and `read_sas()` for SAS. `readxl::read_excel()` opens
spreadsheets, and `writexl::write_xlsx()` writes them. For data too big for
CSV to be sensible, `arrow::write_parquet()` writes a columnar file that
loads far faster and keeps its types.

| Code | When |
|:--|:--|
| `write.csv(d, f, row.names = FALSE)` | A results table anyone should be able to open |
| `saveRDS(d, f)` · `readRDS(f)` | An R object you want back exactly as it was |
| `haven::read_dta(f)` · `write_dta()` | Stata; also `read_sav()`, `read_sas()` |
| `readxl::read_excel(f)` | Spreadsheets, one sheet at a time |
| `arrow::write_parquet(d, f)` | Large tables, typed and fast |

:::

::: {.content-hidden when-format="revealjs"}

## Practice {#practice}

::: {.panel-tabset}

### Practice

```{r}
#| label: practice
#| exercise: true
#| script: "Practice"
#| eval: false
# Everything below uses la, frm and tb, built earlier in this script.

# Matrices
#  1. Build a 3 x 4 matrix of the numbers 1 to 12, filled across the rows.
#  2. Same matrix: what is the sum of each row? Of each column?
#  3. Same matrix: divide every cell by its own column's total.
#  4. Add a fourth row of zeros with rbind(). Then add a fifth column.
#  5. Run rbind(c(1, 2), c("a", "b")). What class is the result, and why?

# Lists
#  6. Build a list holding your name, the numbers 1 to 5, and TRUE.
#  7. Same list: pull the numeric vector out two different ways.
#  8. What class does l["readings"] return? Check with class().
#  9. Split "2025-01-07" on the dash and get back a vector of three pieces.
# 10. unlist() the list from question 6. What happened to the numbers?

# apply and friends
# 11. Which column of la came in as character? (One line, all 22 at once.)
# 12. How many rows does each site have in la? (One line.)
# 13. Which month had the highest county-wide mean, and what was it?
# 14. Which site had the single dirtiest month of 2025?
# 15. Rebuild tb with median instead of mean. Where do the two disagree
#     most, and what does that tell you about those days?
# 16. One line: the number of rows in every CSV in your data folder.
```

### Solutions

```{r}
#| label: practice-solutions
#| purl: false
#| eval: false
# 1
mm <- matrix(1:12, nrow = 3, byrow = TRUE)

# 2
apply(mm, 1, sum); apply(mm, 2, sum)

# 3
sweep(mm, 2, colSums(mm), "/")

# 4
mm <- rbind(mm, 0); mm <- cbind(mm, 0)

# 5  character: one vector underneath, so one type throughout
class(rbind(c(1, 2), c("a", "b")))

# 6
l3 <- list(name = "Ariel", nums = 1:5, ok = TRUE)

# 7
l3$nums; l3[["nums"]]

# 8  a list of length one, not the vector: single brackets stay wrapped
class(l["readings"])

# 9
strsplit("2025-01-07", "-")[[1]]

# 10  everything became text: one type, and "TRUE" is now a word
unlist(l3)

# 11  a data frame is a list of columns, so sapply walks them
names(which(sapply(la, class) == "character"))

# 12
table(la$Local.Site.Name)

# 13
month_means <- apply(tb, 2, mean, na.rm = TRUE)
month_means[which.max(month_means)]

# 14
which(tb == max(tb, na.rm = TRUE), arr.ind = TRUE)

# 15  the mean sits above the median wherever a few days ran very high
tb_med <- tapply(frm$Daily.Mean.PM2.5.Concentration,
                 list(frm$Local.Site.Name, frm$month), median)
round(sort(apply(tb - tb_med, 1, mean, na.rm = TRUE), decreasing = TRUE), 2)

# 16
sapply(list.files("data", pattern = "\\.csv$"),
       function(f) nrow(read.csv(file.path("data", f))))
```

:::

:::

## Quick reference {#quick-reference}

::: {.content-hidden when-format="revealjs"}

**Matrices**

| Code | What it does |
|:--|:--|
| `matrix(x, nrow =, ncol =, byrow =)` | Build one; `byrow = TRUE` fills across; short input is recycled |
| `dim(m)`, `nrow(m)`, `ncol(m)` | Its shape |
| `m[i, j]`, `m[i, ]`, `m[, j]` | Cell, row, column; blank means all |
| `m[, j, drop = FALSE]` | Keep it a matrix instead of dropping to a vector |
| `rownames(m)`, `colnames(m)` | Names on the margins |
| `m["a", "Feb"]`, `m[, c("a", "b")]` | Subset by name instead of position |
| `m[m > 15]` | Subset by logical; returns a vector, not a matrix |
| `which(m > 15, arr.ind = TRUE)` | The row and column of every cell that passes |
| `t(m)`, `m %*% v` | Transpose; matrix multiplication |
| `diag(m)` | Pull out the diagonal |
| `diag(3)`, `diag(c(4, 5, 6))` | Build an identity matrix; build a diagonal one |
| `Matrix::bdiag(a, b)` | Blocks down the diagonal, zeros elsewhere |
| `rbind(a, b)`, `cbind(a, b)` | Stack; widen. Matched by position, not by name |
| `head(m, 3)`, `tail(m)` | First or last rows; works on frames too |
| `rowSums(m)`, `rowMeans(m)` | One number per row, fast |
| `colSums(m)`, `colMeans(m)` | One number per column, fast |
| `sweep(m, 2, v, "/")` | Divide every column by its own entry in `v` |
| `Matrix(m, sparse = TRUE)` | Store only the non-zeros; `library(Matrix)` |
| `object.size(x)`, `system.time(expr)` | How big it is; how long it took |

**Lists**

| Code | What it does |
|:--|:--|
| `list(a = , b = )` | Build one; anything, any length |
| `l[["a"]]`, `l$a` | The element itself |
| `l["a"]` | A list of length one, not the element |
| `names(l)`, `is.list(x)` | The element names; is this a list at all |
| `str(l)`, `lengths(l)` | What is in there; how long each part is |
| `unlist(l)` | Flatten to one vector, one type |
| `do.call(rbind, l)` | Stack the elements into a matrix or frame |
| `lm(y ~ x)`, `fitted(fit)` | A fitted model is a list; the values on the line |
| `plot(x, y)`, `abline(fit)`, `segments()` | Draw the points, the fitted line, the residuals |

**One function, many things**

| Code | What it does |
|:--|:--|
| `apply(m, 1, f)` | Once per row; `2` for columns |
| `apply(m, 2, function(x) ...)` | Write the function inline when it is small |
| `split(x, g)` | A list, one element per group |
| `lapply(l, f)` | A list in, a list of the same length out |
| `sapply(l, f)` | The same, simplified to a vector or matrix |
| `tapply(x, list(g1, g2), f)` | A matrix of `f` over two groupings |
| `reshape(d, direction = "wide", idvar =, timevar =, v.names =)` | Long to wide |
| `reshape(d, direction = "long", varying =, v.names =, times =)` | Wide back to long |
| `sapply(d, class)` | The class of every column at once |

**Dates**

| Code | What it does |
|:--|:--|
| `as.Date(x, format = "%m/%d/%Y")` | Text to a real date; wrong format gives `NA`, not an error |
| `format(d, "%m")`, `"%b"`, `"%Y"`, `"%j"` | Month, short month name, year, day of the year |
| `weekdays(d)`, `months(d)` | The day and month names |
| `d + 30`, `d2 - d1` | Dates are numbers, so arithmetic works |
| `seq(d, by = "week", length.out = 4)` | A sequence of dates; also `"day"`, `"month"`, `"year"` |
| `range(d)`, `sort(d)` | The span; in order |

**Vectors you will keep reaching for**

| Code | What it does |
|:--|:--|
| `unique(x)` | Drop repeats; `length(unique(x))` counts distinct values |
| `duplicated(d)` | `TRUE` for each row that repeats an earlier one |
| `anyDuplicated(d)` | Position of the first repeat, or `0` if there are none |
| `which.max(x)`, `which.min(x)` | The *position* of the extreme value |
| `x[which.max(x)]` | ...so this is the value itself |
| `max(x)`, `min(x)`, `range(x)` | The extremes |
| `sort(x)`, `order(x)` | Sorted values; the positions that would sort them |
| `strsplit(x, "-")` | Split text; returns a list, so `[[1]]` to get the pieces |
| `set.seed(1)`, `sample(x, n)` | A reproducible random draw |
| `rnorm(n)` | `n` draws from a standard normal |
| `sum(x)`, `mean(x)`, `median(x)` | The obvious ones; `sum()` of a logical counts `TRUE`s |
| `round(x, 1)` | Cut decimals before you write, not after |
| `data.frame(a = , b = )` | Build one from columns |
| `paste0("a", "b")` | Glue text with no separator; `paste()` adds a space |

**Files**

| Code | What it does |
|:--|:--|
| `read.csv(f)` | Read a CSV, guessing every column's type |
| `colClasses = c(id = "character")` | Stop it guessing; protects identifiers |
| `write.csv(d, f, row.names = FALSE)` | Write a table anyone can open |
| `saveRDS(d, f)`, `readRDS(f)` | One R object, saved exactly as it is |
| `haven::read_dta(f)`, `write_dta()` | Stata; also `read_sav()`, `read_sas()` |
| `readxl::read_excel(f)` | Spreadsheets |
| `arrow::write_parquet(d, f)` | Large tables, typed and fast |
| `list.files(dir, pattern =)` | The names, ready for `lapply()` |
| `file.path("data", f)` | Build a path that works on any machine |
| `dir.create("data")` | Make a folder from the script, not by hand |
| `download.file(url, dest)` | Fetch a file so the script is self-contained |
| `merge(a, b, by = "id")` | Glue two tables on a shared key |
| `merge(a, b, by =, all.x = TRUE)` | Keep every row of `a`; fill the rest with `NA` |

**Links**

- Session materials: [slides](03-r-essentials-ii-slides.html) ·
  [live script](03-r-essentials-ii.R) ·
  [EPA data](data/epa_pm25_la_county_2025.csv) ·
  [where it came from](data/SOURCE.md) ·
  [EPA's download tool](https://www.epa.gov/outdoor-air-quality-data/download-daily-data)
- [Syllabus](../syllabus.qmd) · [full schedule](../index.qmd#schedule)

:::

::: {.content-visible when-format="revealjs"}

Selected commands from today. The session page has all of them, grouped by
topic.

| Code | What it does |
|:--|:--|
| `matrix(x, nrow =, byrow =)` · `diag()` | Build one; the diagonal |
| `m[i, j]` · `m[, j, drop = FALSE]` | Index; keep it a matrix |
| `rbind()` · `cbind()` | Stack; widen |
| `l[["a"]]` vs `l["a"]` | The element vs a list of one |
| `unlist(l)` · `do.call(rbind, l)` | Flatten; stack |
| `apply(m, 1, f)` · `lapply()` · `sapply()` | A margin; a list; simplified |
| `split(x, g)` · `tapply(x, list(g1, g2), f)` | Groups; two groupings, one matrix |
| `as.Date(x, format =)` · `seq(d, by =)` | Text to date; a run of dates |
| `anyDuplicated(d)` · `unique(x)` | Any repeats; drop them |
| `merge(a, b, by =, all.x =)` | Join, keeping every row of `a` |
| `colClasses = c(id = "character")` | Protect identifiers at read time |
| `saveRDS()` · `readRDS()` | An R object, saved exactly as it was |

:::

```{r}
#| purl: false
#| include: false
source("_purl.R")
purl_session("03-r-essentials-ii.qmd", "03-r-essentials-ii.R")
```
