---
title: "4 · Basic plots I"
subtitle: "Thursday, September 3, 2026"
pagetitle: "4 · Basic plots I"
title-prefix: ""
execute:
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
  revealjs:
    output-file: 04-basic-plots-slides.html
    theme: [default, slides.scss]
    template-partials:
      - title-slide.html
    footer: "AEM 6850 · Session 4"
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    code-overflow: scroll
    history: false
    link-external-newwindow: true
---

```{r}
#| purl: false
#| include: false
knitr::opts_chunk$set(
  comment = "#>",
  collapse = FALSE,
  fig.width = 8,
  fig.height = 4,
  fig.align = "center",
  dpi = 150
)
```

::: {.content-hidden when-format="revealjs"}

[Slides](04-basic-plots-slides.html) · [Live script](04-basic-plots.R) ·
[Source](https://github.com/ArielOrtizBobea/aem6850/blob/main/fall-2026/sessions/04-basic-plots.qmd)

Four plots, the settings that work on all of them, panels, and how to get a
figure out of R and onto disk.

:::

## Objectives {#objectives}

By the end of this session you can:

- Name the plot that matches a question, and say what each one is bad at.
- Draw one series through time, and several series on shared axes.
- Read a distribution with `hist()`, and say when a box plot answers better.
- Move the same argument — `col`, `main`, `las`, `ylim` — between `plot()`,
  `hist()`, `boxplot()` and `barplot()`.
- Split a device into panels, control which panel fills first, and put it back.
- Write a figure to a file at a resolution someone else can use.

## Why plot {#thesis}

> **Let your imagination be the limit to what you can ask and explore, not the tools you happen to
> know.**

- Humans are visual animals. Vision is a fundamental way we interact with the world.
- Seeing a physical representation of numbers is often more impactful and easier to understand than the numbers themselves.
- As a result, visualization is one of the most important skills in research (conducting it and communicating its results)

## Visualization and the questions you ask {#thesis-2}

- There is often more to learn from staring at the data than looking at numeric summaries of such data.
- My invitation is that you cultivate a visual way to interact with data so that you can think "outside the menu" of visualization tools at your disposal.
- The figures you can build can influence the hypothesis you come up with. They empower your curiosity.

::: {.content-hidden when-format="revealjs"}

Most of what goes wrong in an empirical project is visible in the data long
before it reaches a regression table, and a plot is the fastest way to see
it. A summary answers the question you asked; a plot answers questions you
did not think to ask.

The second half of that matters just as much. Every plotting system hands
you a menu of ready-made figures, and it is easy to end up asking only the
questions that menu can answer. Base R does not work that way: a plot is a
blank device you draw on, so the shape of the figure is your decision rather
than the package's. That is why this session spends time on the settings and
the canvas rather than on a catalogue of chart types.

Polishing — fonts, palettes, figures built for a reader — is a different job
and gets its own lecture, with publication-ready output. Nothing today needs
to be pretty. Everything today needs to be fast, because a plot you can
produce in one line is a plot you will actually make, and a plot that takes
twenty minutes is one you will skip.

:::

## Getting the data {#data}

One file: every PM2.5 monitor in Los Angeles County for 2025. This chunk
fetches it if you do not already have it, and does nothing if you do.

```{r}
#| label: get-data
#| script: "Get the data"
dir.create("data", showWarnings = FALSE)

f <- "data/epa_pm25_la_county_2025.csv"
if (!file.exists(f)) {
  download.file(paste0("https://arielortizbobea.github.io/aem6850/",
                       "fall-2026/sessions/", f), f)
}

file.exists(f)
```

::: {.content-hidden when-format="revealjs"}

`dir.create()` makes a folder and `showWarnings = FALSE` keeps it quiet when
the folder is already there. `download.file()` fetches a URL to a path. The
`if (!file.exists(f))` wrapper means you can run this line every time without
re-downloading, which is the habit worth copying: a script that fetches its
own inputs runs on anyone's machine, including yours six months from now.

[The data folder's provenance notes](data/SOURCE.md) record where the file
came from, byte for byte.

:::

## Building the working table {#build}

We take four monitors on a line across the basin.

```{r}
#| label: build
#| script: "Build the working table"
pm <- read.csv("data/epa_pm25_la_county_2025.csv",
               colClasses = c("Site.ID" = "character"))

names(pm)[names(pm) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"
names(pm)[names(pm) == "Local.Site.Name"]                <- "site"
pm$date <- as.Date(pm$Date, format = "%m/%d/%Y")
```

## Four stations on a transect {#transect}

```{r}
#| label: build-transect
#| script: "Cut to four stations"
transect <- c("Pasadena", "Compton",
              "Long Beach-Route 710 Near Road", "Lancaster - Fairgrounds")

d <- pm[pm$site %in% transect &
        pm$POC == 1 & pm$AQS.Parameter.Code == 88101, ]

table(d$site)
```

::: {.content-hidden when-format="revealjs"}

*Pasadena* sits in the foothills, *Compton* and *Long Beach* down toward the
coast, *Lancaster* out over the mountains in the high desert. Four places,
one air basin, one year, and a reader can check why those four.

Three things in that chunk are new:

- `pm$site %in% transect` asks, for every row, *is this site one of these
  four*. It saves writing four `==` tests joined by `or`.
- A subset can sit on the **left** of the arrow.
  `names(pm)[names(pm) == "old"] <- "new"` writes into the one position that
  matches, and `pm$date <- ...` names a column that does not exist yet, so it
  creates it.
- `table()` counts how often each value appears. Four names, four counts —
  that is the check.

And the check is not decoration. Lancaster reported all 365 days; Pasadena
reported 115. Those two monitors are not on the same calendar, and every plot
below has to survive that.


:::

::: {.content-visible when-format="revealjs"}

Lancaster reported **365** days. Pasadena reported **115**. Every plot below
has to survive that.

:::

## Choosing a plot: good for / bad for {#menu}

::: {.tight-cols}

| Plot | Good for | Bad for |
|:--|:--|:--|
| `plot(x, y)` | the relationship between two variables | more points than pixels — they pile into a blob |
| `plot(x, y, type = "l")` | one variable in its natural order, usually time | missing days: the line joins straight across them, so a gap looks like data |
| `plot()` + `lines()`, `points()` | a few series on shared axes | more than about six |
| `hist(x)` | the distribution of one variable | comparing groups |
| `boxplot(y ~ g)` | the distribution of one variable, by group | groups of two or three rows |
| `barplot(table(g))` | how often each category occurs | anything continuous — bin it, or use `hist()` |

:::

::: {.content-hidden when-format="revealjs"}

On the line plot's weakness: `plot(type = "l")` draws a segment from each
row to the next, so a month with no readings comes out as one long straight
stretch rather than as a hole. Nothing warns you. `type = "o"` or
`type = "p"` shows where the observations actually are, which is why the
missing-data question is usually asked with points and not with a line.

**Start from the question, not the plot type.** *Did something happen in
time?* is a line. *Do these two move together?* is a scatter. *Is this variable
distributed the way I think?* is a histogram. *Does this group differ from that
one?* is a box plot. *Is everything here?* is a bar of counts.

:::

## `plot()`: one series through time {#one-series}

```{r}
#| label: line-one
#| script: "One series through time"
#| fig-alt: "Daily PM2.5 at Compton across all of 2025 as a single line. A tall cluster in early January, a sharp isolated spike in early July, a flat gap in April and May, and a high ragged run through December."
comp_y <- d[d$site == "Compton", ]
plot(comp_y$date, comp_y$pm25, type = "l",
     xlab = "", ylab = "Daily PM2.5 (ug/m3)")

comp_y$date[which.max(comp_y$pm25)]
```

::: {.content-hidden when-format="revealjs"}

Three new things:

- `type` picks how the points are drawn: `"p"` marks only, `"l"` a line only,
  `"b"` both with a gap at each mark, `"o"` both, overlaid. `"l"` is the one
  you want for a series in time.
- `xlab` and `ylab` are the axis labels. The y label is where the units go,
  every time; a number with no unit is not a reading.
- `which.max(x)` hands back the **position** of the largest value, not the
  value — which is what you want when the question is *which day*.

Now read it. 301 days in one glance. Two isolated spikes that are not
weather — January 1 at 53.2 and July 4 at 50.7 — a flat stretch through April
and May, and a high ragged run through December that holds the year's
maximum, 56.0 on the 13th. A six-number summary gives you none of that.

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
Two spikes that are not weather: Jan 1 at 53.2, Jul 4 at 50.7. A flat stretch in April–May. The year's maximum is 56.0, on December 13.
:::
:::

## One `plot()` call is one series {#one-call}

The same call on the whole four-station frame draws three long diagonal
strokes.

```{r}
#| label: line-all-four
#| script: "The same call on the whole frame"
#| fig-alt: "The whole four-station frame drawn as one line: a dense band of daily variation crossed by three long diagonal strokes running from the right edge back to the left."
plot(d$date, d$pm25, type = "l",
     xlab = "", ylab = "Daily PM2.5 (ug/m3)")
```

::: {.content-hidden when-format="revealjs"}

`d` is stored station by station — Compton, then Pasadena, then Long Beach,
then Lancaster. So the line runs Compton January to December, snaps back to
January for Pasadena, and again twice more. Each snap-back is one of those
diagonals.

`type = "l"` connects row 1 to row 2 to row 3 **in the order the rows sit in
the frame**. It has no idea what a station is, and it will not warn you.

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
type = "l" connects row 1 to row 2 to row 3 in the order the rows sit in the frame. It has no idea what a station is. One plot(type = "l") call means one series.
:::
:::

## Cutting to January {#jan}

```{r}
#| label: jan
#| script: "Cut to January"
jan  <- d[d$date < as.Date("2025-02-01"), ]
comp <- jan[jan$site == "Compton", ]
lb   <- jan[jan$site == "Long Beach-Route 710 Near Road", ]
lanc <- jan[jan$site == "Lancaster - Fairgrounds", ]
pas  <- jan[jan$site == "Pasadena", ]

xr   <- as.Date(c("2025-01-01", "2025-01-31"))   # one window for every panel
```

## `lines()` and `points()`: several series {#several-series}

```{r}
#| label: lines-four-bad
#| script: "Four stations, first attempt"
#| fig-alt: "January PM2.5 at four stations. Compton in red and Long Beach in grey track each other. The blue Lancaster line is cut off below the bottom of the frame for most of the month. The orange Pasadena line leaves the top of the frame twice and runs as long straight diagonals between its few observations."
plot(comp$date, comp$pm25, type = "l", col = "#b31b1b",
     xlab = "", ylab = "Daily PM2.5 (ug/m3)")
lines(lb$date,   lb$pm25,   col = "grey30")
lines(lanc$date, lanc$pm25, col = "#1f6fb4")
lines(pas$date,  pas$pm25,  col = "darkorange")
```

::: {.content-hidden when-format="revealjs"}

`lines()` draws onto the plot that is already there. It never makes a new
one, and it never resizes anything. `plot()` sized both axes from the
**first** series and everything after obeys them — anything outside is simply
not drawn. No error, no warning.

Compton's January runs 5.7 to 53.2, so that is the window all four series
got. Three separate failures follow from it.

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
Three failures. Clipped at the BOTTOM: Lancaster spends January between 0.4
and 5 and falls out of the frame. Clipped at the TOP: Pasadena peaks at 72.5
and leaves it twice. Both because plot() sized the axes from Compton alone,
5.7 to 53.2. And INVENTED: Pasadena has 11 observations, so the line draws
straight diagonals over days it never measured, including across its own peak.
:::
:::

## Setting the window and the key {#repair}

```{r}
#| label: lines-four-good
#| script: "The repair"
#| fig-alt: "The corrected figure. The y axis runs 0 to 75 so every series fits. Compton is a thick red line, Long Beach grey, Lancaster blue, and Pasadena appears as eleven orange dots, the highest at 72.5 on January 10. A key with no box sits in the top right."
plot(comp$date, comp$pm25, type = "l", col = "#b31b1b", lwd = 2,
     ylim = c(0, 75), xlab = "", ylab = "Daily PM2.5 (ug/m3)")
lines(lb$date,   lb$pm25,   col = "grey30")
lines(lanc$date, lanc$pm25, col = "#1f6fb4")
points(pas$date, pas$pm25,  col = "darkorange", pch = 16)
legend("topright", c("Compton", "Long Beach", "Lancaster", "Pasadena"),
       col = c("#b31b1b", "grey30", "#1f6fb4", "darkorange"),
       lty = c(1, 1, 1, NA), pch = c(NA, NA, NA, 16), bty = "n")
```

::: {.content-hidden when-format="revealjs"}

- `col` sets the colour of the marks. R takes an English name
  (`"darkorange"`), a grey with a number (`"grey30"` dark, `"grey85"` light),
  or a hex code (`"#b31b1b"`); `colors()` prints all 657 names. And `c()`
  glues values into one vector — nearly every setting today wants one.
- `ylim = c(0, 75)` fixes the vertical window. Set it on the **first** call,
  from everything you intend to draw, because nothing after that call resizes
  the axes. `xlim` does the same horizontally.
- `points()` adds marks without joining them — the right answer when the gaps
  between observations are real. `pch` picks the mark: `16` is a filled
  circle, `1` the open one `plot()` draws by default. `lwd` is line width, and
  `lwd = 2` is how you say *this is the series I want you to look at*.
- `legend()` puts a key at a named corner. `lty` is line type — `1` solid, `2`
  dashed — and it means the same thing in `plot()`, `lines()`, `abline()` and
  in the key. `bty = "n"` drops the box. `NA` inside the `lty` or `pch` vector
  means *no sample for this entry*: that is how Pasadena gets a dot and no
  line.

Pasadena is dots and not a line because Pasadena reports one day in three. A
line through it would have drawn straight across its own peak.

### Reading it {#read-four}

Pasadena, nearest the Eaton fire, records January's highest reading: **72.5
on the 10th**, and it was already at **64.1 on the 7th**, the day the fires
started. Compton and Long Beach peak a day earlier and lower — **47.7 and
40.7, both on the 9th** — as the smoke reaches the coast. Lancaster, over the
mountains in the high desert, stays flat through the whole episode and spikes
on its **own** day, January 7, at **24.9**: blowing dust, not smoke.

Four monitors, four stories, one screen. And Compton's tallest January bar,
53.2, is January 1 — fireworks, six days before any fire. That reading is why
the axis had to be set by hand.

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
Read it off the figure. Pasadena, nearest the Eaton fire, is highest: 72.5 on
the 10th, already 64.1 on the 7th, the day the fires start. Compton and Long
Beach peak a day earlier and lower, 47.7 and 40.7 on the 9th, as the smoke
reaches the coast. Lancaster stays flat through the whole episode and spikes
on its OWN day, Jan 7, at 24.9 -- blowing dust, not smoke. Four monitors,
four stories, one screen. And Compton's tallest January bar, 53.2, is Jan 1:
fireworks, six days before any fire. That is why the axis had to be set by
hand.
:::
:::

## `hist()`: the distribution of one variable {#hist}

```{r}
#| label: hist-breaks
#| script: "The distribution of one variable"
#| fig-alt: "Histogram of daily PM2.5 across the four stations with forty bins requested. The mass piles up between 0 and 15 with a long thin tail past 70. The leftmost bar sits left of zero, separated from the main mass, and a dashed red vertical line marks zero."
hist(d$pm25, breaks = 40, col = "grey85",
     main = "", xlab = "Daily mean PM2.5 (ug/m3)")
abline(v = 0, col = "#b31b1b", lty = 2)
```

::: {.content-hidden when-format="revealjs"}

- `hist()` chops one numeric column into bins and draws how many rows fell in
  each. One column, one call.
- `breaks = 40` is a **suggestion** for how many bins; R picks round numbers
  near it. Ask for 40 here and you get 38.
- `main` is the title above the plot; `main = ""` removes the one R writes
  for you.
- On `plot()` and `lines()`, `col` colours the marks. On `hist()`,
  `barplot()` and `boxplot()` it **fills** them.
- `seq(-5, 80, by = 5)` builds −5, 0, 5 … 80: every bin edge written out by
  hand instead of guessed at. That is how you take control of `breaks` when
  the suggestion is not good enough.

:::

## What the histogram says {#hist-read}

```{r}
#| label: hist-summary
#| script: "The same column as six numbers"
summary(d$pm25)
```

::: {.content-hidden when-format="revealjs"}

Right-skewed: the mean, 9.947, sits above the median, 8.400. That is what
skew looks like in numbers — a few large values dragging the average up past
the middle.

And there is a bar left of zero. Nobody asked for that, and no summary line
above draws attention to it except the minimum, which you have to notice.

:::

::: {.content-visible when-format="revealjs"}

Right-skewed: **mean 9.947 above median 8.400**.

And there is a bar **left of zero**. Nobody asked for that.

:::

## Four columns with the same mean and sd {#quartet}

```{r}
#| label: quartet
#| script: "Four columns with the same mean and the same sd"
set.seed(1)
std <- function(x, m = 20, s = 8) (x - mean(x)) / sd(x) * s + m
q   <- list("Normal"       = std(rnorm(500)),
            "Two humps"    = std(c(rnorm(250, -2), rnorm(250, 2))),
            "Right-skewed" = std(rlnorm(500, 0, 1)),
            "Flat"         = std(runif(500)))

round(sapply(q, function(x) c(mean = mean(x), sd = sd(x), max = max(x))), 2)
```

## The same four columns as histograms {#quartet-plot}

```{r}
#| label: quartet-plot
#| script: "The same four columns as histograms"
#| fig-height: 4.4
#| fig-alt: "Four histograms in a two-by-two grid: a bell, two separated humps, a hard right skew with a long thin tail, and a flat block. A red vertical line marks the mean in each, at the same place."
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2.5, 1))
invisible(lapply(names(q), function(nm) {
  hist(q[[nm]], breaks = 30, xlim = c(-10, 55), col = "grey80",
       border = "white", main = nm, xlab = "", ylab = "")
  abline(v = mean(q[[nm]]), col = "#b31b1b", lwd = 2)
}))
par(op)
```

::: {.content-hidden when-format="revealjs"}

All four columns have a mean of 20.00 and a standard deviation of 8.00, to
two decimal places. You have just computed a mean and an sd on a real column;
here are four columns that agree on both and share almost nothing else.

The maxima give the game away — 49.94, 37.72, **138.53**, 32.97 — but only
because you went looking. The two numbers you actually reported cannot
separate a bell from two humps from a hard skew.

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
All four: mean 20.00, sd 8.00. Maxima: 49.94, 37.72, 138.53, 32.97.
:::
:::

## Anscombe's quartet: four identical summaries {#anscombe-numbers}

```{r}
#| label: anscombe-numbers
#| script: "Anscombe's quartet: the summaries"
round(colMeans(anscombe), 2)

round(c(cor(anscombe$x1, anscombe$y1), cor(anscombe$x2, anscombe$y2),
        cor(anscombe$x3, anscombe$y3), cor(anscombe$x4, anscombe$y4)), 3)
```

## Anscombe's quartet: four different pictures {#anscombe}

```{r}
#| label: anscombe
#| script: "Anscombe's quartet: the pictures"
#| fig-height: 4.6
#| fig-alt: "Anscombe's four datasets as four scatterplots: a loose linear cloud, a smooth arch, a tight line with one far outlier, and a vertical stack of identical x values with one point far to the right."
op <- par(mfrow = c(2, 2), mar = c(4, 4, 1, 1))
plot(anscombe$x1, anscombe$y1)
plot(anscombe$x2, anscombe$y2)
plot(anscombe$x3, anscombe$y3)
plot(anscombe$x4, anscombe$y4)
par(op)
```

::: {.content-hidden when-format="revealjs"}

Francis Anscombe built these four datasets in 1973 to make exactly this
point, and they ship inside R as `anscombe`. Every x mean is 9.0, every y
mean is 7.5, and every correlation is 0.816 or 0.817.

The pictures are a loose linear cloud; a smooth arch with nothing linear
about it; a straight line plus one outlier that drags the fit; and ten
identical x values with one far point that invents the entire correlation.

The previous slide made the point about **shapes**. This one makes it about
**relationships**. Same lesson: a summary is a compression, and compression
throws away structure you needed.

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
Every x mean 9.0, every y mean 7.5, every correlation 0.816. A cloud, an arch, a line with an outlier, and one point inventing a correlation.
:::
:::

## `boxplot()`: one column by groups {#boxplot}

```{r}
#| label: box-site
#| script: "Many distributions at once"
#| fig-alt: "Box plots of daily PM2.5 for four stations in alphabetical order: Compton, Lancaster, Long Beach, Pasadena. Lancaster's box, second from the left, is much lower and tighter than the other three, which carry long clouds of outlier dots above them. Only two of the four labels are drawn."
boxplot(pm25 ~ site, data = d)
```

::: {.content-hidden when-format="revealjs"}

- Read `pm25 ~ site` aloud as "pm25 **by** site". The tilde is R's word for
  "split by".
- `data = d` says which frame those two bare, unquoted names live in.
- The heavy line is the median; the box holds the middle half; the whiskers
  reach the ordinary range; anything past them is drawn as its own dot. A dot
  is a candidate for a look, not a verdict.
- `boxplot()` and `table()` put the groups in **alphabetical** order — not the
  order you named them, and not the order of the values. Read the labels, not
  the positions.

Medians, in the order the boxes appear: Compton 10.70, Lancaster 4.20, Long
Beach 10.65, Pasadena 9.30. The three basin stations agree to within a
twentieth of a microgram. Lancaster sits at less than half. Whisker tops run
26.9, 11.9, 21.9, 21.6, and the outlier counts are 28, 8, 28, 10.

The box plot answers a question the histogram cannot: not *what shape is this
column* but *does this group differ from that one*. A histogram holds one
column, so four columns would be four histograms.

And a box built from three rows looks exactly as confident as a box built
from three hundred. Count first.

:::

::: {.content-visible when-format="revealjs"}

Medians, left to right: **10.70 · 4.20 · 10.65 · 9.30**. Three basin stations
agree to within a twentieth of a microgram; Lancaster sits at less than half.

:::

## Labels R drops without warning {#dropped-labels}

::: {.content-hidden when-format="revealjs"}

Look again at the axis on that figure. R drew **two** labels and silently
dropped the other two. Nothing warned you.

The mechanism: the plot region is 6.76 inches wide, so each of the four slots
gets 1.69 inches. `"Lancaster - Fairgrounds"` needs 1.77 inches and
`"Long Beach-Route 710 Near Road"` needs 2.57. When labels do not fit, R
falls back to drawing every second one rather than overlapping them.

This is not a `boxplot()` problem, and no argument to `boxplot()` fixes it.
It is a problem with the canvas the plot was drawn on, and we repair it two
sections from here.

:::

::: {.content-visible when-format="revealjs"}

R drew **two** labels and silently dropped two. Each slot is 1.69 in;
`"Long Beach-Route 710 Near Road"` needs **2.57 in**.

Not a `boxplot()` problem. A **canvas** problem.

:::

## `barplot(table())`: counts of categories {#barplot}

```{r}
#| label: port-bar
#| script: "Counts of categories"
#| fig-height: 4.5
#| fig-alt: "Horizontal bar chart of days reported in 2025, one bar per station in alphabetical order from the bottom: Compton 301, Lancaster 365, Long Beach 338, Pasadena 115. All four station names are legible on the left."
op <- par(mar = c(4, 15, 1, 1))
barplot(table(d$site), horiz = TRUE, las = 1, col = "grey85",
        xlab = "Days with a reading in 2025")
par(op)
```

::: {.content-hidden when-format="revealjs"}

- `barplot()` draws a count table as bars. `table()` counts, `barplot()`
  draws — that pair is how you ask *is everything here?*
- `horiz = TRUE` lays the bars sideways, which is what you do when the
  category names are long. On `boxplot()` the same idea is spelled
  `horizontal = TRUE`.
- `las` turns the axis text. `0` is R's default, parallel to its axis; `1`
  makes everything horizontal and readable; `2` turns it perpendicular, which
  is how long category names fit.

365, 338, 301, 115, alphabetical from the bottom. Pasadena's 115 is not a gap
in a working monitor — it is a one-day-in-three sampling schedule, and it is
why Pasadena had to be drawn as dots.

That `par(mar = ...)` line is doing the work that fixes the box plot's
missing labels. Two sections from now it stops being a magic incantation.

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
Pasadena's 115 is not a broken monitor. It is a one-day-in-three schedule — which is why Pasadena had to be dots.
:::
:::

## Exercise 1 {#exercise-1}

::: {.panel-tabset}

### Exercise

```{r}
#| label: exercise-1
#| exercise: true
#| script: "Exercise 1 (3 minutes)"
#| eval: false
# Three minutes. The histogram had a bar left of zero.
#
# 1. Print the readings below zero.
# 2. Print which station filed them.
# 3. One sentence: keep them or delete them?
#
# Two answers to compare with the room: a count, and one station name.
```

### Solution

```{r}
#| label: exercise-1-solution
#| purl: false
d$pm25[d$pm25 < 0]
unique(d$site[d$pm25 < 0])
```

Eight readings, all Lancaster. **Keep them.** They are real published
measurements, and the file says why if you ask it two more questions:

```{r}
#| label: exercise-1-why
#| purl: false
unique(d$Method.Description[d$pm25 < 0])       # which instrument
round(tapply(d$pm25, d$site, mean), 2)         # how clean the air is there
```

Two conditions have to hold at once, and Lancaster is the only site where
both do. **It is the one monitor on a different instrument.** A BAM measures
mass by firing beta particles through an hour's worth of dust on a filter
tape and reading how much got absorbed; that is a difference between two
noisy counts, so near the detection limit it can come out below zero. The
other seven sites weigh a filter instead, and every one of them has a
positive minimum. **And it is the cleanest site**, at 4.48 µg/m³ against
11–13 elsewhere — the only place where the true concentration sits close
enough to zero for the noise to cross it.

EPA publishes the negative on purpose. AQS accepts values down to the
negative of the method detection limit and asks submitters *not* to
substitute zero, because
[zero-substitution biases every average computed from the data](https://www.epa.gov/sites/default/files/2017-02/documents/negative_values_reporting_to_aqs_10_6_16.pdf).
Deleting these eight would do the same thing to yours. Note that
`Daily.AQI.Value` reads 0 on those days: the *index* is floored at zero, the
*measurement* is not.

Three more days read exactly 0.0, which is a legal reading — *below* zero is
the finding, zero is not.

:::

## Arguments that work on every plot {#port-args}

::: {.tight-cols}

| Argument | Means the same in `plot()`, `hist()`, `boxplot()`, `barplot()` |
|:--|:--|
| `main =` | title above the plot |
| `xlab =`, `ylab =` | axis labels; units go in `ylab` |
| `col =` | colour of the marks, or the fill |
| `las =` | rotation of the axis numbers and labels |
| `xlim =`, `ylim =` | the window on each axis |
| `lwd =`, `lty =` | line width, line type |

:::

## The same arguments on a histogram {#port-hist-slide}

```{r}
#| label: port-hist
#| script: "The same arguments on a histogram"
#| fig-alt: "The four-station histogram with a grey fill, a written title, a labelled x axis, and horizontal y-axis numbers."
hist(d$pm25, col = "grey85", las = 1,
     main = "Four LA County monitors, 2025",
     xlab = "Daily mean PM2.5 (ug/m3)")
```

::: {.content-hidden when-format="revealjs"}

The bar chart two sections back used `col`, `las` and `xlab` on a different
function with a different data shape. Same words, same meaning. That is the
whole point of learning them once.

:::

## Layers: what draws on top {#port-layers}

> A plot is a **stack**. `plot()`, `hist()`, `boxplot()` and `barplot()` each
> start a new one. Everything else draws **on top of what is already there**,
> in the order you call it — so the last call wins where they overlap.

::: {.tight-cols}

| Layer | What it adds |
|:--|:--|
| `points(x, y)` | marks |
| `lines(x, y)` | a joined line |
| `abline(h =, v =)` | a horizontal or vertical reference line |
| `text(x, y, "label")` | a label at data coordinates |
| `legend("topright", ...)` | the key |
| `axis(side, at =, labels =)` | an axis you write yourself |
| `mtext("...", side =)` | text out in the margin, outside the frame |

:::

::: {.content-hidden when-format="revealjs"}

- `text(x, y, "label")` writes at **data** coordinates: a position in the
  units of the plot, not in inches, not in pixels.
- `axis()` draws an axis where R's default was wrong. Sides are numbered from
  the bottom, clockwise: 1 bottom, 2 left, 3 top, 4 right. Suppress the
  default first with `xaxt = "n"` or `yaxt = "n"`, then draw yours with `at`
  and `labels`, which must be the same length.
- `mtext()` writes into the margin instead of into the plot; `side = 3` is the
  top.

Call a layer before any plot exists and R says
`Error in plot.xy(...) : plot.new has not been called yet`. A layer needs
something to land on.

:::

## `par()`: the device, not the plot {#par}

```{r}
#| label: par-sticky
#| script: "par() is sticky"
par("mar")                       # 5.1 4.1 4.1 2.1 -- R's default
op <- par(mar = c(3, 3, 1, 1))   # set it, and keep the old value in op
par("mar")                       # 3 3 1 1 -- still, and for every plot after
par(op)                          # put it back
```

::: {.content-hidden when-format="revealjs"}

- `par("mar")`, with a name in quotes, **reads** a setting.
  `par(mar = c(...))`, with an `=`, **writes** it — and hands you back the old
  value.
- The idiom is `op <- par(...)` then `par(op)` at the end. Type the second
  line before you type the plot. Every mysterious plot you will ever draw is
  a `par()` somebody forgot to put back.
- `mar` is the margin around **this** panel: bottom, left, top, right,
  measured in lines of text. The default is `c(5.1, 4.1, 4.1, 2.1)`.

:::

::: {.content-visible when-format="revealjs"}

`par()` does not change *this* plot. It changes **the device**, and it stays
changed until you change it back.

:::

## `par(mar)`: giving labels room {#mar}

```{r}
#| label: box-fixed
#| script: "Give the labels room"
#| fig-height: 4.5
#| fig-alt: "The four-station box plots turned on their side. All four station names, including Long Beach-Route 710 Near Road, are fully legible along the left. Lancaster's box is visibly lower and tighter than the other three."
op <- par(mar = c(4, 15, 1, 1))
boxplot(pm25 ~ site, data = d, horizontal = TRUE, las = 1,
        xlab = "Daily mean PM2.5 (ug/m3)", ylab = "", col = "grey85")
par(op)
```

::: {.content-hidden when-format="revealjs"}

One margin line is `par("csi")` = 0.2 inches, and the longest label is 2.567
inches wide — just under 13 lines. Fifteen leaves room for the gap.

Three settings did three jobs: `mar` bought the space, `las = 1` turned the
text upright, and `horizontal = TRUE` moved the long names onto the axis that
has room for them.

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
mar bought the space, las = 1 turned the text upright, horizontal = TRUE put the long names on the axis that has room.
:::
:::

## `par(mfrow)` and `par(mfcol)`: panels {#fill-order}

> **`mfrow` fills across the rows. `mfcol` fills down the columns.** Same
> grid, same size, different order — so the fourth plot you draw lands
> somewhere else.

::: {.content-hidden when-format="revealjs"}

`par(mfrow = c(2, 3))` and `par(mfcol = c(2, 3))` split the device into
exactly the same six boxes. The only difference is the order R walks them,
and that order is the order your `plot()` calls land in.

|  | `mfrow = c(2, 3)` | `mfcol = c(2, 3)` |
|:--|:--|:--|
| top row | 1 · 2 · 3 | 1 · 3 · 5 |
| bottom row | 4 · 5 · 6 | 2 · 4 · 6 |

`par("mfrow")` and `par("mfcol")` both report `2 3` whichever one you set.
The grid is readable back out of `par()`; the fill order is not. Set it once,
at the top, and do not guess later.

:::

::: {.content-visible when-format="revealjs"}

|  | `mfrow = c(2, 3)` | `mfcol = c(2, 3)` |
|:--|:--|:--|
| top row | 1 · 2 · 3 | 1 · 3 · 5 |
| bottom row | 4 · 5 · 6 | 2 · 4 · 6 |

:::

## Four stations, four panels {#panels}

```{r}
#| label: panels-four
#| script: "Four stations, four panels"
#| fig-height: 4.6
#| fig-alt: "Four panels in a two-by-two grid, all on the same January window and the same 0 to 75 vertical scale, each showing one station, each with a dashed red line on January 7. Pasadena peaks highest at 72.5 on January 10; Compton and Long Beach peak lower and a day earlier; Lancaster is flat except for one spike on January 7. A single title runs across the top of all four."
op <- par(mfrow = c(2, 2), mar = c(3, 4, 2, 1), oma = c(0, 0, 3, 0))
invisible(lapply(transect, function(s) {
  x <- jan[jan$site == s, ]
  plot(x$date, x$pm25, type = "o", ylim = c(0, 75), xlim = xr, main = s,
       xlab = "", ylab = "PM2.5 (ug/m3)")
  abline(v = as.Date("2025-01-07"), col = "#b31b1b", lty = 2)
}))
mtext("January 2025, four LA County monitors", outer = TRUE, line = 1)
par(op)
```

::: {.content-hidden when-format="revealjs"}

- `mar` belongs to a panel; `oma` belongs to the **device**.
  `oma = c(0, 0, 3, 0)` reserves three lines across the top of the whole
  device, and `mtext(..., outer = TRUE, line = 1)` writes into the second of
  them. A per-panel `main` could never span four panels. 
- `type = "o"` draws the line **and** the points, so you can still see where
  the real observations are. That is why Pasadena's panel is honest: eleven
  circles, and the line between them is visibly interpolation.

`xlim = xr` is load-bearing. Long Beach's January starts on the 4th, so
without it the January 7 rule would sit at a different horizontal position in
that panel than in the other three, and the four panels would stop being
comparable.

Three things to check on any panel figure: every panel on the same `ylim` and
`xlim`, one outer title rather than four inner ones, and `par(op)` on the
last line.

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
- Shared ylim and xlim, or the panels are four plots pretending to be a comparison. oma + mtext(outer = TRUE) is how one title covers four panels. par(op) on the last line, every time.
:::
:::

## `png()` and `dev.off()`: writing to disk {#png}

```{r}
#| label: png-write
#| script: "Writing a figure to disk"
png("compton.png", width = 1600, height = 900, pointsize = 26)

plot(comp_y$date, comp_y$pm25, type = "l", col = "grey30",
     xlab = "", ylab = "Daily PM2.5 (ug/m3)")
abline(h = 35, col = "#b31b1b", lty = 2)

dev.off()                    # without this the file is unusable

file.exists("compton.png")
```

::: {.content-hidden when-format="revealjs"}

`png()` opens a **file** as the drawing device. From that line on, nothing
appears on screen — every plotting call goes into the file instead.
`dev.off()` closes it and writes it out.

**You must call `dev.off()`.** Forget it and the file is truncated or empty,
and every plot you draw afterwards keeps going into it while your screen
stays blank. If your plots have mysteriously stopped appearing, you have an
open device: call `dev.off()` until R says `null device`.

`width` and `height` are in pixels here. `pointsize` is the base text size,
and it is the setting people forget: enlarge a figure without enlarging the
type and the labels come out unreadably small.

:::

::: {.content-visible when-format="revealjs"}

`png()` opens a **file** as the device. Nothing appears on screen until
`dev.off()` closes it.

Forget `dev.off()` and the file is unusable — and your plots keep vanishing
into it.

:::

## Resolution, bitmap, and vector {#formats}

```{r}
#| label: png-formats
#| script: "Resolution, bitmap and vector"
#| eval: false
# Same figure, three ways.

png("fig-draft.png", width = 800,  height = 600)              # screen draft
plot(comp$date, comp$pm25, type = "l"); dev.off()

png("fig-print.png", width = 2400, height = 1800, res = 300)  # print quality
plot(comp$date, comp$pm25, type = "l"); dev.off()

cairo_pdf("fig-final.pdf", width = 7, height = 5)             # vector, in inches
plot(comp$date, comp$pm25, type = "l"); dev.off()
```

::: {.content-hidden when-format="revealjs"}

**Bitmap** formats — `png()`, `jpeg()`, `tiff()` — store a grid of pixels.
`res` sets how many pixels make an inch. At the default 72 the figure looks
fine on screen and blurry the moment it is printed or enlarged, which is why
a plot that looked sharp in RStudio turns soft in Word. Raising `res` without
raising `width` and `height` shrinks the figure; raise all three together.

**Vector** formats — `cairo_pdf()`, `cairo_ps()` — store the lines and shapes
themselves, so there is no resolution to get wrong and no pixellation at any
size. Their width and height are in **inches**, not pixels. Journals normally
want vector for anything that is not a photograph.

Decide early. Converting a folder of bitmaps to vector after a paper is
accepted means redrawing every figure.

:::

::: {.content-visible when-format="revealjs"}

- **Bitmap** (`png`, `jpeg`): pixels. `res` is pixels per inch; the default 72
  is why figures go soft in Word. Raise `width`, `height` and `res` together.
- **Vector** (`cairo_pdf`): lines, no resolution, no pixellation. Sized in
  **inches**. What journals ask for.

:::

## Exercise 2 {#exercise-2}

::: {.panel-tabset}

### Exercise

```{r}
#| label: exercise-2
#| exercise: true
#| script: "Exercise 2 (4 minutes)"
#| eval: false
# Four minutes.
#
# 1. Add ONE line above the four plot() calls so they fill a 2 x 2 grid.
# 2. Run it. Which station landed in the BOTTOM-LEFT panel?
# 3. Change ONE word in your line so that Pasadena lands TOP-RIGHT.
#
# Two answers to compare with the room: a station name, and one word.

# your line goes here

plot(comp$date, comp$pm25, type="l", lwd=2, ylim=c(0,75), xlim=xr, main="Compton")
plot(lb$date,   lb$pm25,   type="l", lwd=2, ylim=c(0,75), xlim=xr, main="Long Beach")
plot(pas$date,  pas$pm25,  type="l", lwd=2, ylim=c(0,75), xlim=xr, main="Pasadena")
plot(lanc$date, lanc$pm25, type="l", lwd=2, ylim=c(0,75), xlim=xr, main="Lancaster")

# put the device back
```

### Solution

```{r}
#| label: exercise-2-solution
#| purl: false
#| eval: false
op <- par(mfrow = c(2, 2), mar = c(3, 4, 3, 1))
# ... the four plot() calls ...
par(op)
```

|  | `mfrow = c(2,2)` | `mfcol = c(2,2)` |
|:--|:--|:--|
| top-left | Compton | Compton |
| top-right | Long Beach | **Pasadena** |
| bottom-left | **Pasadena** | Long Beach |
| bottom-right | Lancaster | Lancaster |

Bottom-left is **Pasadena**. The one word is **`mfrow` → `mfcol`**.

:::

::: {.content-hidden when-format="revealjs"}

## `plot(x, y)`: two columns against each other {#scatter}

```{r}
#| label: scatter
#| script: "Two columns against each other"
#| fig-alt: "Scatterplot of daily AQI against daily mean PM2.5 for the four stations. The points lie on a single bent curve: steep to about 9 micrograms, shallower to about 35, shallower again beyond."
plot(d$pm25, d$Daily.AQI.Value,
     xlab = "Daily mean PM2.5 (ug/m3)", ylab = "Daily AQI", col = "grey40")
```

1,119 points on one bent curve, with visible knees near 9 and near 35. The
AQI-to-PM2.5 ratio falls from about 5.6 at the low end to 3.55 at 20 and 2.25
at 72.5. AQI is not PM2.5 times a constant — it is a piecewise-linear lookup,
and the scatter draws the breakpoints without anyone telling it they exist.

```{r}
#| label: scatter-cor
#| script: "The correlation that hides the kinks"
round(cor(d$pm25, d$Daily.AQI.Value), 3)
```

A correlation of 0.947 says "these move together" and hides the kinks
completely. **The scatter is the only plot here that shows a shape of
relationship rather than a level.** Its bad-for: 1,119 points read fine;
100,000 would be a solid blob.

:::

## Practice {#practice}

::: {.content-visible when-format="revealjs"}

Fourteen exercises with solutions, on the session page and in the live script.

:::

::: {.content-hidden when-format="revealjs"}

::: {.panel-tabset}

### Practice

```{r}
#| label: practice
#| exercise: true
#| script: "Practice"
#| eval: false
# Everything below uses d, comp_y, jan, comp, lb, lanc, pas and xr, all
# built earlier in this script. Every plot needs one written sentence that
# reads it.

# Choosing the plot
#  1. Name the function you would reach for, one word each:
#     (a) How did PM2.5 at Compton move through 2025?
#     (b) How does the spread of daily readings compare across the four?
#     (c) What does the distribution of every reading look like?
#     (d) How many days did each station report?

# One series
#  2. Draw Lancaster's whole year as a line. What is its highest reading,
#     and on what day?
#  3. Draw Compton's December only, as a line. Add a dashed horizontal
#     reference line at 35 ug/m3. How many December days cleared it?

# Distributions
#  4. Histogram of d$pm25 with default bins. How many bins did you get,
#     and where does the first one start?
#  5. Draw it again with breaks = 5, then breaks = 40. Which one changes
#     the shape you would report, and which only the resolution?
#  6. Histogram of Lancaster's readings only, with breaks = seq(-5, 30, 2.5).
#     One sentence on how its shape differs from the pooled column.

# Groups and counts
#  7. Box plots of pm25 by station, on their side, all four names legible.
#     Which station has the widest box, and is that the same as the highest?
#  8. Box plots of Compton's pm25 by month. Two months look strange --
#     count the days in each month and say why.
#  9. Bar chart of how many days each station reported, sorted from most to
#     fewest. (Hint: sort() works on a table.)

# Several series
# 10. Replicate this figure. January 2025, one panel: Compton in red, Long
#     Beach in grey, Lancaster in blue, all lwd = 2, y from 0 to 60, a
#     dashed horizontal line at 35, and a legend with no box in the top
#     right. Title: "January 2025, three stations". Then say what Lancaster
#     does that the other two do not.
# 11. Compton's whole year twice, stacked in two panels: once type = "l",
#     once type = "p". The line tells one lie the points do not. Find it.

# Panels and the canvas
# 12. Four histograms, one per station, in a 2 x 2 grid, all on the same
#     breaks and the same x range. Put the device back when you are done.
# 13. Take exercise 10's figure and give it a left margin of 6 lines, a
#     y-axis label written with mtext(), and horizontal axis numbers.

# To disk
# 14. Write exercise 12's figure to "stations.png" at 2000 x 1500 pixels
#     with pointsize = 30. Open the file. Then write the same figure to
#     "stations.pdf" with cairo_pdf() at 8 x 6 inches, and open that.
#     Zoom both to 400%. Which one still has clean edges?
```

### Solutions

```{r}
#| label: practice-solutions
#| purl: false
#| eval: false
# 1  (a) plot   (b) boxplot   (c) hist   (d) barplot

# 2  the year's high is 24.9, on 2025-01-07 -- the windstorm day itself.
#    Lancaster's worst day of the whole year is blowing dust, not smoke
lan_y <- d[d$site == "Lancaster - Fairgrounds", ]
plot(lan_y$date, lan_y$pm25, type = "l", xlab = "", ylab = "PM2.5 (ug/m3)")
max(lan_y$pm25); lan_y$date[which.max(lan_y$pm25)]

# 3  five days clear 35; December is the year's worst month at this monitor
dec <- comp_y[comp_y$date >= as.Date("2025-12-01"), ]
plot(dec$date, dec$pm25, type = "l", xlab = "", ylab = "PM2.5 (ug/m3)")
abline(h = 35, col = "#b31b1b", lty = 2)
sum(dec$pm25 > 35)

# 4  16 bins, first break at -5: the negatives hide inside a bar that looks
#    like it starts at zero
h <- hist(d$pm25)
length(h$counts); h$breaks[1]

# 5  breaks = 5 gives five bins twenty units wide: everything above 20
#    collapses into three near-empty bars and the skew stops being
#    readable. breaks = 40 gives 38 bins two units wide -- more
#    resolution on the same shape, not a different shape
hist(d$pm25, breaks = 5)
hist(d$pm25, breaks = 40)

# 6  a much tighter, less skewed column: Lancaster's mass sits under 8 with
#    a short tail, where the pooled column runs past 70
hist(d$pm25[d$site == "Lancaster - Fairgrounds"], breaks = seq(-5, 30, 2.5),
     main = "", xlab = "PM2.5 (ug/m3)")

# 7  Compton has the widest box (7.9 to 15.5) and also the highest median,
#    but Long Beach is within a twentieth of it -- spread and level are
#    different questions
op <- par(mar = c(4, 15, 1, 1))
boxplot(pm25 ~ site, data = d, horizontal = TRUE, las = 1,
        xlab = "PM2.5 (ug/m3)", ylab = "")
par(op)

# 8  April and May: the regulatory sampler was down, so those boxes are
#    built from 2 and 1 readings and mean nothing
comp_y$month <- format(comp_y$date, "%m")
boxplot(pm25 ~ month, data = comp_y)
table(comp_y$month)

# 9
op <- par(mar = c(4, 15, 1, 1))
barplot(sort(table(d$site)), horiz = TRUE, las = 1,
        xlab = "Days with a reading in 2025")
par(op)

# 10  Lancaster stays flat through the fire week and spikes on Jan 7 itself
plot(comp$date, comp$pm25, type = "l", col = "#b31b1b", lwd = 2,
     ylim = c(0, 60), xlab = "", ylab = "PM2.5 (ug/m3)",
     main = "January 2025, three stations")
lines(lb$date,   lb$pm25,   col = "grey30",  lwd = 2)
lines(lanc$date, lanc$pm25, col = "#1f6fb4", lwd = 2)
abline(h = 35, lty = 2)
legend("topright", c("Compton", "Long Beach", "Lancaster"),
       col = c("#b31b1b", "grey30", "#1f6fb4"), lwd = 2, bty = "n")

# 11  the line draws straight across April and May, where there are almost
#     no readings; the points show the gap for what it is
op <- par(mfrow = c(2, 1), mar = c(3, 4, 2, 1))
plot(comp_y$date, comp_y$pm25, type = "l", xlab = "", ylab = "PM2.5", main = "type = l")
plot(comp_y$date, comp_y$pm25, type = "p", xlab = "", ylab = "PM2.5", main = "type = p")
par(op)

# 12  same skew everywhere, but Lancaster's is compressed against zero
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
invisible(lapply(sort(unique(d$site)), function(s)
  hist(d$pm25[d$site == s], breaks = seq(-5, 75, 2.5), xlim = c(-5, 75),
       main = s, xlab = "", cex.main = 0.8)))
par(op)

# 13
op <- par(mar = c(4, 6, 2, 1), las = 1)
plot(comp$date, comp$pm25, type = "l", col = "#b31b1b", lwd = 2,
     ylim = c(0, 60), xlab = "", ylab = "",
     main = "January 2025, three stations")
lines(lb$date,   lb$pm25,   col = "grey30",  lwd = 2)
lines(lanc$date, lanc$pm25, col = "#1f6fb4", lwd = 2)
abline(h = 35, lty = 2)
mtext("Daily PM2.5 (ug/m3)", side = 2, line = 4, las = 0)
par(op)

# 14  the PDF. A bitmap has a fixed pixel grid, so zooming shows the grid;
#     a vector file stores the lines themselves and redraws them at any size
png("stations.png", width = 2000, height = 1500, pointsize = 30)
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
invisible(lapply(sort(unique(d$site)), function(s)
  hist(d$pm25[d$site == s], breaks = seq(-5, 75, 2.5), xlim = c(-5, 75),
       main = s, xlab = "")))
par(op)
dev.off()

cairo_pdf("stations.pdf", width = 8, height = 6)
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
invisible(lapply(sort(unique(d$site)), function(s)
  hist(d$pm25[d$site == s], breaks = seq(-5, 75, 2.5), xlim = c(-5, 75),
       main = s, xlab = "")))
par(op)
dev.off()
```

:::

:::

## Quick reference {#quick-reference}

::: {.content-visible when-format="revealjs"}

On the session page: the four plots, the arguments that port between them,
the layers, the canvas, and writing to disk.

:::

::: {.content-hidden when-format="revealjs"}

**The four plots**

| Code | What it draws |
|:--|:--|
| `plot(x, y)` | two columns against each other, one point per row |
| `plot(x, y, type = "l")` | one series in its natural order; `"p"` `"b"` `"o"` are the other marks |
| `hist(x, breaks = 40)` | the distribution of one variable; `breaks` is a suggestion |
| `boxplot(y ~ g, data = d)` | `y` split by the groups in `g`; read `~` as "by" |
| `barplot(table(g))` | how often each category shows up |

**Arguments that mean the same thing everywhere**

| Code | What it does |
|:--|:--|
| `main =`, `xlab =`, `ylab =` | title and axis labels; units go in `ylab` |
| `col =` | colour of the marks, or the fill on `hist`/`bar`/`box` |
| `xlim =`, `ylim =` | the window; set on the **first** call |
| `lwd =`, `lty =` | line width; line type, `1` solid `2` dashed |
| `pch =` | point symbol; `16` filled circle, `1` open |
| `las =` | axis text: `0` parallel, `1` horizontal, `2` perpendicular |
| `horiz =` / `horizontal =` | sideways; `barplot` spells it `horiz`, `boxplot` `horizontal` |

**Layers on top of an open plot**

| Code | What it adds |
|:--|:--|
| `points(x, y)`, `lines(x, y)` | marks, a joined line |
| `abline(h =, v =)` | a horizontal or vertical reference line |
| `text(x, y, "label")` | a label at data coordinates |
| `legend("topright", labels, col =, lty =, pch =, bty = "n")` | the key; `NA` means no sample for that entry |
| `axis(1, at =, labels =)` | your own axis; suppress R's with `xaxt = "n"` |
| `mtext("x", side = 2, line = 3)` | text in the margin; sides 1–4 from the bottom, clockwise |

**The canvas**

| Code | What it does |
|:--|:--|
| `par("mar")` | **read** a setting |
| `op <- par(mar = c(4, 15, 1, 1))` | **write** it and keep the old value; bottom, left, top, right, in lines |
| `par(op)` | put it back — type this line before you type the plot |
| `par(mfrow = c(2, 2))` | panels, filled across the rows |
| `par(mfcol = c(2, 2))` | same grid, filled down the columns |
| `par(oma = c(0, 0, 3, 0))` | outer margin, around **all** the panels |
| `mtext("title", outer = TRUE, line = 1)` | one title across every panel |
| `layout(m, heights =, widths =)` | panels of unequal size; reset with `layout(1)` |

**Writing to disk**

| Code | What it does |
|:--|:--|
| `png("f.png", width =, height =, pointsize =)` | open a bitmap file as the device; pixels |
| `png(..., res = 300)` | pixels per inch; the default 72 is why figures go soft in print |
| `cairo_pdf("f.pdf", width = 7, height = 5)` | vector; no resolution, sized in **inches** |
| `dev.off()` | close the device and write the file — **required** |

**Links**

- Session materials: [slides](04-basic-plots-slides.html) ·
  [live script](04-basic-plots.R) ·
  [EPA data](data/epa_pm25_la_county_2025.csv) ·
  [where it came from](data/SOURCE.md)
- [Syllabus](../syllabus.qmd) · [full schedule](../index.qmd#schedule)

:::

```{r}
#| purl: false
#| include: false
# Single-source build: the .R live script is generated from this file every
# time the page renders, so it cannot fall out of step with the prose.
source("_purl.R")
purl_session("04-basic-plots.qmd", "04-basic-plots.R")
```
