---
title: "4 · Basic plots"
subtitle: "Thursday, September 3, 2026"
pagetitle: "AEM 6850 · Session 4 · Basic plots"
execute:
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
  revealjs:
    output-file: 04-basic-plots-slides.html
    theme: [default, slides.scss]
    template-partials:
      - title-slide.html
    footer: "AEM 6850 · Session 4"
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    code-overflow: scroll
    history: false
    link-external-newwindow: true
---

```{r}
#| purl: false
#| include: false
knitr::opts_chunk$set(
  comment = "#>",
  collapse = FALSE,
  fig.width = 8,
  fig.height = 4,
  fig.align = "center",
  dpi = 150
)
```

::: {.content-hidden when-format="revealjs"}

[Slides](04-basic-plots-slides.html) · [Live script](04-basic-plots.R) ·
[Source](https://github.com/ArielOrtizBobea/aem6850/blob/main/fall-2026/sessions/04-basic-plots.qmd)

This session adds the checks `str()` and `summary()` cannot do: four
base-R plots, one line each, used as instruments rather than
illustrations.

:::

## Objectives {#objectives}

By the end of this session you can:

- Produce one-line diagnostic plots with base R.
- Choose the plot that matches a suspicion about a column.
- Read a distribution for skew, outliers, and impossible values.
- Say what a plot shows that a summary table cannot.

::: {.content-hidden when-format="revealjs"}

## Before class {#before-class}

Today uses `data/epa_pm25_la_county_2025.csv` and a Los Angeles weather
file. If either is missing, the [rebuild section](#rebuild) has a download
fallback that takes thirty seconds.

:::

## What a plot is for {#thesis}

> **A summary answers the question you asked. A plot answers questions you
> did not think to ask.**

Four instruments, one line each, over the data frames you already know.

## Rebuild the table {#rebuild}

Every monitor in the county, all of 2025.

```{r}
#| label: rebuild
#| script: "Rebuild the table"
pm <- read.csv("data/epa_pm25_la_county_2025.csv",
               colClasses = c("Site.ID" = "character"))
names(pm)[names(pm) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"
pm$date  <- as.Date(pm$Date, format = "%m/%d/%Y")
pm$month <- format(pm$date, "%m")

frm <- pm[pm$AQS.Parameter.Code == 88101 & pm$POC == 1, ]   # one instrument per site
tb  <- tapply(frm$pm25, list(frm$Local.Site.Name, frm$month), mean)
dim(tb)                                                     # 8 sites, 12 months
```

And one site's winter, which most of today's instruments read:

```{r}
#| label: rebuild-one
#| script: "One site, one winter"
one <- frm[frm$Local.Site.Name == "Compton" &
           frm$date <= as.Date("2025-02-28"), ]
nrow(one)   # 59 -- one row per day
```

::: {.content-hidden when-format="revealjs"}

The read protects the identifier column, renames the long one, converts the
date, and pulls the month out. `frm` keeps the federal reference method and
the first instrument at each site, so a site contributes one number a day.
`tb` is a matrix: eight sites down, twelve months across.

`one` narrows that to a single monitor over January and February, which is
small enough to see every day at once. Today draws both.

::: {.callout-tip collapse="true"}
## If `data/` is missing a file

Run these once and you are caught up. They fetch the same snapshots these
pages use:

```{r}
#| label: data-fallback
#| script: "If data/ is missing a file"
#| eval: false
# dir.create("data", showWarnings = FALSE)
# download.file(paste0("https://arielortizbobea.github.io/aem6850/",
#                      "fall-2026/sessions/data/epa_pm25_la_county_2025.csv"),
#               "data/epa_pm25_la_county_2025.csv")
# download.file(paste0("https://arielortizbobea.github.io/aem6850/",
#                      "fall-2026/sessions/data/la-weather-dec2024-feb2025.csv"),
#               "data/la-weather-dec2024-feb2025.csv")
```

[The data folder's provenance notes](data/SOURCE.md) document where both
files came from, byte for byte.
:::

:::

## Four instruments {#instruments}

::: {.tight-cols}

| The instrument | What it draws |
|:--|:--|
| `hist(x)` | The shape of one numeric column |
| `plot(x, y)` | Two columns against each other, or one through time |
| `boxplot(y ~ g)` | One column, split by the groups in another |
| `barplot(table(g))` | How often each category shows up |

:::

::: {.content-hidden when-format="revealjs"}

Call them instruments, not figures. A figure serves readers and takes an
hour to polish. An instrument serves you. It takes one line and ten
seconds, it can stay ugly, and it never leaves your screen. You make dozens
of them on a normal working day.

Nothing new sits underneath: every one of these takes ordinary vectors and
data-frame columns. The only new notation is the `~` in `boxplot()`. We
will read it aloud when we get there. Later we add three that build *on* a
plot rather than making one: `points()`, `legend()` and `axis()`.

:::

::: {.content-visible when-format="revealjs"}

These are instruments for you, not figures for readers. Each takes one line
and ten seconds, and it never leaves your screen.

:::

## `hist()`: the shape of one column {#hist}

```{r}
#| label: hist-default
#| script: "The first instrument: hist()"
#| fig-alt: "Histogram of daily PM2.5 at Compton for January and February 2025, drawn with default settings. Most days fall between 5 and 25 micrograms per cubic meter, and a long right tail stretches past 50."
hist(one$pm25)
```

::: {.content-hidden when-format="revealjs"}

The call takes one line, default title and all. R chops the column into
bins and draws how many days fell in each. You now see all fifty-nine days
at once, where `summary()` would compress them into six numbers.

Do not clean this plot up. It is an instrument reading, and it can stay
ugly. The moment you polish, you are making a figure. The cost jumps from
seconds to minutes, and you will stop making these plots.

:::

## Reading the shape {#hist-read}

```{r}
#| label: hist-breaks
#| script: "Turning up the resolution"
#| fig-alt: "The same histogram with twenty bins and a dashed red line at 35 micrograms per cubic meter. The bulk of days sits between 5 and 25, and a separated island of four days lies beyond 40."
hist(one$pm25, breaks = 20,
     main = "", xlab = "Daily PM2.5 (ug/m3)")
abline(v = 35, col = "#b31b1b", lty = 2)   # the EPA 24-hour standard
```

::: {.content-hidden when-format="revealjs"}

`breaks = 20` asks for more bins, which means more resolution on the same
column. Read three things off this plot, in this order:

- **Where the mass sits.** Most winter days at Compton land between 5 and 25
  µg/m³. That is what ordinary air looks like at this monitor.
- **The long right tail.** The distribution is *right-skewed*: the column
  cannot go below zero, so it piles up low and stretches right. Skew is why
  the mean (17.6) sits above the median (15.3). A few bad days pull the
  average up.
- **The island.** Four separated days sit past the red line, the US EPA's
  24-hour standard of 35 µg/m³: January 1, and January 8, 9, 10. A gap
  between the mass and a clump of stragglers signals *events*, distinct
  from the ordinary process that made the rest.

:::

::: {.content-visible when-format="revealjs"}

- **The mass:** ordinary days, 5–25 µg/m³.
- **The tail:** right skew; mean 17.6 sits above median 15.3.
- **The island past 35:** four days. January 1, then 8, 9, 10.

:::

## The sentinel {#sentinel}

::: {.content-hidden when-format="revealjs"}

Our file comes from AQS, EPA's air-quality archive, and it marks a missing
day by *absence*: the row is not there. Plenty of files instead put a number
where the gap is: `-999`, `9999`, `0`. That number is called a **sentinel**.
EPA's own real-time feed, AirNow, ships `-999` to this day. Sooner or later
you will open a file with sentinels in it.

::: {.column-margin}
**Why −999?** Old data formats could not leave a field blank. Every cell had
to hold a number, so agencies picked one no real measurement could produce.
The formats died; the convention did not.
:::

This file is clean, so we corrupt a copy ourselves and watch the failure
happen before it costs you anything:

:::

```{r}
#| label: sentinel
#| script: "The sentinel"
bad <- one$pm25
bad[c(40, 45, 50, 55)] <- -999   # four missing days, the AirNow way
mean(bad)
summary(bad)
```

::: {.content-hidden when-format="revealjs"}

A stray `"n/a"` in a numeric column at least produces an `NA` and a
warning. This produced nothing. `-999` is a legitimate number, so R raises no error and no warning.
`mean()` ran clean and returned garbage. If −51 went into a difference or
a ratio instead of onto your screen, nothing downstream would flag it.

`summary()` does show the damage: `Min. -999`. But only if you read all six
numbers, and only if you know the physical floor is zero.

:::

::: {.content-visible when-format="revealjs"}

- No error, no warning, no `NA`. It ran clean and it is wrong.
- `summary()` shows it, if you read all six numbers and know the floor is
  zero.

:::

## The histogram catches it {#sentinel-hist}

```{r}
#| label: sentinel-hist
#| script: "The histogram catches it"
#| fig-alt: "Histogram of the corrupted column: a lone short bar at minus 1000, a wide empty gap, and all the real data crushed into a single tall bar near zero."
hist(bad, main = "", xlab = "Daily PM2.5 (ug/m3)")
```

```{r}
#| label: sentinel-repair
#| script: "The repair"
bad[bad == -999] <- NA        # sentinel to honest missing
mean(bad, na.rm = TRUE)       # back to 17.6
```

::: {.content-hidden when-format="revealjs"}

You do not need to know what PM2.5 is to see that something is wrong here.
One bar sits a thousand units from everything else. The axis stretched to
fit the garbage and crushed the real data into a single tower.
**A ruined histogram is itself a reading.** When everything you care about
collapses into one bin, the axis made room for a value that should not
exist. Go find that value.

The repair is one line of logical subsetting, and the mean returns to
17.6. The order matters: *look, then repair, then compute*.
Never go straight to the computing.

:::

## Exercise 1 {#exercise-1}

::: {.panel-tabset}

### Exercise

```{r}
#| label: exercise-1
#| exercise: true
#| script: "Exercise 1 (5 minutes)"
#| eval: false
# 1. hist(frm$pm25) -- every monitor, every day of 2025, not one winter.
#    How does the shape differ from hist(one$pm25), and would you have
#    expected that?
#
# 2. Your turn to corrupt. Copy one$pm25 and set three days to 0 instead
#    of -999. Does hist() catch it? Does summary()? Why is 0 the harder
#    one?
#
# 3. tb is a matrix, and hist(tb) works anyway. What is it drawing, and
#    how many values went into it?
```

### Solution

```{r}
#| label: exercise-1-solution
#| purl: false
#| eval: false
# 1  same right skew, longer tail: eight sites and all four seasons
hist(frm$pm25)

# 2  neither one flags it clearly
z <- one$pm25
z[c(10, 20, 30)] <- 0
hist(z)                 # the bars barely move
summary(z)              # Min. 0.00 -- legal-looking

min(one$pm25)           # 1.7: the real floor at this monitor
# -999 is physically impossible and blows up the axis. 0 is a plausible
# reading, so it hides until you know what this monitor's floor is.

# 3  the matrix read as one vector: 96 monthly means
hist(tb)
length(tb)
```

:::

## `plot()`: a column through time {#plot-time}

```{r}
#| label: plot-time
#| script: "A column through time"
#| fig-alt: "Daily PM2.5 at Compton drawn as vertical spikes from January through February 2025. The tallest spike stands alone on January 1. A red vertical line marks January 7, followed immediately by a three-day cluster of high readings. February is much calmer."
plot(one$date, one$pm25, type = "h", col = "grey30",
     xlab = "", ylab = "Daily PM2.5 (ug/m3)")
abline(v = as.Date("2025-01-07"), col = "#b31b1b", lwd = 2)
```

::: {.content-hidden when-format="revealjs"}

`type = "h"` draws one spike per day. When a column has a natural order,
usually time, plot it in that order. The whole winter then shows in one
glance: a lone spike on January 1, the red line at January 7, a broad
three-day shoulder as the fires burn, then a February back to normal.

Now rank the same days instead of drawing them:

```{r}
#| label: plot-time-sorted
#| script: "The same days, ranked"
head(one[order(-one$pm25), c("date", "pm25")], 5)
```

The ranking puts January 9, 8 and 10 at places two, three and four: three
separate rows, with the order destroyed. The plot shows what they actually
are: **one event**, three days wide, starting the day after ignition. It
also shows January 1 the way a ranking never could. That spike stands
*alone*, with quiet days on either side. Fireworks last one night; fires
last days.

:::

::: {.content-visible when-format="revealjs"}

- The sort gave ranks 1, 2, 3, 4. The plot shows a one-day spike, then a
  **three-day event** starting the day after ignition.
- Fireworks last one night; fires last days.

:::

## `plot()`: two columns together {#plot-scatter}

A Los Angeles weather file, ninety winter days, saved alongside these
pages.

```{r}
#| label: weather-load
#| script: "Load the weather file"
wx <- read.csv("data/la-weather-dec2024-feb2025.csv", skip = 3)
names(wx) <- c("date", "gust", "wind", "tmax", "rh_min", "precip")
wx$date <- as.Date(wx$date)
```

```{r}
#| label: plot-scatter
#| script: "Two columns together"
#| fig-alt: "Scatterplot of daily maximum gust against daily maximum sustained wind for ninety winter days in Los Angeles. The points form one tight upward ray. The windiest days sit far up the ray but on it."
plot(wx$wind, wx$gust,
     xlab = "Max sustained wind (km/h)", ylab = "Max gust (km/h)")
```

::: {.content-hidden when-format="revealjs"}

A scatterplot draws a *relationship*, and these two columns must have one.
A gust is a burst on top of the sustained wind, so gusts should run above
wind speed by a believable factor. They do: one tight ray, with gusts at
roughly two to four times the sustained speed, every day of the winter.

Look at where January 7 and 8 sit: far up the ray, *and on it*. Both columns
are extreme; the relationship is ordinary. A real windstorm looks like this.
**Extreme is not the same as wrong.** The point to worry about is never the
one far along the ray. It is the one *off* the ray.

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
Read it live: one ray, gusts at 2–4× the sustained wind, every day. Point
at Jan 7–8, far up the ray and ON it: extreme, but consistent. The point to
worry about is never far along the ray; it is off it. That sentence sets up
the next slide.
:::
:::

## The unit error {#unit-error}

::: {.content-hidden when-format="revealjs"}

This instrument exists to catch one classic failure. Weather feeds serve
wind in km/h, m/s, mph, or knots, and the choice sits in a header field or
a query parameter; Open-Meteo names its parameter `wind_speed_unit`. Stitch
together files pulled with different settings and you get one column with
two units inside. Again we corrupt a copy ourselves, to watch it happen:

:::

```{r}
#| label: unit-error
#| script: "The unit error"
mixed <- wx$wind
jan   <- format(wx$date, "%Y-%m") == "2025-01"
mixed[jan] <- round(wx$wind[jan] / 3.6, 1)   # January delivered in m/s

summary(wx$wind)    # the honest column
summary(mixed)      # the corrupted one
```

::: {.content-hidden when-format="revealjs"}

Read the corrupted summary and try to reject it. The minimum, 0.9, reads as
a calm day. The maximum, 21.8, reads as a windy one. The mean, 6.5, runs
low for winter, and you would need LA wind climatology in your head
to doubt it. **Every number in it is a plausible wind speed in km/h.**
There is no sentinel to spot and no impossible value to catch. The
corruption lives in how thirty-one values relate to the rest, and single
numbers cannot show a relationship.

:::

::: {.content-visible when-format="revealjs"}

Min 0.9, max 21.8, mean 6.5: **every number is a plausible wind speed.**
Nothing is impossible; the corruption lives in the *relationships*.

:::

## The scatterplot catches it {#unit-error-plot}

```{r}
#| label: unit-error-plot
#| script: "The scatterplot catches it"
#| fig-alt: "The same gust-versus-wind scatterplot after January's wind speeds are converted to meters per second: the single ray splits into two distinct rays with clearly different slopes."
plot(mixed, wx$gust,
     xlab = "Wind speed, as delivered", ylab = "Max gust (km/h)")
```

::: {.content-hidden when-format="revealjs"}

The plot shows two rays. The right-hand ray is December and February, with
gusts at two to four times the wind, as before. The steep left-hand ray is
January: the same relationship, times 3.6, because its x-values shrank by
exactly the km/h-to-m/s factor. Thirty-one points do not land on a second
straight line by accident. **A ray that splits in two means two regimes in
one column**, and a clean split at a constant factor means units.

This is the strongest version of today's claim. `summary()` showed the
sentinel, barely. It never showed the unit error. No single number was
wrong, so no list of single numbers could catch it. Only the picture of the
*relationship* could.

:::

::: {.content-visible when-format="revealjs"}

- Two rays mean two regimes in one column. A clean constant factor means
  units.
- `summary()` had nothing to show: no individual number was wrong.
- Only the picture of the relationship could catch it.

:::

## `boxplot()`: a column by groups {#boxplot}

```{r}
#| label: boxplot-month
#| script: "A column by groups: boxplot()"
#| fig-alt: "Box plots of daily PM2.5 for January and February 2025. January's box sits higher and stretches wider, with a single outlier dot far above its whisker. February's box is lower and tighter, with no outliers."
one$month <- format(one$date, "%Y-%m")
boxplot(pm25 ~ month, data = one)
```

::: {.content-hidden when-format="revealjs"}

Read `pm25 ~ month` aloud as "pm25, *by* month". The `~` is R's word for
"split by," and `data = one` says which data frame the two names live in.
(Use `"%Y-%m"` rather than `"%m"` so the labels sort in time order when a
window crosses New Year. The weather file's December would otherwise sort
after February.)

One box per group, five readings per box: the heavy line marks the median;
the box holds the middle half of the days; the whiskers cover the ordinary
range; the plot draws anything beyond them as its own dot. A dot is a
candidate for a closer look, not automatically an error.

Now read this plot against the histogram, because the two instruments
disagree in an instructive way. The histogram showed an island of four days.
Here only one of them survives as a dot: January 1. The three fire days sit
at the top of January's whisker, *inside* the ordinary range. A boxplot
judges "beyond" against each group's own spread, and January's spread is
enormous, so three bad days read as the high end of normal. Neither
instrument is wrong. They answer different questions: "extreme for this
data?" versus "extreme for this group?"

The medians hold the quieter finding: 16.9 in January against 12.6 in
February. Take out the four event days and January's median still reads
14.5 to February's 12.6. The baseline shifted, not just a few days. The
plot does not say why. Lingering smoke from a fire month? Stronger winter
inversions? The plot raises the question and leaves it open. You knew about
the events. You probably did not know about the shifted baseline. That
second kind of finding is why you make these plots.

:::

::: {.content-visible when-format="revealjs"}

- `pm25 ~ month` reads "pm25, **by** month."
- One dot survives: Jan 1. The fire days land *inside* January's whisker,
  because a boxplot judges "extreme" against each group's own spread.
- The quieter finding: January's *median* (16.9) sits above February's (12.6).
  The baseline shifted, not just the events.

:::

## `barplot()`: drawing what you computed {#barplot}

`tb` is a matrix of means. One margin of it is already a named vector, which
is exactly what `barplot()` wants.

```{r}
#| label: barplot-margin
#| script: "Drawing a matrix margin"
#| fig-height: 4.5
#| fig-alt: "Horizontal bar chart of mean PM2.5 for eight Los Angeles County monitors in 2025. Lancaster is much the lowest at about 4.5 micrograms per cubic meter; the seven basin monitors cluster between 9 and 13, with Compton highest."
par(mar = c(4, 14, 1, 1))                      # room for the site names
barplot(sort(apply(tb, 1, mean)), horiz = TRUE, las = 1, cex.names = 0.85,
        xlab = "Mean PM2.5, 2025 (ug/m3)")
par(mar = c(5, 4, 4, 2) + 0.1)                 # put the margins back
```

::: {.content-hidden when-format="revealjs"}

`apply(tb, 1, mean)` gives one number per site, carrying the row names with
it, and `barplot()` draws named vectors directly. `horiz = TRUE` with
`las = 1` turns the labels upright so you can read them; `par(mar = ...)`
widens the left margin to fit them, and the last line puts it back. Widen
too little and R silently clips the longest name, so look at the plot.

Lancaster, out in the high desert past the mountains, reads about a third of
what the basin monitors do. The other seven sit within a few micrograms of
each other. The county is not one airshed, and the map is doing the work:
geography sets the level, and the basin traps what is emitted into it.

:::

::: {.content-visible when-format="revealjs"}

Lancaster, out in the desert, reads a third of the basin monitors. Seven
basin sites sit within a few µg/m³ of each other.

:::

## How many days is each bar? {#coverage}

Every cell of `tb` is filled. None of them says how many days went into it.

```{r}
#| label: barplot-coverage
#| script: "Counts of categories: barplot()"
#| fig-height: 4.5
#| fig-alt: "Horizontal bar chart of the number of days with a reading per monitor in 2025. Signal Hill has 57 and Lancaster has 365, a more than sixfold range across the eight monitors."
par(mar = c(4, 14, 1, 1))
barplot(sort(table(frm$Local.Site.Name)), horiz = TRUE, las = 1, cex.names = 0.85,
        xlab = "Days with a reading, 2025")
par(mar = c(5, 4, 4, 2) + 0.1)
```

::: {.content-hidden when-format="revealjs"}

`table()` counts and `barplot()` draws the counts. Any categorical column
goes through that pair and comes out as a picture of coverage.

Signal Hill reported 57 days. Lancaster reported 365. Both produced a
complete row of twelve monthly means, and **the matrix cannot tell you
which is which.** A mean built from four days looks exactly like a mean
built from thirty-one.

And the bars fall into two clumps with a gap between them: four monitors
near a full year, four near a third of it. Two clumps mean two populations,
the same reading you would take off a histogram. Filter-based samplers run
one day in six by design; continuous monitors run daily.

That is not an error in the file. It is a fact about the data that only the
count plot shows, and it decides how much weight each of the bars above can
carry.

:::

::: {.content-visible when-format="revealjs"}

- Signal Hill: **57 days**. Lancaster: **365**. Both gave a full row of
  monthly means.
- A mean from four days looks exactly like a mean from thirty-one.

:::

## The missing spring {#missing-spring}

```{r}
#| label: missing-spring
#| script: "The missing spring: site or sampler?"
compton <- frm[frm$Local.Site.Name == "Compton", ]
table(compton$month)              # the regulatory sampler, month by month

other <- pm[pm$Local.Site.Name == "Compton" &
            pm$POC == 3 & pm$AQS.Parameter.Code == 88502, ]
table(other$month)                # the site's OTHER instrument
```

::: {.content-hidden when-format="revealjs"}

Compton is one of the well-covered sites, at 301 days. Ask where they fell
and the answer has a hole in it: **April has two days. May has one.**

So check against an independent measurement. The site carries one. The
continuous instrument, POC 3, reported 30 days in April and 31 in May. The
*site* was fine; the *regulatory sampler* was down for two months. (AQS
validation can also invalidate readings after the fact. Either way, the
regulatory series has no spring.)

Here is what the gap costs. Compute "Compton's 2025 annual mean" from the
regulatory series and nothing errors, nothing warns. The number silently
skips almost all of spring, a mild and cleaner season, so it lands biased.
Every mean is a mean *of the days that are there*, and that includes the
mean sitting in row "Compton" of `tb`.

:::

::: {.content-visible when-format="revealjs"}

- Compton has 301 days — but **April: 2. May: 1.**
- POC 3 ran 30 and 31 those months: the **sampler** was down, not the site.
- Every mean is a mean *of the days that are there.*

:::

## Two series on one plot {#layering}

The counts said the sampler was down. Draw both instruments and you can see
it.

```{r}
#| label: layering
#| script: "Two series on one plot"
#| fig-alt: "Daily PM2.5 at Compton for all of 2025. A grey line traces the continuous instrument every day of the year. Red dots mark the regulatory sampler, and they disappear entirely through April and May while the grey line continues."
reg <- frm[frm$Local.Site.Name == "Compton", ]
con <- pm[pm$Local.Site.Name == "Compton" &
          pm$POC == 3 & pm$AQS.Parameter.Code == 88502, ]

plot(con$date, con$pm25, type = "l", col = "grey60",
     xlab = "", ylab = "Daily PM2.5 (ug/m3)")
points(reg$date, reg$pm25, pch = 16, cex = 0.5, col = "#b31b1b")
legend("topright", c("continuous (POC 3)", "regulatory (POC 1)"),
       col = c("grey60", "#b31b1b"), lty = c(1, NA), pch = c(NA, 16),
       bty = "n")
```

::: {.content-hidden when-format="revealjs"}

`plot()` opens a window and sets the axes. Everything after it *adds* to
that window: `points()` for markers, `lines()` for a connecting line,
`abline()` for a straight reference. They draw on whatever `plot()` already
scaled, so the first call has to be the one with the widest range, or the
later ones fall off the edge.

Three arguments do the distinguishing. `pch` picks the point symbol (`16` is
a filled circle), `cex` scales it, and `col` colors it. `legend()` then says
which is which: a position, the labels, and the same `col`/`pch`/`lty` you
drew with, repeated so the key matches the plot. `lty = c(1, NA)` gives the
first entry a line and no symbol; `pch = c(NA, 16)` does the reverse.
`bty = "n"` drops the box around the key.

Now read it. The two instruments track each other all year, which is the
reassuring part: two different measurement methods, same air, same story.
Then **the red dots stop.** They vanish through April and May and pick back
up in June, while the grey line runs straight through. That is the gap the
count plot found, in the form you would actually notice while working.

There is a second thing in here that nobody went looking for: the sharp
spike in early July, on both instruments. Fireworks again, six months after
the New Year's one.

:::

::: {.content-visible when-format="revealjs"}

- `plot()` scales the window; `points()`, `lines()` and `abline()` add to it.
- `pch` symbol, `cex` size, `col` color, and `legend()` says which is which.
- The red dots **stop** through April and May. The grey line does not.

:::

## When the default axis is wrong {#axis}

You could see the gap. You could not say which months it covered, because R
labelled the axis by quarter.

```{r}
#| label: axis
#| script: "When the default axis is wrong"
#| fig-alt: "The same two-instrument plot of Compton 2025, now with a tick and an abbreviated month name at the first of every month. The gap in the red regulatory dots is now readable as April and May."
plot(con$date, con$pm25, type = "l", col = "grey60",
     xaxt = "n",                     # draw no x-axis: we will do it
     xlab = "", ylab = "Daily PM2.5 (ug/m3)")
points(reg$date, reg$pm25, pch = 16, cex = 0.5, col = "#b31b1b")

firsts <- as.Date(paste0("2025-", sprintf("%02d", 1:12), "-01"))
axis(1, at = firsts, labels = month.abb)
```

::: {.content-hidden when-format="revealjs"}

`xaxt = "n"` tells `plot()` to skip the x-axis, and `axis()` then draws the
one you want. Its first argument is the side, numbered clockwise from the
bottom: 1 bottom, 2 left, 3 top, 4 right. `at` says where the ticks go and
`labels` says what to write on them; the two have to be the same length.
`month.abb` is a vector R ships with, the twelve abbreviated month names.

Read the gap now. A couple of red dots in early April, then nothing at all
until June. That is April and May, which is what the counts said, and now
you can say it off the picture instead of trusting a table.

This is the one piece of axis work worth doing on an instrument. Not because
it is prettier, but because the default was **unreadable for the question**:
quarterly ticks cannot resolve a two-month gap. Reach for `axis()` when the
default hides what you are looking at, and leave it alone otherwise. Add
`las = 2` to turn the labels sideways when they collide.

:::

::: {.content-visible when-format="revealjs"}

- `xaxt = "n"` suppresses the axis; `axis(1, at =, labels =)` draws yours.
- Sides are numbered from the bottom, clockwise: 1, 2, 3, 4.
- Now the gap is nameable: **April and May.**

:::

## Choosing the instrument {#choosing}

Start from the suspicion, not from the plot type.

::: {.tight-cols}

| The instrument | The suspicion it answers |
|:--|:--|
| `hist(x)` | Anything impossible in here? Skew? A gap? A second hump? |
| `plot(d$date, y, type = "h")` | Did something happen in time: a spike, a break, a dead stretch? |
| `plot(x, y)` | Do these two columns relate the way they are supposed to? |
| `boxplot(y ~ g, data = d)` | Does one group differ in a way the pooled column averages away? |
| `barplot(table(g))` | Is every category, month, or site there, as often as it should be? |

:::

::: {.content-hidden when-format="revealjs"}

Suspicions come from three places. `str()` and `summary()` supply some: a
character column that should be numeric, a minimum that grazes an
impossible value. The data's own documentation supplies more, like the
two-instrument structure in the POC column. The rest come from the world:
you knew when the fires started before you opened the file, and that
knowledge made the January 1 spike worth chasing. The instruments do not
replace knowing things. They test the file against what you know.

:::

## Same summaries, different data {#anscombe}

```{r}
#| label: anscombe
#| script: "Same summaries, different data: Anscombe 1973"
#| fig-height: 5
#| fig-alt: "Anscombe's four datasets as four scatterplots: a loose linear cloud, a smooth arch, a tight line with one far outlier, and a vertical stack of identical x values with one point far to the right."
par(mfrow = c(2, 2), mar = c(4, 4, 1, 1))
plot(anscombe$x1, anscombe$y1)
plot(anscombe$x2, anscombe$y2)
plot(anscombe$x3, anscombe$y3)
plot(anscombe$x4, anscombe$y4)
par(mfrow = c(1, 1))
```

::: {.content-hidden when-format="revealjs"}

In 1973 the statistician Francis Anscombe built four small datasets to make
today's point, and they ship inside R as `anscombe`. All four pairs have
the same x mean, the same y mean, the same variances, and the same
correlation of 0.82. Every summary statistic you have met says these four
are the same data.

The plots disagree. The first is a loose linear cloud. The second is a
smooth curve with nothing linear about it. The third is a perfect line plus
one outlier that wrecks the fit. The fourth stacks ten identical x values,
and one far point creates the whole correlation. Fifty years on, this
is still the clearest demonstration that **a summary is a compression, and
compression discards structure you needed to see.**

(`par(mfrow = c(2, 2))` splits the plot window into panels, and the setting
sticks until you set it back. The last line resets it. `cor()` here is one
more summary that cannot tell these apart.)

::: {.callout-tip collapse="true"}
## Check the claim

```{r}
#| label: anscombe-check
#| script: "Anscombe: check the claim"
colMeans(anscombe)
cor(anscombe$x1, anscombe$y1)
cor(anscombe$x2, anscombe$y2)
cor(anscombe$x3, anscombe$y3)
cor(anscombe$x4, anscombe$y4)
```
:::

:::

::: {.content-visible when-format="revealjs"}

Four datasets share the same means, variances, and correlation, to two
decimal places. Every summary you have met calls them identical.

:::

## Exercise 2 {#exercise-2}

Five minutes, on the weather frame `wx`. Stubs are in the live script.

```{r}
#| label: exercise-2
#| exercise: true
#| script: "Exercise 2 (5 minutes)"
#| eval: false
# 1. hist(wx$rh_min) -- each day's LOWEST relative humidity. Read it in one
#    sentence. Twelve days sit in single digits; wx$date[wx$rh_min < 10]
#    says when they were. What do those dates line up with?
#
# 2. plot(wx$date, wx$precip, type = "h") -- when did the drought break?
#    Now run summary(wx$precip). Could you have read the break off that?
#
# 3. wx$month <- format(wx$date, "%Y-%m")
#    boxplot(gust ~ month, data = wx)
#    Which month has the extremes -- and are they a shifted box, or dots
#    beyond the whisker? Those are different claims about January.
```

::: {.content-visible when-format="revealjs"}
::: {.notes}
Debriefs, one sentence each: (1) the single-digit days run Dec 10-Jan 24:
Santa Ana air, the fire weather itself; (2) Jan 26 breaks the drought with
17 mm after seven dry weeks, and the biggest storm is Feb 13 at 38 mm;
summary() shows a median of 0 and a max, never *when*; (3) January's
extremes are dots (Jan 7 at 66, far beyond the whisker) while its median
(22.7) barely beats February's (21.6): extreme days, not a windier month
throughout.
:::
:::

## Verify {#verify}

::: {.content-hidden when-format="revealjs"}

The quick checks so far have all been numeric: the right number of rows,
the right classes, nothing impossible in the summary. This one is not:

:::

> **Plot the column before you trust the column.**

::: {.content-hidden when-format="revealjs"}

Plot every column that feeds a number you will report, once, before you
believe anything computed from it. The habit costs ten seconds per column.
It is the only check in the kit that works when every value is plausible
and only the shape is wrong.

**From HW2 on, homeworks grade this habit.** They ask for diagnostic plots,
and each plot needs one written sentence that reads it, the way this page
read each plot above. An unread plot checks nothing.

The quick checks, now five:

| Check | Question it answers |
|:--|:--|
| `dim()` | Did I get the number of rows I expected? |
| `str()` | Is every column the class it should be? |
| `head()` | Does row one look like a row of data? |
| `summary()` | Is anything impossible, or off by a factor of a thousand? |
| `hist()` / `plot()` | Is anything wrong that no single number can show? |

:::

::: {.content-visible when-format="revealjs"}

The quick checks, now five:<br>
`dim()`: the rows I expected?<br>
`str()`: the classes I expected?<br>
`head()`: does row one look like data?<br>
`summary()`: anything impossible?<br>
`hist()` / `plot()`: anything wrong that **no single number can show?**

From HW2 on, every diagnostic plot needs one written sentence that reads
it. HW2 grades the habit.

:::

## Practice {#practice}

::: {.content-visible when-format="revealjs"}

Everything below is in the live script, with solutions.

:::

::: {.content-hidden when-format="revealjs"}

::: {.panel-tabset}

### Practice

```{r}
#| label: practice
#| exercise: true
#| script: "Practice"
#| eval: false
# Everything below uses pm, frm, tb, one and wx, built earlier in this
# script. Every plot needs one written sentence that reads it.

# One column
#  1. hist(wx$tmax). Describe the shape in one sentence.
#  2. Compton's whole year through time, not just the winter. Add the
#     35 ug/m3 line. How many days cleared it?
#  3. Which month is worst covered across ALL monitors? One line.

# Two columns
#  4. plot(wx$tmax, wx$rh_min). What relationship, and does it make
#     physical sense?
#  5. Pick two monitors. Draw one as a line, add the other with
#     points(), and label them with legend().

# Groups
#  6. boxplot(pm25 ~ Local.Site.Name, data = frm). Which monitor has the
#     widest spread, and is that the same as the dirtiest?
#  7. barplot(apply(tb, 2, mean)) -- the county month by month. Which
#     month is worst, and does it match what you saw at Compton?
#  8. table() then barplot() the AQS.Parameter.Code column. What are the
#     two numbers, and why does the file carry both?

# Panels
#  9. par(mfrow = c(2, 2)): four monitors, one histogram each. Do they
#     have the same shape? Reset mfrow when you are done.
# 10. All four Anscombe correlations are 0.82. Which dataset would that
#     number mislead you about most, and which instrument shows it?
```

### Solutions

```{r}
#| label: practice-solutions
#| purl: false
#| eval: false
# 1  a rough bell, mildly right-skewed: LA winter highs cluster near 20 C
hist(wx$tmax)

# 2
cmp <- frm[frm$Local.Site.Name == "Compton", ]
plot(cmp$date, cmp$pm25, type = "h", xlab = "", ylab = "PM2.5 (ug/m3)")
abline(h = 35, col = "#b31b1b", lty = 2)
sum(cmp$pm25 > 35)            # 10 days

# 3
table(frm$month)              # 04 and 05, at 124 each

# 4  strong negative: warm air holds more moisture, so minimum RH falls
plot(wx$tmax, wx$rh_min)

# 5
a <- frm[frm$Local.Site.Name == "Compton", ]
b <- frm[frm$Local.Site.Name == "Reseda", ]
plot(a$date, a$pm25, type = "l", col = "grey60", xlab = "", ylab = "PM2.5")
points(b$date, b$pm25, pch = 16, cex = 0.5, col = "#b31b1b")
legend("topright", c("Compton", "Reseda"), col = c("grey60", "#b31b1b"),
       lty = c(1, NA), pch = c(NA, 16), bty = "n")

# 6  spread and level are different questions: Lancaster is lowest and
#    also tightest, but the widest box is not the highest one
par(mar = c(4, 14, 1, 1))
boxplot(pm25 ~ Local.Site.Name, data = frm, horizontal = TRUE,
        las = 1, cex.axis = 0.7, xlab = "Daily PM2.5 (ug/m3)", ylab = "")
par(mar = c(5, 4, 4, 2) + 0.1)

# 7
barplot(apply(tb, 2, mean), xlab = "Month of 2025", ylab = "County mean")

# 8  88101 is the federal reference method, 88502 the continuous
#    instrument: two ways of measuring the same pollutant
barplot(table(pm$AQS.Parameter.Code))

# 9
par(mfrow = c(2, 2))
for (s in rownames(tb)[1:4]) hist(frm$pm25[frm$Local.Site.Name == s], main = s)
par(mfrow = c(1, 1))

# 10  the fourth: ten identical x values and one far point invent the
#     whole correlation. A scatterplot shows it instantly.
plot(anscombe$x4, anscombe$y4)
```

:::

:::

## Quick reference {#quick-reference}

::: {.content-visible when-format="revealjs"}

On the session page: the four instruments with their options, a
what-shapes-mean table, and every command from today.

:::

::: {.content-hidden when-format="revealjs"}

**The four instruments**

| Code | What it draws |
|:--|:--|
| `hist(x)` | The shape of one numeric column |
| `hist(x, breaks = 20)` | Same, with more resolution |
| `plot(d$date, y, type = "h")` | A column through time, one spike per day |
| `plot(x, y)` | Two columns against each other |
| `boxplot(y ~ g, data = d)` | `y` split by the groups in `g`; read `~` as "by" |
| `barplot(table(g))` | How often each category shows up |

**Dressing an instrument (rarely needed)**

| Code | What it does |
|:--|:--|
| `main = "", xlab = "", ylab = ""` | Title and axis labels |
| `abline(v = 35)`, `abline(h = 10)` | A vertical or horizontal reference line |
| `col =`, `lty =`, `lwd =` | Color, line type, line width |
| `points(x, y, pch = 16, cex = 0.5)` | Add markers to an open plot; `pch` symbol, `cex` size |
| `lines(x, y)` | Add a connecting line to an open plot |
| `legend("topright", labels, col =, pch =, lty =)` | Say which series is which; `bty = "n"` drops the box |
| `xaxt = "n"` then `axis(1, at =, labels =)` | Replace an axis R got wrong; sides 1–4 from the bottom, clockwise |
| `par(mfrow = c(2, 2))` | Panels in one window; reset with `c(1, 1)` |
| `par(mar = c(4, 14, 1, 1))` | Widen a margin for long labels; too narrow clips silently |

**What a shape is telling you**

::: {.tight-cols}

| You see | It usually means |
|:--|:--|
| A bar far from the mass | A sentinel or impossible value; go look at those rows |
| Everything crushed into one bin | The axis stretched to fit a bad value; go find it |
| Two humps | Two populations in one column: units, instruments, or groups |
| A long right tail | Skew; mean and median will disagree, and the mean is the fragile one |
| A ray that splits in two | Two regimes; a clean constant factor means units |
| Dots beyond the whiskers | Outliers to inspect: candidates, not confirmed errors |
| A short bar in a count plot | A coverage gap: days or categories missing |

:::

**Links**

- Session materials: [slides](04-basic-plots-slides.html) ·
  [live script](04-basic-plots.R) ·
  [EPA data](data/epa_pm25_la_county_2025.csv) ·
  [weather data](data/la-weather-dec2024-feb2025.csv) ·
  [where they came from](data/SOURCE.md)
- [Syllabus](../syllabus.qmd) · [full schedule](../index.qmd#schedule)

:::

## After class {#after-class}

Point today's instruments at your own work before you hand any of it in:
`hist()` the concentration column, and `plot()` it through time. If your
peak day is not visibly the peak of the plot, one of the two is wrong. Find
that out before the grader does.

::: {.content-hidden when-format="revealjs"}

**HW2 posts after class today**: *What did the fire
destroy?*, due September 10. The data
comes from CAL FIRE, California's state fire agency: structure-by-structure
damage inspections from the Eaton and Palisades fires. Each row is an
inspected building, with its damage category and its year built. The
homework asks for three diagnostic plots, each read in one written
sentence. HW2 grades the habit from today. One of its columns has a
surprise in it that `summary()` will not show you. Plot first.

:::

::: {.content-visible when-format="revealjs"}

- HW2 posts after class today: *What did the fire destroy?* It asks for three
  diagnostic plots, each read in one sentence. It grades today's habit.

:::

```{r}
#| purl: false
#| include: false
# Single-source build: the .R live script is generated from this file every
# time the page renders, so it cannot fall out of step with the prose.
source("_purl.R")
purl_session("04-basic-plots.qmd", "04-basic-plots.R")
```
