---
title: "Basic plots II"
subtitle: "Thursday, September 3, 2026 · not taught in class"
pagetitle: "Basic plots II"
title-prefix: ""
execute:
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
  revealjs:
    output-file: 04b-basic-plots-ii-slides.html
    theme: [default, slides.scss]
    template-partials:
      - title-slide.html
    footer: "AEM 6850 · Basic plots II"
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    code-overflow: scroll
    history: false
    link-external-newwindow: true
---

```{r}
#| purl: false
#| include: false
knitr::opts_chunk$set(
  comment = "#>",
  collapse = FALSE,
  fig.width = 8,
  fig.height = 4,
  fig.align = "center",
  dpi = 150
)
```

::: {.content-hidden when-format="revealjs"}

[Slides](04b-basic-plots-ii-slides.html) · [Live script](04b-basic-plots-ii.R) ·
[Source](https://github.com/ArielOrtizBobea/aem6850/blob/main/fall-2026/sessions/04b-basic-plots-ii.qmd)

The half of base graphics that turns an instrument into a figure: shading,
colour scales, hand-built axes, unequal panels, heat maps, maps, animation.

:::

## What this page is {#what-this-is}

> **This page is not on the calendar.** It is written, runnable, and about
> seventy minutes of material — roughly one more class meeting, if one were
> ever free.

::: {.content-hidden when-format="revealjs"}

Session 4 stops at the point where a plot is good enough to read. This is
the rest of the toolkit. It is here in full because the material does not
get easier by being withheld, and because if a meeting ever frees up it
should be teachable off this page without anyone rebuilding it first.

Nothing here is assumed by any later session, and nothing here is assessed.
Read it when you need one of the nine things below.

:::

## What is here {#contents}

::: {.tight-cols}

| Section | The commands it turns on | Min |
|:--|:--|--:|
| Unequal panels | `layout()`, `layout.show()` | 9 |
| Shaded areas | `rect()`, `adjustcolor()` | 6 |
| Confidence bands | `polygon()` | 10 |
| Colour scales | `brewer.pal`, `colorRampPalette`, `findInterval` | 10 |
| Custom axes | `axes = FALSE`, `axis()`, `box()`, `mtext()` | 9 |
| Fills, shapes, options | `density=`, `beside=`, `notch=` | 6 |
| Heat maps | `image()` | 7 |
| Maps | `map()`, projections | 8 |
| Animation | the `png()`-in-a-loop recipe | 5 |

:::

::: {.content-hidden when-format="revealjs"}

The colour and map sections want `RColorBrewer` and `maps`; install them
once with `install.packages()`.

:::

## Building the table {#build}

Everything below runs off one data frame, built here so this page and its
script stand on their own.

```{r}
#| label: build
#| script: "Build the table"
pm <- read.csv("data/epa_pm25_la_county_2025.csv",
               colClasses = c("Site.ID" = "character"))
names(pm)[names(pm) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"
names(pm)[names(pm) == "Local.Site.Name"]                <- "site"
pm$date <- as.Date(pm$Date, format = "%m/%d/%Y")

transect <- c("Pasadena", "Compton",
              "Long Beach-Route 710 Near Road", "Lancaster - Fairgrounds")
d      <- pm[pm$site %in% transect &
             pm$POC == 1 & pm$AQS.Parameter.Code == 88101, ]
comp_y <- d[d$site == "Compton", ]
jan    <- d[d$date < as.Date("2025-02-01"), ]
comp   <- jan[jan$site == "Compton", ]
xr     <- as.Date(c("2025-01-01", "2025-01-31"))

dim(d)
```

## `layout()`: panels that are not the same size {#layout}

`mfrow` gives you equal boxes. When you want one wide panel over two small
ones, hand `layout()` a **matrix of panel numbers**:

```{r}
#| label: layout-matrix
#| script: "A layout matrix"
matrix(c(1, 1, 2, 3), nrow = 2, byrow = TRUE)
```

The number is the **order the panel is filled**. Repeat a number and that
panel spans those cells; use `0` to leave a cell empty.

## `layout()`: the five things to know {#layout-rules}

- `matrix()` fills **down the first column** unless you say otherwise.
  `byrow = TRUE` says otherwise.
- `layout()` does what `mfrow` does, except the panels need not be the same
  size.
- `heights =` and `widths =` are ratios, not units: `heights = c(2, 1)` makes
  the top row twice as tall.
- `layout.show(n)` draws the empty numbered regions so you can check a layout
  before spending any plots on it.
- `layout(1)` is the reset. **`par(op)` does not undo a `layout()`.**

## `layout.show()`: checking a layout {#layout-show}

```{r}
#| label: layout-show
#| script: "What that matrix makes"
#| fig-height: 4
#| fig-alt: "Three empty numbered regions: one wide panel across the top labelled 1, and two equal panels beneath labelled 2 and 3, the top row twice as tall as the bottom."
layout(matrix(c(1, 1, 2, 3), nrow = 2, byrow = TRUE), heights = c(2, 1))
layout.show(3)
layout(1)
```

## A layout in use {#layout-real}

```{r}
#| label: layout-real
#| script: "A layout in use"
#| fig-height: 5.5
#| fig-alt: "One wide panel across the top showing Compton's January as a line with circles at each day, over two smaller panels holding histograms of Compton's and Lancaster's full-year readings on identical bins. Compton's tail runs past 55; Lancaster's mass sits three bars to the left."
op <- par(mar = c(3, 4, 2, 1))
layout(matrix(c(1, 1, 2, 3), nrow = 2, byrow = TRUE), heights = c(2, 1))

plot(comp$date, comp$pm25, type = "o", xlim = xr,
     xlab = "", ylab = "PM2.5 (ug/m3)", main = "Compton, January 2025")
hist(d$pm25[d$site == "Compton"], breaks = seq(-5, 60, by = 5),
     main = "Compton, all year", xlab = "")
hist(d$pm25[d$site == "Lancaster - Fairgrounds"], breaks = seq(-5, 60, by = 5),
     main = "Lancaster, all year", xlab = "")

layout(1)
par(op)
```

Two settings were changed, so two get put back — `layout(1)` first, then
`par(op)`. The two histograms share `breaks = seq(-5, 60, by = 5)`, which
spans both stations and keeps them comparable.

## Shaded areas: `rect()` {#rect}

```{r}
#| label: gf-rect
#| script: "Shaded areas: rect()"
#| fig-alt: "Compton's 2025 series as a line with a translucent grey band covering the first three weeks of January."
plot(comp_y$date, comp_y$pm25, type = "l", col = "grey30",
     xlab = "", ylab = "PM2.5 (ug/m3)")
rect(xleft = as.Date("2025-01-01"), xright = as.Date("2025-01-21"),
     ybottom = -5, ytop = 80,
     col = adjustcolor("#b31b1b", alpha.f = 0.15), border = NA)
```

`rect()` takes the four edges, and `adjustcolor(col, alpha.f = )` makes any
colour translucent so the data still shows through. All four edge arguments
are vectorised, so one call can shade many windows.

## Confidence bands: `polygon()` {#polygon}

```{r}
#| label: gf-polygon
#| script: "Confidence bands: polygon()"
#| fig-alt: "A scatter of simulated points with a red fitted cubic curve through them and a translucent red band around the curve widening at both ends."
set.seed(2)
x <- runif(200, 0, 10)
y <- 20 + 3*x - 1.5*x^2 + 0.11*x^3 + 5*rnorm(200)
fit  <- lm(y ~ x + I(x^2) + I(x^3))
X    <- cbind(1, x, x^2, x^3)
se   <- sqrt(diag(X %*% vcov(fit) %*% t(X)))
i    <- order(x)

plot(x, y, pch = 21, col = "grey40", bg = "grey85", xlab = "x", ylab = "y")
polygon(c(x[i], rev(x[i])),
        c(fitted(fit)[i] + 1.96*se[i], rev(fitted(fit)[i] - 1.96*se[i])),
        col = adjustcolor("#b31b1b", alpha.f = 0.25), border = NA)
lines(x[i], fitted(fit)[i], col = "#b31b1b", lwd = 2)
```

`polygon()` takes one closed ring of coordinates. The idiom is *x forwards
then x backwards*, *upper forwards then lower backwards* — that traces the
band's outline in one loop. Every confidence band, event-study plot and
dose-response figure you will ever draw is this shape.

## Colour scales {#colours}

```{r}
#| label: gf-colors
#| script: "Colour scales"
#| fig-height: 3.6
#| fig-alt: "A strip of 200 points coloured along a red-to-blue spectral ramp from left to right."
library(RColorBrewer)
cols <- colorRampPalette(brewer.pal(11, "Spectral"))(100)
v    <- runif(200, 0, 100)
plot(v, seq_along(v), pch = 16, col = cols[findInterval(v, seq(0, 100, length.out = 100))],
     xlab = "value", ylab = "")
```

`brewer.pal(n, name)` gives a designed palette (`display.brewer.all()` shows
them all), `colorRampPalette()` interpolates it to any length, and
`findInterval()` turns a numeric column into positions in that palette. Those
three lines are how every choropleth and heat map gets its colours.

## Custom axes {#axes}

```{r}
#| label: gf-axes
#| script: "Custom axes"
#| fig-alt: "Compton's 2025 series with no default axes: month abbreviations along the bottom, minor ticks between them, horizontal y numbers, a box around the plot, and a margin label reading PM2.5."
op <- par(mar = c(3, 4, 1, 1))
plot(comp_y$date, comp_y$pm25, type = "l", col = "grey30",
     axes = FALSE, xlab = "", ylab = "")
firsts <- as.Date(paste0("2025-", sprintf("%02d", 1:12), "-01"))
axis(1, at = firsts, labels = month.abb)
axis(1, at = firsts + 15, tck = -0.01, lwd = 0, lwd.tick = 1, labels = FALSE)
axis(2, las = 2)
box()
mtext("PM2.5 (ug/m3)", side = 2, line = 2.5)
par(op)
```

`axes = FALSE` suppresses both axes so you can draw your own. A second
`axis()` call on the same side with `labels = FALSE` and a small `tck` gives
minor ticks. `box()` puts the frame back.

## Fills, shapes, and options {#variants}

```{r}
#| label: gf-variants
#| script: "Fills, shapes and options"
#| fig-height: 5
#| fig-alt: "Four panels: a hatched histogram, a stacked bar chart, a side-by-side bar chart, and notched box plots."
op <- par(mfrow = c(2, 2), mar = c(3, 3, 2, 1))
hist(d$pm25, breaks = 30, col = "#b31b1b", density = 25, angle = 45,
     border = "#b31b1b", main = "density = / angle =", xlab = "")

m <- table(format(d$date, "%m"), d$site)[1:4, ]
barplot(m, col = grey.colors(4), main = "stacked", las = 2, cex.names = 0.5)
barplot(m, col = grey.colors(4), beside = TRUE, space = c(0, 2),
        main = "beside = TRUE", las = 2, cex.names = 0.5)

boxplot(pm25 ~ site, data = d, notch = TRUE, boxwex = 0.5,
        main = "notch = / boxwex =", xlab = "", ylab = "", names = rep("", 4))
par(op)
```

A matrix handed to `barplot()` stacks by default and sits side by side with
`beside = TRUE`. `notch = TRUE` cuts a wedge around the median whose width is
a rough confidence interval: non-overlapping notches are a hint the medians
differ. `hist()` also **returns** what it drew — `h <- hist(x, plot = FALSE)`
gives you `h$breaks` and `h$counts` as ordinary vectors.

## Heat maps: `image()` {#image}

```{r}
#| label: gf-image
#| script: "Heat maps with image()"
#| fig-height: 3.6
#| fig-alt: "A four-row heat map, one row per station and one column per month, with Lancaster's row consistently pale and the basin rows darker in January and December."
mm <- tapply(d$pm25, list(d$site, format(d$date, "%m")), mean)
op <- par(mar = c(3, 12, 2, 1))
image(t(mm), col = colorRampPalette(c("white", "#b31b1b"))(20), axes = FALSE)
axis(1, at = seq(0, 1, length.out = 12), labels = month.abb, tick = FALSE)
mtext(rownames(mm), side = 2, at = seq(0, 1, length.out = 4), las = 2,
      cex = 0.7, line = 0.5)
box()
par(op)
```

`image()` draws a matrix as a grid of coloured cells. Note the `t()`: `image()`
puts the **first** matrix dimension on the x axis, so a rows-are-sites matrix
has to be transposed to get sites down the side.

## Maps {#maps}

```{r}
#| label: gf-maps
#| script: "Maps"
#| fig-height: 4
#| fig-alt: "Two panels: an outline map of the lower 48 US states, and an orthographic globe centred on North America with state outlines drawn on it."
library(maps)
library(mapproj)
op <- par(mfrow = c(1, 2), mar = c(0, 0, 2, 0))
map("state"); title("map(\"state\")")
map("world", proj = "orthographic", orientation = c(15, 260, 0))
map("state", proj = "orthographic", orientation = c(15, 260, 0), add = TRUE)
title("orthographic")
par(op)
```

`maps` ships outlines for `"world"`, `"usa"`, `"state"` and `"county"`.
`fill = TRUE` with a vector of colours in the same order as the map's own
region names gives you a choropleth; `add = TRUE` layers one map onto
another, exactly like `lines()` onto a plot.

The orthographic call warns `projection failed for some data`. That is the
projection saying it cannot draw the half of the globe facing away from you,
which is the correct answer. Nothing is broken.

## Animation {#animation}

```{r}
#| label: gf-animation
#| script: "Animation"
#| eval: false
# A GIF is just a folder of PNGs shown in order. Write the frames, then
# stitch them. Needs the magick package, or ImageMagick on the command line.

# invisible(lapply(200:300, function(angle) {
#   png(sprintf("frame_%04d.png", angle), width = 600, height = 600)
#   par(mar = c(0, 0, 0, 0))
#   map("world", proj = "orthographic", orientation = c(15, angle, 0))
#   dev.off()
# }))

# magick::image_write(magick::image_animate(
#   magick::image_read(list.files(pattern = "^frame_.*png$")), fps = 10),
#   "globe.gif")
```

Everything in that loop is `png()` and `dev.off()` from the session, plus a
counter. Any figure you can draw once you can animate.


## Practice {#practice}

::: {.content-visible when-format="revealjs"}

Fourteen exercises with solutions, on this page and in the live script.

:::

::: {.content-hidden when-format="revealjs"}

::: {.panel-tabset}

### Practice

```{r}
#| label: practice
#| exercise: true
#| script: "Practice"
#| eval: false
# Everything below uses d, comp_y, jan, comp and xr, built at the top of
# this script.

# Unequal panels
#  1. layout(matrix(c(1, 2, 1, 3), nrow = 2)) -- before you run anything,
#     say which cells panel 1 covers. Then check with layout.show(3).
#  2. Fill that layout: Compton's whole year in the tall panel, and
#     histograms of Compton and Lancaster in the two small ones. Put the
#     device back afterwards -- both settings.
#  3. Make the top row of a c(1,1,2,3) layout three times as tall as the
#     bottom. Which argument, and is it in inches?

# Shaded areas
#  4. Redraw Compton's January and shade Jan 1 to Jan 21 in translucent
#     red behind the line. The shading must not hide the data.
#  5. Compton's monthly means: shade every month whose mean is above
#     12 ug/m3. How many bands do you get? Do it with ONE rect() call.
#  6. Overlay Compton's and Lancaster's full-year histograms on one plot
#     with translucent fills. (Hint: hist(..., add = TRUE), and both need
#     the same breaks.)

# Confidence bands
#  7. Fit Compton's pm25 on day-of-year and its square. Draw the fitted
#     curve over a scatter of the raw days, with a 95% band. What does the
#     curve say about the shape of the year?

# Colour scales
#  8. Colour the AQI-versus-PM2.5 scatter by month, using a 12-colour ramp
#     built from a Brewer palette. Add a legend.
#  9. Bucket the four annual means into 5 classes with findInterval() and
#     draw a bar chart coloured by bucket.

# Custom axes
# 10. Compton's year with NO default axes: month names along the bottom,
#     small unlabelled ticks at mid-month, horizontal y numbers, a box,
#     and the y label written with mtext().

# Fills, shapes, options
# 11. Days reported per month per station, once stacked and once beside.
#     Which of the two answers "did every station report every month?"
# 12. Box plots by station with notch = TRUE. Which stations' medians are
#     clearly different, and which are not?

# Heat maps and maps
# 13. Site-by-month heat map of mean PM2.5, with the sites sorted by their
#     annual mean. Which site is the outlier, and which month is worst?
# 14. Draw California and put the four monitors on it, sized by annual
#     mean. The coordinates are already in the file.
```

### Solutions

```{r}
#| label: practice-solutions
#| purl: false
#| eval: false
# 1  panel 1 covers BOTH cells of column 1 -- matrix() fills down the
#    first column, so c(1, 2, 1, 3) with nrow = 2 is column 1 = (1, 2)
#    and column 2 = (1, 3). Panel 1 is repeated, so it spans the rows.
layout(matrix(c(1, 2, 1, 3), nrow = 2)); layout.show(3); layout(1)

# 2
op <- par(mar = c(3, 4, 2, 1))
layout(matrix(c(1, 2, 1, 3), nrow = 2))
plot(comp_y$date, comp_y$pm25, type = "l", xlab = "", ylab = "PM2.5",
     main = "Compton 2025")
hist(comp_y$pm25, breaks = seq(-5, 60, 5), main = "Compton", xlab = "")
hist(d$pm25[d$site == "Lancaster - Fairgrounds"], breaks = seq(-5, 60, 5),
     main = "Lancaster", xlab = "")
layout(1); par(op)

# 3  heights =, and they are RATIOS, not inches
layout(matrix(c(1, 1, 2, 3), nrow = 2, byrow = TRUE), heights = c(3, 1))
layout.show(3); layout(1)

# 4  draw the rectangle FIRST, or use a translucent fill, or the shading
#    covers the line. Here: line, then translucent rect, then line again.
plot(comp$date, comp$pm25, type = "l", lwd = 2, xlab = "", ylab = "PM2.5")
rect(as.Date("2025-01-01"), -5, as.Date("2025-01-21"), 80,
     col = adjustcolor("#b31b1b", alpha.f = 0.15), border = NA)
lines(comp$date, comp$pm25, lwd = 2)

# 5  five months: 01, 02, 05, 11, 12. rect() is vectorised, so one call
#    with vectors of edges draws all five.
cm    <- tapply(comp_y$pm25, format(comp_y$date, "%m"), mean)
hot   <- names(cm)[cm > 12]
left  <- as.Date(paste0("2025-", hot, "-01"))
right <- left + 31
plot(comp_y$date, comp_y$pm25, type = "l", xlab = "", ylab = "PM2.5")
rect(left, -5, right, 80, col = adjustcolor("#b31b1b", 0.12), border = NA)
lines(comp_y$date, comp_y$pm25)
length(hot)

# 6
br <- seq(-5, 60, 2.5)
hist(comp_y$pm25, breaks = br, col = adjustcolor("#b31b1b", 0.5),
     border = NA, main = "", xlab = "PM2.5 (ug/m3)")
hist(d$pm25[d$site == "Lancaster - Fairgrounds"], breaks = br,
     col = adjustcolor("#1f6fb4", 0.5), border = NA, add = TRUE)
legend("topright", c("Compton", "Lancaster"), bty = "n",
       fill = adjustcolor(c("#b31b1b", "#1f6fb4"), 0.5))

# 7  a U: high in January, a trough through spring and summer, rising
#    again into December. The band is narrow in the middle where the data
#    is dense and flares at both ends. R-squared is only 0.21 -- the
#    seasonal shape is real but it explains a fifth of the variation.
day <- as.numeric(comp_y$date - as.Date("2025-01-01"))
fit <- lm(comp_y$pm25 ~ day + I(day^2))
X   <- cbind(1, day, day^2)
se  <- sqrt(diag(X %*% vcov(fit) %*% t(X)))
i   <- order(day)
plot(comp_y$date, comp_y$pm25, pch = 16, cex = 0.5, col = "grey60",
     xlab = "", ylab = "PM2.5 (ug/m3)")
polygon(c(comp_y$date[i], rev(comp_y$date[i])),
        c((fitted(fit) + 1.96 * se)[i], rev((fitted(fit) - 1.96 * se)[i])),
        col = adjustcolor("#b31b1b", 0.25), border = NA)
lines(comp_y$date[i], fitted(fit)[i], col = "#b31b1b", lwd = 2)
summary(fit)$r.squared

# 8
library(RColorBrewer)
mon  <- as.numeric(format(d$date, "%m"))
cols <- colorRampPalette(brewer.pal(11, "Spectral"))(12)
plot(d$pm25, d$Daily.AQI.Value, pch = 16, cex = 0.6, col = cols[mon],
     xlab = "Daily mean PM2.5 (ug/m3)", ylab = "Daily AQI")
legend("bottomright", month.abb, col = cols, pch = 16, ncol = 2,
       bty = "n", cex = 0.7)

# 9  Lancaster 4.48, Pasadena 11.67, Long Beach 12.21, Compton 13.38 --
#    Lancaster lands in its own bucket and the other three share one
am  <- sort(tapply(d$pm25, d$site, mean))
brk <- seq(0, 15, length.out = 6)
pal <- colorRampPalette(c("white", "#b31b1b"))(5)
op  <- par(mar = c(4, 15, 1, 1))
barplot(am, horiz = TRUE, las = 1, col = pal[findInterval(am, brk)],
        xlab = "Mean PM2.5, 2025 (ug/m3)")
par(op)

# 10
op <- par(mar = c(3, 5, 1, 1))
plot(comp_y$date, comp_y$pm25, type = "l", col = "grey30",
     axes = FALSE, xlab = "", ylab = "")
firsts <- as.Date(paste0("2025-", sprintf("%02d", 1:12), "-01"))
axis(1, at = firsts, labels = month.abb)
axis(1, at = firsts + 15, tck = -0.01, lwd = 0, lwd.tick = 1, labels = FALSE)
axis(2, las = 2)
box()
mtext("Daily PM2.5 (ug/m3)", side = 2, line = 3)
par(op)

# 11  BESIDE answers it. A stacked bar shows the month's total across all
#     four stations, so a station reporting nothing is invisible inside
#     someone else's bar; side by side, a missing station is a missing bar.
m  <- table(format(d$date, "%m"), d$site)
op <- par(mfrow = c(2, 1), mar = c(3, 4, 2, 1))
barplot(t(m), col = grey.colors(4), main = "stacked")
barplot(t(m), col = grey.colors(4), beside = TRUE, main = "beside = TRUE")
par(op)

# 12  Lancaster (4.20) is clearly different from all three others. Compton
#     (10.70) and Long Beach (10.65) are not different at all -- their
#     notches overlap almost completely. Pasadena (9.30) is borderline
#     against both, and it has the fewest days, which is why its notch is
#     the widest.
boxplot(pm25 ~ site, data = d, notch = TRUE, names = c("Com", "Lan", "LB", "Pas"),
        xlab = "", ylab = "PM2.5 (ug/m3)")

# 13  Lancaster is the outlier row, pale all year. December and January
#     are the worst columns -- and December is worse than January at
#     three of the four sites.
mm <- tapply(d$pm25, list(d$site, format(d$date, "%m")), mean)
mm <- mm[order(rowMeans(mm)), ]
op <- par(mar = c(3, 12, 2, 1))
image(t(mm), col = colorRampPalette(c("white", "#b31b1b"))(20), axes = FALSE)
axis(1, at = seq(0, 1, length.out = 12), labels = month.abb, tick = FALSE)
mtext(rownames(mm), side = 2, at = seq(0, 1, length.out = 4), las = 2,
      cex = 0.7, line = 0.5)
box()
par(op)

# 14  Lancaster sits north of the mountains; the other three cluster in
#     the basin, which is the whole reason its readings differ
library(maps)
co <- unique(d[, c("site", "Site.Latitude", "Site.Longitude")])
co$mean <- tapply(d$pm25, d$site, mean)[co$site]
map("state", region = "california")
points(co$Site.Longitude, co$Site.Latitude, pch = 21, bg = "#b31b1b",
       cex = co$mean / 4)
text(co$Site.Longitude, co$Site.Latitude, round(co$mean, 1), pos = 4, cex = 0.7)
```

:::

:::

## Quick reference {#quick-reference}

::: {.content-visible when-format="revealjs"}

On the page: every command on this page in one table.

:::

::: {.content-hidden when-format="revealjs"}

| Code | What it does |
|:--|:--|
| `layout(m, heights =, widths =)` | panels of unequal size; `m` is a matrix of fill positions, a repeated number spans cells, `0` leaves a cell empty |
| `layout.show(n)` | draw the empty numbered regions before spending plots on them |
| `layout(1)` | the reset — `par(op)` does **not** undo a layout |
| `rect(xleft, ybottom, xright, ytop)` | a rectangle; all four edges are vectorised |
| `adjustcolor(col, alpha.f = 0.2)` | make any colour translucent |
| `polygon(x, y)` | one closed ring; for a band, x forwards then backwards, upper then lower |
| `brewer.pal(n, "Spectral")` | a designed palette; `display.brewer.all()` shows them all |
| `colorRampPalette(cols)(100)` | interpolate a palette to any length |
| `findInterval(x, breaks)` | turn a numeric column into palette positions |
| `axes = FALSE` | suppress both axes so you can draw your own |
| `axis(1, at =, labels =, tck =, lwd.tick =)` | your own axis; `labels = FALSE` with a small `tck` gives minor ticks |
| `box()` | put the frame back after `axes = FALSE` |
| `hist(..., density =, angle =, border =)` | hatched fills |
| `hist(x, plot = FALSE)$counts` | the bin heights without drawing |
| `barplot(m, beside = TRUE, space =)` | a matrix side by side instead of stacked |
| `boxplot(..., notch = TRUE, boxwex =)` | notches as a rough median CI; box width |
| `image(t(m))` | a matrix as coloured cells — note the transpose |
| `map("state")`, `map(..., fill =, add =, proj =)` | outlines, choropleths, projections |
| `png()` in a loop, then stitch | a GIF is a folder of PNGs in order |

:::

```{r}
#| purl: false
#| include: false
source("_purl.R")
purl_session("04b-basic-plots-ii.qmd", "04b-basic-plots-ii.R")
```
