---
title: "5 · Wrangling with dplyr"
subtitle: "Tuesday, September 8, 2026"
pagetitle: "AEM 6850 · Session 5 · Wrangling with dplyr"
execute:
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
  revealjs:
    output-file: 05-wrangling-slides.html
    theme: [default, slides.scss]
    template-partials:
      - title-slide.html
    footer: "AEM 6850 · Session 5"
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    code-overflow: scroll
    history: false
    link-external-newwindow: true
---

```{r}
#| purl: false
#| include: false
knitr::opts_chunk$set(
  comment = "#>",
  collapse = FALSE,
  fig.width = 8,
  fig.height = 4,
  fig.align = "center",
  dpi = 150
)
```

::: {.content-hidden when-format="revealjs"}

[Slides](05-wrangling-slides.html) · [Live script](05-wrangling.R) ·
[Source](https://github.com/ArielOrtizBobea/aem6850/blob/main/fall-2026/sessions/05-wrangling.qmd)

Today the subsetting work gets better grammar: the dplyr verbs, chained
with a pipe into lines you can read aloud. By the end, the two-month hole
in the regulatory series is measured exactly, in three lines.

:::

## Objectives {#objectives}

By the end of this session you can:

- Translate the base-R subsetting you already know into dplyr verbs.
- Build grouped summaries with `group_by()` and `summarize()`.
- Read a pipe chain aloud as a sentence.
- Predict row and group counts before running.

::: {.content-hidden when-format="revealjs"}

## Before class {#before-class}

**HW1 is due today at 8:40 am**, when class begins. Submit before you walk
in.

Have the `data/` folder from sessions 2 and 3 in place; today reuses the EPA
file and the weather file. Nothing to install ahead of time: we install the
course's first package together, in class.

:::

## The first package {#first-package}

::: {.content-hidden when-format="revealjs"}

Everything so far ran on base R, the toolkit every R installation ships
with. A package is R code somebody else wrote, tested, and shared:
functions, help pages, sometimes data. [dplyr](https://dplyr.tidyverse.org)
is the standard package for wrangling data frames, and it is the first one
this course needs.

::: {.column-margin}
dplyr belongs to a coordinated family of packages called the tidyverse.
The course adds members as it needs them.
:::

One command fetches it from [CRAN](https://cran.r-project.org), the archive
R installs packages from. Type it in the console, not in a script:

:::

::: {.content-visible when-format="revealjs"}

A package is R code somebody else wrote, tested, and shared. dplyr is the
standard one for wrangling data frames. In the console:

:::

```{r}
#| label: install
#| purl: false
#| eval: false
install.packages("dplyr")
```

```{r}
#| label: install-script
#| script: "The first package (install once, in the console)"
#| include: false
# install.packages("dplyr")   # run once per machine, then leave commented
```

::: {.content-hidden when-format="revealjs"}

The console fills with a minute of download messages and ends with a line
saying the package was installed. That is the whole install. You do it once
per machine, the way you installed R itself once. The command does not
belong in a script: a script that carries an install line reinstalls the
package on every run.

:::

::: {.content-visible when-format="revealjs"}

- A minute of console messages is normal. It ends with one line saying the
  package was installed.
- Once per machine, like installing R itself. Never in a script.

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
Give the room two minutes to run it and narrate the console spew while it
scrolls. Anyone whose install fails switches to posit.cloud now and brings
the laptop to office hours. Classroom wifi is the usual suspect.
:::
:::

## Load it, every session {#library}

Installing put dplyr on the machine. `library()` puts it into the current R
session:

```{r}
#| label: library
#| script: "Load it (top of every script, every session)"
#| message: true
library(dplyr)
```

::: {.content-hidden when-format="revealjs"}

The block about masked objects is information, not an error. dplyr provides
functions named `filter()` and `lag()`, and base R already had functions
with those names. Loading dplyr puts the new ones in front, and the message
lists what moved. You will see it at every `library(dplyr)`.

The rule from today on: install once per machine, load in every session,
and every script opens with its `library()` lines.

:::

::: {.content-visible when-format="revealjs"}

- The masked-objects block is information, not an error. dplyr's `filter()`
  and `lag()` now stand in front of the base ones.
- Install once per machine. `library()` at the top of every script.

:::

## Rebuild the frame {#rebuild}

Reading stays base R: the same read, the same repairs, the same checks.

```{r}
#| label: rebuild
#| script: "Rebuild the frame"
pm <- read.csv("data/epa_pm25_compton_2025.csv",
               colClasses = c("Site.ID"          = "character",
                              "State.FIPS.Code"  = "character",
                              "County.FIPS.Code" = "character"))
names(pm)[names(pm) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"
pm$date <- as.Date(pm$Date, format = "%m/%d/%Y")

nrow(pm)   # the whole year, both instruments: 662 rows
```

::: {.content-hidden when-format="revealjs"}

dplyr changes nothing about reading a file. `read.csv()`, the identifier
protection, the rename, the date conversion, and the quick checks are
unchanged. What dplyr replaces is everything that came after this point
with brackets and `$`. If `data/` is missing the file, the
[data folder](data/) has it, alongside the provenance notes.

:::

## `filter()`: keep rows {#filter}

Base R cuts the winter window with brackets. Same subset, first verb:

```{r}
#| label: filter
#| script: "filter(): keep rows"
# base R:  pm[pm$date <= as.Date("2025-02-28") &
#                pm$POC == 1 & pm$AQS.Parameter.Code == 88101, ]

one <- filter(pm,
              date <= as.Date("2025-02-28"),
              POC == 1,
              AQS.Parameter.Code == 88101)

nrow(one)   # must be 59: one row per calendar day
```

::: {.content-hidden when-format="revealjs"}

`filter()` keeps the rows where every condition holds. Three things changed
from the base line. The data frame comes first, then the conditions. Column
names stand bare, because `filter()` looks them up inside `pm`, so no `pm$`
and no brackets. Commas join conditions the way `&` did; `|` still spells
or, inside a single condition. The answer did not change: 59 rows, one per
calendar day of the window.

:::

::: {.content-visible when-format="revealjs"}

- Data frame first, then conditions. Bare column names: no `pm$`.
- Commas mean and. `|` still spells or, inside one condition.
- Same rows out: 59, one per day.

:::

## `select()`: keep columns {#select}

Base needed two coordinates and quoted names. `select()` takes bare names,
in the order you want them:

```{r}
#| label: select
#| script: "select(): keep columns"
# base R:  one[, c("date", "pm25")]

head(select(one, date, pm25), 3)
```

::: {.content-hidden when-format="revealjs"}

A minus sign drops instead of keeps: `select(one, -Date)` returns every
column except the raw text date. The result is still a data frame, which is
why `head()` works on it unchanged.

:::

## `mutate()`: add columns {#mutate}

::: {.content-hidden when-format="revealjs"}

`mutate()` adds columns computed from existing ones. Two at once here, and
both get used today. Before running it, set the expectation:
four days in this window clear 35 µg/m³, the 24-hour standard set by EPA
(the US environmental agency) — January 1, 8, 9, and 10. So a TRUE/FALSE column for `pm25 > 35` must sum
to 4.

:::

```{r}
#| label: mutate
#| script: "mutate(): add columns"
# base R:  one$above35 <- one$pm25 > 35

one <- mutate(one,
              above35 = pm25 > 35,
              month   = format(date, "%b"))

head(select(one, date, pm25, above35, month), 3)
sum(one$above35)   # must be 4
```

::: {.content-hidden when-format="revealjs"}

Each new column is `name = expression`, computed inside the data frame, so
`pm25` and `date` stand bare. `format(date, "%b")` pulls the short month
name, `"Jan"`, out of a date: the same format strings, run in reverse.
The sum confirms the count, with `TRUE` counting as 1.

:::

::: {.content-visible when-format="revealjs"}

- `name = expression`, computed inside the frame: bare column names again.
- `"%b"` is the short month name: format strings, in reverse.

:::

## `arrange()`: sort rows {#arrange}

`arrange()` sorts rows by a column. `desc()` flips it to descending:

```{r}
#| label: arrange
#| script: "arrange(): sort rows"
# base R:  head(one[order(-one$pm25), c("date", "pm25")], 5)

worst <- arrange(one, desc(pm25))
head(select(worst, date, pm25), 3)
```

::: {.content-hidden when-format="revealjs"}

January 1 is back on top at 53.2, exactly where the base sort put it.
Notice the cost of working verb by verb: `worst` is a stored
intermediate that nothing else will ever use. The base line avoided
the intermediate by nesting the subset inside `head()`, and paid in
readability. The pipe removes that trade-off.

:::

::: {.content-visible when-format="revealjs"}

- January 1 on top at 53.2, the same answer base sorting gave.
- The cost so far: throwaway objects like `worst`, or nested calls you
  read inside out. The pipe fixes this.

:::

## The pipe {#pipe}

`|>` passes whatever is on its left to the function on its right, as that
function's first argument:

```{r}
#| label: pipe
#| script: "The pipe"
nrow(filter(pm, POC == 1))         # nested: read inside out

pm |> filter(POC == 1) |> nrow()   # piped: read left to right
```

::: {.content-hidden when-format="revealjs"}

Both lines run the same computation and print the same 301. The piped one
reads in the order things happen: start from `pm`, keep the regulatory
instrument's rows, count them. Read `|>` aloud as "then". Every dplyr verb
takes the data frame as its first argument precisely so chains like this
work.

Two practical notes. RStudio types the pipe for you with Cmd+Shift+M (Mac)
or Ctrl+Shift+M (Windows), after you tick "Use native pipe operator" once
under Tools, Global Options, Code. And you will meet an older pipe, `%>%`,
all over existing code and AI-generated code. It does nearly the same job.
Read it the same way; write `|>`.

:::

::: {.content-visible when-format="revealjs"}

- Same computation, same 301. The piped line reads in the order things
  happen.
- Read `|>` as "then". Verbs take the data frame first so chains work.
- Older code writes `%>%`. Read it the same way; write `|>`.

:::

## The chain, end to end {#chain}

```{r}
#| label: chain
#| script: "The chain, end to end"
# base R:  head(one[order(-one$pm25), c("date", "pm25")], 5)

pm |>
  filter(date <= as.Date("2025-02-28"),
         POC == 1, AQS.Parameter.Code == 88101) |>
  arrange(desc(pm25)) |>
  select(date, pm25) |>
  head(5)
```

::: {.content-hidden when-format="revealjs"}

Read it aloud: take the year, keep the winter window from the regulatory
sampler, sort worst first, keep the date and the reading, show five. That is
one sentence. The five rows match the base sort to the decimal: 53.2,
47.7, 44.6, 42.2, 33.6.

The chain also changes how you edit. Each verb sits on its own line, so
adding a step later means inserting a line, not rebuilding a bracket
expression from the inside out.

:::

::: {.content-visible when-format="revealjs"}

- Aloud: take the year, keep the winter window, sort worst first, keep two
  columns, show five.
- Same five rows as the base sort: 53.2, 47.7, 44.6, 42.2, 33.6.
- The whole thing reads as one sentence, one verb per line.

:::

## `group_by()` and `summarize()` {#group-by}

::: {.content-hidden when-format="revealjs"}

`group_by()` declares groups; every `summarize()` after it computes once
per group. Before running the block below, write down the six numbers it
should print. You already have the evidence. Days: the window has no
gaps, so January contributes 31 rows and February 28. Medians: 16.9 for
January and 12.6 for February. Means: all four days above the standard
belong to January, and in a right-skewed month the mean sits above the
median, so January's mean should clear 16.9 with room. Write your six down, then run.

:::

::: {.content-visible when-format="revealjs"}

Write down the six numbers this should print. Evidence: no gaps in the
window, so 31 and 28 days. Boxplot medians were 16.9 and 12.6. January has
all four event days, so its mean should clear its median with room.

:::

```{r}
#| label: group-by
#| script: "group_by() + summarize()"
one |>
  group_by(month) |>
  summarize(days        = n(),
            mean_pm25   = mean(pm25),
            median_pm25 = median(pm25))
```

## Reading the grouped result {#grouped-result}

::: {.content-hidden when-format="revealjs"}

Reconcile line by line. The days column reads 31 and 28: `n()` takes no
arguments and counts the rows of whatever group it stands in, and it only
works inside `summarize()` and friends. The medians come back at 16.9 and
12.6, as written down. January's mean, 21.3, sits far above its
own median because the four event days pull it up; February's, 13.4, barely
moves off its median. (`summarise()` with an s does the same thing; both
spellings work.)

The header row is new. A tibble is dplyr's flavor of data frame: it behaves
like the data frames you know and prints with its dimensions and column
types attached. Grouped summaries come back as tibbles.

And one thing nobody predicted: February printed before January. That is
not a dplyr bug. `month` is text, and text sorts alphabetically. It is
`"10"` before `"9"`, back again.

:::

::: {.content-visible when-format="revealjs"}

- Days 31 and 28. `n()` counts the rows of its group; it works only inside
  `summarize()`.
- Medians 16.9 and 12.6: the boxplot, as numbers. January's mean sits at
  21.3, pulled up by the four event days.
- "A tibble": dplyr's flavor of data frame. Same behavior, more careful
  printing.
- Nobody predicted this part: February printed first. Text sorts
  alphabetically.

:::

## Factors put months in order {#factor-order}

```{r}
#| label: factor-order
#| script: "Factors put the months in order"
month.abb   # a vector R ships with

one <- mutate(one, month = factor(month, levels = month.abb))

one |>
  group_by(month) |>
  summarize(days = n(), median_pm25 = median(pm25))
```

::: {.content-hidden when-format="revealjs"}

A factor is a categorical column with a level table. Here the level table does real work: `levels = month.abb` stores the
categories in calendar order, and every grouped result now follows that
order. This is what factors are for. Categories usually have an order that
is not alphabetical: months, damage classes, education levels, income
brackets. Set the levels once and every table downstream prints in the
order you meant. (The standing warning: `as.numeric()` on a factor
returns level codes, not the labels.)

:::

::: {.content-visible when-format="revealjs"}

- The level table carries the order you meant. Every grouped result now
  follows it.
- This is the factor's job: categories with a real order, recorded once.

:::

## `count()`: the counting shorthand {#count}

```{r}
#| label: count
#| script: "count(): the counting shorthand"
# base R:  table(pm$POC, pm$AQS.Parameter.Code)

pm |> count(POC, AQS.Parameter.Code)
```

::: {.content-hidden when-format="revealjs"}

`count(x)` is `group_by(x)` plus `summarize(n = n())` in one verb. Give it
two columns and it counts every combination that occurs. Base R answers
this question with `table()`; `count()` answers with a data frame, so the
answer itself can go through more verbs. The numbers match it:
301 regulatory days, 7 stray rows under the wrong code, 354 days from the
continuous instrument. They sum to 662, every row in the file.

:::

::: {.content-visible when-format="revealjs"}

- `count(x)` = `group_by(x)` + `summarize(n = n())`.
- Same three cells as `table()`: 301, 7, 354. Sum: 662, the
  whole file.
- The difference: the answer is a data frame, so it can go through more
  verbs.

:::

## The missing spring, measured {#missing-spring}

A bar chart finds the hole. Measuring it takes a chain you can now read:

::: {.callout-important appearance="simple"}
**Write down the twelve counts this should print.** The evidence: the
`count()` above says the regulatory series has 301 rows. A full month has
28 to 31 days. A count plot of this series shows ten tall bars, April near
two days, May near one. Your twelve numbers must sum to 301.
:::

```{r}
#| label: missing-spring
#| script: "The missing spring, measured"
year <- pm |>
  filter(POC == 1, AQS.Parameter.Code == 88101) |>
  mutate(month = factor(format(date, "%b"), levels = month.abb))
```

## The twelve counts {#coverage-table}

```{r}
#| label: coverage-table
#| script: "The twelve counts"
year |> count(month)
```

::: {.content-visible when-format="revealjs"}

- April 2 and May 1, exactly as the bars showed.
- The counts sum to 301. A complete year has 365 days: 64 are missing.
- The part the bars could not show: March logged 30 of 31 days, June 28 of
  30, September 27 of 30.

:::

## Reading the coverage table {#coverage-read}

::: {.content-hidden when-format="revealjs"}

Reconcile against your prediction. April and May land at 2 days and 1, exactly
where a count plot puts them. The counts sum to 301, and a
complete year has 365 days, so the regulatory series is missing 64. Of
those, 58 are the spring outage: April is short 28 days and May 30.

The other six days are the news. If you predicted a full 31 for March, the
table corrects you by one day: March logged 30. June logged 28 of 30,
September 27 of 30. A bar drawn one day short of thirty looks full; a
count is exact. The plot found the hole. The count measured it, and turned
up six more missing days nobody had noticed.

One consequence, computed. The series' "2025 annual mean" is an average of
the 301 days that exist:

```{r}
#| label: annual-mean
#| script: "What the gap does to an annual mean"
mean(year$pm25)   # "the 2025 annual mean" -- of the days that exist
```

Is 13.4 high, low, or fine? The missing days decide. The site's continuous
instrument ran through the spring, and its numbers say April and May were
the cleanest stretch of its year:

```{r}
#| label: annual-mean-check
#| script: "The other instrument's spring"
other <- pm |> filter(POC == 3, AQS.Parameter.Code == 88502)

mean(other$pm25)   # its full-year mean
other |>
  mutate(month = format(date, "%b")) |>
  filter(month == "Apr" | month == "May") |>
  summarize(days = n(), mean_pm25 = mean(pm25))
```

The continuous instrument averages 12.3 over its year and 10.1 over April
and May. The two instruments measure by different methods, so the levels
are not exactly comparable. The direction is: the regulatory series is
missing precisely the months that would have pulled its average down, so
its 13.4 leans high. Every mean is a mean of the days that are there. Today you can put numbers on what that costs.

:::

::: {.content-visible when-format="revealjs"}

- 58 of the 64 missing days are the spring outage. The other six are
  single days in March, June, and September that the bars could not show.
- The cost, quantified on the page: the missing months were the cleanest
  stretch of the year by the site's other instrument. An annual mean over
  the 301 days leans high.

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
If someone asks why the sampler was down: this file does not say, and AQS
validation can also remove readings after the fact (session 4's page). The
point stands either way: the series has no spring.
:::
:::

## The one-line stopwatch {#timing}

One more habit joins the kit today. Wrap any line in `system.time()` and R
reports how long it took:

```{r}
#| label: timing
#| script: "system.time(): the one-line stopwatch"
system.time(read.csv("data/epa_pm25_compton_2025.csv"))
```

::: {.content-hidden when-format="revealjs"}

Read the `elapsed` column: the seconds you actually waited. For this file
the answer is effectively zero, and it will be for everything in Part I.
Part II changes that. Satellite rasters, statewide parcel tables, and
folders of hourly files make single lines take minutes. The habit costs one
line: when something feels slow, wrap it once, note the number, and compare
after you change something.

:::

::: {.content-visible when-format="revealjs"}

- Read `elapsed`: the seconds you waited. Today, effectively zero.
- The habit: when a line feels slow, wrap it once and note the number.

:::

## You try {#you-try}

Five minutes, on the weather frame. Stubs are in the live script.

```{r}
#| label: weather-load
#| script: "Bring back the weather frame"
la <- read.csv("data/la-weather-dec2024-feb2025.csv", skip = 3)
names(la) <- c("date", "gust", "wind", "tmax", "rh_min", "precip")
la$date <- as.Date(la$date)
```

```{r}
#| label: you-try
#| exercise: true
#| script: "You try (5 minutes)"
#| eval: false
# 1. The windstorm, top three: chain arrange(), select(), and head() to
#    print the three biggest gust days with their dates. Write down which
#    date must come first.
#
# 2. Rain days by month: make month a factor with
#    levels = c("Dec", "Jan", "Feb"), then group_by(month) |>
#    summarize(days = n(), rain_days = sum(precip > 0)).
#    Write down which month must have the most rain before you run it.
#
# 3. One line: la |> count(precip > 0). Write both numbers down first.
#    Your table from task 2 already fixes what they must be.
```

::: {.content-visible when-format="revealjs"}
::: {.notes}
Debriefs, one sentence each: (1) Jan 7 at 65.9, then Jan 8 at 50.8, then
Feb 13 at 49.0, the day of session 4's biggest storm; session 1's "66" was
this 65.9 rounded. (2) Dec 2, Jan 3, Feb 8: February has the most rain,
matching the drought's Jan 26 break; without the factor the months print
Dec, Feb, Jan. (3) 13 wet days and 77 dry, because 2 + 3 + 8 = 13 and the
frame has 90 rows.
:::
:::

## Verify {#verify}

::: {.content-hidden when-format="revealjs"}

Each session so far installed one habit:

| Session | The habit |
|:--|:--|
| 1 | Write down what the data should show, then run, then reconcile. |
| 2 | Run `str()` before you compute. |
| 3 | Plot the column before you trust the column. |
| 4 | Predict every count from evidence before you run it. |

:::

> **Write down the count you expect, and where the expectation comes from.
> Then run, and reconcile.**

::: {.content-hidden when-format="revealjs"}

The expectation always has a source: the calendar, the sampling schedule, a
plot you already read, a total that must be conserved. Today the habit ran
four times: 59 rows from the calendar, 4 days from a histogram, six summary
numbers from a boxplot, and twelve counts that had to sum to 301. When the
number that returns is not the one you wrote down, stop and resolve the
difference before moving on. Today the difference was six missing days
nobody had noticed.

Joins need this habit most: they silently drop rows and silently duplicate
them, and the row count you write down before each join is the check that
catches both.

:::

::: {.content-visible when-format="revealjs"}

The expectation always has a source: the calendar, the sampling schedule, a
plot you already read, a total that must be conserved. When the returned
number is not the written-down number, stop and resolve it.

:::

## Quick reference {#quick-reference}

::: {.content-visible when-format="revealjs"}

On the session page: the verb table with base translations, the pipe and
its shortcut, the factor recipe, and the things that bite.

:::

::: {.content-hidden when-format="revealjs"}

**The verbs, with the base R you already know**

::: {.tight-cols}

| dplyr | It keeps or makes | Base equivalent |
|:--|:--|:--|
| `filter(d, x > 1, g == "a")` | Rows where all conditions hold | `d[d$x > 1 & d$g == "a", ]` |
| `select(d, date, pm25)` | The named columns; `-name` drops | `d[, c("date", "pm25")]` |
| `mutate(d, new = expr)` | Adds computed columns | `d$new <- expr` |
| `arrange(d, desc(x))` | Rows sorted; `desc()` descending | `d[order(-d$x), ]` |
| `group_by(d, g)` | Declares groups for what follows | new today |
| `summarize(d, m = mean(x))` | One row per group | new today |
| `count(d, g)` | Grouped row counts | `table(d$g)` |

:::

**The pipe and the helpers**

The pipe itself: `x |> f() |> g()` passes each left side into the next
call as its first argument. Read it as "then".

| Code | What it does |
|:--|:--|
| `n()` | Rows in the current group; only inside `summarize()` and friends |
| `desc(x)` | Descending order inside `arrange()` |
| `head(d, 5)` | Still base R, still works at the end of a chain |
| `x %in% c("a", "b")` | Is each value one of these; tidier than or-chains of `==` |
| `factor(x, levels = month.abb)` | Categories in the order you mean, not alphabetical |

**Things that bite**

| Symptom | Cause |
|:--|:--|
| "Masked objects" message at `library(dplyr)` | Normal. dplyr's `filter()` and `lag()` stand in front of the base ones |
| `filter()` errors before `library(dplyr)` | The package is not loaded in this session |
| `filter(d, x = 1)` errors | `=` assigns; the test is `==` |
| The chain printed, but nothing changed | You never assigned; put `d <-` in front of the chain |
| Reinstalls on every run | `install.packages()` is in the script; it belongs in the console |
| Months print Apr, Aug, Dec, ... | Text sorts alphabetically; make a factor with `levels = month.abb` |
| `n()` errors on its own | It only counts inside `summarize()` and friends |

**Links**

- Session materials: [slides](05-wrangling-slides.html) ·
  [live script](05-wrangling.R) ·
  [EPA data](data/epa_pm25_compton_2025.csv) ·
  [weather data](data/la-weather-dec2024-feb2025.csv) ·
  [where they came from](data/SOURCE.md)
- [dplyr documentation](https://dplyr.tidyverse.org) and the [data
  transformation
  cheatsheet](https://rstudio.github.io/cheatsheets/html/data-transformation.html)
- [R for Data Science, ch. 3](https://r4ds.hadley.nz/data-transform.html),
  the book-length treatment of today
- [Syllabus](../syllabus.qmd) · [full schedule](../index.qmd#schedule)

:::

## After class {#after-class}

**HW2 is out: What did the fire
destroy?** Due Thursday, September
10, 8:40 am.

::: {.content-hidden when-format="revealjs"}

The data is CAL FIRE's building-by-building damage inspections from the
Eaton and Palisades fires (CAL FIRE is California's fire agency): every
inspected structure, its damage category, its year built. The damage
column is categorical, so `count()` is the tool for it. Plot the columns
before you trust them, and read each plot in one written sentence.

**Re-run today's script before you start.** Run it line by line and change
things to see what breaks.

Re-run today's script at home, top to bottom, in a fresh session.

:::

::: {.content-visible when-format="revealjs"}

- Damage categories from the fires: `count()` is the tool. Plot first.
- Re-run today's script at home, top to bottom, in a fresh session.

:::

```{r}
#| purl: false
#| include: false
# Single-source build: the .R live script is generated from this file every
# time the page renders, so it cannot fall out of step with the prose.
source("_purl.R")
purl_session("05-wrangling.qmd", "05-wrangling.R")
```
