---
title: "6 · How to code"
subtitle: "Thursday, September 10, 2026"
pagetitle: "6 · How to code"
title-prefix: ""
execute:
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
  revealjs:
    output-file: 06-functions-debugging-slides.html
    theme: [default, slides.scss]
    template-partials:
      - title-slide.html
    footer: "AEM 6850 · Session 6"
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    code-overflow: scroll
    history: false
    link-external-newwindow: true
---

```{r}
#| purl: false
#| include: false
knitr::opts_chunk$set(
  comment = "#>",
  collapse = FALSE,
  fig.width = 8,
  fig.height = 4,
  fig.align = "center",
  dpi = 150
)
```

::: {.content-hidden when-format="revealjs"}

[Slides](06-functions-debugging-slides.html) · [Live script](06-functions-debugging.R) ·
[Source](https://github.com/ArielOrtizBobea/aem6850/blob/main/fall-2026/sessions/06-functions-debugging.qmd)

Functions, one function over many files, the habits that make a script
readable and checkable, and what to do when a number is wrong.

:::

## Objectives {#objectives}

By the end of this session you can:

- Write a function and call it with its own arguments.
- Run one function over every file in a folder with `lapply()`.
- Trace a `for` loop by hand.
- Structure a script: sections, names, comments that say why, named
  constants, no duplicated code.
- Stop a script with `stopifnot()` when a number is not what you wrote down.
- Localize a defect by bisection, and write the diagnosis down: line,
  mechanism, fix, proof.

## Building the Compton frame {#build}

Everything below runs off the Compton file, read here so this page and its
script stand on their own.

```{r}
#| label: build
#| script: "Build the Compton frame"
pm <- read.csv("data/epa_pm25_compton_2025.csv",
               colClasses = c("Site.ID" = "character"))
names(pm)[names(pm) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"
names(pm)[names(pm) == "Local.Site.Name"]                <- "site"
pm$date <- as.Date(pm$Date, format = "%m/%d/%Y")
pm <- pm[pm$POC == 1 & pm$AQS.Parameter.Code == 88101, ]

nrow(pm)   # one regulatory instrument, the days it ran
```

::: {.content-hidden when-format="revealjs"}

Five lines: read the file with the identifier protected, rename the
concentration and site columns, convert the date, keep one instrument.
Every EPA (the US Environmental Protection Agency) file needs the same five
lines, and every script that reads one starts by typing them again. A
function lets you type them once.

:::

::: {.content-visible when-format="revealjs"}

- Read with the identifier protected, rename, convert the date, keep one
  instrument. 301 rows.
- Every EPA file needs these five lines. A function lets you type them once.

:::

## `function()`: naming a block of code {#function}

```{r}
#| label: function
#| script: "function(): naming a block of code"
days_above <- function(x, standard = 35) {
  sum(x > standard)
}

dec <- pm[format(pm$date, "%m") == "12", ]
days_above(dec$pm25)                  # December, at the 35 standard
days_above(dec$pm25, standard = 50)   # the same days, a higher bar
days_above(c(1, 50))                  # a case you can check by eye
```

::: {.content-hidden when-format="revealjs"}

`function()` builds a function. The names in its parentheses are the
arguments; `standard = 35` gives one of them a default, so a call can leave
it out. The body runs when you call the function, with `x` standing for
whatever you passed in. The value of a function is its last expression.
`return(x)` says the same thing explicitly, and is only needed to leave a
function early.

December had 5 days above the standard and 3 above 50. The last call is
the one to keep in the habit: two numbers you can check by eye, and the
answer must be 1.

:::

::: {.content-visible when-format="revealjs"}

- `function(x, standard = 35)`: the arguments. `= 35` is a default.
- The value is the last expression. `return()` is for leaving early.
- December: 5 days above 35, 3 above 50. `c(1, 50)`: 1, as it must be.

:::

## Local names and global names {#scope}

```{r}
#| label: scope
#| script: "Local names and global names"
#| error: true
count_pos <- function(x) {
  n <- sum(x > 0)   # n exists inside the call
  n
}
count_pos(c(-1, 2, 3))
n                     # and not outside it
```

::: {.content-hidden when-format="revealjs"}

A name made inside a function dies at the closing brace. Outside the call,
`n` never existed. That is what lets a function use short names like `n`
and `d` without trampling anything in the workspace.

:::

::: {.content-visible when-format="revealjs"}

- Names made inside a function die at the closing brace.
- So a function can use short names without touching the workspace.

:::

## A function can read a global name {#scope-global}

```{r}
#| label: scope-global
#| script: "A function can read a global name"
#| error: true
standard <- 35
share_above <- function(x) sum(x > standard) / length(x)
share_above(dec$pm25)

rm(standard)
share_above(dec$pm25)   # worked until the global was gone
```

::: {.content-hidden when-format="revealjs"}

The other direction is the trap. A function can read a name from outside
itself, and `share_above()` does: it uses `standard` without ever being
given it. It works while `standard` happens to sit in the workspace and
fails in a fresh session with `object 'standard' not found`. Everything a
function needs should come in through its arguments. Then it works
anywhere, and a two-number test tells you it reads its own input.

:::

::: {.content-visible when-format="revealjs"}

- A function can read a global name. It works until the global is gone.
- Everything a function needs comes in through its arguments.

:::

## `read_epa()`: the five lines, once {#read-epa}

```{r}
#| label: read-epa
#| script: "read_epa(): the five lines, once"
read_epa <- function(path) {
  d <- read.csv(path, colClasses = c("Site.ID" = "character"))
  names(d)[names(d) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"
  names(d)[names(d) == "Local.Site.Name"]                <- "site"
  d$date <- as.Date(d$Date, format = "%m/%d/%Y")
  d[d$POC == 1 & d$AQS.Parameter.Code == 88101, ]
}

nrow(read_epa("data/epa_pm25_compton_2025.csv"))
nrow(read_epa("data/epa_pm25_la_county_2025.csv"))
```

::: {.content-hidden when-format="revealjs"}

The five lines from the top of the page, with `path` in place of the file
name and the frame as the value. Copy-pasting code is a red flag, and the
function is the fix: a change to the rename or to the instrument filter now
happens in one place, and every file gets it. The county file holds eleven
sites and 4,757 rows; 1,738 of them are on a regulatory instrument.

:::

::: {.content-visible when-format="revealjs"}

- The five lines, with `path` for the file name and the frame as the value.
- Copy-pasting code is a red flag. A function is the fix.
- 301 rows at Compton; 1,738 across the county's regulatory instruments.

:::

## `lapply()` over files {#lapply-files}

```{r}
#| label: lapply-files
#| script: "lapply() over files: a list in, a list out"
files <- list.files("data", pattern = "^epa_.*\\.csv$", full.names = TRUE)
files

frames <- lapply(files, read_epa)
class(frames); length(frames)
names(frames) <- basename(files)
sapply(frames, nrow)
```

::: {.content-hidden when-format="revealjs"}

In: a character vector of paths. Out: a list, one data frame per file, in
the same order. `list.files()` finds the files; the pattern keeps the EPA
ones and `full.names = TRUE` returns paths you can hand to `read_epa()`.
`sapply()` is `lapply()` with the result simplified when every element has
the same shape, which a row count does.

This is the shape of most real data collection: one file per year, per
state, or per month. Get one file right, then hand the function the list.

:::

::: {.content-visible when-format="revealjs"}

- In: a character vector of paths. Out: a list, one data frame per file.
- `sapply()` simplifies when every element has the same shape.
- Get one file right, then hand the function the list.

:::

## `split()`: one data frame per site {#split}

```{r}
#| label: split
#| script: "split(): one data frame per site"
la     <- frames[["epa_pm25_la_county_2025.csv"]]
la_jan <- la[format(la$date, "%m") == "01", ]

by_site <- split(la_jan, la_jan$site)
sapply(by_site, nrow)
```

::: {.content-hidden when-format="revealjs"}

`split()` cuts a data frame into a list, one piece per value of the
grouping column. `by_site[["Compton"]]` is a plain data frame with
Compton's 31 January rows. Eight sites, not the county's eleven: three
sites run no regulatory instrument, so the filter inside `read_epa()`
removed them.

Count rows before you trust a mean. Four sites report 6 to 11 January days;
whatever they average rests on very little. `do.call(rbind, by_site)` puts
the pieces back together, 157 rows, as before the split.

:::

::: {.content-visible when-format="revealjs"}

- A list, one data frame per site. Eight, not eleven: three run no
  regulatory instrument.
- Rows per site first. Four sites have 6 to 11 January days.

:::

## The same function, every site {#lapply-groups}

```{r}
#| label: lapply-groups
#| script: "The same function, every site"
sapply(by_site, function(s) days_above(s$pm25))
```

::: {.content-hidden when-format="revealjs"}

`function(s) days_above(s$pm25)` is a function with no name, written where
it is used. `s` stands for one piece of the list at a time, and the body
says what to do with it. Compton had 4 days above the standard in January,
Lancaster - Fairgrounds none.

:::

::: {.content-visible when-format="revealjs"}

- `function(s) ...`: a function with no name, written where it is used.
- `s` is one piece of the list at a time.
- Compton 4, Lancaster - Fairgrounds 0.

:::

## `for` loops, traced by hand {#for}

```{r}
#| label: for
#| script: "for loops, traced by hand"
x   <- c(10, 40, 36)
out <- numeric(length(x))        # one slot per answer, made first
for (i in seq_along(x)) {
  out[i] <- x[i] > 35
}
out
```

| pass | `i` | `x[i]` | `x[i] > 35` | `out` after |
|:--|:--|:--|:--|:--|
| 1 | 1 | 10 | FALSE | 0 0 0 |
| 2 | 2 | 40 | TRUE | 0 1 0 |
| 3 | 3 | 36 | TRUE | 0 1 1 |

::: {.content-hidden when-format="revealjs"}

`for` runs the body once per value of `i`; `seq_along(x)` is 1, 2, 3. The
table is the loop traced by hand: fill it in before running, then run and
compare. `out` is made first, at full size, so each pass fills a slot
instead of growing a vector.

`lapply()` does this job in one line. `for` is the tool when order matters
or when each pass does something besides return a value, like writing one
file per site.

:::

::: {.content-visible when-format="revealjs"}

- Trace it by hand before running. `out` is made first, full size.
- `for` when order matters or a pass has a side effect. `lapply()` otherwise.

:::

## Exercise 1 {#exercise-1}

::: {.panel-tabset}

### Exercise

```{r}
#| label: exercise-1
#| exercise: true
#| script: "Exercise 1 (5 minutes)"
#| eval: false
# 1. Write mean_above(x, standard = 35): the mean of the readings in x
#    that are above the standard.
# 2. Run it on every site in by_site with sapply().
# 3. One number to compare with the room: Pasadena.
# 4. One site returns NaN. Say why in one sentence, and whether that is
#    wrong.
```

### Solution

```{r}
#| label: exercise-1-solution
#| purl: false
mean_above <- function(x, standard = 35) mean(x[x > standard])
round(sapply(by_site, function(s) mean_above(s$pm25)), 1)
```

::: {.content-visible when-format="revealjs"}

Pasadena 58.1. Lancaster - Fairgrounds `NaN`: no reading above 35, so the
mean of nothing. Honest.

:::

::: {.content-hidden when-format="revealjs"}

Pasadena 58.1. Lancaster - Fairgrounds returns `NaN`: no January reading
there was above 35, so `x[x > standard]` is empty and the mean of nothing
is not a number.
That is the honest answer. A function that quietly returned 0 would be
lying.

:::

:::

## Which software practices pay for research code {#research-code}

::: {.content-hidden when-format="revealjs"}

Software has users, ships, and is maintained for years. A research script
has one author, runs a handful of times, and has to produce a number
someone can defend. So the practices of software engineering divide in
three at this scale:

:::

| Pays now | Pays sometimes | Overhead at this scale |
|:--|:--|:--|
| Names that mean something | Unit tests | Design patterns |
| Sections | Code review | Interfaces for other programs |
| No duplicated code | Refactoring | Coverage targets |
| Small, tested pieces | | Continuous integration |
| Version control | | |

::: {.content-visible when-format="revealjs"}

- One author, a handful of runs, one number someone has to defend.
- Borrow deliberately, not by reputation.

:::

::: {.content-hidden when-format="revealjs"}

Borrow deliberately, not by reputation. The first column is the rest of
this page.

:::

## Naming things and named constants {#naming}

```{r}
#| label: naming
#| script: "Naming things and named constants"
STANDARD <- 35   # US EPA 24-hour PM2.5 standard, ug/m3

days_above <- function(x, standard = STANDARD) sum(x > standard)
days_above(dec$pm25)
```

::: {.content-hidden when-format="revealjs"}

Names are short and informative, lower case, with underscores: `pm25`,
`by_site`, `read_epa`. Verbs for functions. Not `MyModified.PM.var`, not
`T` for `TRUE`.

A number that means something gets a name at the top of the script, in
capitals, and is used by that name everywhere. `35` typed in six places is
six chances to change five of them.

Comments say why, not what. `# EPA 24-hour standard` tells a reader
something the code cannot; `# set x to 35` repeats the line below it.

:::

::: {.content-visible when-format="revealjs"}

- Short, lower case, underscores. Verbs for functions. Never `T` for `TRUE`.
- A number that means something gets a name at the top, in capitals.
- Comments say why. The code already says what.

:::

## `# section ----` structure {#sections}

```{r}
#| label: sections
#| script: "A script with sections"
#| eval: false
# air.R -- days above the PM2.5 standard at one monitor, 2025
# Your Name, NetID

# 0) Setup ----
STANDARD <- 35                        # US EPA 24-hour standard, ug/m3
pm <- read_epa("data/epa_pm25_compton_2025.csv")

# 1) One instrument, one month ----

# 2) The numbers ----

# 3) Write results.csv ----
```

::: {.content-hidden when-format="revealjs"}

A title block says what the script does and who wrote it. A setup section
holds packages, constants, and the data coming in. Then one section per
task, numbered in the order they run, and nothing from setup reappears
lower down. A comment ending in four dashes is a section: RStudio's outline
(Ctrl+Shift+O, Cmd+Shift+O on a Mac) lists them, and you can jump between
them. A script does one job. Two jobs, two scripts.

:::

::: {.content-visible when-format="revealjs"}

- Title block, setup, then one numbered section per task.
- `# Name ----` is a section. RStudio's outline lists them.
- A script does one job. Two jobs, two scripts.

:::

## Sketch the sections first {#sketch}

::: {.content-hidden when-format="revealjs"}

You are writing code for others, and that includes yourself in t+1. Six
months from now you will open this script with no memory of writing it,
and the section headers are what you will read first.

"Would you tell me, please, which way I ought to go from here?" "That
depends a good deal on where you want to get to," said the Cat (*Alice in
Wonderland*, Lewis Carroll). Before the first line of code, write the
section headers: what comes in, what each step produces, what goes out.
Then fill them in, interactively, one at a time.

:::

::: {.content-visible when-format="revealjs"}

- You are writing code for others. That includes yourself in t+1.
- "That depends a good deal on where you want to get to," said the Cat.
- Write the section headers first. Then fill them in, one at a time.

:::

## Write code that is easy to check {#checkable}

::: {.content-hidden when-format="revealjs"}

Whether code can be checked is a property of how it is built, not of how
hard you look at it afterwards. Small pieces with one job each. Intermediate
values that land in named objects, so `nrow(pm)` after the filter and
`range(pm$date)` after the date conversion are one line away. No step that
changes three things at once. Results that come out where a check can reach
them. Code you cannot check is code you cannot defend, whatever it
computed.

:::

::: {.content-visible when-format="revealjs"}

- Small pieces, one job each.
- Intermediate values in named objects you can print.
- No step that changes three things at once.
- Code you cannot check is code you cannot defend.

:::

## `stopifnot()`: a check that stops the script {#stopifnot}

```{r}
#| label: stopifnot
#| script: "stopifnot(): a check that stops the script"
jan <- pm[format(pm$date, "%m") == "01", ]

stopifnot(nrow(pm) == 301)
stopifnot(nrow(jan) == 31, abs(mean(jan$pm25) - 21.28) < 0.05)
```

::: {.content-hidden when-format="revealjs"}

`stopifnot()` takes conditions. If every one is `TRUE` it does nothing;
if any is `FALSE` it stops the script and names the first that failed.
Several conditions fit in one call, one per line when they are long, and a
condition may be a whole vector, in which case every element must be
`TRUE`. A check that sits commented out is turned on by selecting its lines
and pressing Ctrl+Shift+C (Cmd+Shift+C on a Mac). Put the check right
after the step it guards, with the number you wrote down before running.

A mean is compared with a tolerance, `abs(x - 21.28) < 0.05`, never with
`==`. The computed mean is 21.2806..., and 21.28 is what you wrote down;
`==` would call them different.

:::

::: {.content-visible when-format="revealjs"}

- Silent when every condition holds. Stops the script when one fails.
- Several conditions in one call. Right after the step it guards.
- A mean gets a tolerance, `abs(x - 21.28) < 0.05`, never `==`.

:::

## What a failed check looks like {#stopifnot-fail}

```{r}
#| label: stopifnot-fail
#| script: "What a failed check looks like"
#| error: true
stopifnot(nrow(jan) == 30)
stopifnot("one row per calendar day" = nrow(jan) == length(unique(jan$date)))
```

::: {.content-hidden when-format="revealjs"}

The message is the condition that failed, so a reader knows what was
expected without opening the script. A named condition prints its name
instead; the second call passes and prints nothing.

A check you edited so that it passes is worse than a check that fails.
The failing check still tells the truth; the edited one hides a number you
could not explain.

:::

::: {.content-visible when-format="revealjs"}

- The message is the condition that failed. A named condition prints its name.
- A check edited to pass is worse than a check that fails honestly.

:::

## Test small, then run big {#test-small}

```{r}
#| label: test-small
#| script: "Test small, then run big"
days_above(c(1, 50))                              # a case you can check by eye
nrow(read_epa(files[1]))                          # one file before the folder
system.time(lapply(files, read_epa))              # the whole folder, timed
```

::: {.content-hidden when-format="revealjs"}

Never run the big job untested. A function gets a case you can check by
eye. A loop runs on one file before the folder, on one site before all of
them. `system.time()` around the big call tells you what the next run will
cost: nothing here, and minutes when a folder holds a thousand files.

:::

::: {.content-visible when-format="revealjs"}

- Never run the big job untested. A function gets a case you can check by eye.
- One file before the folder. `system.time()` around the big call.

:::

## Restart R and source the script {#restart}

```{r}
#| label: restart
#| script: "Restart R and source the script"
#| eval: false
#| purl: false
# Session > Restart R   (Ctrl+Shift+F10 on Windows, Cmd+Shift+0 on a Mac)
ls()                    # character(0), or it was not a clean restart
getwd()                 # where "data/..." is looked for
list.files("data")      # what is there
source("air.R")         # your script, top to bottom, in a clean session
```

::: {.content-hidden when-format="revealjs"}

A script that runs only because of what is already in your environment is
not a script yet. Restart R: packages unload, and the workspace empties
once Tools > Global Options > General has "Save workspace to .RData on
exit" set to Never and "Restore .RData into workspace at startup"
unchecked. On the default settings RStudio puts the old objects back after
a restart, so set both once, and expect `ls()` to print `character(0)`.
Then `source()` the file. It runs top to bottom and stops at the first
error. `rm(list = ls())` clears objects but not packages or options; it is
not a restart.

`source()` reads `"data/..."` relative to the working directory. Opening a
project's `.Rproj` file sets the working directory to the project folder,
which is why every path in a script is relative: `"data/file.csv"`, never
`"/Users/you/Desktop/file.csv"`, which runs on one laptop and nowhere else.
When R says a file does not exist, check `getwd()` and `list.files("data")`
before blaming the file.

:::

::: {.content-visible when-format="revealjs"}

- Restart R, check `ls()` is empty, then `source()`. `rm()` is not a restart.
- Paths are relative to `getwd()`. The `.Rproj` file sets it when opened.
- "File does not exist": check `getwd()` and `list.files("data")` first.

:::

## Debugging as a strategy {#debugging}

::: {.content-hidden when-format="revealjs"}

Read the error before changing anything. It names the call that failed and
usually the cause. Reproduce the failure on the smallest input that still
fails: one file, one site, two numbers. Form one hypothesis at a time and
test it. Change one thing per run, so that when the output changes you know
why.

Then the tool, matched to the failure. `traceback()` for where it died.
`print()` for how a value evolved. Bisection for a wrong number that never
errored.

:::

::: {.content-visible when-format="revealjs"}

- Read the error before changing anything.
- Reproduce it on the smallest input that still fails.
- One hypothesis at a time. Change one thing per run.
- Then the tool: `traceback()`, `print()`, bisection.

:::

## `traceback()`: where it died {#traceback}

```{r}
#| label: traceback
#| script: "traceback(): where it died"
#| eval: false
frames <- lapply(c("data/epa_pm25_compton_2025.csv", "data/nope.csv"), read_epa)
traceback()
```

```
Error in file(file, "rt") : cannot open the connection
In addition: Warning message:
In file(file, "rt") :
  cannot open file 'data/nope.csv': No such file or directory

5: file(file, "rt")
4: read.table(file = file, header = header, sep = sep, quote = quote, 
       dec = dec, fill = fill, comment.char = comment.char, ...)
3: read.csv(path, colClasses = c(Site.ID = "character"))
2: FUN(X[[i]], ...)
1: lapply(c("data/epa_pm25_compton_2025.csv", "data/nope.csv"), 
       read_epa)
```

::: {.content-hidden when-format="revealjs"}

`traceback()` prints the calls that were running when the error happened,
innermost first. Read it from the bottom: your `lapply()` (1) called your
function (2, `FUN` is `read_epa`), which called `read.csv()` (3), which
could not open the file (5). The warning names the file. The error is in
the line you wrote at 1, not in `read.csv()`.

:::

::: {.content-visible when-format="revealjs"}

- Read it from the bottom: your call, your function, then R's.
- The error is in the line you wrote at 1. The warning names the file.

:::

## `print()` and `cat()`: how a value evolved {#print}

```{r}
#| label: print
#| script: "print() and cat(): how a value evolved"
for (site in names(by_site)[1:3]) {
  s <- by_site[[site]]
  cat(site, ": ", nrow(s), " days, ", days_above(s$pm25), " above\n", sep = "")
}

print(c(days = nrow(jan), above = days_above(jan$pm25)))
```

::: {.content-hidden when-format="revealjs"}

Inside a loop or a function, a `print()` or `cat()` line shows what a value
was on each pass. `print()` shows an object the way the console would;
`cat()` pastes text and numbers into one line, and `"\n"` ends it. A named
vector, `c(days = , above = )`, prints label and value together. Take the
line out once the question is answered.

`browser()` is the same idea with a pause. Put `browser()` inside a
function and the run stops there with a `Browse[1]>` prompt; you can print
any local name, `n` steps to the next line, `c` continues, `Q` quits.

:::

::: {.content-visible when-format="revealjs"}

- `cat()` pastes text and numbers into one line. `print()` shows an object.
- A named vector prints label and value together.
- Take the line out once the question is answered.

:::

## Bisection: a wrong number that never errored {#bisection}

::: {.content-hidden when-format="revealjs"}

No error, no warning, and the number is wrong. `traceback()` has nothing
to point at. Where could the wrong number have entered? Every step that
changes a row count or a value is a suspect.

Write down what each step should produce: rows after the read, rows after
the filter, the date range, the count. Comment out the second half of the
script and run the first half. Compare what printed with what you wrote
down. Halve the half that disagrees. Repeat until the defect has nowhere to
hide. It is usually one line, and the line is usually not the one that
printed the number.

:::

::: {.content-visible when-format="revealjs"}

- No error, no warning, and the number is wrong. Nothing to trace.
- Write down what each step should produce: rows, dates, a count.
- Comment out half. Run. Compare. Halve the half that disagrees.
- Repeat until the defect has nowhere to hide. Usually one line.

:::

## Documenting a diagnosis {#diagnosis}

```{r}
#| label: diagnosis
#| script: "Documenting a diagnosis: line, mechanism, fix, proof"
# Line:      share_above <- function(x) sum(x > standard) / length(x)
# Mechanism: standard is not an argument; the function reads it from the
#            workspace and fails in a fresh session
# Fix:       the standard comes in as an argument, with a default
share_above <- function(x, standard = 35) sum(x > standard) / length(x)
# Proof:
stopifnot(share_above(c(1, 50)) == 0.5)
```

::: {.content-hidden when-format="revealjs"}

Four lines, as comments, next to the fix. The line that was wrong. The
mechanism, meaning why it produced what it produced. The fix. And the
proof: a small case that now gives the right answer, written as a check
that runs. That is what a fixed script carries. When a check fails and you
cannot fix it, the same four lines with the fix left blank are the honest
note, and they are worth more than an edited check.

:::

::: {.content-visible when-format="revealjs"}

- Line. Mechanism. Fix. Proof, as a check that runs.
- Cannot fix it? The same four lines, fix left blank. Never an edited check.

:::

## The bug hunt {#bug-hunt}

::: {.content-hidden when-format="revealjs"}

No AI. The script below runs clean and prints a wrong number. Find the
line, then write the four-line diagnosis: line, mechanism, fix, proof.
Copy the script into its own file if you want to halve it.

:::

::: {.panel-tabset}

### Exercise

```{r}
#| label: bug-hunt
#| exercise: true
#| script: "The bug hunt (15 minutes)"
# No AI. Runs clean, prints a wrong number. Find the line; write the
# diagnosis: line, mechanism, fix, proof.
# 0) Setup ----
STANDARD <- 35
air <- read.csv("data/epa_pm25_compton_2025.csv",
                colClasses = c("Site.ID" = "character"))
names(air)[names(air) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"
air$date <- as.Date(air$Date, format = "%m/%d/%Y")
# 1) The month ----
jan_air <- air[format(air$date, "%m") == "01", ]
# 2) One instrument ----
air_one <- air[air$POC == 1 & air$AQS.Parameter.Code == 88101, ]
# 3) The count ----
count_above <- function(x, standard = STANDARD) sum(x > standard)
cat("January:", count_above(jan_air$pm25), "days above the standard\n")
```

### Solution

```{r}
#| label: bug-hunt-solution
#| purl: false
nrow(jan_air)         # 62 rows for a 31-day month
table(jan_air$POC)    # two instruments, 31 rows each
```

```{r}
#| label: bug-hunt-diagnosis
#| purl: false
# Line:      jan_air <- air[...] (section 1) runs before the filter (section 2)
# Mechanism: both instruments in January, 62 rows for 31 days; each day
#            above the standard counted twice, 8 instead of 4
# Fix:       cut the month from the one-instrument frame
jan_air <- air_one[format(air_one$date, "%m") == "01", ]
# Proof:
stopifnot(nrow(jan_air) == 31, count_above(jan_air$pm25) == 4)
```

::: {.content-hidden when-format="revealjs"}

Nothing errored, nothing warned, and 8 is a believable number for the
month of the January 2025 Los Angeles fires. The defect is the order of two sections: the month was
cut before the instrument was chosen. Bisection finds it at the first
halving, because the row count written down for section 1 was 31.

:::

:::

## Your style sheet {#style-sheet}

::: {.content-visible when-format="revealjs"}

On the page: a table with eleven rows, one per dimension of how you write
code. Copy it into a file in your project and fill in one rule per row.
Keep it with your code.

:::

::: {.content-hidden when-format="revealjs"}

Every choice below is one you make anyway, by habit, every time you write
a script. Making them once, on purpose, and writing them down is what
turns habits into a style. Copy the table into a file in your project and
fill in one rule per row. Keep it with your code.

| Dimension | A rule from this page | Your rule |
|:--|:--|:--|
| Project structure | Code, data, and output in separate folders; raw data never edited | |
| File-level conventions | A title block; one script, one job | |
| Naming | `snake_case`; verbs for functions; capitals for constants | |
| Formatting | Spaces around `<-` and `=`; one statement per line; indent inside braces | |
| When to write a function | The second time you paste the same lines | |
| Function design | Everything the function needs comes in through its arguments; the value is the last expression | |
| Control flow and idioms | `lapply()` when each pass returns a value; `for` when order or a side effect matters | |
| Errors, messages, logging | `stopifnot()` after every step that changes a row count | |
| Comments and documentation | Comments say why; a diagnosis is four lines | |
| Interactive workflow | Test small, then run big; restart R and source before you trust a result | |
| Reproducibility hygiene | Relative paths from the project folder; never `setwd()`; a fresh session runs top to bottom | |

:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
The sheet comes back when students start briefing an assistant: the
filled-in table is the brief. Do not say so today.
:::
:::

## Practice {#practice}

::: {.content-visible when-format="revealjs"}

Nine exercises with solutions, on the session page and in the live script.

:::

::: {.content-hidden when-format="revealjs"}

::: {.panel-tabset}

### Practice

```{r}
#| label: practice
#| exercise: true
#| script: "Practice"
#| eval: false
# Everything below uses pm, dec, jan, files, frames, la, by_site,
# read_epa() and days_above(), all built earlier in this script.

# Functions
#  1. Write peak_day(d): the date of the highest pm25 in a data frame d.
#     Run it on pm. Check the answer against the row that which.max()
#     points at.
#  2. Run peak_day() on every site in by_site. Which site peaked last in
#     January? (sapply() turns dates into numbers; wrap the call in
#     format() to keep them readable.)
#  3. From memory, without scrolling up: write share_above(x, standard = 35)
#     so that it works in a fresh session, with nothing read from the
#     workspace. Run it on dec$pm25 and on c(1, 50).

# Lists and loops
#  4. How many sites does each frame in frames hold? One sapply() call.
#  5. Trace this by hand, writing the value of total after each pass, then
#     run it:
#       x <- c(3, 8, 1); total <- 0
#       for (i in seq_along(x)) total <- total + x[i]
#  6. One plot per site, written to disk: for each site in by_site, a
#     line plot of pm25 against date saved as a PNG named after the site.
#     Do not run it until it works for one site.

# Checks
#  7. After read_epa() on the county file, write one stopifnot() with two
#     conditions: exactly 8 sites, and no date outside 2025.
#  8. This check fails on la. Say why, and whether the data or the check
#     is wrong. Then write the check that is right for this frame.
#       stopifnot(nrow(la) == length(unique(la$date)))
#  9. Time read_epa() on the county file alone, then lapply() over both
#     files. Does the second take more than the first? By roughly how much?
```

### Solutions

::: {.callout-tip collapse="true"}
## Solutions

```{r}
#| label: practice-solutions
#| purl: false
# 1.
peak_day <- function(d) d$date[which.max(d$pm25)]
peak_day(pm)
pm[which.max(pm$pm25), c("date", "pm25")]

# 2. Pasadena, on 2025-01-10. Four sites peaked on January 1.
sapply(by_site, function(s) format(peak_day(s)))

# 3.
share_above <- function(x, standard = 35) sum(x > standard) / length(x)
round(share_above(dec$pm25), 3)
share_above(c(1, 50))

# 4.
sapply(frames, function(d) length(unique(d$site)))

# 5. total is 3, then 11, then 12.
x <- c(3, 8, 1); total <- 0
for (i in seq_along(x)) total <- total + x[i]
total

# 7. The second condition is a whole vector; every element must be TRUE.
stopifnot(length(unique(la$site)) == 8,
          format(la$date, "%Y") == "2025")

# 8. The check is wrong for this frame, not the data. Several sites
#    report on the same day, so dates repeat by design. The right check
#    is one row per site and day.
nrow(la) == length(unique(la$date))
stopifnot(nrow(la) == nrow(unique(la[, c("site", "date")])))

# 9. Barely. The Compton file is about a seventh the size of the county
#    file, so the second call costs about a seventh more.
system.time(read_epa("data/epa_pm25_la_county_2025.csv"))
system.time(lapply(files, read_epa))
```

```{r}
#| label: practice-solution-6
#| purl: false
#| eval: false
# 6. Test on one site first: for (site in names(by_site)[1]). Then all of them.
for (site in names(by_site)) {
  s <- by_site[[site]]
  png(paste0(site, ".png"), width = 1600, height = 800, res = 150)
  plot(s$date, s$pm25, type = "l", main = site,
       xlab = "", ylab = "PM2.5 (ug/m3)")
  dev.off()
}
```

:::

:::

:::

## Quick reference {#quick-reference}

::: {.content-visible when-format="revealjs"}

On the page: every command on this page in one table, and the habits next
to it.

:::

::: {.content-hidden when-format="revealjs"}

**Functions and lists**

| Code | What it does |
|:--|:--|
| `f <- function(x, k = 35) { ... }` | A function; `k = 35` is a default; the value is the last expression |
| `return(x)` | Leave the function now with `x`; only needed to leave early |
| `function(s) ...` inside `lapply()` | A function with no name, written where it is used |
| `list.files("data", pattern =, full.names = TRUE)` | Paths you can hand to a function |
| `lapply(files, f)` | A list in, a list out: `f` on each element |
| `sapply(x, f)` | The same, simplified when every element has the same shape |
| `split(d, d$g)` | A list, one data frame per value of `g` |
| `do.call(rbind, pieces)` | Stack a list of data frames back into one |
| `for (i in seq_along(x)) out[i] <- ...` | A loop; make `out` first, at full size |

**Checks and habits**

| Code | What it does |
|:--|:--|
| `stopifnot(a, b)` | Stop the script and name the first condition that is not `TRUE` |
| `stopifnot("why" = a)` | The same, printing your message instead |
| `abs(x - 21.28) < 0.05` | Compare a computed number with a tolerance, never `==` |
| `STANDARD <- 35` | A named constant at the top; used by name everywhere |
| `# Name ----` | A section RStudio's outline can jump to |
| `system.time(expr)` | Seconds the line took; wrap the big call once |
| Session > Restart R, then `source("air.R")` | The only test that a script runs on its own |
| `getwd()`, `list.files("data")` | Where relative paths are looked for, and what is there |

**When something is wrong**

| Symptom | Tool |
|:--|:--|
| An error | Read it. Then `traceback()`: your call is at the bottom |
| A loop or function misbehaves on some passes | `print()` or `cat()` inside it; `browser()` to pause there |
| A wrong number and no error | Bisection: write down what each step should give, halve, compare |
| `object 'x' not found` inside a function | The function reads a global; make `x` an argument |
| `NaN` from a mean | The subset was empty; that is the honest answer |
| A file "does not exist" | `getwd()` and `list.files()` before blaming the file |

:::

```{r}
#| purl: false
#| include: false
source("_purl.R")
purl_session("06-functions-debugging.qmd", "06-functions-debugging.R")
```
