---
title: "6 · How to code"
subtitle: "Thursday, September 10, 2026"
pagetitle: "6 · How to code"
title-prefix: ""
execute:
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
  revealjs:
    output-file: 06-how-to-code-slides.html
    theme: [default, slides.scss]
    template-partials:
      - title-slide.html
    footer: "AEM 6850 · Session 6"
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    code-overflow: scroll
    history: false
    link-external-newwindow: true
---

```{r}
#| purl: false
#| include: false
knitr::opts_chunk$set(
  comment = "#>",
  collapse = FALSE,
  fig.width = 8,
  fig.height = 4,
  fig.align = "center",
  dpi = 150
)
```

::: {.content-hidden when-format="revealjs"}

[Slides](06-how-to-code-slides.html) · [Live script](06-how-to-code.R) ·
[Source](https://github.com/ArielOrtizBobea/aem6850/blob/main/fall-2026/sessions/06-how-to-code.qmd)

How a project and a script are organized, the habits that make a script
readable and checkable, writing functions instead of copy-pasting, and what
to do when a number is wrong.

:::

## Objectives {#objectives}

By the end of this session you can:

- Lay out a project: folders for code, data and output, one script per
  task, and a master script that runs them in order.
- Structure a script: sections, names, comments that say why, named
  constants, no duplicated code.
- Write a function and call it with its own arguments.
- Run one function over every file in a folder with `lapply()`.
- Stop a script with `stopifnot()` when a number is not what you wrote down.
- Localize a mistake by halving the script, and write the diagnosis down:
  line, mechanism, fix, proof.

## Plan for today {#plan}

1. Motivation and philosophy
2. Organizing projects, scripts, and code
3. Guiding principles
4. Writing functions
5. Checking a script as you write it
6. Finding a mistake when the number is wrong

## Disclaimers {#disclaimers}

- I'm not a computer scientist.
- The ideas here reflect my experience and that of others.
- Although our work is in R, many of these ideas apply to other languages.

## Motivation {#motivation}

Good code...

- Increases confidence in your work
- Increases the chances other researchers will reproduce, replicate and
  cite your work
- Saves you (and others) time
- Reduces the chances of making mistakes
- Is easy to understand
- May be portable

## Coding philosophy {#philosophy}

::: {.content-hidden when-format="revealjs"}

You are writing code for others. That includes yourself at $t+1$.

A script is written once and read many times. You read it again when a
referee asks where a number came from, when a coauthor wants the same
table on a different sample, when you reopen a project after a semester
on something else. Each of those readers works from the file. None of
them works from what you remember about the file.

Most of the time that reader is you. Not the you who wrote it, who knew
which rows were dropped and why. The you six months from now, who does
not. What you knew at the keyboard is not in the file; only what you
wrote down is. Everything in this session follows from that.

:::

::: {.content-visible when-format="revealjs"}

- You are writing code for others. That includes yourself at $t+1$.
- A script is written once and read many times.
- What you knew when you wrote it is not in the file. Only what you
  wrote down is.

:::

## Two kinds of efficiency {#efficiency}

::: {.content-hidden when-format="revealjs"}

Efficiency has two meanings and they pull against each other.

One is the machine's: how many operations, how much memory, how long the
run takes. The other is yours: how long it takes you to read the script,
change it, and work out why a number moved.

Compact, fast code can be hard to follow. Code you can trace line by line
often does more work than it strictly needs to. Neither is always right,
but at the scale of a research script the machine's time is cheap and
yours is not. A script that takes four seconds instead of one has cost
you nothing. A script you cannot follow has cost you an afternoon.

This session is about the second kind.

:::

::: {.content-visible when-format="revealjs"}

- Fast for the *machine*: fewer operations, less memory, shorter runtime.
- Fast for *you*: you can read it, change it, and find the mistake.
- There can be a tradeoff, but in our context writing code that is efficient for you to read and understand dominates.

:::

## Organizing a project {#organizing-project .smaller}

- A project comes with general instructions

    - It could be a `readme.txt` file
    - It should specify the authors, what the project is about, and
      describe the files

- Folders for different types of files

    - Don't mix raw data with "cleaned" data

- No single best way of organizing, but this is a good start:

    - Folder for code (no subfolders)
    - Folder(s) for input(s): raw data you never modify, and
      cleaned data your scripts create
    - Folder(s) for output(s): figures, tables
    - Folder for packages (optional)

## Example of folder structure {#folder-structure}

- `code` — all script files, including the RStudio project

- `data`

    - `raw` — data as downloaded from the web, can be large
    - `clean` — data needed for the analysis

- `output` — what shows up in the paper

    - `figures`
    - `tables`

- `paper`

- `presentations`

- Others: meeting notes, documentation

## Organizing your script files {#organizing-scripts}

- A script file should accomplish a specific and well-identified task

    - Cleaning data, regression analysis, and so on

- Never have all your code in a single script file, unless it is a tiny
  project — your homework, for instance

- Script files should have intuitive names and follow a workflow:

    - `1.1_clean_economic_data.R`
    - `1.2_clean_pollution_data.R`
    - `1.3_prepare_regression_data.R`
    - `2.1_regression_analysis.R`
    - `2.2_post_estimation_analysis.R`

## Good practice to create special script files {#special-scripts}

- A master script file, `master.R`, that runs all the project's scripts
  in the correct order

    - Installs and loads all necessary packages
    - Simplifies life for whoever reproduces your work
    - Typically written at the very end of the project

- A file containing user-defined functions, `functions.R`

    - Improves automation when the same function is used in several
      scripts
    - Invites you to write more portable functions for future projects

## Code within a script file {#sections}

```{r}
#| label: sections
#| eval: false
#| purl: false
# air.R -- days above the PM2.5 standard at one monitor, 2025
# Your Name, NetID

# 0) Setup ----
STANDARD <- 35                        # US EPA 24-hour standard, ug/m3
pm <- read.csv("data/epa_pm25_compton_2025.csv",
               colClasses = c("Site.ID" = "character"))

# 1) One instrument, one month ----

# 2) The numbers ----

# 3) Write results.csv ----
```

::: {.content-hidden when-format="revealjs"}

A title block says what the script does and who wrote it. A setup section
holds packages, constants, and the data coming in. Then one section per
task, numbered in the order they run, and nothing from setup reappears
lower down. A comment ending in four dashes is a section: RStudio's outline
(Ctrl+Shift+O, Cmd+Shift+O on a Mac) lists them, and you can jump between
them. A script does one job. Two jobs, two scripts.

:::

::: {.content-visible when-format="revealjs"}

- Title block, setup, then one numbered section per task.
- `# Name ----` is a section. RStudio's outline lists them.
- A script does one job. Two jobs, two scripts.

:::

## Two versions of the same script {#clarity}

Both count the days above the standard at one monitor. Both are correct.

```{r}
#| label: clarity-legible
#| eval: false
#| purl: false
STANDARD <- 35                    # US EPA 24-hour standard, ug/m3

pm <- read.csv("data/epa_pm25_compton_2025.csv")
names(pm)[names(pm) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"
pm <- pm[pm$POC == 1, ]           # one instrument

sum(pm$pm25 > STANDARD)
```

::: {style="height:1.1em"}
:::

```{r}
#| label: clarity-clever
#| eval: false
#| purl: false
compose <- function(...) { fs <- list(...); function(x) Reduce(\(a, f) f(a), fs, x) }

run <- compose(read.csv, \(d) subset(d, POC == 1),
               \(d) d$Daily.Mean.PM2.5.Concentration, \(v) sum(v > 35))

run("data/epa_pm25_compton_2025.csv")
```

::: {.content-visible when-format="revealjs"}

- The second is shorter and more reusable. It is also unreadable at a
  glance, and the `35` has disappeared into an argument.
- Short and clever are not the same thing as clear.

:::

::: {.content-hidden when-format="revealjs"}

The second version is shorter, and by the standards of software
engineering it is better: nothing is repeated, each step is a separate
piece, and the whole thing could be pointed at another file tomorrow. It
is also unreadable at a glance. The `35` has disappeared into the last
argument as a bare number with no name, there is no object between the
file and the answer, and you have to read it inside out to work out what
happens first.

Research code is read far more often than it is reused. These are the
dimensions that decide whether a reader can follow it:

| | The first version | The second |
|:--|:--|:--|
| **Names** | `STANDARD` says what `35` is | `35` is a bare number in the last argument |
| **Visible values** | every number is on its own line | values are buried inside calls |
| **Intermediate objects** | `pm` exists; you can print it | nothing between the file and the answer |
| **Somewhere to put a check** | after any step, on `pm` | nowhere |
| **Layers to hold at once** | one line at a time | `compose`, `Reduce`, and four unnamed functions |
| **Reading order** | top to bottom, in the order it happens | inside out |

The same list applies to a single function: a function is easy to read
when its name says what it returns, its arguments are visible at the call,
and its body is short enough to hold in your head at once.

:::

## Guiding principles 1/2 {#principles-1 .smaller}

- Think constantly about reproducibility

    - Your script files should run entirely without errors — start with
      your own computer
    - Rely on relative paths, no hard coding
    - RStudio projects help. Use them.

- Do not repeat yourself

    - Do not copy-paste code within a script file. Write a function and
      use loops.
    - Do not copy-paste code across script files. Put the function in its
      own file and load it when needed.
    - Copy-pasting code is a red flag.

- Give objects and folders simple and intuitive names

    - Good names are short and informative
    - Limit upper case and odd characters: `MyModified.GDP.var` or
      `log_gdp`?

## Guiding principles 2/2 {#principles-2 .smaller}

- Annotate your code concisely but generously

    - Better to err on the side of too many annotations
    - They should help a reader understand the purpose of any significant
      piece of your code
    - Annotate single lines when the code is relatively complex

- Make your code look good, which is to say easy to read

    - Keep consistent indentation and spacing
    - Exploit RStudio's ability to fold your code
    - Avoid deep nesting when possible
    - Let your code breathe — leave some blank space
    - Favor several lines with short expressions over one long line with
      several expressions

## General tips on the process of writing code {#process-tips .smaller}

- Before starting, visualize the workflow

    - *"If you don't know where you're going, any road'll take you
      there"* — after *Alice in Wonderland*, Lewis Carroll
    - Have a clear idea of the repetitive tasks
    - Think about how the project and the code should be structured when
      it is done

- Then start writing your code interactively

    - Working in the console and the editor together
    - But do not lose sight of the forest for the trees

- Do not run a big task without testing it on a small dataset

- Constantly re-check that your code runs *entirely*, not just in pieces

- Leave breadcrumbs so you can find a mistake later

## Getting the data {#data .smaller}

Two files: one monitor in Compton, and every PM2.5 monitor in Los Angeles
County, both for 2025. This chunk fetches them if you do not already have
them, and does nothing if you do.

```{r}
#| label: get-data
#| script: "Get the data"
dir.create("data", showWarnings = FALSE)

f <- "data/epa_pm25_compton_2025.csv"
if (!file.exists(f)) {
  download.file(paste0("https://arielortizbobea.github.io/aem6850/",
                       "fall-2026/sessions/", f), f)
}

f <- "data/epa_pm25_la_county_2025.csv"
if (!file.exists(f)) {
  download.file(paste0("https://arielortizbobea.github.io/aem6850/",
                       "fall-2026/sessions/", f), f)
}

list.files("data", pattern = "^epa_.*\\.csv$")
```

::: {.content-hidden when-format="revealjs"}

`dir.create()` makes the folder and stays quiet when it is already there.
`download.file()` fetches a URL to a path. The `if (!file.exists(f))`
wrapper means running this a second time costs nothing.

A script that fetches its own inputs runs on a machine that has never seen
the data. That includes a coauthor's, a referee's, and your own six months
from now.

:::

::: {.content-visible when-format="revealjs"}

- `dir.create()`, then `download.file()` only if the file is missing.
- A script that fetches its own inputs runs on a machine that has never
  seen the data.

:::

## Building the Compton dataframe {#build}

Everything below runs off the Compton file, read here so this page and its
script stand on their own.

```{r}
#| label: build
#| script: "Build the Compton frame"
pm <- read.csv("data/epa_pm25_compton_2025.csv",
               colClasses = c("Site.ID" = "character"))
names(pm)[names(pm) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"
names(pm)[names(pm) == "Local.Site.Name"]                <- "site"
pm$date <- as.Date(pm$Date, format = "%m/%d/%Y")
pm <- pm[pm$POC == 1 & pm$AQS.Parameter.Code == 88101, ]

nrow(pm)   # one regulatory instrument, the days it ran
```

::: {.content-hidden when-format="revealjs"}

Five lines: read the file with the identifier protected, rename the
concentration and site columns, convert the date, keep one instrument.
Every EPA (the US Environmental Protection Agency) file needs the same five
lines, and every script that reads one starts by typing them again. A
function lets you type them once.

:::

::: {.content-visible when-format="revealjs"}

- Read with the identifier protected, rename, convert the date, keep one
  instrument. 301 rows.
- Every EPA file needs these five lines. A function lets you type them once.

:::

## Naming things and named constants {#naming}

```{r}
#| label: naming
#| script: "Naming things and named constants"
STANDARD <- 35   # US EPA 24-hour PM2.5 standard, ug/m3

sum(pm$pm25 > STANDARD)   # days above the standard, 2025
```

::: {.content-hidden when-format="revealjs"}

Names are short and informative, lower case, with underscores: `pm25`,
`site`, `days_above`. Verbs for functions. Not `MyModified.PM.var`, not
`T` for `TRUE`.

A number that means something gets a name at the top of the script, in
capitals, and is used by that name everywhere. `35` typed in six places is
six chances to change five of them.

Comments say why, not what. `# EPA 24-hour standard` tells a reader
something the code cannot; `# set x to 35` repeats the line below it.

:::

::: {.content-visible when-format="revealjs"}

- Short, lower case, underscores. Verbs for functions. Never `T` for `TRUE`.
- A number that means something gets a name at the top, in capitals.
- Comments say why. The code already says what.

:::

## Write code that is easy to check {#checkable}

::: {.content-hidden when-format="revealjs"}

Whether code can be checked is a property of how it is built, not of how
hard you look at it afterwards. Small pieces with one job each. Intermediate
values that land in named objects, so `nrow(pm)` after the filter and
`range(pm$date)` after the date conversion are one line away. No step that
changes three things at once. Results that come out where a check can reach
them. Code you cannot check is code you cannot defend, whatever it
computed.

:::

::: {.content-visible when-format="revealjs"}

- Small pieces, one job each.
- Intermediate values in named objects you can print.
- No step that changes three things at once.
- Code you cannot check is code you cannot defend.

:::

## `function()`: naming a block of code {#function}

```{r}
#| label: function
#| script: "function(): naming a block of code"
days_above <- function(x, standard = 35) {
  sum(x > standard)
}

dec <- pm[format(pm$date, "%m") == "12", ]
days_above(dec$pm25)                  # December, at the 35 standard
days_above(dec$pm25, standard = 50)   # the same days, a higher bar
days_above(c(1, 50))                  # a case you can check by eye
```

::: {.content-hidden when-format="revealjs"}

`function()` builds a function. The names in its parentheses are the
arguments; `standard = 35` gives one of them a default, so a call can leave
it out. The body runs when you call the function, with `x` standing for
whatever you passed in. The value of a function is its last expression.
`return(x)` says the same thing explicitly, and is only needed to leave a
function early.

December had 5 days above the standard and 3 above 50. The last call is
the one to keep in the habit: two numbers you can check by eye, and the
answer must be 1.

:::

::: {.content-visible when-format="revealjs"}

- `function(x, standard = 35)`: the arguments. `= 35` is a default.
- The value is the last expression. `return()` is for leaving early.
- December: 5 days above 35, 3 above 50. `c(1, 50)`: 1, as it must be.

:::

## Two versions of the same function {#function-clarity}

Both return the number of days above the standard.

```{r}
#| label: fn-legible
#| eval: false
#| purl: false
days_above <- function(x, standard = 35) {
  sum(x > standard)
}
```

::: {style="height:1.1em"}
:::

```{r}
#| label: fn-clever
#| eval: false
#| purl: false
da <- function(v, s = 35, na = TRUE, f = NULL) {
  if (!is.null(f)) v <- f(v)
  length(which(if (na) v[!is.na(v)] > s else v > s))
}
```

::: {.content-visible when-format="revealjs"}

- The name does not say what it returns. Four arguments where one is
  needed, three of them hidden at the call.
- A function is easy to read when its name says what it returns and its
  body fits in your head.

:::

::: {.content-hidden when-format="revealjs"}

The second returns the same answer, and it does more: it drops missing
values, and it will apply a transformation first if you hand it one. Every
one of those additions was free to write and is paid for on every reading.

`da` does not say what it returns. `v`, `s`, `na` and `f` do not say what
they take. Three of the four arguments have defaults nobody at the call
site will see, so `da(x)` hides three decisions. `length(which(...))`
counts what `sum()` counts, in more steps.

The same dimensions from the two script versions apply here: does the name
say what comes back, are the values visible at the call, and is the body
short enough to hold at once.

:::

## Local names and global names {#scope}

```{r}
#| label: scope
#| script: "Local names and global names"
#| error: true
count_pos <- function(x) {
  n <- sum(x > 0)   # n exists inside the call
  n
}
count_pos(c(-1, 2, 3))
n                     # and not outside it
```

::: {.content-hidden when-format="revealjs"}

A name made inside a function dies at the closing brace. Outside the call,
`n` never existed. That is what lets a function use short names like `n`
and `d` without trampling anything in the workspace.

:::

::: {.content-visible when-format="revealjs"}

- Names made inside a function die at the closing brace.
- So a function can use short names without touching the workspace.

:::

## A function can read a global name {#scope-global}

```{r}
#| label: scope-global
#| script: "A function can read a global name"
#| error: true
standard <- 35
share_above <- function(x) sum(x > standard) / length(x)
share_above(dec$pm25)

rm(standard)
share_above(dec$pm25)   # worked until the global was gone
```

::: {.content-hidden when-format="revealjs"}

The other direction is the trap. A function can read a name from outside
itself, and `share_above()` does: it uses `standard` without ever being
given it. It works while `standard` happens to sit in the workspace and
fails in a fresh session with `object 'standard' not found`. Everything a
function needs should come in through its arguments. Then it works
anywhere, and a two-number test tells you it reads its own input.

:::

::: {.content-visible when-format="revealjs"}

- A function can read a global name. It works until the global is gone.
- Everything a function needs comes in through its arguments.

:::

## Writing the five lines once {#setup-objects .smaller}

```{r}
#| label: setup-objects
#| script: "Writing the five lines once"
read_epa <- function(path) {
  d <- read.csv(path, colClasses = c("Site.ID" = "character"))
  names(d)[names(d) == "Daily.Mean.PM2.5.Concentration"] <- "pm25"
  names(d)[names(d) == "Local.Site.Name"]                <- "site"
  d$date <- as.Date(d$Date, format = "%m/%d/%Y")
  d[d$POC == 1 & d$AQS.Parameter.Code == 88101, ]
}

files  <- list.files("data", pattern = "^epa_.*\\.csv$", full.names = TRUE)
frames <- lapply(files, read_epa)
names(frames) <- basename(files)

la      <- frames[["epa_pm25_la_county_2025.csv"]]
la_jan  <- la[format(la$date, "%m") == "01", ]
by_site <- split(la_jan, la_jan$site)

sapply(by_site, nrow)
```

::: {.content-hidden when-format="revealjs"}

The five lines from the build, with `path` in place of the file name and
the frame as the value. Copy-pasting code is a red flag, and the function
is the fix: a change to the rename or to the instrument filter now happens
in one place, and every file gets it.

`lapply()` hands the function every EPA file in the folder. `split()` cuts
January into one frame per site. Eight sites, not the county's eleven:
three run no regulatory instrument, so the filter inside `read_epa()`
removed them. Count rows before you trust a mean — four sites report
between 6 and 11 January days.

:::

::: {.content-visible when-format="revealjs"}

- The five lines, once, with `path` as the argument. A change happens in
  one place.
- `lapply()` over every EPA file, `split()` into one frame per site.
- Eight sites, not eleven: three run no regulatory instrument.

:::

## Exercise {#exercise-1}

::: {.panel-tabset}

### Exercise

```{r}
#| label: exercise-1
#| exercise: true
#| script: "Exercise (5 minutes)"
#| eval: false
# 1. Write mean_above(x, standard = 35): the mean of the readings in x
#    that are above the standard.
# 2. Run it on every site in by_site with sapply().
# 3. One number to compare with the room: Pasadena.
# 4. One site returns NaN. Say why in one sentence, and whether that is
#    wrong.
```

### Solution

```{r}
#| label: exercise-1-solution
#| purl: false
mean_above <- function(x, standard = 35) mean(x[x > standard])
round(sapply(by_site, function(s) mean_above(s$pm25)), 1)
```

::: {.content-visible when-format="revealjs"}

Pasadena 58.1. Lancaster - Fairgrounds `NaN`: no reading above 35, so the
mean of nothing. Honest.

:::

::: {.content-hidden when-format="revealjs"}

Pasadena 58.1. Lancaster - Fairgrounds returns `NaN`: no January reading
there was above 35, so `x[x > standard]` is empty and the mean of nothing
is not a number.
That is the honest answer. A function that quietly returned 0 would be
lying.

:::

:::

## `stopifnot()`: a check that stops the script {#stopifnot}

```{r}
#| label: stopifnot
#| script: "stopifnot(): a check that stops the script"
jan <- pm[format(pm$date, "%m") == "01", ]
stopifnot(nrow(pm) == 301)
stopifnot(nrow(jan) == 31)
```

::: {.content-hidden when-format="revealjs"}

`stopifnot()` takes conditions. If every one is `TRUE` it does nothing; if
any is `FALSE` it stops the script and names the first that failed, so a
reader knows what was expected without opening the script. Several
conditions fit in one call, one per line when they are long. A check that
sits commented out is turned on by selecting its lines and pressing
Ctrl+Shift+C (Cmd+Shift+C on a Mac).

Put the check right after the step it guards. A step that changes a row
count is where a mistake enters without announcing itself, and a check
there turns a wrong number into a stopped script.

:::

::: {.content-visible when-format="revealjs"}

- Silent when every condition holds. Stops the script when one fails.
- Several conditions in one call. Right after the step it guards.
- The message names the condition that failed.

:::

## Test small, then run big {#test-small}

```{r}
#| label: test-small
#| script: "Test small, then run big"
days_above(c(1, 50))                              # a case you can check by eye
nrow(read_epa(files[1]))                          # one file before the folder
system.time(lapply(files, read_epa))              # the whole folder, timed
```

::: {.content-hidden when-format="revealjs"}

Never run the big job untested. A function gets a case you can check by
eye. A loop runs on one file before the folder, on one site before all of
them. `system.time()` around the big call tells you what the next run will
cost: nothing here, and minutes when a folder holds a thousand files.

:::

::: {.content-visible when-format="revealjs"}

- Never run the big job untested. A function gets a case you can check by eye.
- One file before the folder. `system.time()` around the big call.

:::

## Restart R and source the script {#restart}

```{r}
#| label: restart
#| script: "Restart R and source the script"
#| eval: false
#| purl: false
# Session > Restart R   (Ctrl+Shift+F10 on Windows, Cmd+Shift+0 on a Mac)
ls()                    # character(0), or it was not a clean restart
getwd()                 # where "data/..." is looked for
list.files("data")      # what is there
source("air.R")         # your script, top to bottom, in a clean session
```

::: {.content-hidden when-format="revealjs"}

A script that runs only because of what is already in your environment is
not a script yet. Restart R: packages unload, and the workspace empties
once Tools > Global Options > General has "Save workspace to .RData on
exit" set to Never and "Restore .RData into workspace at startup"
unchecked. On the default settings RStudio puts the old objects back after
a restart, so set both once, and expect `ls()` to print `character(0)`.
Then `source()` the file. It runs top to bottom and stops at the first
error. `rm(list = ls())` clears objects but not packages or options; it is
not a restart.

`source()` reads `"data/..."` relative to the working directory. Opening a
project's `.Rproj` file sets the working directory to the project folder,
which is why every path in a script is relative: `"data/file.csv"`, never
`"/Users/you/Desktop/file.csv"`, which runs on one laptop and nowhere else.
When R says a file does not exist, check `getwd()` and `list.files("data")`
before blaming the file.

:::

::: {.content-visible when-format="revealjs"}

- Restart R, check `ls()` is empty, then `source()`. `rm()` is not a restart.
- Paths are relative to `getwd()`. The `.Rproj` file sets it when opened.
- "File does not exist": check `getwd()` and `list.files("data")` first.

:::

## Debugging as a strategy {#debugging}

::: {.content-hidden when-format="revealjs"}

Read the error before changing anything. It names the call that failed and
usually the cause. Reproduce the failure on the smallest input that still
fails: one file, one site, two numbers. Form one hypothesis at a time and
test it. Change one thing per run, so that when the output changes you know
why.

Then the tool, matched to the failure. `traceback()` for where it died.
`print()` for how a value evolved. Halving for a wrong number that never
errored.

:::

::: {.content-visible when-format="revealjs"}

- Read the error before changing anything.
- Reproduce it on the smallest input that still fails.
- One hypothesis at a time. Change one thing per run.
- Then the tool: `traceback()`, `print()`, halving.

:::

## `traceback()`: where it died {#traceback}

```{r}
#| label: traceback
#| script: "traceback(): where it died"
#| eval: false
frames <- lapply(c("data/epa_pm25_compton_2025.csv", "data/nope.csv"), read_epa)
traceback()
```

```
Error in file(file, "rt") : cannot open the connection
In addition: Warning message:
In file(file, "rt") :
  cannot open file 'data/nope.csv': No such file or directory

5: file(file, "rt")
4: read.table(file = file, header = header, sep = sep, quote = quote, 
       dec = dec, fill = fill, comment.char = comment.char, ...)
3: read.csv(path, colClasses = c(Site.ID = "character"))
2: FUN(X[[i]], ...)
1: lapply(c("data/epa_pm25_compton_2025.csv", "data/nope.csv"), 
       read_epa)
```

::: {.content-hidden when-format="revealjs"}

`traceback()` prints the calls that were running when the error happened,
innermost first. Read it from the bottom: your `lapply()` (1) called your
function (2, `FUN` is `read_epa`), which called `read.csv()` (3), which
could not open the file (5). The warning names the file. The error is in
the line you wrote at 1, not in `read.csv()`.

:::

::: {.content-visible when-format="revealjs"}

- Read it from the bottom: your call, your function, then R's.
- The error is in the line you wrote at 1. The warning names the file.

:::

## `print()` and `cat()`: how a value evolved {#print}

```{r}
#| label: print
#| script: "print() and cat(): how a value evolved"
for (site in names(by_site)[1:3]) {
  s <- by_site[[site]]
  cat(site, ": ", nrow(s), " days, ", days_above(s$pm25), " above\n", sep = "")
}

print(c(days = nrow(jan), above = days_above(jan$pm25)))
```

::: {.content-hidden when-format="revealjs"}

Inside a loop or a function, a `print()` or `cat()` line shows what a value
was on each pass. `print()` shows an object the way the console would;
`cat()` pastes text and numbers into one line, and `"\n"` ends it. A named
vector, `c(days = , above = )`, prints label and value together. Take the
line out once the question is answered.

`browser()` is the same idea with a pause. Put `browser()` inside a
function and the run stops there with a `Browse[1]>` prompt; you can print
any local name, `n` steps to the next line, `c` continues, `Q` quits.

:::

::: {.content-visible when-format="revealjs"}

- `cat()` pastes text and numbers into one line. `print()` shows an object.
- A named vector prints label and value together.
- Take the line out once the question is answered.

:::

## Halving: a wrong number that never errored {#halving}

::: {.content-hidden when-format="revealjs"}

No error, no warning, and the number is wrong. `traceback()` has nothing
to point at. Where could the wrong number have entered? Every step that
changes a row count or a value is a suspect.

Write down what each step should produce: rows after the read, rows after
the filter, the date range, the count. Comment out the second half of the
script and run the first half. Compare what printed with what you wrote
down. Halve the half that disagrees. Repeat until the mistake has nowhere to
hide. It is usually one line, and the line is usually not the one that
printed the number.

:::

::: {.content-visible when-format="revealjs"}

- No error, no warning, and the number is wrong. Nothing to trace.
- Write down what each step should produce: rows, dates, a count.
- Comment out half. Run. Compare. Halve the half that disagrees.
- Repeat until the mistake has nowhere to hide. Usually one line.

:::

## Documenting a diagnosis {#diagnosis}

```{r}
#| label: diagnosis
#| script: "Documenting a diagnosis: line, mechanism, fix, proof"
# Line:      share_above <- function(x) sum(x > standard) / length(x)
# Mechanism: standard is not an argument; the function reads it from the
#            workspace and fails in a fresh session
# Fix:       the standard comes in as an argument, with a default
share_above <- function(x, standard = 35) sum(x > standard) / length(x)
# Proof:
stopifnot(share_above(c(1, 50)) == 0.5)
```

::: {.content-hidden when-format="revealjs"}

Four lines, as comments, next to the fix. The line that was wrong. The
mechanism, meaning why it produced what it produced. The fix. And the
proof: a small case that now gives the right answer, written as a check
that runs. That is what a fixed script carries. When a check fails and you
cannot fix it, the same four lines with the fix left blank are the honest
note, and they are worth more than an edited check.

:::

::: {.content-visible when-format="revealjs"}

- Line. Mechanism. Fix. Proof, as a check that runs.
- Cannot fix it? The same four lines, fix left blank. Never an edited check.

:::

## Practice {#practice}

::: {.content-visible when-format="revealjs"}

Eight exercises with solutions, on the session page and in the live script.

:::

::: {.content-hidden when-format="revealjs"}

::: {.panel-tabset}

### Practice

```{r}
#| label: practice
#| exercise: true
#| script: "Practice"
#| eval: false
# Everything below uses pm, dec, jan, files, frames, la, by_site,
# read_epa() and days_above(), all built earlier in this script.

# Functions
#  1. Write peak_day(d): the date of the highest pm25 in a data frame d.
#     Run it on pm. Check the answer against the row that which.max()
#     points at.
#  2. Run peak_day() on every site in by_site. Which site peaked last in
#     January? (sapply() turns dates into numbers; wrap the call in
#     format() to keep them readable.)
#  3. From memory, without scrolling up: write share_above(x, standard = 35)
#     so that it works in a fresh session, with nothing read from the
#     workspace. Run it on dec$pm25 and on c(1, 50).

# Lists and loops
#  4. How many sites does each frame in frames hold? One sapply() call.
#  5. One plot per site, written to disk: for each site in by_site, a
#     line plot of pm25 against date saved as a PNG named after the site.
#     Do not run it until it works for one site.

# Checks
#  6. After read_epa() on the county file, write one stopifnot() with two
#     conditions: exactly 8 sites, and no date outside 2025.
#  7. This check fails on la. Say why, and whether the data or the check
#     is wrong. Then write the check that is right for this frame.
#       stopifnot(nrow(la) == length(unique(la$date)))
#  8. Time read_epa() on the county file alone, then lapply() over both
#     files. Does the second take more than the first? By roughly how much?
```

### Solutions

::: {.callout-tip collapse="true"}
## Solutions

```{r}
#| label: practice-solutions
#| purl: false
# 1.
peak_day <- function(d) d$date[which.max(d$pm25)]
peak_day(pm)
pm[which.max(pm$pm25), c("date", "pm25")]

# 2. Pasadena, on 2025-01-10. Four sites peaked on January 1.
sapply(by_site, function(s) format(peak_day(s)))

# 3.
share_above <- function(x, standard = 35) sum(x > standard) / length(x)
round(share_above(dec$pm25), 3)
share_above(c(1, 50))

# 4.
sapply(frames, function(d) length(unique(d$site)))

# 6. The second condition is a whole vector; every element must be TRUE.
stopifnot(length(unique(la$site)) == 8,
          format(la$date, "%Y") == "2025")

# 7. The check is wrong for this frame, not the data. Several sites
#    report on the same day, so dates repeat by design. The right check
#    is one row per site and day.
nrow(la) == length(unique(la$date))
stopifnot(nrow(la) == nrow(unique(la[, c("site", "date")])))

# 8. Barely. The Compton file is about a seventh the size of the county
#    file, so the second call costs about a seventh more.
system.time(read_epa("data/epa_pm25_la_county_2025.csv"))
system.time(lapply(files, read_epa))
```

```{r}
#| label: practice-solution-6
#| purl: false
#| eval: false
# 5. Test on one site first: for (site in names(by_site)[1]). Then all of them.
for (site in names(by_site)) {
  s <- by_site[[site]]
  png(paste0(site, ".png"), width = 1600, height = 800, res = 150)
  plot(s$date, s$pm25, type = "l", main = site,
       xlab = "", ylab = "PM2.5 (ug/m3)")
  dev.off()
}
```

:::

:::

:::

## Quick reference {#quick-reference}

::: {.content-visible when-format="revealjs"}

On the page: every command on this page in one table, and the habits next
to it.

:::

::: {.content-hidden when-format="revealjs"}

**Functions and lists**

| Code | What it does |
|:--|:--|
| `f <- function(x, k = 35) { ... }` | A function; `k = 35` is a default; the value is the last expression |
| `return(x)` | Leave the function now with `x`; only needed to leave early |
| `function(s) ...` inside `lapply()` | A function with no name, written where it is used |
| `list.files("data", pattern =, full.names = TRUE)` | Paths you can hand to a function |
| `lapply(files, f)` | A list in, a list out: `f` on each element |
| `sapply(x, f)` | The same, simplified when every element has the same shape |
| `split(d, d$g)` | A list, one data frame per value of `g` |

**Checks and habits**

| Code | What it does |
|:--|:--|
| `stopifnot(a, b)` | Stop the script and name the first condition that is not `TRUE` |
| `stopifnot("why" = a)` | The same, printing your message instead |
| `STANDARD <- 35` | A named constant at the top; used by name everywhere |
| `# Name ----` | A section RStudio's outline can jump to |
| `system.time(expr)` | Seconds the line took; wrap the big call once |
| Session > Restart R, then `source("air.R")` | The only test that a script runs on its own |
| `getwd()`, `list.files("data")` | Where relative paths are looked for, and what is there |

**When something is wrong**

| Symptom | Tool |
|:--|:--|
| An error | Read it. Then `traceback()`: your call is at the bottom |
| A loop or function misbehaves on some passes | `print()` or `cat()` inside it; `browser()` to pause there |
| A wrong number and no error | Halve the script: write down what each step should give, compare |
| `object 'x' not found` inside a function | The function reads a global; make `x` an argument |
| `NaN` from a mean | The subset was empty; that is the honest answer |
| A file "does not exist" | `getwd()` and `list.files()` before blaming the file |

:::

```{r}
#| purl: false
#| include: false
source("_purl.R")
purl_session("06-how-to-code.qmd", "06-how-to-code.R")
```
